### 2024-12-30 tex/main.tex
[ new ] tex file to start with

\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle
\title{Type theory as a quotient inductive-recursive type}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{[funding]}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen and T.-C.~Tsai}
\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{The formal implementation is hosted at \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko.}%optional

\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{macro.tex}

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Type theory can be formally defined in a type theory as a quotient inductive-inductive type (QIIT for short), but terms are littered with explicit transports along equality constructors, typically omitted and retouched in presentation, in proofs or even in the definition of type theory.
  In this paper, we argue that QIITs are sufficient but infeasible to define type theories in practice and propose to use quotient inductive-inductive-recursive types (\emph{QIIRT}s for short) and definitions by overlapping patterns altogether, to reduce the use of transports.
  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as QIIRTs and prove meta-theoretic properties and compare our definitions with typical definitions using QIITs.
\end{abstract}

\section{Introduction} \label{sec:intro}
Points to elaborate:
\begin{enumerate}
  \item QIITs are littered with transports, causing too much pains while formalising a type theory.
  \item Rewrite rules are popular to make structural rules definitional but requires efforts to justify meta-theoretic properties in general.
  \item Further, the confluence check adopted by \Agda is based on complete development and requires one-step parallel reductions to have a confluent term, making a definition more complicated than necessary.
  \item Moreover, the use of rewrite rules is typically \emph{not} presented and discussed in literature. We are ignoring the elephant in the room.
  \item Instead, definitions by overlapping patterns (in theory) requires strong normalisation and local confluence, making definitions simpler to design.
  \item The rules for type substitutions are \emph{structural} shared with other type theories based on parallel substitution (in particular cwf).

\end{enumerate}

\paragraph*{Contributions}
\begin{itemize}
  \item Exploring the potential use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give
    \begin{itemize}
      \item a definition of parallel substitution calculus in type theory, whose type substitution rules are definitional and confluent, using quotient inductive-inductive-recursive types and definitions by overlapping patterns;
      \item a definition of type theory with Coquand universes and $\Pi$-types in type theory, extending the above substitution calculus.
    \end{itemize}
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

\paragraph*{Plan of the paper}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006}
\paragraph*{Single substitution calculus}
\cite{Kaposi2023,Kaposi2024a}
\paragraph*{Schemata of inductive types}
\cite{Altenkirch2016a,Kaposi2019}

\section{Metatheory and formalisation}
\cite{Pujet2022,Pujet2024}

\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}
\subsection{Definitions by overlapping patterns}
\cite{Cockx2014}
\subsection{Rewrite rules}
\cite{Cockx2020,Cockx2021}
\paragraph*{Local confluence}
\paragraph*{Strong normalisation}

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\subsection{Parallel substitution calculus} \label{subsec:SC}
\subsection{... with Coquand universes} \label{subsec:SC+U}
\cite{Coquand2013}
\subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}

\section{Metatheoretic properties}

\section{Future work}

\paragraph*{Type theory in (observational) type theory}

\paragraph*{Formalising definitions by overlapping patterns}

\paragraph*{A general schema of quotient inductive-inductive-recursive types}

\IfFileExists{./reference.bib}{
  \bibliography{reference}
}{
  \bibliography{ref}
}

\appendix

\begin{conjecture}\label{testenv-conjecture}
\end{conjecture}

\begin{claim}\label{testenv-claim}
\end{claim}

\begin{claim*}\label{testenv-claim2}
\end{claim*}

\begin{claimproof}
\end{claimproof}
\begin{lstlisting}[caption={Formal definition in \Agda},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
for i:=maxint to 0 do 
begin 
    j:=square(root(i));
end;
\end{lstlisting}

\end{document}

### 2024-12-31 tex/main.tex
Pen-and-paper presentation

@@
 \title{Type theory as a quotient inductive-recursive type}
-\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{[funding]}
+\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding]}
 \author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
@@
 
-\supplement{The formal implementation is hosted at \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g.
+\supplement{The formal development is hosted at the GitHub repository: \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g.
 related research data, source code, ...
@@
   \item Exploring the potential use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
-    In particular, we give
+    In particular, we give in a type theory:
     \begin{itemize}
-      \item a definition of parallel substitution calculus in type theory, whose type substitution rules are definitional and confluent, using quotient inductive-inductive-recursive types and definitions by overlapping patterns;
-      \item a definition of type theory with Coquand universes and $\Pi$-types in type theory, extending the above substitution calculus.
+      \item a definition of parallel substitution calculus;
+      \item a definition of type theory with Coquand universes and $\Pi$-types,
     \end{itemize}
+   whose type substitution rules are definitional using quotient inductive-inductive-recursive types and definitions by overlapping patterns.
   \item Comparison with other definitions using quotient inductive-inductive types.
@@
 
-\paragraph*{Plan of the paper}
+\subsection{Plan of the paper}
 \subsection{Related work}
@@
 \cite{Altenkirch2016a,Kaposi2019}
+\cite{Dybjer2003,Dybjer2000,Dybjer1999}
 
 \section{Metatheory and formalisation}
-\cite{Pujet2022,Pujet2024}
+\cite{Pujet2022,Pujet2024,Pujet2022a}
 
-\subsection{Inductive-recursive types}
-\cite{Dybjer2003,Dybjer2000,Dybjer1999}
+\subsection{Quotient inductive-inductive-recursive types}
 \subsection{Definitions by overlapping patterns}
@@
 \subsection{Parallel substitution calculus} \label{subsec:SC}
+\[
+\begin{alignedat}{10}
+  \Ctx   & : \Set                   \\
+  \Ty    & : \Ctx \to \Set          \\
+  \Sub   & : \Ctx \to \Ctx \to \Set \\
+  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set \\
+  \emptyctx & : \Ctx \\
+  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx \\
+  [\blank]_{\Ty}\blank &: \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+  \emptysub & : \Sub\,\Gamma\,\emptyctx \\
+  \blank,\blank & : \Sub\,\Gamma\,\Delta \to \Tm\,\Gamma\,([ \sigma ] A) \to \Sub\,\Gamma,(\Delta, A) \\
+  \idS_{\blank} & : (\Gamma : \Ctx) \to \Sub\,\Gamma\,\Gamma \\
+  \blank;\blank & : \Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
+  \pi_1 & : \Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
+  \pi_2 & : (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \sigma ] A) \\
+  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma] A) \\
+  [\blank] \blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma] A)
+\end{alignedat}
+\qquad
+\begin{alignedat}{10}
+          & [ \idS ]_{\Ty}\,A           && = A \\
+          & [ \idS ]_{\Tm}\,t           && = t \\
+          & [ \sigma ; \tau ]_{\Ty}\,A  && = [ \sigma ]_{\Ty}\,[ \tau ]_{\Ty} A \\
+          & [ \sigma ; \tau ]_{\Tm}\,t  && = [ \sigma ]_{\Tm}\,[ \tau ]_{\Tm} t \\
+          & [ \sigma ]_{\Tm}\,t         && = [ \sigma ]\,t & \text{otherwise}  \\
+  [\idS]t & : [\,\idS\,]\,t             && = t \\
+  [;]t    & : [\,\sigma ; \tau\,]\,t    && = [ \sigma ]\,[ \tau ]\,t \\
+\end{alignedat}
+\]
+
 \subsection{... with Coquand universes} \label{subsec:SC+U}

### 2025-01-01 tex/main.tex
[ workflow ] Add a LaTeX action

@@
 
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\bibliographystyle{plainurl}
+
 \title{Type theory as a quotient inductive-recursive type}
-\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding]}
+\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding].}
 \author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
@@
 Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko.}%optional
+\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko, and Meven Lennon-Bertrand.}%optional
 
@@
 
-%TODO mandatory: add short abstract of the document
 \begin{abstract}
-  Type theory can be formally defined in a type theory as a quotient inductive-inductive type (QIIT for short), but terms are littered with explicit transports along equality constructors, typically omitted and retouched in presentation, in proofs or even in the definition of type theory.
-  In this paper, we argue that QIITs are sufficient but infeasible to define type theories in practice and propose to use quotient inductive-inductive-recursive types (\emph{QIIRT}s for short) and definitions by overlapping patterns altogether, to reduce the use of transports.
-  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as QIIRTs and prove meta-theoretic properties and compare our definitions with typical definitions using QIITs.
+  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors, typically omitted in presentation for clarity.
+  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
+  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as QIIRTs with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
 \end{abstract}
@@
 \begin{itemize}
-  \item Exploring the potential use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
+  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
     In particular, we give in a type theory:
@@
 \paragraph*{Formalisation of type theory in type theory}
-\cite{Danielsson2006}
+\cite{Danielsson2006,Altenkirch2016a}
+\cite{Altenkirch2017}
+
 \paragraph*{Single substitution calculus}
@@
 \paragraph*{Schemata of inductive types}
-\cite{Altenkirch2016a,Kaposi2019}
+\cite{Kaposi2019}
 \cite{Dybjer2003,Dybjer2000,Dybjer1999}
@@
 
+\subsection{Definitions by overlapping patterns}
+\cite{Cockx2014,Altenkirch2016a}
 \subsection{Quotient inductive-inductive-recursive types}
-\subsection{Definitions by overlapping patterns}
-\cite{Cockx2014}
-\subsection{Rewrite rules}
+\subsection{Definitions by rewrite rules}
 \cite{Cockx2020,Cockx2021}
+
+Unfortunately, neither general schemata of quotient inductive-inductive-recursive types nor definitions by overlapping patterns have been developed or implemented in any existing proof assistant.
+
 \paragraph*{Local confluence}
@@
 \subsection{Parallel substitution calculus} \label{subsec:SC}
-\[
-\begin{alignedat}{10}
+\begin{alignat*}{3}
   \Ctx   & : \Set                   \\
@@
   [\blank]_{\Ty}\blank &: \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+  U & : \Ty \\
   \emptysub & : \Sub\,\Gamma\,\emptyctx \\
-  \blank,\blank & : \Sub\,\Gamma\,\Delta \to \Tm\,\Gamma\,([ \sigma ] A) \to \Sub\,\Gamma,(\Delta, A) \\
+  \blank,\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
   \idS_{\blank} & : (\Gamma : \Ctx) \to \Sub\,\Gamma\,\Gamma \\
@@
   \pi_1 & : \Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
-  \pi_2 & : (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \sigma ] A) \\
-  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma] A) \\
-  [\blank] \blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma] A)
-\end{alignedat}
-\qquad
-\begin{alignedat}{10}
-          & [ \idS ]_{\Ty}\,A           && = A \\
-          & [ \idS ]_{\Tm}\,t           && = t \\
-          & [ \sigma ; \tau ]_{\Ty}\,A  && = [ \sigma ]_{\Ty}\,[ \tau ]_{\Ty} A \\
-          & [ \sigma ; \tau ]_{\Tm}\,t  && = [ \sigma ]_{\Tm}\,[ \tau ]_{\Tm} t \\
-          & [ \sigma ]_{\Tm}\,t         && = [ \sigma ]\,t & \text{otherwise}  \\
-  [\idS]t & : [\,\idS\,]\,t             && = t \\
-  [;]t    & : [\,\sigma ; \tau\,]\,t    && = [ \sigma ]\,[ \tau ]\,t \\
-\end{alignedat}
-\]
+  \pi_2 & : (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty}\, A) \\
+  [\blank] \blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\, A)
+\end{alignat*}
+
+\begin{alignat*}{3}
+[ \idS ]_{\Ty}\,A          & = A \\
+[ \sigma ; \tau ]_{\Ty}\,A & = [ \sigma ]_{\Ty}\;[ \tau ]_{\Ty}\;A \\
+[ \sigma ]_{\Ty}\,U & = U
+\end{alignat*}
+
+\begin{alignat*}{3}
+  [\idS]t         & : {[\,\idS\,]\,t}          && =_{\Tm\,\Gamma\,A} && t \\
+  [;]t            & : {[\,\sigma ; \tau\,]\,t} && =_{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
+  \mathsf{idr}    & : {\sigma ; \idS_{\Delta}} && =_{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  \mathsf{idl}    & : {\idS_{\Gamma} ; \sigma} && =_{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{assoc} & : (\sigma ; \tau) ; \gamma && =_{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \sigma ; (\tau , t)      && =_{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
+  \emptyctx\eta   & : \sigma                   && =_{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \pi\eta         & : \sigma                   && =_{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  \pi_1\beta      & : \pi_1(\sigma , t)        && =_{\Sub\,\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \pi_2(\sigma , t)        && =_{\Tm\,\Gamma\, A} &&  t
+\end{alignat*}
 
@@
 \cite{Coquand2013}
+\begin{alignat*}{3}
+[ \idS ]_{\Tm}\,t          & = t \\
+[ \sigma ; \tau ]_{\Tm}\,t & = [ \sigma ]_{\Tm}\;[ \tau ]_{\Tm}\;t \\
+[ \sigma ]_{\Tm}\,t        & = [ \sigma ]\,t & \text{otherwise}  \\
+\end{alignat*}
 \subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
@@
 
-\IfFileExists{./reference.bib}{
-  \bibliography{reference}
-}{
-  \bibliography{ref}
-}
+\IfFileExists{./reference.bib}{\bibliography{reference}}{\bibliography{ref}}
 

### 2025-01-01 tex/main.tex
Give an example of transport hell

@@
 
+\usepackage{todonotes}
+
 \input{macro.tex}
@@
 \begin{abstract}
-  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors, typically omitted in presentation for clarity.
+  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors.
+  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in public presentation for clarity, ignoring the elephant in the room.
   In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
@@
 \cite{Pujet2022,Pujet2024,Pujet2022a}
+Points to include:
+\begin{enumerate}
+  \item dependent identity type (heterogeneous equality in \Agda)
+  \item 
+\end{enumerate}
 
@@
 
-Unfortunately, neither general schemata of quotient inductive-inductive-recursive types nor definitions by overlapping patterns have been developed or implemented in any existing proof assistant.
+Unfortunately, neither general schemata of quotient inductive-inductive-recursive types nor definitions by overlapping patterns have been developed or implemented in existing proof assistants.
 
@@
   \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx \\
-  [\blank]_{\Ty}\blank &: \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-  U & : \Ty \\
-  \emptysub & : \Sub\,\Gamma\,\emptyctx \\
-  \blank,\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
-  \idS_{\blank} & : (\Gamma : \Ctx) \to \Sub\,\Gamma\,\Gamma \\
-  \blank;\blank & : \Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
-  \pi_1 & : \Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
-  \pi_2 & : (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty}\, A) \\
-  [\blank] \blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\, A)
+  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+  U & : \implicit{\Gamma}\;\Ty\, \Gamma\\
+  \emptysub & : \implicit{\Gamma}\;\Sub\,\Gamma\,\emptyctx \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
+  \idS & : \implicit{\Gamma}\;\Sub\,\Gamma\,\Gamma \\
+  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
+  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A) \\
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-[ \idS ]_{\Ty}\,A          & = A \\
-[ \sigma ; \tau ]_{\Ty}\,A & = [ \sigma ]_{\Ty}\;[ \tau ]_{\Ty}\;A \\
-[ \sigma ]_{\Ty}\,U & = U
+  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \,A         && =^{\Ty\,\Gamma} A \\
+  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\,A && =^{\Ty\,\Gamma} [ \sigma ]\;([ \tau ]\;A) \\
+  []U      & : \implicit{\Gamma, \Delta, \sigma}                  && [ \sigma ]\,U        && =^{\Ty\,\Gamma} U
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
+  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
+  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
+  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t} \\
+\end{alignat*}
+
+\begin{example}
+  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$.
+  Intuitively, this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ applied by the substitution $\sigma$. 
+  That is, the following equality apparently holds
+  \[
+    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
+  \]
+  by a back-of-the-envelope calculation
+  \begin{equation} \label{eq:pi2-comp-proof}
+    \pi_2\,(\sigma ; \tau) 
+    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
+    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
+    = [\sigma] (\pi_2\,\tau).
+  \end{equation}
+
+  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
+  Hence the above identity does not even make sense, since their types do not match.
+  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, more verbosely,
+  \[
+    \transfib{\Tm\,\Gamma}{p}{\pi_2\,(\sigma ; \tau)} =^{\Tm\,\Gamma\,([\sigma][\pi_1\,\tau]\,A)}  [\sigma] (\pi_2\,\tau)
+  \]
+  where $p$ is a term of the identity $[\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\,\tau]A$.
+  Now that we have a transported term we have to carry out our proof with coercion rules explicitly, and moreover in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ and $\pi_2\beta$ which introduced another transported terms.
+  It turns out that we have to `eliminate' these $\transp$'s to derive the right-hand side of~\eqref{eq:pi2-comp-proof}, making the actual proof barely readable.
+  \todo{write down the actual proof}
+\end{example}
+\begin{alignat*}{3}
+[ \idS ]_{\Ty}\,A             & \jdeq A \\
+[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
+[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
+[ \sigma ]_{\Ty}\,U           & \jdeq U
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  [\idS]t         & : {[\,\idS\,]\,t}          && =_{\Tm\,\Gamma\,A} && t \\
-  [;]t            & : {[\,\sigma ; \tau\,]\,t} && =_{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
-  \mathsf{idr}    & : {\sigma ; \idS_{\Delta}} && =_{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  \mathsf{idl}    & : {\idS_{\Gamma} ; \sigma} && =_{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  ;\text{-}\mathsf{assoc} & : (\sigma ; \tau) ; \gamma && =_{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \sigma ; (\tau , t)      && =_{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
-  \emptyctx\eta   & : \sigma                   && =_{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
-  \pi\eta         & : \sigma                   && =_{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  \pi_1\beta      & : \pi_1(\sigma , t)        && =_{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \pi_2(\sigma , t)        && =_{\Tm\,\Gamma\, A} &&  t
+  \mathsf{idr}    & : {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  \mathsf{idl}    & : {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{assoc} & : (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
+  \emptyctx\eta   & : \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \pi\eta         & : \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  \pi_1\beta      & : \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
+  [\idS]t         & : {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
+  [;]t            & : {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
 \end{alignat*}

### 2025-01-02 tex/main.tex
[ tex ]

@@
   Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors.
-  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in public presentation for clarity, ignoring the elephant in the room.
-  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
-  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as QIIRTs with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
+  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in presentation for clarity, ignoring the elephant in the room.
+  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
+  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
 \end{abstract}
@@
 \begin{enumerate}
-  \item dependent identity type (heterogeneous equality in \Agda)
+  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality in \Agda $x \simeq y$ for $x : A$ and $y : B$
   \item 
@@
   \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
+  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
   [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
@@
 \begin{example}
-  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$.
-  Intuitively, this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ applied by the substitution $\sigma$. 
-  That is, the following equality apparently holds
+  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$.
+  In short, the following equality apparently holds
   \[
@@
   \end{equation}
-
   Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
   Hence the above identity does not even make sense, since their types do not match.
-  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, more verbosely,
+  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
   \[
-    \transfib{\Tm\,\Gamma}{p}{\pi_2\,(\sigma ; \tau)} =^{\Tm\,\Gamma\,([\sigma][\pi_1\,\tau]\,A)}  [\sigma] (\pi_2\,\tau)
+    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
   \]
-  where $p$ is a term of the identity $[\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\,\tau]A$.
-  Now that we have a transported term we have to carry out our proof with coercion rules explicitly, and moreover in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ and $\pi_2\beta$ which introduced another transported terms.
-  It turns out that we have to `eliminate' these $\transp$'s to derive the right-hand side of~\eqref{eq:pi2-comp-proof}, making the actual proof barely readable.
-  \todo{write down the actual proof}
+  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
+  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to eliminate that $\transp$ to derive the right hand side.
+
+  To better illustrate the annoyance, we provide a complete proof.
+  First, we claim each of the following steps holds:
+  \begin{alignat}{3}
+         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \label{eq:pi2-proof-1} \\
+    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
+    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
+    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
+    = {} & (\sub{\sigma;\pi_1\tau}{A} &&, [\sigma] (\pi_2\,\tau)) \nonumber
+  \end{alignat}
+  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$.
 \end{example}
+
 \begin{alignat*}{3}

### 2025-01-17 tex/main.tex
Outline sections

@@
 
-\title{Type theory as a quotient inductive-recursive type}
+\title{Type theories as quotient inductive-recursive types}
 \author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding].}
@@
 \cite{Kaposi2019}
-\cite{Dybjer2003,Dybjer2000,Dybjer1999}
 
@@
 \end{enumerate}
+
+\subsection{Inductive-recursive types}
+\cite{Dybjer2003,Dybjer2000,Dybjer1999}
+
+Mention that the recursion part of induction-recursion are already not defined by elimination rule.
 
@@
   \Sub   & : \Ctx \to \Ctx \to \Set \\
-  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set \\
+  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set
+\end{alignat*}
+\begin{alignat*}{3}
   \emptyctx & : \Ctx \\
-  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx \\
+  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx
+\end{alignat*}
+\begin{alignat*}{3}
   [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-  U & : \implicit{\Gamma}\;\Ty\, \Gamma\\
+  U & : \implicit{\Gamma}\;\Ty\, \Gamma
+\end{alignat*}
+\begin{alignat*}{3}
   \emptysub & : \implicit{\Gamma}\;\Sub\,\Gamma\,\emptyctx \\
@@
   Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
-  \[
+  \begin{equation}\label{eq:pi2-comp-real-proof}
     ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
-  \]
+  \end{equation}
   as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
@@
 
-  To better illustrate the annoyance, we provide a complete proof.
-  First, we claim each of the following steps holds:
-  \begin{alignat}{3}
+  To better illustrate the annoyance, the complete proof of \eqref{eq:pi2-comp-real-proof} requires to show first three equations and the last equation explicitly by Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively:
+  \begin{alignat*}{3}
          & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \label{eq:pi2-proof-1} \\
     = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
-    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
+    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \\
     = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
     = {} & (\sub{\sigma;\pi_1\tau}{A} &&, [\sigma] (\pi_2\,\tau)) \nonumber
-  \end{alignat}
+  \end{alignat*}
   where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$.
@@
 [ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
+[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi\_1\tau]_\Ty \,A) \\
 [ \sigma ]_{\Ty}\,U           & \jdeq U
@@
 
-\subsection{... with Coquand universes} \label{subsec:SC+U}
+\subsection{... with a Coquand universe} \label{subsec:SC+U}
 \cite{Coquand2013}
@@
 [ \sigma ; \tau ]_{\Tm}\,t & = [ \sigma ]_{\Tm}\;[ \tau ]_{\Tm}\;t \\
+[ \pi_1(\sigma, t) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\,t \\
+[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm ([\pi\_1\tau]_\Tm \,t) \\
+[ \sigma ]_{\Ty}\,U           & \jdeq U
 [ \sigma ]_{\Tm}\,t        & = [ \sigma ]\,t & \text{otherwise}  \\
@@
 
+\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
+
+\section{Eliminators and their examples}
+\begin{enumerate}
+  \item The elimination rule is somewhat surprising
+
+\end{enumerate}
+
+\section{Equivalence to type theories as quotient inductive-inductive types}
 \section{Metatheoretic properties}

### 2025-01-18 tex/main.tex
List formal definitions

@@
 \usepackage{todonotes}
+\usepackage{manfnt}
+\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
+\newcommand{\LT}[1]{\todo[inline,author={L-T},caption={}]{#1}}
 
@@
 \begin{abstract}
-  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are painstakingly littered with explicit coercions, i.e.\ \transp's, along equality constructors.
-  This mess has been dubbed the \emph{transport hell} by the community in private but typically suppressed in presentation for clarity, ignoring the elephant in the room.
-  In this paper, we aim for shortening the gap between formal and informal definitions of type theory in type theory using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
-  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types with meta-theoretic properties proved and compare our definitions with quotient inductive-inductive definitions.
+  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs needs to take account of coercions painstakingly alongside the interesting part.
+  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
+  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
+  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
 \end{abstract}
@@
 \section{Introduction} \label{sec:intro}
+\LT{%
 Points to elaborate:
@@
 \end{enumerate}
+}
 
@@
     \end{itemize}
-   whose type substitution rules are definitional using quotient inductive-inductive-recursive types and definitions by overlapping patterns.
+   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
   \item Comparison with other definitions using quotient inductive-inductive types.
@@
 \section{Metatheory and formalisation}
-\cite{Pujet2022,Pujet2024,Pujet2022a}
+\cite{UFP2013}
+\LT{
 Points to include:
 \begin{enumerate}
-  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality in \Agda $x \simeq y$ for $x : A$ and $y : B$
-  \item 
-\end{enumerate}
+  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
+  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
+  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
+\end{enumerate}}
 
@@
 
-Mention that the recursion part of induction-recursion are already not defined by elimination rule.
+\LT{Mention that the recursion part of induction-recursion are already not defined by elimination rule.}
 
@@
 \cite{Cockx2014,Altenkirch2016a}
-\subsection{Quotient inductive-inductive-recursive types}
+\subsection{Quotient inductive(-inductive)-recursive types}
+\LT{Give an example of QIRT: join list?}
 \subsection{Definitions by rewrite rules}
@@
 \section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
+\LT{
+\begin{enumerate}
+  \item Goal: no transports in the definition.
+  \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
+  \item Develop the quotient inductive-inductive-recursive definition of type theory step by step (which needs to be locally confluent and terminating).
+    \begin{enumerate}
+      \item parallel substitution: type substitution
+      \item the type of elements of $A$: term substitution needs to be split into two definitional substitution and explicit term substitution to maintain local confluence.
+      \item $\Pi$-type: because of the substitution lifting $\sigma^+ \defeq (\pi_1\idS, \pi_2\idS)$ is used we need to introduce a definitional lifting to maintain local confluence.
+        We prefer categorical combinator and avoid using $\left< t \right>$ in the definition, which would introduce transports in the definition.
+      \item Other type formers can be introduced as usual (as long as we use categorical combinators).
+    \end{enumerate}
+\end{enumerate}
+}
 \subsection{Parallel substitution calculus} \label{subsec:SC}
@@
   [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-  U & : \implicit{\Gamma}\;\Ty\, \Gamma
+\end{alignat*}
+
+\begin{alignat*}{3}
+  \UU & : \implicit{\Gamma} & \Ty\, \Gamma
 \end{alignat*}
@@
   \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
-  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
   \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
@@
 
-  To better illustrate the annoyance, the complete proof of \eqref{eq:pi2-comp-real-proof} requires to show first three equations and the last equation explicitly by Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively:
+  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
   \begin{alignat*}{3}
@@
   \end{alignat*}
-  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$.
+  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$,
 \end{example}
+explicitly.
+The first three equations and the last equation follow from Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively, which cannot be proved naively as we did in \eqref{eq:pi2-comp-proof}.
 
 \begin{alignat*}{3}
+  [\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
 [ \idS ]_{\Ty}\,A             & \jdeq A \\
@@
 [ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
-[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi\_1\tau]_\Ty \,A) \\
-[ \sigma ]_{\Ty}\,U           & \jdeq U
+[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
+[ \sigma ]_{\Ty}\,\UU           & \jdeq \UU
 \end{alignat*}
 
-\begin{alignat*}{3}
-  \mathsf{idr}    & : {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  \mathsf{idl}    & : {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  ;\text{-}\mathsf{assoc} & : (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
-  \emptyctx\eta   & : \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
-  \pi\eta         & : \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  \pi_1\beta      & : \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
-  [;]t            & : {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
+\begin{alignat*}{5}
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
+  [\idS]t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
+  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
 \end{alignat*}
+
+\begin{proposition}[Local confluence]
+  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+  \danger
+\end{proposition}
+\begin{proposition}[Coherence]
+  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
+  \danger
+\end{proposition}
 
@@
 \begin{alignat*}{3}
-[ \idS ]_{\Tm}\,t          & = t \\
-[ \sigma ; \tau ]_{\Tm}\,t & = [ \sigma ]_{\Tm}\;[ \tau ]_{\Tm}\;t \\
+  \UU  & : \implicit{\Gamma}\;\Ty\, \Gamma \\
+  \El & : \implicit{\Gamma}\;\Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
+  [ \sigma ]_{\Ty}\,(\El\, u) & \jdeq \El\,([\sigma]_{\Tm}{u}) \\
+\end{alignat*}
+\begin{alignat*}{3}
+  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
+[ \idS ]_{\Tm}\,t          & \jdeq t \\
+[ \sigma ; \tau ]_{\Tm}\,t & \jdeq [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
 [ \pi_1(\sigma, t) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\,t \\
-[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm ([\pi\_1\tau]_\Tm \,t) \\
-[ \sigma ]_{\Ty}\,U           & \jdeq U
-[ \sigma ]_{\Tm}\,t        & = [ \sigma ]\,t & \text{otherwise}  \\
+[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
+[ \sigma ]_{\Tm}\,t        & \jdeq [ \sigma ]\,t, \quad \text{otherwise}  \\
 \end{alignat*}
+
+\begin{proposition}[Local confluence]
+  Type and term substitutions $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are locally confluent.
+  \danger
+\end{proposition}
+
+\begin{proposition}[Coherence]
+  The following statements hold:
+  \danger
+  \begin{enumerate}
+    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
+    \item $[\sigma]_{\Tm}\,A = [\tau]_{\Tm}\,A$ whenever $\sigma = \tau$.
+  \end{enumerate}
+\end{proposition}
 \subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
+\begin{alignat*}{3}
+  \Pi     &: \implicit{\Gamma}            && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
+  \lambda &: \implicit{\Gamma, A, B}      && \Tm\,(\Gamma, A)\,B \to \Tm\,\Gamma\,(\Pi\,A\,B) \\
+  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\,\Gamma\,(\Pi\,A\,B) \to \Tm\,(\Gamma, A)\,B \\
+\end{alignat*}
+\begin{alignat*}{3}
+  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
+  \sigma^+ & \defeq (\pi_1\idS ; \sigma, \pi_2\idS) \\
+\end{alignat*}
+\begin{alignat*}{3}
+  \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
+\idS                \uparrow A & \jdeq \idS \\
+\sigma ; \tau       \uparrow A & \jdeq (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
+\pi_1(\sigma, t)    \uparrow A & \jdeq \sigma \uparrow A \\
+\pi_1(\sigma; \tau) \uparrow A & \jdeq \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
+\sigma              \uparrow A & \jdeq \sigma^+ , \quad \text{otherwiese} \\
+  [ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \jdeq \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
+\end{alignat*}
 
+\begin{alignat*}{3}
+  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \\
+  \Pi\beta & : \mathsf{app}(\lambda\,t) && =^{\Tm\,\Gamma,A} {} && t \\
+  \Pi\eta & :  t && =^{\Tm\,\Gamma,A} {} && \lambda(\mathsf{app}\,t)
+\end{alignat*}
+
+%% ⟨_⟩ : Tm Γ A → Sub Γ (Γ , A)
+%% ⟨ t ⟩ = idS , t
+%% 
+%% _$$_
+%%   : (t : Tm Γ (Π A B)) (u : Tm Γ A)
+%%   → Tm Γ ([ ⟨ u ⟩ ] B)
+%% t $$ u = [ ⟨ u ⟩ ]t app t
+\begin{alignat*}{3}
+  \left<\blank\right> & : \Tm\,\Gamma\,A \to \Sub\,\Gamma\,(\Gamma, A) \\
+  \left< t \right> & \defeq (\idS , t) \\
+  \blank \$\blank & : \Tm\,\Gamma\,(\Pi\,A\,B) \to (u : \Tm\,\Gamma\,A) \to \Tm\,\Gamma\,([ \left< u \right> ]\,B) \\
+  t \mathop{\$} u & \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)
+\end{alignat*}
+
+\begin{proposition}
+  For every substitution $\sigma : \Sub\,\Gamma\,\Delta$, type $A : \Ty\,\Delta$, and term $t : \Tm\,\Delta\,A$, the following statements hold:
+  \danger
+  \[
+    \sigma \uparrow A = (\pi_1 \idS ; \sigma , \pi_2 \idS)
+  \]
+  and
+  \[
+    [ \sigma ]_{\Tm}\,t = [ \sigma ]\,t
+  \]
+\end{proposition}
+
+\begin{proposition}
+  The following statements hold:
+  \danger
+  \begin{enumerate}
+    \item $\sigma \uparrow A = \tau \uparrow A$ whenever $\sigma = \tau$.
+    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
+    \item $[\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t$ whenever $\sigma = \tau$.
+  \end{enumerate}
+\end{proposition}
 \subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
+\LT{%
+\begin{enumerate}
+  \item Discuss the unit type.
+  \item Discuss the Boolean type.
+  \item Discuss the type of natural numbers.
+  \item Discuss the extensional identity type.
+\end{enumerate}
+}
 
-\section{Eliminators and their examples}
+\section{The eliminator and the standard model}
+\LT{%
 \begin{enumerate}
-  \item The elimination rule is somewhat surprising
+  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
+  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
+  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
+\end{enumerate}
+}
 
+\section{Equivalence to the quotient inductive-inductive definition of type theory}
+\LT{%
+\begin{enumerate}
+  \item Show that two definitions are equivalent (ongoing).
+  \item Show that methods are equivalent (ongoing).
 \end{enumerate}
-
-\section{Equivalence to type theories as quotient inductive-inductive types}
-\section{Metatheoretic properties}
+}
 
@@
 
-\paragraph*{Formalising definitions by overlapping patterns}
-
 \paragraph*{A general schema of quotient inductive-inductive-recursive types}
+\subparagraph*{Formalising definitions by overlapping patterns}
 
@@
 
-\begin{conjecture}\label{testenv-conjecture}
-\end{conjecture}
+\section{Complete definitions}
+\begin{alignat*}{3}
+  \Ctx      & : && \Set                   \\
+  \Ty       & : && \Ctx \to \Set          \\
+  \Sub      & : && \Ctx \to \Ctx \to \Set \\
+  \Tm       & : && (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set \\
+  \emptyctx & : && \Ctx \\
+  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx \\
+  [\blank]\blank & : \implicit{\Gamma, \Delta} && \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+  \emptysub & : \implicit{\Gamma} && \Sub\,\Gamma\,\emptyctx \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
+  \idS & : \implicit{\Gamma} && \Sub\,\Gamma\,\Gamma \\
+  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
+  \pi_1 & : \implicit{\Gamma, \Delta, A} && \Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
+  \pi_2 & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A) \\
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A) \\
+  \UU     & : \implicit{\Gamma} && \Ty\, \Gamma \\
+  \El     & : \implicit{\Gamma} && \Tm\,\Gamma\,U \to \Ty\,\Gamma \\
+  \Pi     & : \implicit{\Gamma} && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
+\end{alignat*}
 
-\begin{claim}\label{testenv-claim}
-\end{claim}
+\begin{alignat*}{5}
+[ \idS ]_{\Ty}\,A             & \jdeq A \\
+[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
+[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
+[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
+[ \sigma ]_{\Ty}\,\UU           & \jdeq \UU \\
+[ \sigma ]_{\Ty}\,(\El\, u) & \jdeq \El\,([\sigma]_{\Tm}{u}) \\
+[ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \jdeq \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
+\idS                \uparrow A & \jdeq \idS \\
+\sigma ; \tau       \uparrow A & \jdeq (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
+\pi_1(\sigma, t)    \uparrow A & \jdeq \sigma \uparrow A \\
+\pi_1(\sigma; \tau) \uparrow A & \jdeq \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
+\sigma              \uparrow A & \jdeq (\pi_1 \idS ; \sigma , \pi_2 \idS) & \text{otherwiese} \\
+ [ \idS ]_{\Tm}\,t                & \jdeq t \\
+ [ \sigma ; \tau ]_{\Tm}\,t       & \jdeq [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
+ [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \jdeq [\sigma]_\Tm\,t \\
+ [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
+ [ \sigma ]_{\Tm}\,t              & \jdeq [ \sigma ]\,t & \text{otherwise}
+\end{alignat*}
 
-\begin{claim*}\label{testenv-claim2}
-\end{claim*}
+\begin{alignat*}{5}
+  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
+  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \pi\eta         & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
+  [\idS]t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
+  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
+\end{alignat*}
 
-\begin{claimproof}
-\end{claimproof}
-\begin{lstlisting}[caption={Formal definition in \Agda},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
-for i:=maxint to 0 do 
-begin 
-    j:=square(root(i));
-end;
-\end{lstlisting}
-
+\section{Formal definitions in \Agda}
 \end{document}

### 2025-01-20 tex/main.tex
[ tex ] Progress on S3.1

@@
 \begin{abstract}
-  Type theory can be defined in a type theory as a quotient inductive-inductive type, but terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs needs to take account of coercions painstakingly alongside the interesting part.
+  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs need to take account of coercions painstakingly alongside the interesting part.
   This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
@@
 \cite{Altenkirch2017}
+\cite{Munoz1998}
+\cite{Dybjer1996,Castellan2021}
+\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
 
+Substitution calculus is closely linked to the categorical model of type theory \emph{categories with families}~\cite{Dybjer1996} 
 \paragraph*{Single substitution calculus}
@@
 \subsection{Parallel substitution calculus} \label{subsec:SC}
+We begin with recalling the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a quotient inductive-inductive-type, emphasising the use of transports in its definition, which will be developed into a more complicated type theory.
+This calculus is a fragment of type theory studied by Altenkirch and Kaposi~\cite{Altenkirch2016a}, which can be defined using the general schema~\cite{Kaposi2019} introduced by Kaposi\ et~al.
+
+Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types under some context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from the domain $\Gamma$ and the codomain $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms under a context $\Gamma$ and its type $A$.
+These types amount to the following types indexed by types being defined (hence inductive-inductive):
 \begin{alignat*}{3}
@@
 \end{alignat*}
+The type $\Ctx$ has two constructors $\emptyctx$ for the empty context and $\blank,\blank$ for context extension:
 \begin{alignat*}{3}
@@
 \end{alignat*}
+where the context extension $\blank,\blank$ requires that the type $\Ty$ indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.
+
+Type substitution takes a substitution $\sigma : \Sub\,\Gamma\,\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
 \begin{alignat*}{3}
-  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma,
+\end{alignat*}
+If we were defining cwfs, it would be enough to require type substitution for $\Ty$.
+However, we need a ground type $\UU$ in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
+otherwise the type $\Ty$ would be empty as well as other types:
+\begin{alignat*}{3}
+  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
 \end{alignat*}
 
-\begin{alignat*}{3}
-  \UU & : \implicit{\Gamma} & \Ty\, \Gamma
-\end{alignat*}
+Note that $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) are quantified implicitly above.
+For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.
+
+Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitutions $\sigma, t$ extended by some term~$t$:
 \begin{alignat*}{3}
   \emptysub & : \implicit{\Gamma}\;\Sub\,\Gamma\,\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]\, A) \to \Sub\,\Gamma,(\Delta, A),
+\end{alignat*}
+Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
+As substitution calculus is closely related to cwfs (more precisely, it is the initial model), substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
+\begin{alignat*}{3}
   \idS & : \implicit{\Gamma}\;\Sub\,\Gamma\,\Gamma \\
   \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
+\end{alignat*}
+satisfying certain laws introduced later.
+Context comprehension is given by projections from \emph{non-empty} substitutions $\Sub\;\Gamma\;(\Delta, A)$:
+\begin{alignat*}{3}
   \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
-  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A) \\
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A)
+\end{alignat*}
+where $\pi_1\,\sigma$ and $\pi_2\,\sigma$ can be intuitively understood as the \emph{tail} and the \emph{head} of a non-empty substitution $\sigma : \Sub\;\Gamma\;(\Delta, A)$ respectively.
+Again, a type substitution $[\pi_1\,\sigma]\,A$ is needed for $\pi_2$, because $A$ is under the context $\Delta$ instead of $\Gamma$.
+Finally, we also have term substitution $[\sigma]\,t$ whose type is $[\sigma]\,A$ for a term $t : \Tm\;\Delta\;A$
+\begin{alignat*}{3}
   [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A)
 \end{alignat*}
+where the symbol $[\blank]\blank$ is overloaded.
 
+We also have equality constructors to stipulate properties for type substitution:
 \begin{alignat*}{3}
@@
   [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\,A && =^{\Ty\,\Gamma} [ \sigma ]\;([ \tau ]\;A) \\
-  []U      & : \implicit{\Gamma, \Delta, \sigma}                  && [ \sigma ]\,U        && =^{\Ty\,\Gamma} U
+  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                  && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma} \UU
 \end{alignat*}
 
+Laws for substitutions are stipulated using equality constructors.
 \begin{alignat*}{5}
@@
   \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+\end{alignat*}
+In particular, $\blank;\blank$ acts like a list concatenation, so conceptually $\sigma; (\tau, t)$ is equal to $(\sigma; \tau), t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
+Moving the term $t$ from the left-hand side to the right-hand side requires us to apply substitution $\sigma$ to the term $t$ to obtain an inhabitant of $\Tm\;\Gamma\;[\sigma][\tau]\,A$.
+By definition of the substitution extension, however, $[\sigma]\;t$ requires to be $\Tm\;\Gamma\;[\sigma; \tau]\,A$ instead.
+Therefore, we need to transport $[\sigma]\;t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.
+
+The tail and the head of a non-empty substitution $(\sigma, t)$ is obviously $\sigma$ and $t$ respectively:
+\begin{alignat*}{5}
   \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
+\end{alignat*}
+whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
+Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.
+
+Similarly, for term substitution, terms are transported along the corresponding rules:
+\begin{alignat*}{5}
   [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}         && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
@@
 \end{alignat*}
+equivalent to the following (homogeneous) identities
+\[
+  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
+  \quad\text{and}\quad
+  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
+\]
+respectively.
+Constructors introduced so far complete the definition of substitution calculus.
 
+The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts.
 \begin{example}
@@
   where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$,
-\end{example}
 explicitly.
 The first three equations and the last equation follow from Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively, which cannot be proved naively as we did in \eqref{eq:pi2-comp-proof}.
+\end{example}
+
+To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we make type substitution \emph{definitional}, since the root cause is the type mismatch after type substitution.
+
+One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution is a recursion.
+Considering that $[ \idS ]_{\Ty}\,A \jdeq A$ and $[ \sigma ; \tau ]_{\Ty}\,A \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may attempt to define it by:
+\begin{alignat*}{3}
+[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+[ \sigma ]_{\Ty}\,\UU           & = \UU
+\end{alignat*}
+Now, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively.
+Still, they are not definitional but propositional, so this recursion does not eliminate any transport in the definition.
+
+Our second attempt is to define type substitution with additional rules:
+\begin{alignat*}{3}
+[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
+[ \idS ]_{\Ty}\,A             & = A \\
+[ \sigma ; \tau ]_{\Ty}\,A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)
+[ \sigma ]_{\Ty}\,\UU         & = \UU \\
+\end{alignat*}
+However, this definition will \emph{not} compute $[\sigma]_{\Ty}\,\UU$ in most type theories.
+\LT{briefly explain how this definition should be translated to an eliminator}
 
@@
 \end{alignat*}
-
+With this type substitution being definitional for each clause, we update other equality constructors without any $\transp$:
 \begin{alignat*}{5}
@@
 
+\LT{Revisit our previous example.}
 \begin{proposition}[Local confluence]

### 2025-01-21 tex/main.tex
Draft S3.1

@@
 
-\usepackage{todonotes}
+\usepackage[obeyFinal]{todonotes}
 \usepackage{manfnt}
 \newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
-\newcommand{\LT}[1]{\todo[inline,author={L-T},caption={}]{#1}}
+\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
 
@@
 
-Substitution calculus is closely linked to the categorical model of type theory \emph{categories with families}~\cite{Dybjer1996} 
+Substitution calculus is closely related to \emph{categories with families}~\cite{Dybjer1996}.
 \paragraph*{Single substitution calculus}
@@
 
-\subsection{Definitions by overlapping patterns}
+\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
 \cite{Cockx2014,Altenkirch2016a}
@@
 \paragraph*{Local confluence}
+\LT{Explain the notion of critical pair; local peak}
 \paragraph*{Strong normalisation}
+\cite{Abel2002}
+\LT{Explain how these two conditions imply global confluence}
 
@@
 }
-\subsection{Parallel substitution calculus} \label{subsec:SC}
+
 We begin with recalling the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a quotient inductive-inductive-type, emphasising the use of transports in its definition, which will be developed into a more complicated type theory.
 This calculus is a fragment of type theory studied by Altenkirch and Kaposi~\cite{Altenkirch2016a}, which can be defined using the general schema~\cite{Kaposi2019} introduced by Kaposi\ et~al.
+
+\subsection{Parallel substitution calculus} \label{subsec:SC}
 
@@
   \Sub   & : \Ctx \to \Ctx \to \Set \\
-  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set
+  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
 \end{alignat*}
@@
   \emptyctx & : \Ctx \\
-  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx
+  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
 \end{alignat*}
@@
 
-Type substitution takes a substitution $\sigma : \Sub\,\Gamma\,\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
+Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
 \begin{alignat*}{3}
-  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma,
+  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
 \end{alignat*}
@@
 
-Note that $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) are quantified implicitly above.
+Note that $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.
 For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.
@@
 \begin{alignat*}{3}
-  \emptysub & : \implicit{\Gamma}\;\Sub\,\Gamma\,\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]\, A) \to \Sub\,\Gamma,(\Delta, A),
+  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
 \end{alignat*}
-Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
+Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
 As substitution calculus is closely related to cwfs (more precisely, it is the initial model), substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
 \begin{alignat*}{3}
-  \idS & : \implicit{\Gamma}\;\Sub\,\Gamma\,\Gamma \\
-  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
+  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
+  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
-  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A)
+  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A)
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \,A         && =^{\Ty\,\Gamma} A \\
-  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\,A && =^{\Ty\,\Gamma} [ \sigma ]\;([ \tau ]\;A) \\
-  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                  && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma} \UU
+  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \;A         && =^{\Ty\,\Gamma}\;A \\
+  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\;A && =^{\Ty\,\Gamma} [ \sigma ]\;([ \tau ]\;A) \\
+  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                && [ \sigma ]\;\UU        && =^{\Ty\,\Gamma} \UU
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}         && =^{\Tm\,\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
-  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t} \\
+  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
+  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
 \end{alignat*}
@@
 The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts.
-\begin{example}
+\begin{example}\label{ex:pi2-comp}
   Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$.
@@
   \begin{alignat*}{3}
-         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \label{eq:pi2-proof-1} \\
+         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
     = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
-    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \\
-    = {} & ([\pi_1(\sigma;\pi_1\,\tau, t)]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, t)) \nonumber \\
-    = {} & (\sub{\sigma;\pi_1\tau}{A} &&, [\sigma] (\pi_2\,\tau)) \nonumber
+    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
+    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
+    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
   \end{alignat*}
-  where $p \judgeq [;]_{\Ty}^{-1} : \sub{\sigma}{\sub{\pi_1\,\tau}{A}} \to \sub{\sigma; \pi_1 \tau}{A}$ and $t \judgeq\transfib{\Tm\,\Gamma}{p}{[\sigma]\,(\pi_2\,\tau)}$,
-explicitly.
-The first three equations and the last equation follow from Lemma~2.3.4 and 2.3.2 in \cite{UFP2013} respectively, which cannot be proved naively as we did in \eqref{eq:pi2-comp-proof}.
+  The first three equations follow from Lemma~2.3.4in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
 \end{example}
@@
 
-One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution is a recursion.
-Considering that $[ \idS ]_{\Ty}\,A \jdeq A$ and $[ \sigma ; \tau ]_{\Ty}\,A \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may attempt to define it by:
+One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function should compute.
+Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
 \begin{alignat*}{3}
-[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-[ \sigma ]_{\Ty}\,\UU           & = \UU
+[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
+[ \sigma ]_{\Ty}\;\UU & = \UU
 \end{alignat*}
-Now, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively.
-Still, they are not definitional but propositional, so this recursion does not eliminate any transport in the definition.
+However, this definition is \emph{not} inductive-recursive in the style of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in the schema of Dybjer and Setzer, the codomain of a simultaneous function does not refer to any inductive type being defined.
+While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006,Chapman2009}.
+Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
+This definition fails to eliminate any transport in the definition of substitution calculus.
 
 Our second attempt is to define type substitution with additional rules:
+\begin{alignat}{3}
+[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
+[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
+[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
+[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
+\end{alignat}
+However, this definition does \emph{not} compute \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
+By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using first-match semantics~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
+\LT[noinline]{How about \Coq and \Lean?}
+As a result, the above definition amounts to the following definitional clause:
 \begin{alignat*}{3}
-[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-[ \idS ]_{\Ty}\,A             & = A \\
-[ \sigma ; \tau ]_{\Ty}\,A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)
-[ \sigma ]_{\Ty}\,\UU         & = \UU \\
+[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
+[ \sigma ]_{\Ty}\;\UU  & \judgeq \UU
 \end{alignat*}
-However, this definition will \emph{not} compute $[\sigma]_{\Ty}\,\UU$ in most type theories.
-\LT{briefly explain how this definition should be translated to an eliminator}
+where $\judgeq$ is used to emphasise the definitional equality: we have done nothing different from our first attempt! 
+%As a result, neither  are definitional.
+We may swap the order of function clauses to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
+\begin{alignat*}{3}
+[ \emptysub ]_{\Ty}\;\UU         & \judgeq \UU \\
+[ \sigma, t ]_{\Ty}\;\UU         & \judgeq \UU \\
+[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \judgeq \UU,
+\end{alignat*}
+making $[\sigma]\;\UU = \UU$ propositional instead.
+Not being able to compute $[ \sigma ]_{\Ty}\;\UU = \UU$ definitionally would make our later extension with a type $\El\,A$ of elements of $A$ require another transport in the definition.
+Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
+That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.
 
-\begin{alignat*}{3}
-  [\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-[ \idS ]_{\Ty}\,A             & \jdeq A \\
-[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
-[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
-[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
-[ \sigma ]_{\Ty}\,\UU           & \jdeq \UU
-\end{alignat*}
-With this type substitution being definitional for each clause, we update other equality constructors without any $\transp$:
-\begin{alignat*}{5}
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
-  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
-\end{alignat*}
-
-\LT{Revisit our previous example.}
+To ensure that type substitution is definitional for all \emph{structural rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional.
+In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
+We then end up with the following definition:
+\begin{alignat}{3}
+[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
+[ \idS ]_{\Ty}\;A                & \jdeq A \label{def:definitional-type-subst-1}\\
+[ \sigma ; \tau ]_{\Ty}\;A       & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{def:definitional-type-subst-2}\\
+[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \jdeq [\sigma]_\Ty\;A                    \label{def:definitional-type-subst-3} \\
+[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \jdeq [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A) \label{def:definitional-type-subst-4} \\
+[ \sigma ]_{\Ty}\;\UU            & \jdeq \UU                                \label{def:definitional-type-subst-5}
+\end{alignat}
+We first check that this definition of type substitution is valid.\footnote{%
+\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
 \begin{proposition}[Local confluence]
@@
 \end{proposition}
+\begin{proof}
+We consider each combination of \eqref{def:definitional-type-subst-5} with \eqref{def:definitional-type-subst-1}--\eqref{def:definitional-type-subst-4}, that is, 
+  \[
+    [\idS]_{\Ty}\;\UU, \quad [\sigma;\tau]_{\Ty}\;\UU, \quad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \quad\text{and}\quad [\pi_1(\sigma; \tau)]_{\Ty}\;\U
+  \]
+which are all local peaks, and show that each applicable clause does result in the same term.
+For example, $[\sigma;\tau]_{\Ty}\;\UU \judgeq [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \judgeq [\sigma]_{\Ty}\;\UU \judgeq \UU$ using \eqref{def:definitional-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \judgeq \UU$ using \eqref{def:definitional-type-subst-5} as the first reduction.
+It is easy to see that every local peak reduces to the same term $\UU$, so $[\blank]_{\Ty}\blank$ is locally confluent.
+
+\end{proof}
+\begin{proposition}[Termination]
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+  \danger
+\end{proposition}
+\begin{proof}
+  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
+
+\end{proof}
+
+With this definition of type substitution, other equality constructors can be introduced without any $\transp$, we list updated constructors as follows.
+\begin{alignat*}{5}
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t) && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), [ \sigma ]\;t \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
+  [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A}] && t \\
+  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
+\end{alignat*}
+
+\begin{example}
+  Revisiting \Cref{ex:pi2-comp}, we can write $\pi_2\;(\sigma; \tau) = [\sigma](\pi_2\;\tau)$ formally instead of \eqref{eq:pi2-comp-real-proof}, since the type $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ of the term on the LHS is definitionally equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
+  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are definitionally equal because of \eqref{def:definitional-type-subst-3} and \eqref{def:definitional-type-subst-4}.
+  Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
+\end{example}
+
+Yet, as we are defining a function on an inductive type with equality constructors, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.\LT{general condition?} 
+This boils down to that $[\blank]_{\Ty}\blank$ preserves equality constructors for quotient types.
 \begin{proposition}[Coherence]
@@
 \end{proposition}
+\begin{proof}
+  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
+  \begin{align*}
+    & [\sigma;\idS]_{\Ty} \;A = [\sigma]_{\Ty}\;A,
+    && [\idS;\sigma]_{\Ty} \;A = [\sigma]_{\Ty}\;A, && [(\sigma;\tau);\gamma]_{\Ty} \;A = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
+    & [\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A \\
+    & [\sigma; (\tau, t)]_{\Ty} \;A = [(\sigma;\tau), t]_{\Ty}\;A && [\sigma]_{\Ty} \;A = [\emptyctx]_{\Ty}\;A, && [\sigma]_{\Ty} \;A = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+  \end{align*}
+  hold (propositionally).
+  The first two identities of the first row hold definitionally by \eqref{def:definitional-type-subst-1} and \eqref{def:definitional-type-subst-2}, while terms of the both sides of the third compute to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{def:definitional-type-subst-2}.
+
+  The sole identity of the second row is just one, i.e.\ \eqref{def:definitional-type-subst-3}, of the clauses of the definition.
+
+  We show the remaining cases by induction on $A$.
+  However, $\UU$ is the only constructor for $\Ty\;\Gamma$, and each of them computes to $\UU$ definitionally.
+\end{proof}
 

### 2025-01-21 tex/main.tex
Polish some paragraphs in S3.1

@@
 \subsection{Plan of the paper}
-\subsection{Related work}
-\paragraph*{Formalisation of type theory in type theory}
-\cite{Danielsson2006,Altenkirch2016a}
-\cite{Altenkirch2017}
-\cite{Munoz1998}
-\cite{Dybjer1996,Castellan2021}
-\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
-
-Substitution calculus is closely related to \emph{categories with families}~\cite{Dybjer1996}.
-\paragraph*{Single substitution calculus}
-\cite{Kaposi2023,Kaposi2024a}
-\paragraph*{Schemata of inductive types}
-\cite{Kaposi2019}
 
@@
 \end{alignat*}
-If we were defining cwfs, it would be enough to require type substitution for $\Ty$.
-However, we need a ground type $\UU$ in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
-otherwise the type $\Ty$ would be empty as well as other types:
+where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.
+
+If we were defining cwfs, type substitution would be enough for $\Ty$.
+However, we need a base case in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
+otherwise the type $\Ty$ would be empty as well as other types.
+Hence another constructor is introduced:
 \begin{alignat*}{3}
@@
 \end{alignat*}
-
-Note that $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.
 For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.
@@
 
-The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts.
+The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for such a simple fact below.
 \begin{example}\label{ex:pi2-comp}
@@
     = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
-    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
-    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
+    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
+    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
     = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
   \end{alignat*}
-  The first three equations follow from Lemma~2.3.4in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
+  The first three equations follow from Lemma~2.3.4 in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
 \end{example}
@@
 \end{alignat*}
-However, this definition is \emph{not} inductive-recursive in the style of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in the schema of Dybjer and Setzer, the codomain of a simultaneous function does not refer to any inductive type being defined.
-While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006,Chapman2009}.
+However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneous recursive function does not refer to any inductive type being defined.
+While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
 Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
@@
     & [\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A \\
-    & [\sigma; (\tau, t)]_{\Ty} \;A = [(\sigma;\tau), t]_{\Ty}\;A && [\sigma]_{\Ty} \;A = [\emptyctx]_{\Ty}\;A, && [\sigma]_{\Ty} \;A = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+    & [\sigma; (\tau, t)]_{\Ty} \;A = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A && [\sigma]_{\Ty} \;A = [\emptyctx]_{\Ty}\;A, && [\sigma]_{\Ty} \;A = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
   \end{align*}
@@
   We show the remaining cases by induction on $A$.
-  However, $\UU$ is the only constructor for $\Ty\;\Gamma$, and each of them computes to $\UU$ definitionally.
+  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of identities with $A \defeq \UU$ computes to $\UU$ definitionally.
 \end{proof}
@@
 
-\section{Future work}
+\section{Discussion and future work}
+\subsection{Related work}
+\paragraph*{Formalisation of type theory in type theory}
+\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
+\cite{Altenkirch2017}
+\cite{Munoz1998}
+\cite{Dybjer1996,Castellan2021}
+\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
 
+Substitution calculus is closely related to \emph{categories with families}~\cite{Dybjer1996}.
+\paragraph*{Single substitution calculus}
+\cite{Kaposi2023,Kaposi2024a}
+\paragraph*{Schemata of inductive types}
+\cite{Kaposi2019}
+
+\subsection{Future work}
 \paragraph*{Type theory in (observational) type theory}
-
+\paragraph*{Formalising definitions by overlapping patterns}
 \paragraph*{A general schema of quotient inductive-inductive-recursive types}
-\subparagraph*{Formalising definitions by overlapping patterns}
 
@@
 
-\section{Formal definitions in \Agda}
+\section{Formal definitions in Agda}
 \end{document}

### 2025-01-22 tex/main.tex
Draft S3.2

@@
 
-\usepackage[obeyFinal]{todonotes}
+\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
 \usepackage{manfnt}
@@
   In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
-  As a case study, we investigate (parallel) substitution calculus and type theory with Coquand universes and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
+  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
 \end{abstract}
@@
       \item a definition of parallel substitution calculus;
-      \item a definition of type theory with Coquand universes and $\Pi$-types,
+      \item a definition of type theory with a universe and $\Pi$-types,
     \end{itemize}
@@
 \section{Metatheory and formalisation}
+
 \cite{UFP2013}
@@
 
+\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive types).}
 \subsection{Inductive-recursive types}
@@
 \cite{Cockx2014,Altenkirch2016a}
+\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
 \subsection{Quotient inductive(-inductive)-recursive types}
@@
 
-Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
+Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
 \begin{alignat*}{3}
@@
 
-If we were defining cwfs, type substitution would be enough for $\Ty$.
-However, we need a base case in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
-otherwise the type $\Ty$ would be empty as well as other types.
+If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
+For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
 Hence another constructor is introduced:
@@
 \end{alignat*}
-For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.
+For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.
 
-Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitutions $\sigma, t$ extended by some term~$t$:
+Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
 \begin{alignat*}{3}
@@
 Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
-As substitution calculus is closely related to cwfs (more precisely, it is the initial model), substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
+As substitution calculus is the initial cwf with a constant type $\UU$, substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
 \begin{alignat*}{3}
@@
 \end{alignat*}
-satisfying certain laws introduced later.
+satisfying certain laws (introduced later).
 Context comprehension is given by projections from \emph{non-empty} substitutions $\Sub\;\Gamma\;(\Delta, A)$:
@@
 where $\pi_1\,\sigma$ and $\pi_2\,\sigma$ can be intuitively understood as the \emph{tail} and the \emph{head} of a non-empty substitution $\sigma : \Sub\;\Gamma\;(\Delta, A)$ respectively.
-Again, a type substitution $[\pi_1\,\sigma]\,A$ is needed for $\pi_2$, because $A$ is under the context $\Delta$ instead of $\Gamma$.
+Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
 Finally, we also have term substitution $[\sigma]\,t$ whose type is $[\sigma]\,A$ for a term $t : \Tm\;\Delta\;A$
@@
 In particular, $\blank;\blank$ acts like a list concatenation, so conceptually $\sigma; (\tau, t)$ is equal to $(\sigma; \tau), t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
-Moving the term $t$ from the left-hand side to the right-hand side requires us to apply substitution $\sigma$ to the term $t$ to obtain an inhabitant of $\Tm\;\Gamma\;[\sigma][\tau]\,A$.
-By definition of the substitution extension, however, $[\sigma]\;t$ requires to be $\Tm\;\Gamma\;[\sigma; \tau]\,A$ instead.
+Moving the term $t$ from the left-hand side to the right-hand side requires us to apply substitution $\sigma$ to the term $t$ to obtain an inhabitant of $\Tm\;\Gamma\;([\sigma][\tau]\,A)$.
+By definition of the substitution extension, however, $[\sigma]\;t$ requires to be $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
 Therefore, we need to transport $[\sigma]\;t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.
@@
 \begin{alignat*}{5}
-  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
-  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
+  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
+  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
 \end{alignat*}
@@
 
-To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we make type substitution \emph{definitional}, since the root cause is the type mismatch after type substitution.
+To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor cannot compute so that type mismatches occur after substitution.
 
-One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function should compute.
+One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function computes substitution at the meta-level.
 Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
@@
 \end{alignat}
-However, this definition does \emph{not} compute \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
-By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using first-match semantics~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
+However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
+By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
 \LT[noinline]{How about \Coq and \Lean?}
-As a result, the above definition amounts to the following definitional clause:
+Using the first-match semantics, the above definition amounts to the following definition:
 \begin{alignat*}{3}
 [\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
-[ \sigma ]_{\Ty}\;\UU  & \judgeq \UU
+[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
 \end{alignat*}
-where $\judgeq$ is used to emphasise the definitional equality: we have done nothing different from our first attempt! 
+where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
 %As a result, neither  are definitional.
-We may swap the order of function clauses to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
+We may swap \eqref{eq:type-sub-at2-2} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to
 \begin{alignat*}{3}
-[ \emptysub ]_{\Ty}\;\UU         & \judgeq \UU \\
-[ \sigma, t ]_{\Ty}\;\UU         & \judgeq \UU \\
-[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \judgeq \UU,
+[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
+[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
+[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
 \end{alignat*}
 making $[\sigma]\;\UU = \UU$ propositional instead.
-Not being able to compute $[ \sigma ]_{\Ty}\;\UU = \UU$ definitionally would make our later extension with a type $\El\,A$ of elements of $A$ require another transport in the definition.
+Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
 Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
@@
 
-To ensure that type substitution is definitional for all \emph{structural rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional.
+To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
 In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
@@
 [\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
-[ \idS ]_{\Ty}\;A                & \jdeq A \label{def:definitional-type-subst-1}\\
-[ \sigma ; \tau ]_{\Ty}\;A       & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{def:definitional-type-subst-2}\\
-[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \jdeq [\sigma]_\Ty\;A                    \label{def:definitional-type-subst-3} \\
-[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \jdeq [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A) \label{def:definitional-type-subst-4} \\
-[ \sigma ]_{\Ty}\;\UU            & \jdeq \UU                                \label{def:definitional-type-subst-5}
+[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
+[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
+[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
+[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
+[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
 \end{alignat}
@@
 \Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
-\begin{proposition}[Local confluence]
+\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
   Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
@@
 \begin{proof}
-We consider each combination of \eqref{def:definitional-type-subst-5} with \eqref{def:definitional-type-subst-1}--\eqref{def:definitional-type-subst-4}, that is, 
+  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
   \[
-    [\idS]_{\Ty}\;\UU, \quad [\sigma;\tau]_{\Ty}\;\UU, \quad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \quad\text{and}\quad [\pi_1(\sigma; \tau)]_{\Ty}\;\U
+    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
   \]
-which are all local peaks, and show that each applicable clause does result in the same term.
-For example, $[\sigma;\tau]_{\Ty}\;\UU \judgeq [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \judgeq [\sigma]_{\Ty}\;\UU \judgeq \UU$ using \eqref{def:definitional-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \judgeq \UU$ using \eqref{def:definitional-type-subst-5} as the first reduction.
-It is easy to see that every local peak reduces to the same term $\UU$, so $[\blank]_{\Ty}\blank$ is locally confluent.
+  and show that each applicable clause does result in the same term $\UU$.
+
+  For example, $[\sigma;\tau]_{\Ty}\;\UU \reduce [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \reduce [\sigma]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
+  It is easy to see that remaining local peaks reduce to the same term $\UU$.
 
@@
   \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A}] && t \\
+  [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
   [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
@@
 
+We revisit \Cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
 \begin{example}
-  Revisiting \Cref{ex:pi2-comp}, we can write $\pi_2\;(\sigma; \tau) = [\sigma](\pi_2\;\tau)$ formally instead of \eqref{eq:pi2-comp-real-proof}, since the type $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ of the term on the LHS is definitionally equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
-  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are definitionally equal because of \eqref{def:definitional-type-subst-3} and \eqref{def:definitional-type-subst-4}.
+  The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
+  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ and is definitionally equal to $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$, i.e.\ the type of the term on the RHS.
+  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are definitionally equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
   Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
@@
 This boils down to that $[\blank]_{\Ty}\blank$ preserves equality constructors for quotient types.
-\begin{proposition}[Coherence]
+\begin{proposition}[Coherence]\label{prop:coherence-1}
   $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
@@
   \begin{align*}
-    & [\sigma;\idS]_{\Ty} \;A = [\sigma]_{\Ty}\;A,
-    && [\idS;\sigma]_{\Ty} \;A = [\sigma]_{\Ty}\;A, && [(\sigma;\tau);\gamma]_{\Ty} \;A = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
-    & [\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A \\
-    & [\sigma; (\tau, t)]_{\Ty} \;A = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A && [\sigma]_{\Ty} \;A = [\emptyctx]_{\Ty}\;A, && [\sigma]_{\Ty} \;A = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
+                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
+                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
+    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
+    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
   \end{align*}
   hold (propositionally).
-  The first two identities of the first row hold definitionally by \eqref{def:definitional-type-subst-1} and \eqref{def:definitional-type-subst-2}, while terms of the both sides of the third compute to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{def:definitional-type-subst-2}.
+  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.
 
-  The sole identity of the second row is just one, i.e.\ \eqref{def:definitional-type-subst-3}, of the clauses of the definition.
+  The sole identity of the second row is just one, i.e.\ \eqref{eq:def-type-subst-3}, of the clauses of the definition.
 
   We show the remaining cases by induction on $A$.
-  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of identities with $A \defeq \UU$ computes to $\UU$ definitionally.
+  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of identities with $A \defeq \UU$ reduce to $\UU$.
+\end{proof}
+\begin{remark}\label{re:coherence-proof}
+  Note that in the above proof the first four identities hold definitionally by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
+  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how congruence rules for type substitution is defined.
+\end{remark}
+
+\LT{We may wonder if the last three identities can be made definitional.
+Can we find a complete set of rules?}
+
+\subsection{... with an empty universe} \label{subsec:SC+U}
+\LT[noinline]{Shall we go for Coquand universes? \cite{Coquand2013}}
+
+In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
+For a QII definition \cite{Altenkirch2016a}, this extension amounts to adding the following constructors 
+\begin{alignat*}{3}
+  \UU   & : \implicit{\Gamma} && \Ty\, \Gamma \\
+  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
+  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}})
+\end{alignat*}
+where the type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.
+
+In our QIIR definition, the transported term $\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}}$ is no longer needed, since $[]\UU$ is definitional. 
+Therefore, we may replace the equality constructor $[]\El$ by a clause for type substitution:
+\begin{alignat}{3}
+  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u}) \label{eq:def-type-subst-6}
+\end{alignat}
+Yet, this naive clause breaks the local confluence!
+Observe that the local peak
+\[
+  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u)
+\]
+cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.
+
+To repair the local confluence, we add a simultaneous recursive function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
+\begin{alignat*}{3}
+  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
+[ \idS ]_{\Tm}\,t          & \reduce t \\
+[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
+[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
+[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
+[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
+\end{alignat*}
+which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can propagate from type to term substitution.
+The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
+Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
+
+Then, \eqref{eq:def-type-subst-6} and the rule $\mathsf{concat}$ are accordingly changed to 
+\begin{align}
+  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}\\
+  \mathsf{concat} : \sigma ; (\tau , t) & =^{\Sub\,\Gamma\,(\Theta, A)} (\sigma ; \tau) , [ \sigma ]_{\Tm} t \nonumber
+\end{align}
+to retain the local confluence for type substitution:
+\begin{proposition}[Local confluence]
+  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+  \danger
+\end{proposition}
+\begin{proof}
+  Again, in addition to those in the proof of \Cref{prop:local-confluence-1}, consider following terms 
+  \[
+    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
+  \]
+  which reduce to the following terms
+  \[
+    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
+    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
+  \]
+  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
+  Therefore, type substitution is locally confluent.
+\end{proof}
+\begin{proposition}[Termination]
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+  \danger
+\end{proposition}
+\begin{proof}
+  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
 \end{proof}
 
-\subsection{... with a Coquand universe} \label{subsec:SC+U}
-\cite{Coquand2013}
-\begin{alignat*}{3}
-  \UU  & : \implicit{\Gamma}\;\Ty\, \Gamma \\
-  \El & : \implicit{\Gamma}\;\Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
-  [ \sigma ]_{\Ty}\,(\El\, u) & \jdeq \El\,([\sigma]_{\Tm}{u}) \\
-\end{alignat*}
-\begin{alignat*}{3}
-  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
-[ \idS ]_{\Tm}\,t          & \jdeq t \\
-[ \sigma ; \tau ]_{\Tm}\,t & \jdeq [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
-[ \pi_1(\sigma, t) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\,t \\
-[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
-[ \sigma ]_{\Tm}\,t        & \jdeq [ \sigma ]\,t, \quad \text{otherwise}  \\
-\end{alignat*}
+However, we also have to show that the recursive term substitution $[\blank]_{\Tm}\blank$ is propositionally equal to the explicit substitution $[\blank]\blank$, so they can be used together consistently.
 
-\begin{proposition}[Local confluence]
-  Type and term substitutions $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are locally confluent.
+\begin{proposition}\label{prop:correctness}
+  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
   \danger
 \end{proposition}
+\begin{proof}
+  We show this statement by induction on $\sigma$.
+  \begin{enumerate}
+    \item For the identity substitution $\idS$, we have $[\idS]t : [\idS]t = t$ but $t \judgeq [\idS]_{\Tm} t$ by definition.
+    \item For a substitution composite $\sigma; \tau$, we have
+      \begin{align*}
+        [\sigma;\tau]t & = [\sigma]\,[\tau] t               && \text{by $[;]t$} \\
+                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis $[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
+                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
+      \end{align*}
+    \item For $\pi_1(\sigma, t)$, we have
+      \begin{align*}
+        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
+                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
+      \end{align*}
+    \item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
+    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ definitionally.
+  \end{enumerate}
+  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
+  \LT[noinline]{Shall we say anything about equality constructors?}
+\end{proof}
 
-\begin{proposition}[Coherence]
+Finally, we have to check that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent with respect to equality constructors.
+\begin{proposition}[Coherence] \label{prop:coherence-2}
   The following statements hold:
@@
 \end{proposition}
+\begin{proof}
+  By \Cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
+  \begin{align*}
+    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+  \end{align*}
+  while other cases hold definitionally.
+  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
+  However, all cases follow easily from \Cref{prop:correctness}.
+  For example, consider the first case:
+  \begin{align*}
+    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by definition \eqref{eq:def-type-subst-7}} \\
+                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness}} \\
+                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
+                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness}} \\
+                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by definition \eqref{eq:def-type-subst-7} }
+  \end{align*}
+  The remaining two cases are omitted.
+
+  As for term substitution $[\blank]_{\Tm}\blank$, each case follows from \cref{prop:correctness} in conjunction with the congruence rule for the identity type.
+\end{proof}
+
+We have seen that the extension of substitution calculus with an (empty) universe breaks the local confluence of type substitution defined in \cref{subsec:SC}, because we have to consider the interaction of the newly added function clause with other existing clauses.
+In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist.
+We will see the same pattern in the next section. 
 \subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
+
 \begin{alignat*}{3}
@@
   \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
-\idS                \uparrow A & \jdeq \idS \\
-\sigma ; \tau       \uparrow A & \jdeq (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
-\pi_1(\sigma, t)    \uparrow A & \jdeq \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A & \jdeq \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
-\sigma              \uparrow A & \jdeq \sigma^+ , \quad \text{otherwiese} \\
-  [ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \jdeq \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
+\idS                \uparrow A & \reduce \idS \\
+\sigma ; \tau       \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
+\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \\
+\pi_1(\sigma; \tau) \uparrow A & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
+\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwiese} \\
+  [ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
 \end{alignat*}
@@
 %% t $$ u = [ ⟨ u ⟩ ]t app t
+\begin{proposition}[Local confluence]
+  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+  \danger
+\end{proposition}
 \begin{alignat*}{3}
@@
 \begin{alignat*}{5}
-[ \idS ]_{\Ty}\,A             & \jdeq A \\
-[ \sigma ; \tau ]_{\Ty}\,A    & \jdeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
-[ \pi_1(\sigma, t) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty\,A \\
-[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \jdeq [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
-[ \sigma ]_{\Ty}\,\UU           & \jdeq \UU \\
-[ \sigma ]_{\Ty}\,(\El\, u) & \jdeq \El\,([\sigma]_{\Tm}{u}) \\
-[ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \jdeq \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
-\idS                \uparrow A & \jdeq \idS \\
-\sigma ; \tau       \uparrow A & \jdeq (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
-\pi_1(\sigma, t)    \uparrow A & \jdeq \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A & \jdeq \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
-\sigma              \uparrow A & \jdeq (\pi_1 \idS ; \sigma , \pi_2 \idS) & \text{otherwiese} \\
- [ \idS ]_{\Tm}\,t                & \jdeq t \\
- [ \sigma ; \tau ]_{\Tm}\,t       & \jdeq [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
- [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \jdeq [\sigma]_\Tm\,t \\
- [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \jdeq [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
- [ \sigma ]_{\Tm}\,t              & \jdeq [ \sigma ]\,t & \text{otherwise}
+[ \idS ]_{\Ty}\,A             & \reduce A \\
+[ \sigma ; \tau ]_{\Ty}\,A    & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
+[ \pi_1(\sigma, t) ]_{\Ty}\,A & \reduce [\sigma]_\Ty\,A \\
+[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \reduce [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
+[ \sigma ]_{\Ty}\,\UU           & \reduce \UU \\
+[ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm}{u}) \\
+[ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
+\idS                \uparrow A & \reduce \idS \\
+\sigma ; \tau       \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
+\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \\
+\pi_1(\sigma; \tau) \uparrow A & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
+\sigma              \uparrow A & \reduce (\pi_1 \idS ; \sigma , \pi_2 \idS) & \text{otherwiese} \\
+ [ \idS ]_{\Tm}\,t                & \reduce t \\
+ [ \sigma ; \tau ]_{\Tm}\,t       & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
+ [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \reduce [\sigma]_\Tm\,t \\
+ [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
+ [ \sigma ]_{\Tm}\,t              & \reduce [ \sigma ]\,t & \text{otherwise}
 \end{alignat*}

### 2025-01-23 tex/main.tex
Draft S3.4 (was S3.3)

@@
 
-We begin with recalling the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a quotient inductive-inductive-type, emphasising the use of transports in its definition, which will be developed into a more complicated type theory.
-This calculus is a fragment of type theory studied by Altenkirch and Kaposi~\cite{Altenkirch2016a}, which can be defined using the general schema~\cite{Kaposi2019} introduced by Kaposi\ et~al.
+We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a quotient inductive-inductive type, emphasising how the use of transports in its very definition complicates formal reasoning.
+Then we introduce its quotient inductive-inductive-recursive counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning and we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
 
-\subsection{Parallel substitution calculus} \label{subsec:SC}
+\subsection{Substitution calculus as a quotient inductive-inductive type} \label{subsec:SC-QIIT}
 
@@
   \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
-  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
+  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
 \end{alignat*}
@@
 
-We also have equality constructors to stipulate properties for type substitution:
+The structural rules for type substitution are stipulated by following equality constructors:
 \begin{alignat*}{3}
@@
 
-Laws for substitutions are stipulated using equality constructors.
+The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the length of terms in $\sigma$, giving rise two $\eta$-laws in the case that $\Delta$ is empty and non-empty:
+\begin{alignat*}{5}
+  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma        && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma)
+\end{alignat*}
+
+The laws for substitution composition are stipulated using equality constructors.
 \begin{alignat*}{5}
@@
   ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}} \\
-  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
-  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
 \end{alignat*}
-In particular, $\blank;\blank$ acts like a list concatenation, so conceptually $\sigma; (\tau, t)$ is equal to $(\sigma; \tau), t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
-Moving the term $t$ from the left-hand side to the right-hand side requires us to apply substitution $\sigma$ to the term $t$ to obtain an inhabitant of $\Tm\;\Gamma\;([\sigma][\tau]\,A)$.
-By definition of the substitution extension, however, $[\sigma]\;t$ requires to be $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
-Therefore, we need to transport $[\sigma]\;t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.
+Substitution composition $\blank;\blank$ acts like a list concatenation, so $\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$.
+However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.
 
@@
 
-The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for such a simple fact below.
+The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
 \begin{example}\label{ex:pi2-comp}
@@
 
+\subsection{Substitution calculus as a quotient inductive-inductive-recursive type} \label{subsec:SC-QIIRT}
+
 To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor cannot compute so that type mismatches occur after substitution.
 
-One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function computes substitution at the meta-level.
+One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneously defined function computes.
 Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
@@
 \end{alignat*}
-However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneous recursive function does not refer to any inductive type being defined.
+However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.
 While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
@@
 %As a result, neither  are definitional.
-We may swap \eqref{eq:type-sub-at2-2} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to
+We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to
 \begin{alignat*}{3}
@@
 \end{alignat}
-We first check that this definition of type substitution is valid.\footnote{%
+We first check that this definition of type substitution is confluent.\footnote{%
 \Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
@@
   \]
-  and show that each applicable clause does result in the same term $\UU$.
+  and show that each applicable clause does result in $\UU$.
 
   For example, $[\sigma;\tau]_{\Ty}\;\UU \reduce [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \reduce [\sigma]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
-  It is easy to see that remaining local peaks reduce to the same term $\UU$.
-
+  It is easy to see that remaining local peaks reduce to the same term.
 \end{proof}
@@
 
-In our QIIR definition, the transported term $\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}}$ is no longer needed, since $[]\UU$ is definitional. 
-Therefore, we may replace the equality constructor $[]\El$ by a clause for type substitution:
+In our QIIR definition, the transported term $\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}}$ is no longer needed, since the equality constructor $[]\UU$ becomes definitional. 
+Therefore, we may replace the equality constructor $[]\El$ by a clause of type substitution:
 \begin{alignat}{3}
@@
 \end{alignat}
-Yet, this naive clause breaks the local confluence!
+Yet, this naive change breaks the local confluence!
 Observe that the local peak
@@
 
-To repair the local confluence, we add a simultaneous recursive function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
+To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
 \begin{alignat*}{3}
@@
 to retain the local confluence for type substitution:
-\begin{proposition}[Local confluence]
+\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
   Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
@@
 \begin{proof}
-  Again, in addition to those in the proof of \Cref{prop:local-confluence-1}, consider following terms 
+  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider following terms 
   \[
@@
 
-\begin{proposition}\label{prop:correctness}
+\begin{proposition}\label{prop:correctness-1}
   For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
@@
   Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
-  However, all cases follow easily from \Cref{prop:correctness}.
+  However, all cases follow easily from \Cref{prop:correctness-1}.
   For example, consider the first case:
@@
     [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by definition \eqref{eq:def-type-subst-7}} \\
-                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness}} \\
+                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
                                       & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
-                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness}} \\
+                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
                                       & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by definition \eqref{eq:def-type-subst-7} }
@@
 
-  As for term substitution $[\blank]_{\Tm}\blank$, each case follows from \cref{prop:correctness} in conjunction with the congruence rule for the identity type.
+  As for term substitution $[\blank]_{\Tm}\blank$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
 \end{proof}
 
-We have seen that the extension of substitution calculus with an (empty) universe breaks the local confluence of type substitution defined in \cref{subsec:SC}, because we have to consider the interaction of the newly added function clause with other existing clauses.
-In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist.
+We have seen that the extension of substitution calculus with an (empty) universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
+In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this desired propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
 We will see the same pattern in the next section.
+
 \subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
 
+In this section, we extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
+For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to add the following constructors for the type formation rule, the term introduction/elimination rules:
 \begin{alignat*}{3}
   \Pi     &: \implicit{\Gamma}            && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
-  \lambda &: \implicit{\Gamma, A, B}      && \Tm\,(\Gamma, A)\,B \to \Tm\,\Gamma\,(\Pi\,A\,B) \\
-  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\,\Gamma\,(\Pi\,A\,B) \to \Tm\,(\Gamma, A)\,B \\
+  \lambda &: \implicit{\Gamma, A, B}      && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
+  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
 \end{alignat*}
-\begin{alignat*}{3}
-  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
-  \sigma^+ & \defeq (\pi_1\idS ; \sigma, \pi_2\idS) \\
+the computation rules
+\begin{alignat*}{5}
+  \Pi\beta           & : \implicit{\Gamma, A, B, t} && \mathsf{app}\,(\lambda\,t) && =^{\Tm\,\Gamma,A} && t \\
+  \Pi\eta            & : \implicit{\Gamma, A, B, t} && t                        && =^{\Tm\,\Gamma,A} && \lambda(\mathsf{app}\,t)
 \end{alignat*}
-\begin{alignat*}{3}
-  \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \\
-\idS                \uparrow A & \reduce \idS \\
-\sigma ; \tau       \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
-\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
-\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwiese} \\
-  [ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
+and the structural rules for substitution
+\begin{alignat}{5}
+  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
+  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t} && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{[]\Pi} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
+\end{alignat}
+where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \transfib{\Tm\;\Gamma}{[;]_T}{\pi_2\idS})$.
+
+For our QIIR definition, $\sigma\uparrow A$ need not a transport since $[;]_T$ is definitional.
+However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck on the lifting $\sigma \uparrow A$.
+For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $[\idS \uparrow A]_{\Ty}B$ is not definitionally equal to $B$, so the local peak $(\Pi\;A\;B) \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined by a confluent term.
+
+To retain the local confluence, we turn $\sigma \uparrow A$ into a simultaneously defined function:
+\begin{alignat}{3}
+  \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
+\idS                \uparrow A & \reduce \idS \nonumber \\
+(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
+\pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
+\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow (\sub{\pi_1 \tau}{A})) ; (\pi_1 \tau \uparrow A)\nonumber \\
+\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwiese} \nonumber
+\end{alignat}
+where $\sigma^+ \defeq (\pi_1\idS ; \sigma, \pi_2\idS)$.
+Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
+\begin{alignat}{5}
+                     & [ \sigma ]_{\Ty}\,(\Pi\,A\,B) && \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \label{eq:def-type-subst-9} \\
+  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \nonumber
+\end{alignat}
+
+\begin{proposition}[Local confluence]\label{prop:local-confluence-3}
+  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+  \danger
+\end{proposition}
+\begin{proof}
+  In addition to previous cases in \cref{prop:local-confluence-1,prop:local-confluence-2}, consider
+  \begin{align*}
+    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
+  \end{align*}
+  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
+  \begin{align*}
+    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}[\tau \uparrow A ]_{\Ty}\,B) \\
+    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}[\pi_1\tau \uparrow A]_{\Ty}\,B)
+  \end{align*}
+  respectively.
+  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+\end{proof}
+\begin{proposition}[Termination]
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+  \danger
+\end{proposition}
+
+Similar to term substitution, we can show that the lifted substitution $\sigma \uparrow A$ is propositionally equal to $\sigma^+$, based on two basic properties\footnote{%
+  The reader is invited to prove them using the QII definition of type theory.}
+about the lifting.
+\begin{lemma} \label{lem:lifting}
+  The following statements hold:
+  \begin{enumerate}
+    \item $\idS\;\{\Gamma, A\} = (\idS\;\{\Gamma\})^+$ and
+    \item $(\sigma; \tau)^+ = \sigma^+ ; \tau^+$.
+  \end{enumerate}
+\end{lemma}
+
+\begin{proposition} \label{prop:correctness-2}
+  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$
+  \danger
+  $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
+\end{proposition}
+\begin{proof}
+  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
+  Every case follows from equational reasoning straightforwardly.
+  For example, if $\sigma = \sigma;\tau$ is a composite, we have
+  \begin{align*}
+    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
+                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
+                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
+  \end{align*}
+  Other cases follow similarly.
+\end{proof}
+
+As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the propositional equality $[\sigma]\,t = [\sigma]_{\Tm}\,t$ but the same argument still applies:
+\begin{proposition} \label{prop:correctness-3}
+  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$,
+  \danger
+  the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
+\end{proposition}
+
+Finally, we show that these simultaneously defined functions are coherent.
+\begin{proposition}[Coherence]
+  The following identities
+  \danger
+  \[
+    \sigma \uparrow A = \tau \uparrow A
+    \qquad\text{and}\qquad
+    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t,
+  \]
+  hold if $\sigma = \tau$ for any substitutions $\sigma$ and $\tau$.
+\end{proposition}
+\begin{proof}
+  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily.
+\end{proof}
+
+\subsubsection{Coherence of type substitution}
+For the coherence of~$[\blank]_{\Ty}\blank$, we have to generalise the desired identity to account for the lifting, since a naive induction does not work: for example, unfolding $[\sigma]_{\Ty}\,(\Pi\;A\;B)$ gives us
+\[
+  \Pi\;([\sigma]_{\Ty}\,A)\;([\sigma\uparrow A]_{\Ty}\,B)
+\]
+where the induction hypothesis $[\sigma]_{\Ty}\,B = [\emptysub]_{\Ty}\,B$ cannot be applied.
+Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
+Then, we show the identity
+$[\sigma \mathop{\upuparrows} \Xi ]_{\Ty}\,A = [\tau \mathop{\upuparrows} \Xi ]_{\Ty}\,A$
+for arbitrary $\Xi$, subsuming the special case $[\sigma ]_{\Ty}\,A = [\tau ]_{\Ty}\,A$ for $\Xi$ being empty.
+
+First, define an inductive-recursive type of \emph{telescopes} under a context $\Gamma$ simultaneously with a concatenation $\blank \dplus \blank$ as follows.
+\begin{alignat*}{5}
+  \Tel               & :                   && \Ctx \to \Set                                     \\
+  \blank\dplus\blank & :                   && (\Gamma : \Ctx) \to (\Xi : \Tel\;\Gamma) \to \Ctx \\
+  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
+  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) \to (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma \\
+  \Gamma \dplus \emptytel & = && \Gamma \\
+  \Gamma \dplus (\Xi, A)  & = && (\Gamma \dplus \Xi) , A
 \end{alignat*}
 
-\begin{alignat*}{3}
-  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \\
-  \Pi\beta & : \mathsf{app}(\lambda\,t) && =^{\Tm\,\Gamma,A} {} && t \\
-  \Pi\eta & :  t && =^{\Tm\,\Gamma,A} {} && \lambda(\mathsf{app}\,t)
+Then, we define the multiple lifting $\blank \upuparrows \blank$ which generalises the single lifting $\blank\uparrow\blank$.
+Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$,
+\[
+\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
+\to \Sub\;(\Gamma \dplus ([\sigma]_{\Tel}\,\Xi))\;(\Delta \dplus \Xi)
+\]
+requires telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$ defined simultaneously.
+Hence, we arrive definitions for telescope substitution and the multiple lifting by a telescope:
+\begin{alignat*}{5}
+  & [ \sigma ]_{\Tel}\,\emptytel && = \emptytel \\
+  & [ \sigma ]_{\Tel}\,(\Xi, A)  && = [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
+  & \sigma \upuparrows \emptytel && = \sigma \\
+  & \sigma \upuparrows (\Xi, A)  && = (\sigma \upuparrows \Xi), A
 \end{alignat*}
+
+Now, we have to prove the coherence property for telescope and type substitution together.
+\begin{proposition}
+  The identities
+  \danger
+  \[
+    [\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi
+    \quad\text{and}\quad
+    [\sigma \upuparrows \Xi ]_{\Ty}\,A = [\tau \upuparrows \Xi ]_{\Ty}\,A
+  \]
+  hold if $\sigma = \tau$ for any substitutions $\sigma$ and $\tau$, and a telescope $\Xi$.
+\end{proposition}
+\begin{proof}
+  We prove these two identities together by induction on the telescope $\Xi$ for the first identity and the type $A$ for the second.
+
+  \LT{
+  By \cref{re:coherence-proof}, we consider the following case
+  \begin{align*}
+    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+  \end{align*}
+  We only consider the case $A = \Pi\;A\;B$ in addition to those considered in \cref{prop:coherence-2}.
+  }
+\end{proof}
 
@@
 %% t $$ u = [ ⟨ u ⟩ ]t app t
-\begin{proposition}[Local confluence]
-  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
-  \danger
-\end{proposition}
+
+We conclude this extension with $\Pi$-types by noting that single term substitution
 \begin{alignat*}{3}
-  \left<\blank\right> & : \Tm\,\Gamma\,A \to \Sub\,\Gamma\,(\Gamma, A) \\
-  \left< t \right> & \defeq (\idS , t) \\
-  \blank \$\blank & : \Tm\,\Gamma\,(\Pi\,A\,B) \to (u : \Tm\,\Gamma\,A) \to \Tm\,\Gamma\,([ \left< u \right> ]\,B) \\
-  t \mathop{\$} u & \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)
+  \left<\blank\right> & : \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
+  \left< t \right> & \defeq (\idS , t)
 \end{alignat*}
+can be introduced without any transport, which would be $(\idS, \transfib{\Tm\;\Gamma}{[id]_T^{-1}}{t})$ in its QII definition, and which can used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$.
+Similarly, the structural rule for substitution on $\mathsf{app}$ can be derived
+\begin{align*}
+  [\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
+                                              & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
+                                              & = \mathsf{app}\,([\sigma]\,t) && \text{by $\Pi\eta^{-1}$.}
+\end{align*}
+without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become definitional.
 
-\begin{proposition}
-  For every substitution $\sigma : \Sub\,\Gamma\,\Delta$, type $A : \Ty\,\Delta$, and term $t : \Tm\,\Delta\,A$, the following statements hold:
-  \danger
-  \[
-    \sigma \uparrow A = (\pi_1 \idS ; \sigma , \pi_2 \idS)
-  \]
-  and
-  \[
-    [ \sigma ]_{\Tm}\,t = [ \sigma ]\,t
-  \]
-\end{proposition}
-
-\begin{proposition}
-  The following statements hold:
-  \danger
-  \begin{enumerate}
-    \item $\sigma \uparrow A = \tau \uparrow A$ whenever $\sigma = \tau$.
-    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
-    \item $[\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t$ whenever $\sigma = \tau$.
-  \end{enumerate}
-\end{proposition}
 \subsection{... and other type formers} \label{subsec:SC+U+Pi+more}

### 2025-01-23 tex/main.tex
Fix the coherence statement in S3.4

@@
 
-One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneously defined function computes.
-Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
+One possibility is to define substitution calculus as a QII\emph{R} type where type substitution, as a simultaneously defined recursion, does compute.
+Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
 \begin{alignat*}{3}
@@
 \end{alignat*}
-However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.
-While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
-Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
+Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
+  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
+Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
 This definition fails to eliminate any transport in the definition of substitution calculus.
@@
 \end{alignat}
-However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
+However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
 By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
@@
 In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
-We then end up with the following definition:
+We end up with the following definition for type substitution:
 \begin{alignat}{3}
@@
 
-Yet, as we are defining a function on an inductive type with equality constructors, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.\LT{general condition?} 
-This boils down to that $[\blank]_{\Ty}\blank$ preserves equality constructors for quotient types.
+Yet, as we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.
+This boils down to that $[\blank]_{\Ty}\blank$ preserves all equality constructors.
+\LT{Should we mention general condition for coherence?} 
 \begin{proposition}[Coherence]\label{prop:coherence-1}
-  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
+  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
   \danger
@@
   \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
-  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}})
+  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
 \end{alignat*}
@@
 
-In our QIIR definition, the transported term $\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}}$ is no longer needed, since the equality constructor $[]\UU$ becomes definitional. 
-Therefore, we may replace the equality constructor $[]\El$ by a clause of type substitution:
+In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since the equality constructor $[]\UU$ becomes definitional. 
+Therefore, the equality constructor $[]\El$ may be replaced by a clause of type substitution:
 \begin{alignat}{3}
@@
 \begin{proof}
-  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider following terms 
+  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider terms 
   \[
@@
   For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
-  \danger
 \end{proposition}
@@
 \begin{proposition}[Coherence] \label{prop:coherence-2}
-  The following statements hold:
+  For every equality constructor $p : \sigma = \tau$, the following identities
   \danger
-  \begin{enumerate}
-    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
-    \item $[\sigma]_{\Tm}\,A = [\tau]_{\Tm}\,A$ whenever $\sigma = \tau$.
-  \end{enumerate}
+  \[
+    [\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A
+    \quad\text{and}\quad
+    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t
+  \]
+  for any type $A$ and any term $t$.
 \end{proposition}
@@
 
-We have seen that the extension of substitution calculus with an (empty) universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
+We have seen that a naive extension of substitution calculus with a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
 In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this desired propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
@@
   []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
-  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t} && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{[]\Pi} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
+  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t} && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
 \end{alignat}
-where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \transfib{\Tm\;\Gamma}{[;]_T}{\pi_2\idS})$.
+where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.
 
@@
   For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$
-  \danger
   $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
@@
   For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$,
-  \danger
   the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
@@
 \begin{proposition}[Coherence]
-  The following identities
+  For every equality constructor $p : \sigma = \tau$, the following identities
   \danger
@@
   \]
-  hold if $\sigma = \tau$ for any substitutions $\sigma$ and $\tau$.
+  hold for any type $A$ and any term $t$.
 \end{proposition}
@@
 
-\subsubsection{Coherence of type substitution}
-For the coherence of~$[\blank]_{\Ty}\blank$, we have to generalise the desired identity to account for the lifting, since a naive induction does not work: for example, unfolding $[\sigma]_{\Ty}\,(\Pi\;A\;B)$ gives us
+\subsubsection{Multiple lifting and the coherence of type substitution}
+For the coherence of~$[\blank]_{\Ty}\blank$, we have to generalise the desired identity to account for the lifting, since a naive induction does not work: for example, unfolding $[\sigma]_{\Ty}\,(\Pi\;A\;B)$ leads to
 \[
@@
 Then, we define the multiple lifting $\blank \upuparrows \blank$ which generalises the single lifting $\blank\uparrow\blank$.
-Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$,
+Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$, should be
 \[
@@
 \]
-requires telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$ defined simultaneously.
-Hence, we arrive definitions for telescope substitution and the multiple lifting by a telescope:
+requiring telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
+Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-
-Now, we have to prove the coherence property for telescope and type substitution together.
+Now, we have to prove the coherence property for the multiple lifting, telescope substitution, and type substitution altogether.
 \begin{proposition}
-  The identities
+  For every equality constructor $p : \sigma = \tau$, the identities
   \danger
   \[
-    [\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi
+    \sigma \upuparrows \Xi = \tau \upuparrows \Xi, \quad
+    [\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi,
     \quad\text{and}\quad
@@
   \]
-  hold if $\sigma = \tau$ for any substitutions $\sigma$ and $\tau$, and a telescope $\Xi$.
+  hold for any telescope $\Xi$ and type $A$.
 \end{proposition}
@@
 \end{alignat*}
-can be introduced without any transport, which would be $(\idS, \transfib{\Tm\;\Gamma}{[id]_T^{-1}}{t})$ in its QII definition, and which can used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$.
+can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition, and which can used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$.
 Similarly, the structural rule for substitution on $\mathsf{app}$ can be derived

### 2025-01-24 tex/main.tex
Fix the argument for the coherence: telescope substitution, multiple lifting, and substitution with lifted substitution need to be proved altogether

@@
 \title{Type theories as quotient inductive-recursive types}
-\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding].}
+\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
 \author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
@@
 \newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
+\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
+\newcommand{\Fredrik}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
 
@@
 
-\section{Metatheory and formalisation}
+\section{Metatheory and formalisation} \label{sec:meta}
 
@@
 \begin{enumerate}
-  \item Goal: no transports in the definition.
+  \item Goal: remove transports in the definition so that less transports during reasoning.
   \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
@@
 
-We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a quotient inductive-inductive type, emphasising how the use of transports in its very definition complicates formal reasoning.
-Then we introduce its quotient inductive-inductive-recursive counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning and we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
+We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
+Then we introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
+Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
 
@@
 \end{alignat*}
-where the context extension $\blank,\blank$ requires that the type $\Ty$ indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.
+where the context extension $\blank,\blank$ requires that $\Ty$ is indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.
 
@@
 \begin{alignat*}{3}
-  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
+  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
 \end{alignat*}
@@
 satisfying certain laws (introduced later).
-Context comprehension is given by projections from \emph{non-empty} substitutions $\Sub\;\Gamma\;(\Delta, A)$:
+Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
+for its \emph{tail} and its \emph{head}:
 \begin{alignat*}{3}
@@
 \end{alignat*}
-where $\pi_1\,\sigma$ and $\pi_2\,\sigma$ can be intuitively understood as the \emph{tail} and the \emph{head} of a non-empty substitution $\sigma : \Sub\;\Gamma\;(\Delta, A)$ respectively.
+respectively.
 Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
-Finally, we also have term substitution $[\sigma]\,t$ whose type is $[\sigma]\,A$ for a term $t : \Tm\;\Delta\;A$
+Finally, we have term substitution:
 \begin{alignat*}{3}
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A)
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
 \end{alignat*}
-where the symbol $[\blank]\blank$ is overloaded.
+such that $[\sigma]\,t$ is of $\Tm\;\Gamma\;([\sigma]\,A)$.
+Note that $[\blank]\blank$ is overloaded for type and term substitution.
 
@@
   as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
-  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to eliminate that $\transp$ to derive the right hand side.
+  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.
 
@@
 
-To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor cannot compute so that type mismatches occur after substitution.
+To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur after substitution.
 
-One possibility is to define substitution calculus as a QII\emph{R} type where type substitution, as a simultaneously defined recursion, does compute.
+One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
 Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
@@
   While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
-Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
-This definition fails to eliminate any transport in the definition of substitution calculus.
+Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional.
+This definition fails to remove any transport in the definition of substitution calculus.
 
@@
 However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
-By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
+By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching.
+In this case, only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
 \LT[noinline]{How about \Coq and \Lean?}
@@
 Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
-That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.
+That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.
 
@@
 This boils down to that $[\blank]_{\Ty}\blank$ preserves all equality constructors.
-\LT{Should we mention general condition for coherence?} 
+\LT[noinline]{Should we mention any general condition for coherence?} 
 \begin{proposition}[Coherence]\label{prop:coherence-1}
@@
 In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
-For a QII definition \cite{Altenkirch2016a}, this extension amounts to adding the following constructors 
+For a QII definition \cite{Altenkirch2016a}, this extension adds following constructors
 \begin{alignat*}{3}
-  \UU   & : \implicit{\Gamma} && \Ty\, \Gamma \\
   \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
-  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
+  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}}
 \end{alignat*}
-where the type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.
+as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
+The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.
 
-In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since the equality constructor $[]\UU$ becomes definitional. 
-Therefore, the equality constructor $[]\El$ may be replaced by a clause of type substitution:
-\begin{alignat}{3}
-  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u}) \label{eq:def-type-subst-6}
-\end{alignat}
+In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes definitional. 
+Likewise, the equality constructor $[]\El$ may be replaced by 
+\begin{alignat*}{3}
+  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u})
+\end{alignat*}
 Yet, this naive change breaks the local confluence!
@@
 \[
-  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u)
+  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u).
 \]
@@
 
-Then, \eqref{eq:def-type-subst-6} and the rule $\mathsf{concat}$ are accordingly changed to 
+Then, the equality constructor $[]\El$ and the rule $\mathsf{concat}$ are accordingly changed to 
 \begin{align}
@@
 
-However, we also have to show that the recursive term substitution $[\blank]_{\Tm}\blank$ is propositionally equal to the explicit substitution $[\blank]\blank$, so they can be used together consistently.
+However, we also have to show that the recursive substitution $[\blank]_{\Tm}\blank$ is propositionally equal to the explicit substitution $[\blank]\blank$, so they can be used together consistently.
 
@@
 
-Finally, we have to check that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent with respect to equality constructors.
+Finally, we have to show that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent.
 \begin{proposition}[Coherence] \label{prop:coherence-2}
-  For every equality constructor $p : \sigma = \tau$, the following identities
+  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
   \danger
@@
   \begin{align*}
-    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by definition \eqref{eq:def-type-subst-7}} \\
+    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by \eqref{eq:def-type-subst-7}} \\
                                       & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
@@
                                       & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
-                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by definition \eqref{eq:def-type-subst-7} }
+                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by \eqref{eq:def-type-subst-7} }
   \end{align*}
@@
 
-We have seen that a naive extension of substitution calculus with a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
-In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this desired propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
-We will see the same pattern in the next section.
+\begin{remark}
+We have seen that a naive extension with a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
+In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
+\end{remark}
 
@@
 
-For our QIIR definition, $\sigma\uparrow A$ need not a transport since $[;]_T$ is definitional.
-However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck on the lifting $\sigma \uparrow A$.
-For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $[\idS \uparrow A]_{\Ty}B$ is not definitionally equal to $B$, so the local peak $(\Pi\;A\;B) \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined by a confluent term.
+For our QIIR definition, $\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
+However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
+For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositional but not definitional equal to $\idS$, so the local peak $(\Pi\;A\;B) \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
 
@@
   The following statements hold:
-  \begin{enumerate}
-    \item $\idS\;\{\Gamma, A\} = (\idS\;\{\Gamma\})^+$ and
-    \item $(\sigma; \tau)^+ = \sigma^+ ; \tau^+$.
-  \end{enumerate}
+  \[
+    \idS^+ =^{\Sub\;(\Gamma, A)\;(\Gamma, A)} \idS
+    \qquad\text{and}\qquad
+    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, [\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;(\Theta, A)} \sigma^+ ; \tau^+.
+  \]
 \end{lemma}
@@
 \begin{proposition}[Coherence]
-  For every equality constructor $p : \sigma = \tau$, the following identities
+  For every $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
   \danger
   \[
-    \sigma \uparrow A = \tau \uparrow A
+    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
     \qquad\text{and}\qquad
-    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t,
+    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
   \]
@@
 \begin{proof}
-  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily.
+  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
 \end{proof}
@@
   \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) \to (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma \\
-  \Gamma \dplus \emptytel & = && \Gamma \\
-  \Gamma \dplus (\Xi, A)  & = && (\Gamma \dplus \Xi) , A
+  \Gamma \dplus \emptytel \; & \reduce && \Gamma \\
+  \Gamma \dplus (\Xi, A)  \; & \reduce && (\Gamma \dplus \Xi) , A
 \end{alignat*}
@@
 \]
-requiring telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
+asking for telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
 Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
+\begin{minipage}{.5\textwidth}
 \begin{alignat*}{5}
-  & [ \sigma ]_{\Tel}\,\emptytel && = \emptytel \\
-  & [ \sigma ]_{\Tel}\,(\Xi, A)  && = [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
-  & \sigma \upuparrows \emptytel && = \sigma \\
-  & \sigma \upuparrows (\Xi, A)  && = (\sigma \upuparrows \Xi), A
+  & [ \sigma ]_{\Tel}\,\emptytel && \reduce \emptytel \\
+  & [ \sigma ]_{\Tel}\,(\Xi, A)  && \reduce [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A
 \end{alignat*}
-Now, we have to prove the coherence property for the multiple lifting, telescope substitution, and type substitution altogether.
-\begin{proposition}
-  For every equality constructor $p : \sigma = \tau$, the identities
+\end{minipage}
+\begin{minipage}{.5\textwidth}
+\begin{alignat*}{5}
+  & \sigma \upuparrows \emptytel && \reduce \sigma \\
+  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi), A
+\end{alignat*}
+\end{minipage}
+\vskip\belowdisplayskip
+
+To prove the coherence property for the multiple lifting, we would need the properties $(\idS\;\{\Gamma\}) \upuparrows \Xi = \idS\;\{\Gamma \dplus  \Xi\}$ and $(\sigma; \tau) \upuparrows \Xi = \sigma; \tau^+$, similar to \cref{lem:lifting}, but which cannot be typed: the type of $\idS \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
+
+Instead of resorting to use \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding following clauses:
+\begin{minipage}{.48\textwidth}
+\begin{alignat*}{5}
+  & [ \idS   ]_{\Tel}       \,\Xi && \reduce \Xi \\
+  & [ \sigma ; \tau ]_{\Tel}\,\Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
+\end{alignat*}
+\end{minipage}
+\begin{minipage}{.48\textwidth}
+\begin{alignat*}{5}
+  & \idS          \upuparrows \Xi && \reduce \idS \\
+  & (\sigma;\tau) \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
+\end{alignat*}
+\end{minipage}\vskip\belowdisplayskip\noindent
+whose confluence property is easy to see.
+Now, we proceed with the coherence property for the multiple lifting, telescope substitution, and type substitution altogether.
+\begin{proposition}[Coherence]
+  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the identities
   \danger
-  \[
-    \sigma \upuparrows \Xi = \tau \upuparrows \Xi, \quad
-    [\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi,
-    \quad\text{and}\quad
-    [\sigma \upuparrows \Xi ]_{\Ty}\,A = [\tau \upuparrows \Xi ]_{\Ty}\,A
-  \]
-  hold for any telescope $\Xi$ and type $A$.
+  \begin{align*}
+  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
+                         & [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi, \\
+  [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A
+  \end{align*}
+  hold for any telescope $\Xi$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
 \end{proposition}
 \begin{proof}
-  We prove these two identities together by induction on the telescope $\Xi$ for the first identity and the type $A$ for the second.
-
   \LT{
-  By \cref{re:coherence-proof}, we consider the following case
+    Consider the following cases 
   \begin{align*}
+    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
+                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
+                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
+    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
     [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
@@
   \end{align*}
-  We only consider the case $A = \Pi\;A\;B$ in addition to those considered in \cref{prop:coherence-2}.
+  Identities in the firs row are straightforward (not definitional, though).
+  We shall demonstrate the proof for $[\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A$ only.
   }
 \end{proof}
-
-%% ⟨_⟩ : Tm Γ A → Sub Γ (Γ , A)
-%% ⟨ t ⟩ = idS , t
-%% 
-%% _$$_
-%%   : (t : Tm Γ (Π A B)) (u : Tm Γ A)
-%%   → Tm Γ ([ ⟨ u ⟩ ] B)
-%% t $$ u = [ ⟨ u ⟩ ]t app t
+\begin{corollary}[Coherence]
+  the identity $[\sigma]_{\Ty} A = [\tau]_{\Ty}A$ holds for every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$.
+  \danger
+\end{corollary}
 
@@
   \left<\blank\right> & : \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
-  \left< t \right> & \defeq (\idS , t)
+  \left< t \right> & \defeq (\idS , t),
 \end{alignat*}
-can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition, and which can used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$.
+used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
+can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
 Similarly, the structural rule for substitution on $\mathsf{app}$ can be derived
@@
 
-\section{The eliminator and the standard model}
+\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
 \LT{%
@@
 
-\section{Equivalence to the quotient inductive-inductive definition of type theory}
 \LT{%
@@
 
-\section{Discussion and future work}
+\section{Discussion}
+\subsection{Formal implementation}
 \subsection{Related work}
@@
 Substitution calculus is closely related to \emph{categories with families}~\cite{Dybjer1996}.
-\paragraph*{Single substitution calculus}
+\paragraph*{Quotient inductive-recursive definition of single substitution calculus}
 \cite{Kaposi2023,Kaposi2024a}
@@
 \cite{Kaposi2019}
+\LT{How does the general schema in \cite{Kaposi2019} model equality constructors?
+extensionally?
+Can their schemata manifest the difference of our QIIR definition and the QII definition in \cite{Altenkirch2016a}?}
 
@@
 \paragraph*{Type theory in (observational) type theory}
+\cite{Pujet2022,Pujet2024,Pujet2022a}
+\LT{OTT is a concrete instance of a type theory with UIP and function extensionality.
+It also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types for OTT are underdeveloped.}
 \paragraph*{Formalising definitions by overlapping patterns}
+\LT{Overlapping patters are discussed by Cocks \cite{Cockx2014}, but we would like to integrate it with the schema of QIIRT.}
 \paragraph*{A general schema of quotient inductive-inductive-recursive types}
+\LT{What a general schema should have?  inductive types should be allowed as the codomain of a recursion.}
 

### 2025-01-25 tex/main.tex
Some progress on the coherence proofs for `π₁,`  in `SC+U+PI+Id.QIIRT`
Fix the coherence proof for `_⇈_`, telescope substitution, and type substitution

@@
 \begin{abstract}
-  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's, along equality constructors and proofs need to take account of coercions painstakingly alongside the interesting part.
+  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
   This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
@@
 \cite{Dybjer2003,Dybjer2000,Dybjer1999}
+\LT{Introduce abbreviations here}
 
@@
   Hence the above identity does not even make sense, since their types do not match.
-  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
+  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
   \begin{equation}\label{eq:pi2-comp-real-proof}
@@
   \end{alignat*}
-  The first three equations follow from Lemma~2.3.4 in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
+  The first three equations corresponding to \eqref{eq:pi2-comp-proof} follow with Lemma~2.3.4 in \cite{UFP2013}, while the last has nothing to do with substitution calculus but the path lifting property (\cite[Lemma~2.3.2]{UFP2013}).
 \end{example}
@@
 
-To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur after substitution.
+To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur with substitution.
 
@@
 By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching.
-In this case, only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
+In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered.
 \LT[noinline]{How about \Coq and \Lean?}
@@
 We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to
-\begin{alignat*}{3}
-[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
-[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
-[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
-\end{alignat*}
+\[
+[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
+[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
+[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
+\]
 making $[\sigma]\;\UU = \UU$ propositional instead.
@@
 We first check that this definition of type substitution is confluent.\footnote{%
-\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
+\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
+This option is turned on globally in our formal implementation.}
 \begin{proposition}[Local confluence] \label{prop:local-confluence-1}
@@
 \end{alignat*}
-which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can propagate from type to term substitution.
+which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
 The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
@@
 
-Then, the equality constructor $[]\El$ and the rule $\mathsf{concat}$ are accordingly changed to 
+Then, the equality constructor $[]\El$ is accordingly changed to 
 \begin{align}
-  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}\\
-  \mathsf{concat} : \sigma ; (\tau , t) & =^{\Sub\,\Gamma\,(\Theta, A)} (\sigma ; \tau) , [ \sigma ]_{\Tm} t \nonumber
+  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
 \end{align}
@@
 In this section, we extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
-For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to add the following constructors for the type formation rule, the term introduction/elimination rules:
+For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
 \begin{alignat*}{3}
@@
 However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
-For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositional but not definitional equal to $\idS$, so the local peak $(\Pi\;A\;B) \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
+For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositional but not definitional equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
 
@@
 \pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
-\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow (\sub{\pi_1 \tau}{A})) ; (\pi_1 \tau \uparrow A)\nonumber \\
-\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwiese} \nonumber
+\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
+\sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
 \end{alignat}
@@
   \begin{align*}
-    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}[\tau \uparrow A ]_{\Ty}\,B) \\
-    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}[\pi_1\tau \uparrow A]_{\Ty}\,B)
+    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
+    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
   \end{align*}
@@
 \begin{proposition} \label{prop:correctness-2}
-  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$
+  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
   $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
@@
   Every case follows from equational reasoning straightforwardly.
-  For example, if $\sigma = \sigma;\tau$ is a composite, we have
+  For example, if $\sigma$ is a composite $\sigma;\tau$, we have
   \begin{align*}
@@
 First, define an inductive-recursive type of \emph{telescopes} under a context $\Gamma$ simultaneously with a concatenation $\blank \dplus \blank$ as follows.
+
+\begin{minipage}{.5\textwidth}
 \begin{alignat*}{5}
   \Tel               & :                   && \Ctx \to \Set                                     \\
-  \blank\dplus\blank & :                   && (\Gamma : \Ctx) \to (\Xi : \Tel\;\Gamma) \to \Ctx \\
   \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
-  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) \to (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma \\
-  \Gamma \dplus \emptytel \; & \reduce && \Gamma \\
-  \Gamma \dplus (\Xi, A)  \; & \reduce && (\Gamma \dplus \Xi) , A
+  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
 \end{alignat*}
-
-Then, we define the multiple lifting $\blank \upuparrows \blank$ which generalises the single lifting $\blank\uparrow\blank$.
+\end{minipage}
+\begin{minipage}{.5\textwidth}
+\begin{alignat*}{5}
+  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
+  \Gamma & \dplus \emptytel && \reduce \Gamma \\
+  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
+\end{alignat*}
+\end{minipage}
+\vskip\belowdisplayskip\noindent
+Then, we define the multiple lifting $\blank \upuparrows \blank$.
 Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$, should be
@@
 Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.5\linewidth}
 \begin{alignat*}{5}
@@
 \end{minipage}
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.5\linewidth}
 \begin{alignat*}{5}
@@
 
-To prove the coherence property for the multiple lifting, we would need the properties $(\idS\;\{\Gamma\}) \upuparrows \Xi = \idS\;\{\Gamma \dplus  \Xi\}$ and $(\sigma; \tau) \upuparrows \Xi = \sigma; \tau^+$, similar to \cref{lem:lifting}, but which cannot be typed: the type of $\idS \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
+To prove the coherence property for the multiple lifting, we would need the properties $(\idS\;\{\Gamma\}) \upuparrows \Xi = \idS\;\{\Gamma \dplus  \Xi\}$ and $(\sigma; \tau) \upuparrows \Xi = \sigma; \tau^+$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
 
-Instead of resorting to use \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding following clauses:
-\begin{minipage}{.48\textwidth}
+Instead of resorting to \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding following clauses:
+
+\begin{minipage}{.5\textwidth}
 \begin{alignat*}{5}
@@
 \end{minipage}
-\begin{minipage}{.48\textwidth}
+\begin{minipage}{.5\textwidth}
 \begin{alignat*}{5}
@@
   \begin{align*}
+  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
+                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
 \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
-                         & [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi, \\
-  [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A
+%    &
+%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
   \end{align*}
@@
   \end{align*}
-  Identities in the firs row are straightforward (not definitional, though).
+  Identities in the firs row are straightforward (not completely trivial, though).
   We shall demonstrate the proof for $[\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A$ only.
+
+  By induction on $A$ first then prove other identities by induction on $\Xi$.
   }
 \end{proof}
+
 \begin{corollary}[Coherence]
@@
 
+\section{Formal implementation}
+\LT{Mention the NbE for SC here}
+\LT{Compare some formal proofs with pen-and-paper proofs here}
 \section{Discussion}
-\subsection{Formal implementation}
 \subsection{Related work}

### 2025-01-31 tex/main.tex
draft intro

@@
 \author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
+\author{Fredrik Nordvall Forsberg}{Department of Computer and Information Sciences, University of Strathclyde, UK \and \url{https://fredriknf.com}}{fredrik.nordvall-forsberg@strath.ac.uk}{https://orcid.org/0000-0001-6157-9288}{Supported by the Engineering and Physical Sciences Research Council [EP/Y000455/1].}
 \author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
-\authorrunning{L.-T.~Chen and T.-C.~Tsai}
-\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
+\authorrunning{L.-T.~Chen, F.~Nordvall Forsberg, and T.-C.~Tsai}
+\Copyright{Liang-Ting Chen, Fredrik Nordvall Forsberg, and Tzu-Chun Tsai}
 \ccsdesc[500]{Theory of computation~Type theory}
@@
 Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko, and Meven Lennon-Bertrand.}%optional
+\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional
 
@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventEditors{YYY}
+\EventNoEds{1}
+\EventLongTitle{10th International Conference on Formal Structures for Computation and Deduction (FSCD 2025)}
+\EventShortTitle{FSCD 2025}
+\EventAcronym{FSCD}
+\EventYear{2025}
+\EventDate{July 14--20, 2025}
+\EventLocation{Birmingham, United Kingdom}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{XX}
+\ArticleNo{XX}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
-\newcommand{\Fredrik}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
+\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
 
@@
 \section{Introduction} \label{sec:intro}
+
+Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
+%
+In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas, etc.
+%
+Doing this effectively seems to require several sophisticated features of the host type theory, most notably advanced inductive definitions.
+%
+For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{NordvallForsbergPhD} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
+%
+This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2000}, the resulting development would be considerable harder to work with, and most likely would have exhausted the resources of both the proof assistant and its user.
+
+Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{NordvallForsberg2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to quotient constructors.
+%
+In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
+%
+The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
+%
+As a result, more equations will hold strictly, and we can (at least partially) avoid the so-called ``transport hell'' that arises when constructions need to be transported across equality constructors.
+%
+This is a general phenomenon when working with quotient inductive-inductive types: to make use of equalities in later constructors and when using the eliminator for the type, the equalities need to be transported across.
+%
+Any method that reduces the friction this introduces seem worth exploring.
+
+%However, we found that using quotient inductive-inductive-recursive types alone were not sufficient for a smooth development.
+%%
+%%Not only do we want the action of substitutions on types to compute definitionally, but we also want as many equations to hold strictly as possible:
+%%
+%%on the one hand, we want substitutions to distribute over type formers, but on the other hand we want substitutions to satisfy composition and unit laws of the category of context and substitutions up to definitional equality.
+%%
+%Instead, we also needed to make use of functions defined by overlapping patterns.
+%%
+%In a proof assistant such as Agda, this can be emulated by Agda's experimental support for rewrite rules, but with an additional proof burden on the user to show (informally) that their use is justified meta-theoretically.
+%%
+%To get some help with this task by Agda's confluence checker, it is often necessary to needlessly complicate a definition to convince the checker that the proposed rewrite rules are indeed confluent.
+
+Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
+%
+This seems to be needed in order to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
+%
+We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not well understood or foundationally well motivated.
+Nevertheless, Agda can be convinced to accept our development, using some experimental flags.
+Indeed, we also see this paper as motivation to study quotient inductive-inductive-recursive types and overlapping patterns further.
+
+\Fredrik{I'm not sure if I have included all points here. I tried to stay more positive re overlapping patterns.}
 \LT{%

### 2025-02-03 tex/main.tex
minor revision on intro

@@
 %
-For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{NordvallForsbergPhD} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
+For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
 %
-This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2000}, the resulting development would be considerable harder to work with, and most likely would have exhausted the resources of both the proof assistant and its user.
+This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerable harder to work with, and most likely would have exhausted the resources of both the proof assistant and its user.
 
-Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{NordvallForsberg2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to quotient constructors.
+Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to quotient constructors.
 %
+Yet, there is a general phenomenon, called `transport hell', when working with quotient inductive-inductive types: to make use of equalities in later constructors and when using the eliminator for the type, the equalities need to be transported across.
+%
+Any method that reduces the friction this introduces seem worth exploring.
+%
+
 In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
 %
-The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
+The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
 %
-As a result, more equations will hold strictly, and we can (at least partially) avoid the so-called ``transport hell'' that arises when constructions need to be transported across equality constructors.
+Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
 %
-This is a general phenomenon when working with quotient inductive-inductive types: to make use of equalities in later constructors and when using the eliminator for the type, the equalities need to be transported across.
+This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
 %
-Any method that reduces the friction this introduces seem worth exploring.
+As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself and, at least, partially in the proofs of its properties.
+\LT{Strict versus definitional?}
+%
 
@@
 
-Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
 %
-This seems to be needed in order to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
-%
-We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not well understood or foundationally well motivated.
+We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
 Nevertheless, Agda can be convinced to accept our development, using some experimental flags.
-Indeed, we also see this paper as motivation to study quotient inductive-inductive-recursive types and overlapping patterns further.
+Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
 
 \Fredrik{I'm not sure if I have included all points here. I tried to stay more positive re overlapping patterns.}
-\LT{%
-Points to elaborate:
-\begin{enumerate}
-  \item QIITs are littered with transports, causing too much pains while formalising a type theory.
-  \item Rewrite rules are popular to make structural rules definitional but requires efforts to justify meta-theoretic properties in general.
-  \item Further, the confluence check adopted by \Agda is based on complete development and requires one-step parallel reductions to have a confluent term, making a definition more complicated than necessary.
-  \item Moreover, the use of rewrite rules is typically \emph{not} presented and discussed in literature. We are ignoring the elephant in the room.
-  \item Instead, definitions by overlapping patterns (in theory) requires strong normalisation and local confluence, making definitions simpler to design.
-  \item The rules for type substitutions are \emph{structural} shared with other type theories based on parallel substitution (in particular cwf).
-
-\end{enumerate}
-}
+\LT{Thanks!}
 
@@
 However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
-By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching.
-In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered.
+By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching.
 \LT[noinline]{How about \Coq and \Lean?}
-Using the first-match semantics, the above definition amounts to the following definition:
+In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
 \begin{alignat*}{3}
@@
 \subsection{Future work}
-\paragraph*{Type theory in (observational) type theory}
+\paragraph*{Type theory in type theory}
 \cite{Pujet2022,Pujet2024,Pujet2022a}
-\LT{OTT is a concrete instance of a type theory with UIP and function extensionality.
-It also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types for OTT are underdeveloped.}
-\paragraph*{Formalising definitions by overlapping patterns}
+\LT{OTT is a concrete instance of a type theory with UIP and function extensionality.}
+
+\paragraph*{Semantics and design for definitions by overlapping patterns}
 \LT{Overlapping patters are discussed by Cocks \cite{Cockx2014}, but we would like to integrate it with the schema of QIIRT.}
-\paragraph*{A general schema of quotient inductive-inductive-recursive types}
+\LT{\cite{Eremondi2025} has recently hinted a possible direction for overlapping patterns using sheaf theory.}
+
+\paragraph*{A schema of quotient inductive-inductive-recursive types in observation type theory}
 \LT{What a general schema should have?  inductive types should be allowed as the codomain of a recursion.}
+
+\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.}
 

### 2025-02-04 tex/main.tex
Draft the coherence proof in S3.4 and some intro in S4

@@
 %
-As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself and, at least, partially in the proofs of its properties.
-\LT{Strict versus definitional?}
+As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
+\LT{terminology: strict vs.\ definitional vs.\ judgemental?}
 %
@@
 We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
-Nevertheless, Agda can be convinced to accept our development, using some experimental flags.
+Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags.
 Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
@@
 \LT{Give an example of QIRT: join list?}
+\LT{How the congruence rule for the identity type can be used in QII(R)T?}
 \subsection{Definitions by rewrite rules}
@@
   \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
   \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
@@
 Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
+\begin{remark}
+  The function $[\blank]_{\Tm}\blank$ can be understood as a \emph{smart constructor} for the explicit term substitution $[\blank]\blank$, a well-known idiom in functional programming, since the function $[\blank]_{\Tm}\blank$ (mostly) builds an explicit substitution but computes certain cases away.
+\end{remark}
 
@@
   & \sigma \upuparrows \emptytel && \reduce \sigma \\
-  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi), A
+  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
 \end{alignat*}
@@
 
-To prove the coherence property for the multiple lifting, we would need the properties $(\idS\;\{\Gamma\}) \upuparrows \Xi = \idS\;\{\Gamma \dplus  \Xi\}$ and $(\sigma; \tau) \upuparrows \Xi = \sigma; \tau^+$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
+To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows [\tau]_{\Tel} \Xi); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
 
@@
 whose confluence property is easy to see.
-Now, we proceed with the coherence property for the multiple lifting, telescope substitution, and type substitution altogether.
-\begin{proposition}[Coherence]
-  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the identities
+
+Now, we proceed with the coherence property for the recursive functions introduced so far.
+The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
+\begin{proposition}[Coherence, I]
+  \danger
+  For the following equality constructors $p : \sigma = \tau$
+  \[
+    \mathsf{idr} : \sigma;\idS = \sigma, \quad
+    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
+    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
+  \]
+  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
+  hold definitionally for any telescope $\Xi$ and any type $A$ and
+  \[
+    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
+    \tau   \upuparrows \Xi
+  \]
+  holds propositionally for any telescope $\Xi$.
+\end{proposition}
+\begin{proof}
+  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds definitionally by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
+  Take $\mathsf{idr}$ as an example, we can see 
+  \begin{align*}
+    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
+                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
+                                   & = \sigma \upuparrows \Xi
+  \end{align*}
+  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.
+
+  For other two equality constructors, the arguments follow similarly.
+\end{proof}
+The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
+\begin{proposition}[Coherence, II]
+  For the following equality constructors $p : \sigma = \tau$
+  \begin{alignat*}{5}
+    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
+    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
+  \end{alignat*}
+   the identities
   \danger
@@
                          & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
-\sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
+  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
 %    &
@@
   \end{align*}
-  hold for any telescope $\Xi$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
+  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
 \end{proposition}
 \begin{proof}
-  \LT{
-    Consider the following cases 
+  We demonstrate the proof for $\pi_1\beta: \pi_1(\sigma, t) = t$ and other three cases follow similarly.
+
+  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$, assuming the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$, which are then proved by induction on $\Xi$.
+  \begin{enumerate}
+    \item By induction on $A$, consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$.
+      \begin{description}
+        \item[$A$ is $\UU$:] By definition
+          \[
+            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
+          \]
+          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by the congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
+          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
+          \LT[noinline]{I think the congruence for $\UU$ is true but not sure how precisely.}
+
+        \item[$A$ is $\El\,u$:]
+          Similar to \Cref{prop:coherence-2}, we use \Cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.
+
+        \item[$A$ is $\Pi\,B\,C$:]
+          Note that one of the induction hypothesises is
+          \[
+            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
+          \]
+          for \emph{arbitrary} $\Xi$, so we can apply the induction hypothesis to $B$ and $C$ below:
           \begin{align*}
-    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
-                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
-                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
-    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
-    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
+            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
+            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
+            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by the induction hypothesises} \\
+            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
+            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
           \end{align*}
-  Identities in the firs row are straightforward (not completely trivial, though).
-  We shall demonstrate the proof for $[\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A$ only.
+          Note that 
+      \end{description}
+    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
+      \begin{description}
+        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
+        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
+          \begin{align*}
+            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
+                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
+                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
+          \end{align*}
+      \end{description}
+    \item By induction on $\Xi$, consider the following cases for $\pi_1(\sigma, t) \upuparrows \Xi = \sigma \upuparrows\,\Xi$.
+      \begin{description}
+        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
+        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
+          \begin{align*}
+            \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
+                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
+                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
+                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
+                                                  & \judgeq \sigma \upuparrows (\Xi , A)
+          \end{align*}
+      \end{description}
+  \end{enumerate}
 
-  By induction on $A$ first then prove other identities by induction on $\Xi$.
-  }
 \end{proof}
@@
 \begin{corollary}[Coherence]
-  the identity $[\sigma]_{\Ty} A = [\tau]_{\Ty}A$ holds for every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$.
+  The identity $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ holds for every equality constructor $p : \sigma = \tau$.
   \danger
@@
 }
-
 \LT{%
@@
 }
+After giving the QIIR definition of a type theory, we have to define how to \emph{eliminate} types $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ altogether to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
+Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).
 
+\subsection{Eliminator} \label{subsec:eliminator}
+As the same as the QII definition~\cite{Altenkirch2016a}, the motives for the QIIR definition are
+\begin{align*}
+  \Ctx^A & : \Ctx \to \Set &
+  \Sub^A & : \implicit{\Gamma, \Delta}\,\Ctx^A\;\Gamma \to \Ctx^A\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
+  \Ty^A  & : \implicit{\Gamma}\,\Ctx^A\;\Gamma \to \Ty\;\Gamma \to \Set & 
+  \Tm^A  & : \implicit{\Gamma, A}\; (\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\;A \to \Tm\;\Gamma\;A \to \Set
+\end{align*}
+and each method corresponds to each constructor, e.g., 
+\[
+  \emptyctx^A : \Ctx^A\;\emptyctx 
+  \quad
+  \blank ,^A \blank : \implicit{\Gamma, A}\;(\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\; A \to \Ctx^A\;(\Gamma, A)
+\]
+\subsection{The standard model} \label{subsec:std-model}
+\subsection{The equivalence} \label{subsec:equivalence}
 \section{Formal implementation}
@@
 \subsection{Future work}
+\LT{Remember that it is for further developments.}
 \paragraph*{Type theory in type theory}

### 2025-02-05 tex/main.tex
Draft S4.1

@@
 \subsection{... with an empty universe} \label{subsec:SC+U}
-\LT[noinline]{Shall we go for Coquand universes? \cite{Coquand2013}}
-
 In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
@@
 \begin{proof}
-  We demonstrate the proof for $\pi_1\beta: \pi_1(\sigma, t) = t$ and other three cases follow similarly.
-
-  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$, assuming the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$, which are then proved by induction on $\Xi$.
+  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.
+
+  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$ and the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
   \begin{enumerate}
-    \item By induction on $A$, consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$.
+    \item Consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$ by induction on $A$.
       \begin{description}
@@
           \]
-          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by the congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
+          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
           Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
-          \LT[noinline]{I think the congruence for $\UU$ is true but not sure how precisely.}
+          \LT[noinline]{To fix: Clarify this.}
 
@@
             \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
-            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by the induction hypothesises} \\
+            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by induction hypothesises} \\
             \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
@@
           \end{align*}
-          Note that 
       \end{description}
@@
 \begin{enumerate}
+  \item Discuss the extensional identity type.
+  \item Discuss Coquand universes.
   \item Discuss the unit type.
@@
   \item Discuss the type of natural numbers.
-  \item Discuss the extensional identity type.
 \end{enumerate}
@@
 
-\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
+\section{Elimination and models}
 \LT{%
@@
 }
-After giving the QIIR definition of a type theory, we have to define how to \emph{eliminate} types $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ altogether to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
+With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
 Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).
 
-\subsection{Eliminator} \label{subsec:eliminator}
-As the same as the QII definition~\cite{Altenkirch2016a}, the motives for the QIIR definition are
-\begin{align*}
-  \Ctx^A & : \Ctx \to \Set &
-  \Sub^A & : \implicit{\Gamma, \Delta}\,\Ctx^A\;\Gamma \to \Ctx^A\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
-  \Ty^A  & : \implicit{\Gamma}\,\Ctx^A\;\Gamma \to \Ty\;\Gamma \to \Set & 
-  \Tm^A  & : \implicit{\Gamma, A}\; (\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\;A \to \Tm\;\Gamma\;A \to \Set
-\end{align*}
-and each method corresponds to each constructor, e.g., 
-\[
-  \emptyctx^A : \Ctx^A\;\emptyctx 
-  \quad
-  \blank ,^A \blank : \implicit{\Gamma, A}\;(\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\; A \to \Ctx^A\;(\Gamma, A)
-\]
+\subsection{The elimination principle} \label{subsec:eliminator}
+To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify four types
+\begin{align*}
+  \Ctx^M & : \Ctx \to \Set &
+  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
+  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
+  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
+\end{align*}
+called the \emph{motivates} and functions indexed by corresponding constructors such as 
+\begin{alignat*}{3}
+  \emptyctx^M     & :                      && \Ctx^M\;\emptyctx \\
+  \blank,^M\blank & : \implicit{\Gamma, A} && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  \idS^M          & : \implicit{\Gamma, \Gamma^M}    && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
+  \blank;\blank   & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \Theta, \Theta^M} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Sub^M\;\Delta^M\;\Theta^M\;\tau \to \Sub^M\;\Gamma^M\;\Theta^M\;(\sigma;\tau)\\
+  [\blank]^M\blank & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \sigma, A, A^M, t} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \\
+  &&& \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
+\end{alignat*}
+as well as recursive functions
+\begin{alignat*}{3}
+  [\blank]_{\Ty}^M\blank & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \sigma, A} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  [\blank]_{\Tm}^M\blank & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \sigma, A, A^M, t} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \\
+                         &&& \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
+\end{alignat*}
+called \emph{methods} subject to the corresponding equations \emph{and function clauses}, e.g., 
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
+  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
+  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
+\end{alignat*}
+To define the elimination principle, we cannot avoid the use of transports in identities that would need transports in its QII definition of syntax, such as $[\idS]_{\Tm}^M$, since identities for substitutions (e.g., $[id]_{\Ty}^M$) are still propositional.
+On the other hand, in the elimination principle for the QII definition, the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the type of $A^M$.
+For brevity, we omit the rest of methods and we refer to the Agda implementation for the complete definition of motives and methods.
+\LT[noinline]{Add a link.}
+
+Given the motives and the methods, the elimination principle is defined as functions
+\begin{alignat*}{5}
+  \elim_{\Ctx} & : (\Gamma : \Ctx) \to \Ctx^M\;\Gamma \\
+  \elim_{\Ty}  & : (A : \Ty\;\Gamma) \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
+  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
+  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t
+\end{alignat*}
+by interpreting each constructor using the corresponding method inductively.
+
+In the QII counterpart, $[\blank]\blank$ is a constructor, so $\elim_{\Ty}\;([\sigma]\,A)$ is eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$ definitionally and likewise for the term substitution.
+In our QIIR definition, however, these become propositional identities:
+\begin{proposition}
+  Given the motives and the methods, we have
+\begin{alignat*}{5}
+  \elim_{\Ty}[] & : \implicit{\Gamma, \Delta, \sigma, A} & \elim_{\Ty}\,([\sigma]_{\Ty}\,A) & = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
+  \elim_{\Tm}[] & : \implicit{\Gamma, \Delta, \sigma, A, t} & \elim_{\Tm}\,([\sigma]_{\Tm}\,t) & =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t)
+\end{alignat*}
+
+\end{proposition}
+These identities are required to interpret terms involving definitional identities.
+For example, consider the case for eliminating an extension $(\sigma, t)$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$:
+\[
+  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M \; t^M
+\]
+The eliminated type of $t^M$ is $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$ instead of the eliminated type $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ of $\elim_{\Tm}\;t$, so $t^M$ above have to be the transported term
+\[
+  \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
+\]
+of $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$. 
+\LT{Add some remark regarding the proof}
+
 \subsection{The standard model} \label{subsec:std-model}

### 2025-02-06 tex/main.tex
Polish S4.1

--- 2025-02-04 tex/main.tex
@@
 \begin{abstract}
-  Type theory can be defined in a type theory as a quotient inductive-inductive type, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
+  \LT[noinline]{Try to interest Thorsten Altenkirch in the abstract}
+  Defining type theory in type theory is a long-standing challenge and a benchmark for the expressiveness of a proof assistant.
+  Type theory consisting of well-typed terms only is defined using quotient inductive-inductive type by Altenkirch and Kaposi, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
   This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
   In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
-  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types as quotient inductive-inductive-recursive types and compare ours with quotient inductive-inductive definitions.
+  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types and compare ours with quotient inductive-inductive definitions.
 \end{abstract}
@@
 %
-The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
+The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
 %
@@
 We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
-Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags.
+Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags to introduce additional rewrite rules.\footnote{%
+  As shown in \cite{Cockx2021}, adding non-terminating rewrite rules does not break consistency as long as these rules are consistent, so consistency is not an issue in our development.
+}
 Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
@@
 \begin{proposition}[Coherence, II]
+  \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
   For the following equality constructors $p : \sigma = \tau$
@@
           \]
-          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by the congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
+          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
           Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
-          \LT[noinline]{I think the congruence for $\UU$ is true but not sure how precisely.}
+          \LT{Justify the use of congruence rule for QIIT, e.g. $\transfib{\Ty}{p}{\UU_{\Gamma}} = \UU_{\Delta}$ for $p : \Gamma = \Delta$}
 
@@
 
-\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
-\LT{%
-\begin{enumerate}
-  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
-  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
-  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
-\end{enumerate}
-}
+\section{The elimination principle and models}
+\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
 \LT{%
@@
 }
-After giving the QIIR definition of a type theory, we have to define how to \emph{eliminate} types $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ altogether to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
-Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).
-
-\subsection{Eliminator} \label{subsec:eliminator}
-As the same as the QII definition~\cite{Altenkirch2016a}, the motives for the QIIR definition are
-\begin{align*}
-  \Ctx^A & : \Ctx \to \Set &
-  \Sub^A & : \implicit{\Gamma, \Delta}\,\Ctx^A\;\Gamma \to \Ctx^A\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
-  \Ty^A  & : \implicit{\Gamma}\,\Ctx^A\;\Gamma \to \Ty\;\Gamma \to \Set & 
-  \Tm^A  & : \implicit{\Gamma, A}\; (\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\;A \to \Tm\;\Gamma\;A \to \Set
-\end{align*}
-and each method corresponds to each constructor, e.g., 
-\[
-  \emptyctx^A : \Ctx^A\;\emptyctx 
-  \quad
-  \blank ,^A \blank : \implicit{\Gamma, A}\;(\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\; A \to \Ctx^A\;(\Gamma, A)
-\]
+With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
+Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
+
+\subsection{The elimination principle} \label{subsec:eliminator}
+To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
+\begin{align*}
+  \Ctx^M & : \Ctx \to \Set &
+  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
+  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
+  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
+\end{align*}
+called the \emph{motivates} as the target types for the elimination principle:
+\begin{alignat*}{5}
+  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma \\
+  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
+  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
+  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
+\end{alignat*}
+
+As inhabitants of these inductive types are formed by constructors, each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$ such as
+\begin{alignat*}{3}
+  \emptyctx_{\Ctx}^M     & :                      && \Ctx^M\;\emptyctx \\
+  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  \emptysub_{\Sub}^M     & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
+  \blank,_{\Sub}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
+  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
+  \blank;\blank   & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Sub^M\;\Delta^M\;\Theta^M\;\tau \to \Sub^M\;\Gamma^M\;\Theta^M\;(\sigma;\tau)\\
+  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
+\end{alignat*}
+and simultaneously defined functions are also given by corresponding functions 
+\begin{alignat*}{3}
+  [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
+\end{alignat*}
+subject to the corresponding equations \emph{and function clauses}, e.g., 
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
+  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
+  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
+\end{alignat*}
+called \emph{methods}.
+For brevity, we omit the rest of methods and refer to the Agda implementation for their complete definition.
+\LT[noinline]{Add a link.}
+
+\begin{definition}
+  An \emph{eliminator} for the QIIR definition of type theory consists of motives and methods above.
+\end{definition}
+
+\begin{remark}
+To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
+However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ \emph{already} requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
+That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
+Nevertheless, the difference is insignificant, as both definitions would be derived from the signature of a quotient inductive-inductive(-recursive) type.
+\end{remark}
+
+Given an eliminator, the computation is given to the following function clauses
+\begin{alignat*}{5}
+  \elim_{\Ctx}\,\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
+  \elim_{\Ctx}\,(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
+  \elim_{\Ty}\,\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
+  \elim_{\Sub}\,\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
+                            & \vdots
+\end{alignat*}
+by interpreting each constructor using the corresponding method inductively, similar to its QII counterpart.
+Here, the difference is that type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
+In contrast, substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions in our QIIR definition that cannot be eliminated inductively, but they become propositional identities:
+\begin{proposition}\label{prop:provable-function-clauses}
+  Given the motives and the methods, we have
+\begin{alignat*}{5}
+  \elim_{\Ty}[] & : \implicit{\dots} & \elim_{\Ty}\,([\sigma]_{\Ty}\,A) & = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
+  \elim_{\Tm}[] & : \implicit{\dots} & \elim_{\Tm}\,([\sigma]_{\Tm}\,t) & =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t)
+\end{alignat*}
+\end{proposition}
+\LT{Add some remark regarding the proof?}
+
+These identities are required to interpret constructors using definitional identities.
+E.g., consider eliminating an extension $(\sigma, t)$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
+\[
+  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M_{\Sub} \; \alert{t^M}.
+\]
+In the QII definition, $t^M$ is just $\elim_{\Tm}\,t$ of type
+\begin{equation} \label{eq:elim-subext}
+  \Tm^M\;\Gamma^M\;(\elim_{\Ty}\;([\sigma]_{\Ty}\,A))\;t\judgeq \Tm^M\;\Gamma^M\;[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)\;t
+\end{equation}
+as the RHS of \eqref{eq:elim-subext} is indeed the type for the second argument of $\blank,^M_{\Sub}\blank$.
+In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ which is not definitionally equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ is in fact
+\[
+  (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
+\]
+requiring a transport of $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$.
+
+\begin{remark}
+Note again that our inductive-recursive definition is not an instance of the general schema of inductive-recursion in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003} where simultaneously defined functions do not require corresponding interpretation as they target meta-theoretic types instead of inductive types being defined.
+\end{remark}
+
 \subsection{The standard model} \label{subsec:std-model}
+Our QIIR definition of type theory has a standard model by giving the following motives and methods.
+
 \subsection{The equivalence} \label{subsec:equivalence}

### 2025-02-06 tex/main.tex
IR background

@@
 
-\Fredrik{I'm not sure if I have included all points here. I tried to stay more positive re overlapping patterns.}
-\LT{Thanks!}
-
 \paragraph*{Contributions}
@@
 
-\cite{UFP2013}
-\LT{
-Points to include:
-\begin{enumerate}
-  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
-  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
-  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
-\end{enumerate}}
-
-\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive types).}
+We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality.
+%
+We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
+%
+We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
+
+We write $x \jdeq y$ for definitional equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$.
+One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$, where $\transfibf{P} : (p : x =^A y) \to P(x) \to P(y)$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
+%
+Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
+As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
+
+We have formally implemented our definitions and proofs in Agda, with the following options: \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
+We simulate quotient inductive types using postulated equations to introduce equality constructors~\cite{Licata2011}.
+Some of our theorem statements are meta-theoretical, and thus cannot be checked by Agda.
+Others can only be checked  partially, because of postulated quotient inductive types.
+We have marked such ``informal'' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
+
+%\LT{
+%Points to include:
+%\begin{enumerate}
+%  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
+%  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
+%  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
+%\end{enumerate}}
+%
+%\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive typeswbe checked ).}
+
 \subsection{Inductive-recursive types}
-\cite{Dybjer2003,Dybjer2000,Dybjer1999}
-\LT{Introduce abbreviations here}
-
-\LT{Mention that the recursion part of induction-recursion are already not defined by elimination rule.}
+
+An ordinary inductive type $A$, such as for example the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, by providing an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$.
+For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\mathsf{suc}\,n)$, where the latter might refer to the already computed value $f(n)$.
+%
+The type is first defined inductively by listing its constructors, and then afterwards recursive functions can be defined out of it.
+
+In contrast, an inductive-recursive type is defined inductively \emph{simultaneously} with a recursively defined function out of it.
+The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a ``decoding'' function $T : U \to \Set$ which maps each code to the type it represents.
+Crucially, the code for dependent function types
+\[
+  \pi : (a : U) \to (T(a) \to U) \to U
+\]
+with $T(\pi(a, b)) = (x : T(a)) \to T(b(x))$ only makes sense because we already have access to the decoding function $T$.
+%
+Inductive-recursive types (IRT) were introduced by Dybjer~\cite{Dybjer2000}, and then extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
+The general schema allows one to define a type $A$ inductively, together with a recursive function $F : A \to D$, for some fixed (possibly large) type $D$.
+
+An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, again reflecting that $A$ is inductively defined.
+%
+However, note that the recursive function $F : A \to D$ is not defined using this elimination rule, but part of the definition of the type itself.
+%
+In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must always target a fixed pre-existing type $D$.
+%
+In this work, we will consider an extension \emph{generalised} IRTs, where also $F : A \to A$ is allowed.
+%
+Such IRTs were previously used by Danielsson~\cite{Danielsson2006} to represent type theory in type theory, and are discussed by Kaposi~\cite{Kaposi2023}.
+%
+Their general metatheory would be interesting future work.
+
+%\LT{Mention that the recursion part of induction-recursion are already not defined by elimination rule.}
 
@@
 
-\section{The elimination principle and models}
+\section{The elimination principle and models} \label{sec:models}
 \LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
@@
 \subsection{The equivalence} \label{subsec:equivalence}
-\section{Formal implementation}
+\section{Formal implementation}\label{sec:formal-implementation}
 \LT{Mention the NbE for SC here}

### 2025-02-06 tex/main.tex
QIITs

@@
 \LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
-\subsection{Quotient inductive(-inductive)-recursive types}
-\LT{Give an example of QIRT: join list?}
+\subsection{Quotient inductive-inductive types}
+% \subsection{Quotient inductive(-inductive)-recursive types}
+
+Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
+%
+Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
+%
+For example, the type of \emph{join lists} $\JList{X}$ can be given by the QIT
+with the following constructors:
+\begin{alignat*}{3}
+  \Jnil & : \JList{X} \\
+  \Jsing{-} & : X \to \JList{X} \\
+  \Jconcat{-}{-} & : \JList{X} \to \JList{X} \to \JList{X} \\
+  \mathsf{idl} & : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{X}} \ell \\
+  \mathsf{idr} & : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{X}} \ell \\
+  \mathsf{assoc} & : \implicit{\ell_0, \ell_1, \ell_2}\; (\Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2}) =^{\JList{X}} (\Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})})
+\end{alignat*}
+
+A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a family of types $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively; indeed, inductive-inductive definitions only make sense when $D = \Set$.
+The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kovacs2018,Kaposi2019,Kovacs2020}.
+
+\Fredrik{Give an actual example of QIIT (or even QIRT)??}
+
+Shulman~\cite{Shulman2014} considered combining higher inductive types and inductive-recursive types, but the foundational status of this combination is underdeveloped.
+We will also make use of such a combination, which gives additional motivation to study its metatheory in future work.
+
 \LT{How the congruence rule for the identity type can be used in QII(R)T?}

### 2025-02-06 tex/main.tex
Draft S4.2 on the standard model

@@
 \usepackage{manfnt}
+\usepackage{underscore}
 \newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
@@
 Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
-As substitution calculus is the initial cwf with a constant type $\UU$, substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
+Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families), 
 \begin{alignat*}{3}
@@
 \end{alignat*}
-satisfying certain laws (introduced later).
+satisfying certain laws (introduced later as equality constructors).
 Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
@@
 One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
-Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
+Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
 \begin{alignat*}{3}
@@
   While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
-Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional.
+Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
 This definition fails to remove any transport in the definition of substitution calculus.
@@
 Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
-That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.
+That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.
 
 To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
-In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
+In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
 We end up with the following definition for type substitution:
@@
 
-With this definition of type substitution, other equality constructors can be introduced without any $\transp$, we list updated constructors as follows.
+With this definition of type substitution, other equality constructors can be introduced without any $\transp$, updated constructors are listed below.
 \begin{alignat*}{5}
@@
 With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
-Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
+Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
 
 \subsection{The elimination principle} \label{subsec:eliminator}
+\LT{We shall say something about the elimination/induction principle in \cref{sec:meta}}
+
 To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
@@
 \end{align*}
-called the \emph{motivates} as the target types for the elimination principle:
+called the \emph{motivates} as the target types for the elimination:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-
-As inhabitants of these inductive types are formed by constructors, each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$ such as
+Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, i.e.\ 
 \begin{alignat*}{3}
   \emptyctx_{\Ctx}^M     & :                      && \Ctx^M\;\emptyctx \\
-  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  \blank,_{\Ctx}^M\blank & : \implicit{\dots} && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
   \emptysub_{\Sub}^M     & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
@@
   \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
-  \blank;\blank   & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Sub^M\;\Delta^M\;\Theta^M\;\tau \to \Sub^M\;\Gamma^M\;\Theta^M\;(\sigma;\tau)\\
-  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
+  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t) \\
+                   && \vdots
 \end{alignat*}
@@
   [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
-  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
+  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
+                         && \vdots
 \end{alignat*}
@@
   [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
-  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
+  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M \\
+                 && \vdots
 \end{alignat*}
 called \emph{methods}.
-For brevity, we omit the rest of methods and refer to the Agda implementation for their complete definition.
+For brevity, we omit the rest of methods and refer to the Agda implementation for them.
 \LT[noinline]{Add a link.}
+We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$.
 
 \begin{definition}
-  An \emph{eliminator} for the QIIR definition of type theory consists of motives and methods above.
+  An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
 \end{definition}
@@
 
-Given an eliminator, the computation is given to the following function clauses
+Given an eliminator, the following clauses are derived to carry out the actual computation
 \begin{alignat*}{5}
@@
 \end{alignat*}
-by interpreting each constructor using the corresponding method inductively, similar to its QII counterpart.
-Here, the difference is that type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
-In contrast, substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions in our QIIR definition that cannot be eliminated inductively, but they become propositional identities:
+interpreting each constructor using the corresponding method inductively.
+
+Here, the key difference from its QII counterpart is that type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
+In contrast, substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions rather than constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
 \begin{proposition}\label{prop:provable-function-clauses}
@@
 
-These identities are required to interpret constructors using definitional identities.
-E.g., consider eliminating an extension $(\sigma, t)$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
+These identities are used to interpret constructors whereas definitional identities are used in syntax.
+Consider eliminating an extension $\sigma, t$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
 \[
@@
 as the RHS of \eqref{eq:elim-subext} is indeed the type for the second argument of $\blank,^M_{\Sub}\blank$.
-In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ which is not definitionally equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ is in fact
+In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ is not definitionally equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
 \[
@@
 \]
-requiring a transport of $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$.
+transporting $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$ to replace \eqref{eq:elim-subext}.
 
@@
 \subsection{The standard model} \label{subsec:std-model}
-Our QIIR definition of type theory has a standard model by giving the following motives and methods.
-
-\subsection{The equivalence} \label{subsec:equivalence}
-\section{Formal implementation}\label{sec:formal-implementation}
+
+The soundness of our definition is justified by the standard model, which is further given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+\begin{align*}
+  \Ctx^M & \defeq \lam{\_} \Set &
+  \Sub^M & \defeq \lam{\Gamma^M}{\Delta^M}{\_} \Gamma^M \to \Delta^M \\
+  \Ty^M  & \defeq \lam{\Gamma^M}{\_}\Gamma^M \to \Set & 
+  \Tm^M  & \defeq \lam{\Gamma^M}{A^M}{\_} (\gamma : \Gamma^M) \to A^M \gamma
+\end{align*}
+with the additional methods to its QII counterpart:
+\begin{align*}
+  \blank\uparrow^M\blank & \judgeq \lam{\sigma^M}{A^M}{(\gamma, x^M)} (\sigma^M\,\gamma , x^M) & \\
+                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
+  [\blank]_{\Tm}^M\blank & \judgeq \lam{\sigma^M}{t^M} (\sigma^M ; t^M)  \\
+                         & : (\Gamma^M \to \Delta^M) \to ((\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
+\end{align*}
+where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
+The additional identities for these methods all hold definitionally.
+For example, 
+\LT{give an example}
+
+Indeed, it is not surprising.
+Additional rules we have used to enable further reduction are derivable already, so they can be justified by the same standard model.
+It is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
+\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
+\section{Formal implementation} \label{sec:formal-implementation}
 \LT{Mention the NbE for SC here}

### 2025-02-06 tex/main.tex
tried to polish abstract

@@
   Defining type theory in type theory is a long-standing challenge and a benchmark for the expressiveness of a proof assistant.
-  Type theory consisting of well-typed terms only is defined using quotient inductive-inductive type by Altenkirch and Kaposi, but well-typed terms are littered with explicit coercions, i.e.\ \transp's along equality constructors, and proofs need to account for coercions painstakingly alongside the interesting part.
-  This mess has been dubbed the `transport hell' but typically suppressed in presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
-  In this paper, we aim for shortening the gap using quotient inductive-inductive-recursive types and definitions by overlapping patterns altogether, reducing the use of \transp's.
-  As a case study, we investigate (parallel) substitution calculus and type theory with a universe and $\Pi$-types and compare ours with quotient inductive-inductive definitions.
+  Altenkirch and Kaposi used quotient inductive-inductive types to represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
+  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
+  This %mess
+  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
+  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
+  As a case study, we investigate the (parallel) substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
 \end{abstract}

### 2025-02-06 tex/main.tex
polish Section 3

@@
   \Jnil & : \JList{X} \\
-  \Jsing{-} & : X \to \JList{X} \\
-  \Jconcat{-}{-} & : \JList{X} \to \JList{X} \to \JList{X} \\
+  \Jsing{\blank} & : X \to \JList{X} \\
+  \Jconcat{\blank}{\blank} & : \JList{X} \to \JList{X} \to \JList{X} \\
   \mathsf{idl} & : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{X}} \ell \\
@@
 
-\Fredrik{Give an actual example of QIIT (or even QIRT)??}
-
-Shulman~\cite{Shulman2014} considered combining higher inductive types and inductive-recursive types, but the foundational status of this combination is underdeveloped.
+We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{X}$ is defined simultaneously with a recursively defined function $\flatten : \JList{X} \to \List{X}$ which flattens a join list into an ordinary list.
+We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
+\[
+\mathsf{nf} : (\ell, \ell' : \JList{X}) \to \flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell' \to \ell =^{\JList{X}} \ell'
+\]
+which states that join lists are equal as soon as their flattenings are equal.
+For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{X}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$.
+
+%Shulman~\cite{Shulman2014} considered combining higher inductive types and inductive-recursive types, but the foundational status of this combination is underdeveloped.
 We will also make use of such a combination, which gives additional motivation to study its metatheory in future work.
@@
 
-We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
-Then we introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
+We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
+We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
 Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
@@
 
-Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types under some context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from the domain $\Gamma$ and the codomain $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms under a context $\Gamma$ and its type $A$.
-These types amount to the following types indexed by types being defined (hence inductive-inductive):
+Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types in context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from context $\Gamma$ to context $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms in context $\Gamma$ of type $A$.
+This amounts to the following type families, all mutually inductively defined (hence inductive-inductive):
 \begin{alignat*}{3}
@@
 \end{alignat*}
-The type $\Ctx$ has two constructors $\emptyctx$ for the empty context and $\blank,\blank$ for context extension:
+The type $\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as follows.
 \begin{alignat*}{3}
@@
 \end{alignat*}
-where the context extension $\blank,\blank$ requires that $\Ty$ is indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.
-
-Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
+Note that a context extension $\Gamma , A$ is only wellformed if $A : \Ty\;\Gamma$, ensuring overall wellformedness of contexts, but requiring contexts and types to be defined simultaneously.
+
+Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
 \begin{alignat*}{3}
@@
 \end{alignat*}
-where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.
-
-If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
-For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
-Hence another constructor is introduced:
+where $\Gamma$ and $\Delta$ are kept implicit.
+\Fredrik{Could we change this to $A[\sigma]$ instead?}
+
+Since $\Ty$ is inductively defined, we add a base case to ensure that $\Ty\;\Gamma$ is not empty for every $\Gamma$:
+%
+%If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
+%For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
+%Hence another constructor is introduced:
 \begin{alignat*}{3}
@@
 
-Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
+Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
 \begin{alignat*}{3}
@@
 \end{alignat*}
-Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
-Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families), 
+Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
+We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$, because the substitution calculus is supposed to be the initial \emph{category} with families:
+%Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
 \begin{alignat*}{3}
@@
 \end{alignat*}
-satisfying certain laws (introduced later as equality constructors).
-Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
-for its \emph{tail} and its \emph{head}:
+These need to satisfy certain laws, which we introduce later as equality constructors.
+Context extension is characterised by its projections, which picks out the rest of the context and the most recent variable
+%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
 \begin{alignat*}{3}
@@
 respectively.
-Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
-Finally, we have term substitution:
+Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
+Finally, we have term substitution, which given a substitution $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in context $\Delta$ to terms in context $\Gamma$, appropriately preserving the type, which is also substituted:
 \begin{alignat*}{3}
@@
 \end{alignat*}
-such that $[\sigma]\,t$ is of $\Tm\;\Gamma\;([\sigma]\,A)$.
 Note that $[\blank]\blank$ is overloaded for type and term substitution.
@@
 
-The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the length of terms in $\sigma$, giving rise two $\eta$-laws in the case that $\Delta$ is empty and non-empty:
-\begin{alignat*}{5}
-  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma        && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
-  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma)
-\end{alignat*}
-
-The laws for substitution composition are stipulated using equality constructors.
+The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the number of terms in $\sigma$, giving rise to two $\eta$-laws --- one for $\Delta$ being empty, and one for $\Delta$ being non-empty:
+\begin{alignat*}{5}
+  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma}\; && \sigma        && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
+  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma}\; && \sigma   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma)
+\end{alignat*}
+
+The laws for substitution composition are stipulated using equality constructors, where, for the first time, we see the need to adjust the types involved with a $\transp$ in order for both sides to have the same type.
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Substitution composition $\blank;\blank$ acts like a list concatenation, so $\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$.
-However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.
-
-The tail and the head of a non-empty substitution $(\sigma, t)$ is obviously $\sigma$ and $t$ respectively:
+%
+Composition of substitutions $\blank;\blank$ acts like a list concatenation, so
+we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), [\sigma]\,t$.
+However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the context extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
+Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ which we can transport along to fix up the equation.
+
+%$\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
+%However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.
+
+The first and second projections of a non-empty substitution $(\sigma, t)$ should  obviously be $\sigma$ and $t$ respectively
 \begin{alignat*}{5}
@@
 \end{alignat*}
-whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
-Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.
-
+but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$.
+%whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
+%Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.
+%
 Similarly, for term substitution, terms are transported along the corresponding rules:
@@
 
-To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur with substitution.
-
-One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
-Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
-\begin{alignat*}{3}
-[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
-[ \sigma ]_{\Ty}\;\UU & = \UU
-\end{alignat*}
-Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
-  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
-Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
-This definition fails to remove any transport in the definition of substitution calculus.
-
-Our second attempt is to define type substitution with additional rules:
+To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.
+
+One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
+%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
+%\begin{alignat*}{3}
+%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
+%[ \sigma ]_{\Ty}\;\UU & = \UU
+%\end{alignat*}
+Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
+  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
+%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
+%This definition fails to remove any transport in the definition of substitution calculus.
+%
+Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
 \begin{alignat}{3}
 [\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
-[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
-[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
-[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
+[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
+[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
+[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
 \end{alignat}
 However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
-By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching.
+By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
 \LT[noinline]{How about \Coq and \Lean?}
-In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
-\begin{alignat*}{3}
-[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
-[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
-\end{alignat*}
-where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
+\Fredrik[noinline]{They also translate to eliminators, so I think we do not need to say anything}
+%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
+%\begin{alignat*}{3}
+%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
+%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
+%\end{alignat*}
+%where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
 %As a result, neither  are definitional.
@@
 \]
-making $[\sigma]\;\UU = \UU$ propositional instead.
-Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
-Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
-That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.
-
-To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
+where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the definitional equality $[\sigma]\;\UU \equiv \UU$.
+%
+%making $[\sigma]\;\UU = \UU$ propositional instead.
+%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
+%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
+%That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.
+
+To ensure that type substitution is definitional for all %\emph
+{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
 In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
@@
 \begin{proof}
-  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
-
+  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
 \end{proof}
 
-With this definition of type substitution, other equality constructors can be introduced without any $\transp$, updated constructors are listed below.
+With this definition of type substitution, other equality constructors can be introduced without any $\transp$, and we can give the updated constructors the types below.
 \begin{alignat*}{5}
@@
   [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
-  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
+  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;([ \tau ]\;t})
 \end{alignat*}
@@
   The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
-  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ and is definitionally equal to $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$, i.e.\ the type of the term on the RHS.
+  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is definitionally equal to $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$, i.e.\ the type of the term on the RHS.
   Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are definitionally equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
@@
 
-Yet, as we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.
-This boils down to that $[\blank]_{\Ty}\blank$ preserves all equality constructors.
+As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
 \LT[noinline]{Should we mention any general condition for coherence?}
+\Fredrik[noinline]{Is there one?}
 \begin{proposition}[Coherence]\label{prop:coherence-1}
@@
 \begin{proof}
-  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
+  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that the following identities
   \begin{align*}
@@
   \end{align*}
-  hold (propositionally).
+  hold propositionally.
   The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.
 
-  The sole identity of the second row is just one, i.e.\ \eqref{eq:def-type-subst-3}, of the clauses of the definition.
+  The sole identity of the second row is clause \eqref{eq:def-type-subst-3} of the definition of $[\blank]_{\Ty}\blank$.
 
   We show the remaining cases by induction on $A$.
-  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of identities with $A \defeq \UU$ reduce to $\UU$.
+  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of the identities with $A \defeq \UU$ reduce to $\UU$.
 \end{proof}
@@
 
-\subsection{... with an empty universe} \label{subsec:SC+U}
+\subsection{Adding a universe} \label{subsec:SC+U}
 In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
-For a QII definition \cite{Altenkirch2016a}, this extension adds following constructors
+For a QII definition \cite{Altenkirch2016a}, this extension adds the following constructors
 \begin{alignat*}{3}
   \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
-  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}}
-\end{alignat*}
-as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
+  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
+\end{alignat*}
+%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
 The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.
@@
 Yet, this naive change breaks the local confluence!
-Observe that the local peak
+The problem is that there is now additional interaction between types and terms. Observe that the local peak
 \[
@@
 \begin{proof}
-  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
+  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
 \end{proof}
 
-However, we also have to show that the recursive substitution $[\blank]_{\Tm}\blank$ is propositionally equal to the explicit substitution $[\blank]\blank$, so they can be used together consistently.
+We also have to show that the recursive substitution $[\blank]_{\Tm}\blank$ is propositionally equal to the explicit substitution $[\blank]\blank$, so they can be used together consistently.
 
@@
   We show this statement by induction on $\sigma$.
+Since we are proving a propositional equality, and we are assuming uniqueness of identity proofs, all equality constructors are automatically respected, so we only need to worry about the point constructors.
   \begin{enumerate}
-    \item For the identity substitution $\idS$, we have $[\idS]t : [\idS]t = t$ but $t \judgeq [\idS]_{\Tm} t$ by definition.
-    \item For a substitution composite $\sigma; \tau$, we have
-      \begin{align*}
-        [\sigma;\tau]t & = [\sigma]\,[\tau] t               && \text{by $[;]t$} \\
-                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
-                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis $[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
+    \item For the identity substitution $\idS$, we have $[\idS]t = t \judgeq [\idS]_{\Tm} t$ by definition.
+    \item For a composite substitution $\sigma; \tau$, we have
+      \begin{align*}
+        [\sigma;\tau]t & = [\sigma]\,([\tau] t)            && \text{by $[;]t$} \\
+%                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
                        & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
@@
         [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
-                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                             & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
       \end{align*}
-    \item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
+    %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
     \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ definitionally.
@@
   Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
-  \LT[noinline]{Shall we say anything about equality constructors?}
 \end{proof}
@@
   \]
-  for any type $A$ and any term $t$.
+  hold for any type $A$ and any term $t$.
 \end{proposition}
@@
 
-\subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
+\subsection{Adding \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
 
@@
 
-For our QIIR definition, $\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
+For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
+we do not need any transport since $[;]_T$ and $[]\Pi$ are definitional.
 However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
-For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositional but not definitional equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
+For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositionally but not definitionally equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
 
@@
   These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
+\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
 \end{proof}
@@
 \subsubsection{Multiple lifting and the coherence of type substitution}
+\Fredrik{This section gets rather technical. Keeping the page limit in mind, I wonder if we should only briefly summarise it in the main text?}
+
 For the coherence of~$[\blank]_{\Ty}\blank$, we have to generalise the desired identity to account for the lifting, since a naive induction does not work: for example, unfolding $[\sigma]_{\Ty}\,(\Pi\;A\;B)$ leads to
@@
 }
+\Fredrik{I think we should keep this brief.
+I started doing the natural numbers in Agda, but there were some problems with the rewrite $[\sigma] \mathsf{Nat} = \mathsf{Nat}$ not firing in the type of the eliminator...}
 

### 2025-02-07 tex/main.tex
Polish S4.2

@@
-\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate]{lipics-v2021}
+\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate,final]{lipics-v2021}
 %This is a template for producing LIPIcs articles. 
@@
 
-\supplement{The formal development is hosted at the GitHub repository: \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g.
+\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/genetsai95/DTT-QIIRT}.\LT[noinline]{Transfer it to my account?}}%optional, e.g.
 related research data, source code, ...
@@
 \usepackage{manfnt}
-\usepackage{underscore}
 \newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
-\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
+\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
 \newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
@@
 \begin{abstract}
-  \LT[noinline]{Try to interest Thorsten Altenkirch in the abstract}
-  Defining type theory in type theory is a long-standing challenge and a benchmark for the expressiveness of a proof assistant.
+  Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
   Altenkirch and Kaposi used quotient inductive-inductive types to represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
@@
   In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
-  As a case study, we investigate the (parallel) substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
+  As a case study, we investigate substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
 \end{abstract}
@@
 %
-This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerable harder to work with, and most likely would have exhausted the resources of both the proof assistant and its user.
+This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.
 
@@
 
-\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-\cite{Cockx2014,Altenkirch2016a}
-\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
 \subsection{Quotient inductive-inductive types}
@@
 A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a family of types $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively; indeed, inductive-inductive definitions only make sense when $D = \Set$.
-The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kovacs2018,Kaposi2019,Kovacs2020}.
+The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
 
@@
 \LT{How the congruence rule for the identity type can be used in QII(R)T?}
+
+\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
+\cite{Cockx2014,Altenkirch2016a}
+\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
 \subsection{Definitions by rewrite rules}
@@
 \section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
-\LT{
-\begin{enumerate}
-  \item Goal: remove transports in the definition so that less transports during reasoning.
-  \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
-  \item Develop the quotient inductive-inductive-recursive definition of type theory step by step (which needs to be locally confluent and terminating).
-    \begin{enumerate}
-      \item parallel substitution: type substitution
-      \item the type of elements of $A$: term substitution needs to be split into two definitional substitution and explicit term substitution to maintain local confluence.
-      \item $\Pi$-type: because of the substitution lifting $\sigma^+ \defeq (\pi_1\idS, \pi_2\idS)$ is used we need to introduce a definitional lifting to maintain local confluence.
-        We prefer categorical combinator and avoid using $\left< t \right>$ in the definition, which would introduce transports in the definition.
-      \item Other type formers can be introduced as usual (as long as we use categorical combinators).
-    \end{enumerate}
-\end{enumerate}
-}
 
@@
 This amounts to the following type families, all mutually inductively defined (hence inductive-inductive):
-\begin{alignat*}{3}
-  \Ctx   & : \Set                   \\
-  \Ty    & : \Ctx \to \Set          \\
-  \Sub   & : \Ctx \to \Ctx \to \Set \\
-  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
-\end{alignat*}
-The type $\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as follows.
+\begin{align*}
+  \Ctx   & : \Set                   & \Sub   & : \Ctx \to \Ctx \to \Set \\
+  \Ty    & : \Ctx \to \Set          & \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
+\end{align*}
+$\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as
 \begin{alignat*}{3}
@@
 \Fredrik{Could we change this to $A[\sigma]$ instead?}
+\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition: }
 
@@
 By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
-\LT[noinline]{How about \Coq and \Lean?}
-\Fredrik[noinline]{They also translate to eliminators, so I think we do not need to say anything}
 %In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
@@
 \Fredrik[noinline]{Is there one?}
+\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
 \begin{proposition}[Coherence]\label{prop:coherence-1}
@@
 \Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
+\LT{How about deferring introducing heterogeneous equality until \cref{sec:formal-implementation}?}
 \end{proof}
@@
 \Fredrik{This section gets rather technical. Keeping the page limit in mind, I wonder if we should only briefly summarise it in the main text?}
+\LT{Move this to the appendix?}
 
@@
 
-\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
+\subsection{Adding other type formers} \label{subsec:SC+U+Pi+more}
 \LT{%
@@
 \end{align*}
-called the \emph{motivates} as the target types for the elimination:
-\begin{alignat*}{5}
-  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma \\
-  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
+called the \emph{motivates} as the target types for the elimination principle:
+\begin{alignat*}{5}
+  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma &
   \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
-  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
-\end{alignat*}
-Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, i.e.\ 
+  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \quad
+               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
+\end{alignat*}
+Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, e.g.,\ 
 \begin{alignat*}{3}
   \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
-  \blank,_{\Ctx}^M\blank & : \implicit{\dots} && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
   \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
@@
   \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
-  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
-  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t) \\
-                   && \vdots
+%  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
+  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
-  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
-                         && \vdots
-\end{alignat*}
-subject to the corresponding equations \emph{and function clauses}, e.g., 
-\begin{alignat*}{5}
-  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
+  [\blank]_{\Ty}^M\blank & : \implicit{\dots} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
+\end{alignat*}
+subject to the identities corresponding to equality constructors and function clauses, e.g., 
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
   [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
-  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M \\
-                 && \vdots
+  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  \elim_{\Ctx}\,\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
-  \elim_{\Ctx}\,(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
-  \elim_{\Ty}\,\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
-  \elim_{\Sub}\,\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
+  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
+  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
+  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
+  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
                             & \vdots
@@
 
-Here, the key difference from its QII counterpart is that type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
-In contrast, substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions rather than constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
+Interestingly, the key difference from its QII counterpart is that substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions rather than constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
+%type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
 \begin{proposition}\label{prop:provable-function-clauses}
@@
 \]
-In the QII definition, $t^M$ is just $\elim_{\Tm}\,t$ of type
+In the QII definition, $\elim_{\Tm}\,t$ is exactly $t^M$ and has the type
 \begin{equation} \label{eq:elim-subext}
@@
 \end{equation}
-as the RHS of \eqref{eq:elim-subext} is indeed the type for the second argument of $\blank,^M_{\Sub}\blank$.
+as the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
 In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ is not definitionally equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
@@
 \]
-transporting $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$ to replace \eqref{eq:elim-subext}.
+transporting $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$.
 
 \begin{remark}
-Note again that our inductive-recursive definition is not an instance of the general schema of inductive-recursion in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003} where simultaneously defined functions do not require corresponding interpretation as they target meta-theoretic types instead of inductive types being defined.
+Note again that our definition is not an instance of Dybjer and Setzer's schema of inductive-recursion where simultaneously defined functions do not require corresponding interpretation as they target meta-theoretic types instead of inductive types being defined.
 \end{remark}
@@
 \subsection{The standard model} \label{subsec:std-model}
-
-The soundness of our definition is justified by the standard model, which is further given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
-\begin{align*}
-  \Ctx^M & \defeq \lam{\_} \Set &
-  \Sub^M & \defeq \lam{\Gamma^M}{\Delta^M}{\_} \Gamma^M \to \Delta^M \\
-  \Ty^M  & \defeq \lam{\Gamma^M}{\_}\Gamma^M \to \Set & 
-  \Tm^M  & \defeq \lam{\Gamma^M}{A^M}{\_} (\gamma : \Gamma^M) \to A^M \gamma
+Our definition has the standard model, justifying its soundness of our definition, given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+\begin{align*}
+  \Ctx^M & \defeq \lam{\nameless} \Set &
+  \Sub^M & \defeq \lam{\Gamma^M}{\Delta^M}{\nameless} \Gamma^M \to \Delta^M \\
+  \Ty^M  & \defeq \lam{\Gamma^M}{\nameless}\Gamma^M \to \Set & 
+  \Tm^M  & \defeq \lam{\Gamma^M}{A^M}{\nameless} (\gamma : \Gamma^M) \to A^M \gamma
 \end{align*}
@@
 \begin{align*}
-  \blank\uparrow^M\blank & \judgeq \lam{\sigma^M}{A^M}{(\gamma, x^M)} (\sigma^M\,\gamma , x^M) & \\
+  \blank\uparrow^M\blank & \judgeq \lam{\sigma^M}{A^M}{\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
                          & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
@@
 where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
-The additional identities for these methods all hold definitionally.
-For example, 
-\LT{give an example}
-
-Indeed, it is not surprising.
-Additional rules we have used to enable further reduction are derivable already, so they can be justified by the same standard model.
-It is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
+
+In the QII counterpart, none of identities hold definitionally, because their indices always involve equality constructors.
+For example, the identity type substitution corresponds to
+\begin{alignat*}{5}
+  [\idS]^M & : [ \idS^M ]^M_{\Ty}\, A^M && =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]}} A^M
+\end{alignat*}
+The transport along $[\idS] : [\idS]\,A = A$ above is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of $A$.
+Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{[\idS]}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and requires a lemma about constant transport~\cite[Lemma~2.3.5]{UFP2013} to show.
+
+In contrast, in our QIIR definition, many identities just hold definitionally.
+Consider
+\begin{alignat*}{5}
+  [\idS]^M                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
+  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
+\end{alignat*}
+where $[\idS]^M$ is just $(\lam{\gamma} A^M\,\gamma) = A^M$ for $A^M : \Gamma^M \to \Set$ in the model and holds definitionally by the $\eta$-law for $\Pi$-types.
+Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$, definitionally equal to $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) = \lam{\gamma}\gamma$, and holds definitionally by the $\eta$-law for $\Sigma$-types.
+
+Additional rules we have used to enable further reduction are derivable already, and they can be justified by the same standard model with simpler proofs.
+Now is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
 \subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
@@
 \IfFileExists{./reference.bib}{\bibliography{reference}}{\bibliography{ref}}
+%\bibliography{ref}
 

### 2025-02-07 tex/main.tex
Some revisions after the meeting 1. definitional to strict 2. comment out heterogeneous identity;

@@
-\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate,final]{lipics-v2021}
+\documentclass[a4paper,UKenglish,numberwithinsect,cleveref,thm-restate]{lipics-v2021}
 %This is a template for producing LIPIcs articles. 
@@
   Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
-  Altenkirch and Kaposi used quotient inductive-inductive types to represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
+  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
   As a result, proofs need to painstakingly account for coercions alongside the interesting part.
@@
   In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
-  As a case study, we investigate substitution calculus and type theory with a universe and $\Pi$-types, and compare ours with quotient inductive-inductive definitions.
+  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our definition with quotient inductive-inductive definitions.
 \end{abstract}
@@
 As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
-\LT{terminology: strict vs.\ definitional vs.\ judgemental?}
 %
@@
 %%
-%%Not only do we want the action of substitutions on types to compute definitionally, but we also want as many equations to hold strictly as possible:
+%%Not only do we want the action of substitutions on types to compute strictly, but we also want as many equations to hold strictly as possible:
 %%
-%%on the one hand, we want substitutions to distribute over type formers, but on the other hand we want substitutions to satisfy composition and unit laws of the category of context and substitutions up to definitional equality.
+%%on the one hand, we want substitutions to distribute over type formers, but on the other hand we want substitutions to satisfy composition and unit laws of the category of context and substitutions up to strict equality.
 %%
@@
 
-We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality.
+We work in intensional Martin-L\"of type theory with uniqueness of identity proofs.
+In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
 %
@@
 
-We write $x \jdeq y$ for definitional equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$.
+We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$.
 One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$, where $\transfibf{P} : (p : x =^A y) \to P(x) \to P(y)$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
 %
-Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
-As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
+%Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
+%As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
 
@@
 
-Since $\Ty$ is inductively defined, we add a base case to ensure that $\Ty\;\Gamma$ is not empty for every $\Gamma$:
+Since $\Ty$ is inductively defined, we add a base case $\UU : \implicit{\Gamma} \Ty\,\Gamma$ to ensure that $\Ty\;\Gamma$ is not empty.
+%for every $\Gamma$:
 %
@@
 %Hence another constructor is introduced:
-\begin{alignat*}{3}
-  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
-\end{alignat*}
+%\begin{alignat*}{3}
+%  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
+%\end{alignat*}
 For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.
@@
   Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$.
-  In short, the following equality apparently holds
+  In short, the following identity apparently holds
   \[
@@
 
-To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.
+To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules strict, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.
 
@@
 \end{alignat}
-However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
+However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
 By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
@@
 %\end{alignat*}
-%where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
-%As a result, neither  are definitional.
-We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to
+%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
+%As a result, neither  are strict.
+We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to
 \[
@@
 \]
-where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the definitional equality $[\sigma]\;\UU \equiv \UU$.
+where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
 %
@@
 %Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
-%That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.
-
-To ensure that type substitution is definitional for all %\emph
-{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
+%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.
+
+To ensure that type substitution is strict for all %\emph
+{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
 In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
@@
 
-We revisit \Cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
+We revisit \cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
 \begin{example}
   The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
-  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is definitionally equal to $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$, i.e.\ the type of the term on the RHS.
-  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are definitionally equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
+  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is strictly equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
+  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are strictly equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
   Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
@@
 \begin{proof}
-  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that the following identities
+  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
   \begin{align*}
@@
   hold propositionally.
-  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.
+  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.
 
@@
 \begin{remark}\label{re:coherence-proof}
-  Note that in the above proof the first four identities hold definitionally by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
+  Note that in the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
   As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how congruence rules for type substitution is defined.
@@
 
-\LT{We may wonder if the last three identities can be made definitional.
+\LT{We may wonder if the last three identities can be made strict.
 Can we find a complete set of rules?}
@@
 
-In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes definitional. 
+In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
 Likewise, the equality constructor $[]\El$ may be replaced by 
@@
 \begin{proof}
-  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider terms 
+  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
   \[
@@
     %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
-    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ definitionally.
+    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ strictly.
   \end{enumerate}
@@
 \begin{proof}
-  By \Cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
+  By \cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
   \begin{align*}
@@
   \end{align*}
-  while other cases hold definitionally.
+  while other cases hold strictly.
   Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
-  However, all cases follow easily from \Cref{prop:correctness-1}.
+  However, all cases follow easily from \cref{prop:correctness-1}.
   For example, consider the first case:
@@
 
-For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are definitional.
-we do not need any transport since $[;]_T$ and $[]\Pi$ are definitional.
+For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
+we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
 However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
-For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositionally but not definitionally equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
+For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
 
@@
 
-As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the propositional equality $[\sigma]\,t = [\sigma]_{\Tm}\,t$ but the same argument still applies:
+As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $[\sigma]\,t = [\sigma]_{\Tm}\,t$ but the same argument still applies:
 \begin{proposition} \label{prop:correctness-3}
@@
   the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
-  hold definitionally for any telescope $\Xi$ and any type $A$ and
+  hold strictly for any telescope $\Xi$ and any type $A$ and
   \[
@@
 \begin{proof}
-  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds definitionally by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
+  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
   Take $\mathsf{idr}$ as an example, we can see 
@@
         \item[$A$ is $\El\,u$:]
-          Similar to \Cref{prop:coherence-2}, we use \Cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.
+          Similar to \cref{prop:coherence-2}, we use \cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.
 
@@
 \end{align*}
-without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become definitional.
+without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.
 
@@
 \section{The elimination principle and models} \label{sec:models}
-\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
 \LT{%
@@
 That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
-Nevertheless, the difference is insignificant, as both definitions would be derived from the signature of a quotient inductive-inductive(-recursive) type.
 \end{remark}
 
-Given an eliminator, the following clauses are derived to carry out the actual computation
+Given an eliminator, the actual computation for the elimination principle is given by
 \begin{alignat*}{5}
@@
 \end{proposition}
-\LT{Add some remark regarding the proof?}
-
-These identities are used to interpret constructors whereas definitional identities are used in syntax.
+
+These identities are needed to interpret constructors whereas strict identities are used in syntax.
 Consider eliminating an extension $\sigma, t$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
@@
 as the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
-In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ is not definitionally equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
+In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
 \[
@@
 
-\begin{remark}
+\begin{remark} \label{re:method-recursion-IIR}
 Note again that our definition is not an instance of Dybjer and Setzer's schema of inductive-recursion where simultaneously defined functions do not require corresponding interpretation as they target meta-theoretic types instead of inductive types being defined.
@@
 
-\subsection{The standard model} \label{subsec:std-model}
-Our definition has the standard model, justifying its soundness of our definition, given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
-\begin{align*}
-  \Ctx^M & \defeq \lam{\nameless} \Set &
-  \Sub^M & \defeq \lam{\Gamma^M}{\Delta^M}{\nameless} \Gamma^M \to \Delta^M \\
-  \Ty^M  & \defeq \lam{\Gamma^M}{\nameless}\Gamma^M \to \Set & 
-  \Tm^M  & \defeq \lam{\Gamma^M}{A^M}{\nameless} (\gamma : \Gamma^M) \to A^M \gamma
-\end{align*}
+\subsection{The \texorpdfstring{$\Set$}{Set}-theoretic model} \label{subsec:std-model}
+Our definition has the standard $\Set$-theoretic model, justifying the soundness of our definition, given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+\begin{alignat*}{5}
+  \Ctx^M & \defeq \lam{\nameless} \; && \Set & \qquad \Sub^M & \defeq \lam{\Gamma^M\,\Delta^M\,\nameless}\; && \Gamma^M \to \Delta^M \\
+  \Ty^M  & \defeq \lam{\Gamma^M\, \nameless} && \Gamma^M \to \Set & 
+  \Tm^M  & \defeq \lam{\Gamma^M\,A^M\,\nameless} && (\gamma : \Gamma^M) \to A^M\,\gamma
+\end{alignat*}
 with the additional methods to its QII counterpart:
 \begin{align*}
-  \blank\uparrow^M\blank & \judgeq \lam{\sigma^M}{A^M}{\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
+  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
                          & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
-  [\blank]_{\Tm}^M\blank & \judgeq \lam{\sigma^M}{t^M} (\sigma^M ; t^M)  \\
+  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M} (\sigma^M ; t^M)  \\
                          & : (\Gamma^M \to \Delta^M) \to ((\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
@@
 
-In the QII counterpart, none of identities hold definitionally, because their indices always involve equality constructors.
-For example, the identity type substitution corresponds to
+In the QII counterpart, none of identities hold strictly, because their indices always involve equality constructors.
+E.g., the equality constructor $[\idS]_{\Ty} : [\idS]\,A =^{\Ty\,\Gamma} A$ gives rise to
 \begin{alignat*}{5}
@@
 \end{alignat*}
-The transport along $[\idS] : [\idS]\,A = A$ above is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of $A$.
-Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{[\idS]}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and requires a lemma about constant transport~\cite[Lemma~2.3.5]{UFP2013} to show.
-
-In contrast, in our QIIR definition, many identities just hold definitionally.
+The transport along $[\idS]$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of $A$.
+Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs a step~\cite[Lemma~2.3.5]{UFP2013} of reasoning about the constant transport to show.
+
+In contrast, in our QIIR definition, many identities just hold strictly.
 Consider
@@
 \end{alignat*}
-where $[\idS]^M$ is just $(\lam{\gamma} A^M\,\gamma) = A^M$ for $A^M : \Gamma^M \to \Set$ in the model and holds definitionally by the $\eta$-law for $\Pi$-types.
-Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$, definitionally equal to $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) = \lam{\gamma}\gamma$, and holds definitionally by the $\eta$-law for $\Sigma$-types.
-
-Additional rules we have used to enable further reduction are derivable already, and they can be justified by the same standard model with simpler proofs.
+where $[\idS]^M$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
+Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$, strictly equal to another strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq \lam{\gamma}\gamma$ by the $\eta$-law for $\Sigma$-types.
+
+The additional rules we have used to enable further reduction are derivable already, and they can be justified by the same standard model with simpler proofs.
 Now is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?

### 2025-02-07 tex/main.tex
Fix some typos in S4.2

@@
 \begin{abstract}
-  Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
+  Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and a proof assistant.
   Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
@@
 \]
-where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
+where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
 %
@@
 
-\LT{We may wonder if the last three identities can be made strict.
-Can we find a complete set of rules?}
+\LT{We may wonder if the last three identities can be made strict.}
 
@@
 To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
-However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ \emph{already} requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
+However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
 That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
@@
 \subsection{The \texorpdfstring{$\Set$}{Set}-theoretic model} \label{subsec:std-model}
-Our definition has the standard $\Set$-theoretic model, justifying the soundness of our definition, given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+The standard $\Set$-theoretic model for our definition, justifying its soundness, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
 \begin{alignat*}{5}
@@
                          & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
-  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M} (\sigma^M ; t^M)  \\
-                         & : (\Gamma^M \to \Delta^M) \to ((\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
+  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
+                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
 \end{align*}
@@
 where $[\idS]^M$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
-Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$, strictly equal to another strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq \lam{\gamma}\gamma$ by the $\eta$-law for $\Sigma$-types.
-
-The additional rules we have used to enable further reduction are derivable already, and they can be justified by the same standard model with simpler proofs.
+Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$ and by definition is a strict equality $\lam{\gamma}(\pi_1\gamma, \pi_2\gamma) \judgeq \lam{\gamma}\gamma$ by the $\eta$-law for $\Sigma$-types.
+
+The additional rules we have used to enable further reduction are derivable in type theory, and they can be justified by the same standard model with simpler proofs.
 Now is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
 \subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
+%\subsubsection{Syntactic equivalence}
+%\subsubsection{Semantics equivalence}
+
 \section{Formal implementation} \label{sec:formal-implementation}
@@
 
-\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.}
+\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.
+It will be nice to have OTT equipped with QII(R)T on a par with HIIT (or Cubical Agda).}
 

### 2025-02-08 tex/main.tex
Minor revisions; fix a typo in S4.2

@@
 We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
+\LT{We write $\Set$ for the type of types employing the `typical ambiguity'.}
 
@@
 We have marked such ``informal'' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
+\LT{Single quote or double quote in British English?}
 
@@
 %
-For example, the type of \emph{join lists} $\JList{X}$ can be given by the QIT
+For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
 with the following constructors:
 \begin{alignat*}{3}
-  \Jnil & : \JList{X} \\
-  \Jsing{\blank} & : X \to \JList{X} \\
-  \Jconcat{\blank}{\blank} & : \JList{X} \to \JList{X} \to \JList{X} \\
-  \mathsf{idl} & : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{X}} \ell \\
-  \mathsf{idr} & : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{X}} \ell \\
-  \mathsf{assoc} & : \implicit{\ell_0, \ell_1, \ell_2}\; (\Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2}) =^{\JList{X}} (\Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})})
+  \Jnil                    & : \implicit{A}      && \JList{A} \\
+  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
+  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
+  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
+  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
+  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
 \end{alignat*}
@@
 
-We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{X}$ is defined simultaneously with a recursively defined function $\flatten : \JList{X} \to \List{X}$ which flattens a join list into an ordinary list.
+We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list.
 We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
 \[
-\mathsf{nf} : (\ell, \ell' : \JList{X}) \to \flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell' \to \ell =^{\JList{X}} \ell'
+  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'
 \]
 which states that join lists are equal as soon as their flattenings are equal.
-For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{X}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$.
+For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
 
@@
 We will also make use of such a combination, which gives additional motivation to study its metatheory in future work.
-
-\LT{How the congruence rule for the identity type can be used in QII(R)T?}
 
@@
 We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
-We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
-Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
+We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning, and develop a type theory with a universe (\cref{subsec:SC+U}) of small types, $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
 
@@
 \Fredrik{Could we change this to $A[\sigma]$ instead?}
-\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition: }
+\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
+\LT{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}
 
@@
 
-Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
-\begin{alignat*}{3}
-  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
-\end{alignat*}
-Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
+Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as (snoc-)lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
+\begin{alignat*}{3}
+  \emptysub & : \implicit{\Gamma}\;                && \Sub\;\Gamma\;\emptyctx \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
+\end{alignat*}
+Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
 We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$, because the substitution calculus is supposed to be the initial \emph{category} with families:
@@
 \begin{alignat*}{3}
-  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
-  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
+  \idS & : \implicit{\Gamma}\;                         && \Sub\;\Gamma\;\Gamma \\
+  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;&& \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
 \end{alignat*}
@@
   \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
-  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\, A)
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\, A).
 \end{alignat*}
@@
 \begin{alignat*}{3}
-  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \;A         && =^{\Ty\,\Gamma}\;A \\
-  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\;A && =^{\Ty\,\Gamma} [ \sigma ]\;([ \tau ]\;A) \\
-  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                && [ \sigma ]\;\UU        && =^{\Ty\,\Gamma} \UU
+  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \,A         && =^{\Ty\,\Gamma}\;A \\
+  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\,A && =^{\Ty\,\Gamma} [ \sigma ]\,([ \tau ]\,A) \\
+  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma} \UU
 \end{alignat*}
@@
 we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), [\sigma]\,t$.
-However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the context extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
+However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
 Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ which we can transport along to fix up the equation.
@@
 However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
-By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
+By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
 %In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
@@
 \begin{alignat*}{3}
-  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
+  [\blank]_{\Tm}\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
 [ \idS ]_{\Tm}\,t          & \reduce t \\
@@
 \begin{alignat}{3}
-  \blank\uparrow \blank & : (\sigma : \Sub\,\Gamma,\Delta) \to (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
+  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
 \idS                \uparrow A & \reduce \idS \nonumber \\
@@
 Similar to term substitution, we can show that the lifted substitution $\sigma \uparrow A$ is propositionally equal to $\sigma^+$, based on two basic properties\footnote{%
-  The reader is invited to prove them using the QII definition of type theory.}
+  The reader is invited to prove them using the QII definition with transports.}
 about the lifting.
@@
     \qquad\text{and}\qquad
-    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, [\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;(\Theta, A)} \sigma^+ ; \tau^+.
+    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, [\sigma; \tau]_{\Ty}\,A)\;(\Theta, A)} \sigma^+ ; \tau^+.
   \]
@@
 \begin{alignat*}{3}
-  \left<\blank\right> & : \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
+  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
   \left< t \right>    & \defeq (\idS , t),
@@
 With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
-Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
+Then, we can give this QIIR definition of type theory a \emph{$\Set$-theoretic model} (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
 
@@
 \begin{equation} \label{eq:elim-subext}
-  \Tm^M\;\Gamma^M\;(\elim_{\Ty}\;([\sigma]_{\Ty}\,A))\;t\judgeq \Tm^M\;\Gamma^M\;[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)\;t
+  \Tm^M\;\Gamma^M\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t\judgeq
+  \Tm^M\;\Gamma^M\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t
 \end{equation}
 as the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
-In the QIIR definition, on the contrary, the type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
+In the QIIR definition, on the contrary, the interpreted type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$, which is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
 \[
@@
 \]
-transporting $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$.
+transporting $\elim_{\Tm}\,t$ along the identity proof $\elim_{\Ty}[]$.
 
@@
 \subsection{The \texorpdfstring{$\Set$}{Set}-theoretic model} \label{subsec:std-model}
-The standard $\Set$-theoretic model for our definition, justifying its soundness, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+The $\Set$-theoretic model (also known as the \emph{standard} model) for our definition, justifying its soundness, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
 \begin{alignat*}{5}
@@
 The transport along $[\idS]$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of $A$.
-Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs a step~\cite[Lemma~2.3.5]{UFP2013} of reasoning about the constant transport to show.
+Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs one step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport to show.
 

### 2025-02-10 tex/main.tex
Describe the typical workaround

@@
   Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and a proof assistant.
-  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
+  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory in intensional type theory, with the price to be paid that terms frequently contain transports along equality constructors.
   As a result, proofs need to painstakingly account for coercions alongside the interesting part.
   This %mess
-  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
+  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation.
   In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
@@
 
-Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to quotient constructors.
+Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to equality constructors.
 %
@@
 %
-Any method that reduces the friction this introduces seem worth exploring.
+
+A typical workaround in practice (as adopted by Kaposi et al.~\cite{Kaposi2019}) is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
+%
+As the host theory is assumed to have the uniqueness of identity proof and function extensionality, this workaround is anticipatively justified by the fact~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
+%
+However, this workaround is computationally ill-behaved, as confluence does not hold any more.
+\LT[noinline]{Do they actually justify this workaround anywhere?}
+%
+Any method that reduces the friction introduced by the use of equality constructors without breaking computational properties seem worth exploring.
 %
@@
 We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
-Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags to introduce additional rewrite rules.\footnote{%
-  As shown in \cite{Cockx2021}, adding non-terminating rewrite rules does not break consistency as long as these rules are consistent, so consistency is not an issue in our development.
-}
+Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags to introduce additional rewrite rules.
 Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
@@
 %
-We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
-%
-We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
-\LT{We write $\Set$ for the type of types employing the `typical ambiguity'.}
-
-We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P(x)$, $u : P(y)$ and $p : x =^A y$.
-One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$, where $\transfibf{P} : (p : x =^A y) \to P(x) \to P(y)$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
+We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
+%
+We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
+\LT{We write $\Set$ for the type of types employing the `typical ambiguity'}
+
+We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$.
+One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
 %

### 2025-02-10 tex/main.tex
Adjust some displayed math to save space

@@
 $\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as
-\begin{alignat*}{3}
-  \emptyctx & : \Ctx \\
-  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
-\end{alignat*}
+\[
+  \emptyctx : \Ctx
+  \qquad\qquad
+  \blank,\blank : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx.
+\]
 Note that a context extension $\Gamma , A$ is only wellformed if $A : \Ty\;\Gamma$, ensuring overall wellformedness of contexts, but requiring contexts and types to be defined simultaneously.
@@
 Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as (snoc-)lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
-\begin{alignat*}{3}
-  \emptysub & : \implicit{\Gamma}\;                && \Sub\;\Gamma\;\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
-\end{alignat*}
+\[
+  \emptysub : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx
+  \qquad\qquad
+  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A).
+\]
 Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
@@
 %Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
-\begin{alignat*}{3}
-  \idS & : \implicit{\Gamma}\;                         && \Sub\;\Gamma\;\Gamma \\
-  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;&& \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
-\end{alignat*}
+\[
+  \idS : \implicit{\Gamma}\; \Sub\;\Gamma\;\Gamma
+  \qquad\qquad \blank;\blank : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta.
+\]
 These need to satisfy certain laws, which we introduce later as equality constructors.
@@
 Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
-Finally, we have term substitution, which given a substitution $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in context $\Delta$ to terms in context $\Gamma$, appropriately preserving the type, which is also substituted:
+Finally, we have term substitution, which given $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in $\Delta$ to terms in $\Gamma$, appropriately preserving the type, which is also substituted:
 \begin{alignat*}{3}
@@
 \begin{alignat*}{3}
-  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \,A         && =^{\Ty\,\Gamma}\;A \\
-  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\,A && =^{\Ty\,\Gamma} [ \sigma ]\,([ \tau ]\,A) \\
-  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma} \UU
+  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && [ \idS ] \,A           && =^{\Ty\,\Gamma}\;A \\
+  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& [ \sigma ; \tau ]\,A  && =^{\Ty\,\Gamma}\;[ \sigma ]\,([ \tau ]\,A) \\
+  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma}\;\UU
 \end{alignat*}
@@
 The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the number of terms in $\sigma$, giving rise to two $\eta$-laws --- one for $\Delta$ being empty, and one for $\Delta$ being non-empty:
-\begin{alignat*}{5}
-  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma}\; && \sigma        && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
-  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma}\; && \sigma   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma)
-\end{alignat*}
+\[
+  \emptyctx\eta : \implicit{\Gamma, A, \sigma}\; \sigma        =^{\Sub\,\Gamma\,\emptyctx} \; \emptysub
+  \qquad\qquad
+  \pi\eta       : \implicit{\Gamma, \Delta, \sigma}\; \sigma   =^{\Sub\,\Gamma\,(\Delta, A)} (\pi_1 \sigma, \pi_2 \sigma).
+\]
 
@@
   \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
+  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
 \end{alignat*}
@@
 
-The first and second projections of a non-empty substitution $(\sigma, t)$ should  obviously be $\sigma$ and $t$ respectively
+The first and second projections of $(\sigma, t)$ should obviously be $\sigma$ and $t$ respectively
 \begin{alignat*}{5}
@@
 Similarly, for term substitution, terms are transported along the corresponding rules:
-\begin{alignat*}{5}
-  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
-  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
-\end{alignat*}
+\[
+  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {[\idS]\,t} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
+  \qquad\qquad
+  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {[\sigma ; \tau]\,t} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{[ \sigma ]\,[ \tau ]\,t}
+\]
 equivalent to the following (homogeneous) identities
@@
 
-To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules strict, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.
+To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules strict, since the root cause is that type substitution does not compute as a constructor, leading to type mismatches occurring with substitution.
 
@@
 With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
-Then, we can give this QIIR definition of type theory a \emph{$\Set$-theoretic model} (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
+Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
 
 \subsection{The elimination principle} \label{subsec:eliminator}
-\LT{We shall say something about the elimination/induction principle in \cref{sec:meta}}
 
@@
 \end{alignat*}
-and simultaneously defined functions are also given by corresponding functions 
+and simultaneously defined functions are also interpreted by corresponding functions 
 \begin{alignat*}{3}
@@
   [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
-  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
+  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && t^M
 \end{alignat*}
@@
   An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
+  A \emph{recursor} is an eliminator whereas motives are constants over syntaxes.
 \end{definition}
@@
 \begin{alignat*}{5}
-  \elim_{\Ty}[] & : \implicit{\dots} & \elim_{\Ty}\,([\sigma]_{\Ty}\,A) & = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
-  \elim_{\Tm}[] & : \implicit{\dots} & \elim_{\Tm}\,([\sigma]_{\Tm}\,t) & =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t)
+  \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
+  \nameless & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
+  \nameless & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
 \end{alignat*}
@@
 
-\subsection{The \texorpdfstring{$\Set$}{Set}-theoretic model} \label{subsec:std-model}
-The $\Set$-theoretic model (also known as the \emph{standard} model) for our definition, justifying its soundness, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+\subsection{The standard model} \label{subsec:std-model}
+The standard model, justifying the soundness of our definition, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
 \begin{alignat*}{5}
@@
 
-In the QII counterpart, none of identities hold strictly, because their indices always involve equality constructors.
-E.g., the equality constructor $[\idS]_{\Ty} : [\idS]\,A =^{\Ty\,\Gamma} A$ gives rise to
-\begin{alignat*}{5}
-  [\idS]^M & : [ \idS^M ]^M_{\Ty}\, A^M && =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]}} A^M
-\end{alignat*}
-The transport along $[\idS]$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of $A$.
+In the QII counterpart, none of identities hold strictly, since their indices always involve equality constructors.
+E.g., the constructor $[\idS]_{\Ty}$ gives rise to
+$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
+and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
 Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs one step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport to show.
@@
 \begin{alignat*}{5}
-  [\idS]^M                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
+  [\idS]^M_{\Ty}                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
   \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
 \end{alignat*}
-where $[\idS]^M$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
-Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$ and by definition is a strict equality $\lam{\gamma}(\pi_1\gamma, \pi_2\gamma) \judgeq \lam{\gamma}\gamma$ by the $\eta$-law for $\Sigma$-types.
-
+where $[\idS]^M_{\Ty}$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
+Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$.
+By definition, it is a strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq (\lam{\gamma}\gamma)$ by the $\eta$-law for $\Sigma$-types.
+
+\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
 The additional rules we have used to enable further reduction are derivable in type theory, and they can be justified by the same standard model with simpler proofs.
-Now is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
-\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
-%\subsubsection{Syntactic equivalence}
-%\subsubsection{Semantics equivalence}
 

### 2025-02-11 tex/main.tex
Use single quotes instead of double quotes

@@
 
-We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the ``dependent'' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$.
+We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the `dependent' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$.
 One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
@@
 Others can only be checked  partially, because of postulated quotient inductive types.
-We have marked such ``informal'' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
+We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
 \LT{Single quote or double quote in British English?}
@@
 In contrast, an inductive-recursive type is defined inductively \emph{simultaneously} with a recursively defined function out of it.
-The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a ``decoding'' function $T : U \to \Set$ which maps each code to the type it represents.
+The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents.
 Crucially, the code for dependent function types

### 2025-02-11 tex/main.tex
S2.3 on overlapping patterns

@@
 
-\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/genetsai95/DTT-QIIRT}.\LT[noinline]{Transfer it to my account?}}%optional, e.g.
+\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/L-TChen/TTasQIIRT}}%optional, e.g.
 related research data, source code, ...
@@
 %
-We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
-\LT{We write $\Set$ for the type of types employing the `typical ambiguity'}
+We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which is not given explicitly, but instead can be figured out from context.
+We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.
 
@@
 
-We have formally implemented our definitions and proofs in Agda, with the following options: \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
-We simulate quotient inductive types using postulated equations to introduce equality constructors~\cite{Licata2011}.
-Some of our theorem statements are meta-theoretical, and thus cannot be checked by Agda.
-Others can only be checked  partially, because of postulated quotient inductive types.
+We have formally implemented our definitions and proofs in \Agda, using the options \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
+Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}.
+Some of our statements are meta-theoretical and thus not checked by \Agda.
+Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types directly.
 We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
-\LT{Single quote or double quote in British English?}
-
-%\LT{
-%Points to include:
-%\begin{enumerate}
-%  \item identity type $x =^{A} y$ for $x, y : A$, dependent identity type, $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P y} u$ for $t : P(x)$ and $u : P(y)$, heterogeneous equality~\cite{McBride1999} in \Agda $x \simeq y$ for $x : A$ and $y : B$
-%  \item We work with intensional type theory with uniqueness of identity proof and function extensionality (only used for NbE and the standard model for the extensional identity type).
-%  \item We use Agda to formally implement our definitions with the following options: \texttt{-{}-with-K}, \texttt{-{}-local-confluence-check}, \texttt{-{}-exact-split}, and \texttt{-{}-rewriting} using postulated equations~\cite{Licata2011} to introduce equality constructors for quotient types.
-%\end{enumerate}}
-%
-%\LT{The symbol \textdbend indicates a property that is proved informally or checked partially with \Agda (because of postulated quotient inductive typeswbe checked ).}
 
@@
 \subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-\cite{Cockx2014,Altenkirch2016a}
-\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
+Functions are typically defined informally using pattern matching for convenience and readability.
+Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
+This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
+For example, the following two definitions are essentially the same
+
+\begin{minipage}[t]{.5\textwidth}
+  \begin{alignat*}{5}
+    \blank  & + \blank     && : \N \to \N \to \N \\
+    0       &  + m         && = m  \\
+  n         &  + 0         && = m  \\
+  (\suc\;n) &  + m         && = \suc\;(n + m) \\
+  n         &  + (\suc\;m) && = \suc\;(n + m)
+\end{alignat*}
+\end{minipage}
+\begin{minipage}[t]{.5\textwidth}
+\begin{alignat*}{5}
+  \blank         & +'\blank     && : \N \to \N \to \N \\
+  0              & +' m         && \defeq m  \\
+  (\suc\;n)      & +' 0         && \defeq \suc\,n \\
+  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
+\end{alignat*}
+\end{minipage}
+\vskip\belowdisplayskip\noindent
+after elaboration in \Agda: each clause of $\blank +' \blank$ is strict (hence the symbol $\defeq$ is used) while clauses of $\blank+\blank$ are not necessarily strict by definition.
+This semantics for function definitions affects how definitions reduce and therefore how identities are proved.
+
+Cockx explores an alternative semantics for pattern matching~\cite{Cockx2014}, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
+Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry of $\blank + \blank$ to be proved commutative directly by induction on $n$ and $m$.
+Thus, when a function $f$ is defined with overlapping patterns, we use the notation $f\;\overrightarrow{p_i} \reduce t_i$ to emphasize its reduction behaviour.
+
+A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
+Cockx shows~\cite[Proposition~2]{Cockx2014} that if a definition of $f$ is
+\begin{enumerate}
+  \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
+  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, their right-hand sides $\sigma t_1$ and $\sigma t_2$ are strictly equal,
+\end{enumerate}
+then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
+
 \subsection{Definitions by rewrite rules}
@@
 [\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
-[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
-[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
-[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
+[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
+[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
+[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
 \end{alignat}
-However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
-By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
+As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
 %In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
@@
 This option is turned on globally in our formal implementation.}
+
+\begin{proposition}[Termination]
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+  \danger
+\end{proposition}
+\begin{proof}
+  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
+\end{proof}
 \begin{proposition}[Local confluence] \label{prop:local-confluence-1}
@@
   It is easy to see that remaining local peaks reduce to the same term.
-\end{proof}
-\begin{proposition}[Termination]
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
-  \danger
-\end{proposition}
-\begin{proof}
-  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
 \end{proof}
@@
 to retain the local confluence for type substitution:
+\begin{proposition}[Termination]
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+  \danger
+\end{proposition}
+\begin{proof}
+  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
+\end{proof}
 \begin{proposition}[Local confluence] \label{prop:local-confluence-2}
@@
   Therefore, type substitution is locally confluent.
-\end{proof}
-\begin{proposition}[Termination]
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
-  \danger
-\end{proposition}
-\begin{proof}
-  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
 \end{proof}
@@
 
+\begin{proposition}[Termination]
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+  \danger
+\end{proposition}
 \begin{proposition}[Local confluence]\label{prop:local-confluence-3}
@@
 \end{proof}
-\begin{proposition}[Termination]
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
-  \danger
-\end{proposition}
 

### 2025-02-11 tex/main.tex
S2.4 on rewrite rules

@@
 Some of our statements are meta-theoretical and thus not checked by \Agda.
-Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types directly.
+Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types.
 We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
@@
   \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
-  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, their right-hand sides $\sigma t_1$ and $\sigma t_2$ are strictly equal,
+  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, terms $\sigma t_1$ and $\sigma t_2$ have the same normal form,
 \end{enumerate}
 then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
-
-\subsection{Definitions by rewrite rules}
-\cite{Cockx2020,Cockx2021}
-
-Unfortunately, neither general schemata of quotient inductive-inductive-recursive types nor definitions by overlapping patterns have been developed or implemented in existing proof assistants.
-
-\paragraph*{Local confluence}
-\LT{Explain the notion of critical pair; local peak}
-\paragraph*{Strong normalisation}
-\cite{Abel2002}
-\LT{Explain how these two conditions imply global confluence}
+We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ a \emph{local peak} informally.
+
+\subsection{Overlapping patterns via adding rewrite rules}
+Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
+To emulate a function by overlapping patterns in \Agda, we declare each function clause an additional rewrite rules \cite{Cockx2020,Cockx2021} (via the option \texttt{-{}-rewriting}) and use the local confluence checker (via \texttt{-{}-local-confluence-check}).
+This is indeed one of the main use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
+The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.
 

### 2025-02-11 tex/main.tex
Section 3.4.1

@@
 \begin{abstract}
-  Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and a proof assistant.
+  Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and proof assistants.
   Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory in intensional type theory, with the price to be paid that terms frequently contain transports along equality constructors.
@@
   In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
-  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our definition with quotient inductive-inductive definitions.
+  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our development with one using quotient inductive-inductive types.
 \end{abstract}
@@
 
-\subsection{Quotient inductive-inductive types}
+\subsection{Quotient inductive-inductive and quotient inductive-recursive types}
 % \subsection{Quotient inductive(-inductive)-recursive types}
@@
 \begin{proposition}[Coherence]
-  For every $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
+  For every constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
   \danger
@@
   \]
-  hold for any type $A$ and any term $t$.
+  for any type $A$ and any term $t$.
 \end{proposition}
@@
 \subsubsection{Multiple lifting and the coherence of type substitution}
-\Fredrik{This section gets rather technical. Keeping the page limit in mind, I wonder if we should only briefly summarise it in the main text?}
-\LT{Move this to the appendix?}
-
-For the coherence of~$[\blank]_{\Ty}\blank$, we have to generalise the desired identity to account for the lifting, since a naive induction does not work: for example, unfolding $[\sigma]_{\Ty}\,(\Pi\;A\;B)$ leads to
-\[
-  \Pi\;([\sigma]_{\Ty}\,A)\;([\sigma\uparrow A]_{\Ty}\,B)
-\]
-where the induction hypothesis $[\sigma]_{\Ty}\,B = [\emptysub]_{\Ty}\,B$ cannot be applied.
+
+We also need to prove the coherence for type substitution, i.e., that $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
+To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
+\begin{align*}
+  [\sigma]_{\Ty}\,(\Pi\;A\;B) & = \Pi\;([\sigma]_{\Ty}\,A)\;([\sigma\uparrow A]_{\Ty}\,B) \\
+  [\tau]_{\Ty}\,(\Pi\;A\;B) & = \Pi\;([\tau]_{\Ty}\,A)\;([\tau\uparrow A]_{\Ty}\,B)
+\end{align*}
+but the induction hypothesis does not give us $[\sigma\uparrow A]_{\Ty}\,B = [\tau\uparrow A]_{\Ty}\,B$.
 Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
-Then, we show the identity
+Then, for each constructor $p : \sigma = \tau$, we show the identity
 $[\sigma \mathop{\upuparrows} \Xi ]_{\Ty}\,A = [\tau \mathop{\upuparrows} \Xi ]_{\Ty}\,A$
-for arbitrary $\Xi$, subsuming the special case $[\sigma ]_{\Ty}\,A = [\tau ]_{\Ty}\,A$ for $\Xi$ being empty.
-
-First, define an inductive-recursive type of \emph{telescopes} under a context $\Gamma$ simultaneously with a concatenation $\blank \dplus \blank$ as follows.
+for arbitrary $\Xi$, subsuming the special case $[\sigma ]_{\Ty}\,A = [\tau ]_{\Ty}\,A$ when $\Xi$ is empty.
+
+The type of telescopes over a ``base'' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.
 
@@
 \vskip\belowdisplayskip\noindent
-Then, we define the multiple lifting $\blank \upuparrows \blank$.
-Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$, should be
+Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
 \[
@@
 \]
-asking for telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
-Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
+which requires us to also simultaneously define telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$.
+
 \begin{minipage}{.5\linewidth}
@@
 To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows [\tau]_{\Tel} \Xi); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
-
-Instead of resorting to \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding following clauses:
+%
+Instead of resorting to \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
 
@@
 
-Now, we proceed with the coherence property for the recursive functions introduced so far.
-The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
-\begin{proposition}[Coherence, I]
+Of course, we also have to prove coherence for the newly introduced recursive functions to verify that they are well defined.
+\begin{proposition}[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
   \danger
-  For the following equality constructors $p : \sigma = \tau$
-  \[
-    \mathsf{idr} : \sigma;\idS = \sigma, \quad
-    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
-    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
-  \]
-  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
-  hold strictly for any telescope $\Xi$ and any type $A$ and
-  \[
-    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
-    \tau   \upuparrows \Xi
-  \]
-  holds propositionally for any telescope $\Xi$.
-\end{proposition}
-\begin{proof}
-  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
-  Take $\mathsf{idr}$ as an example, we can see 
-  \begin{align*}
-    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
-                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
-                                   & = \sigma \upuparrows \Xi
-  \end{align*}
-  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.
-
-  For other two equality constructors, the arguments follow similarly.
-\end{proof}
-The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
-\begin{proposition}[Coherence, II]
-  \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
-  For the following equality constructors $p : \sigma = \tau$
-  \begin{alignat*}{5}
-    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
-    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
-  \end{alignat*}
-   the identities
-  \danger
-  \begin{align*}
-  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
-                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
-  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
-%    &
-%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
-  \end{align*}
-  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
-\end{proposition}
-\begin{proof}
-  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.
-
-  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$ and the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
-  \begin{enumerate}
-    \item Consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$ by induction on $A$.
-      \begin{description}
-        \item[$A$ is $\UU$:] By definition
-          \[
-            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
-          \]
-          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
-          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
-          \LT{Justify the use of congruence rule for QIIT, e.g. $\transfib{\Ty}{p}{\UU_{\Gamma}} = \UU_{\Delta}$ for $p : \Gamma = \Delta$}
-
-        \item[$A$ is $\El\,u$:]
-          Similar to \cref{prop:coherence-2}, we use \cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.
-
-        \item[$A$ is $\Pi\,B\,C$:]
-          Note that one of the induction hypothesises is
-          \[
-            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
-          \]
-          for \emph{arbitrary} $\Xi$, so we can apply the induction hypothesis to $B$ and $C$ below:
-          \begin{align*}
-            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
-            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
-            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
-            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by induction hypothesises} \\
-            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
-            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
-          \end{align*}
-      \end{description}
-    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
-      \begin{description}
-        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
-        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
-          \begin{align*}
-            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
-                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
-                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
-          \end{align*}
-      \end{description}
-    \item By induction on $\Xi$, consider the following cases for $\pi_1(\sigma, t) \upuparrows \Xi = \sigma \upuparrows\,\Xi$.
-      \begin{description}
-        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
-        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
-          \begin{align*}
-            \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
-                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
-                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
-                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
-                                                  & \judgeq \sigma \upuparrows (\Xi , A)
-          \end{align*}
-      \end{description}
-  \end{enumerate}
-
-\end{proof}
+  For any equality constructor $p : \sigma = \tau$, we have
+  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$
+  for any telescope $\Xi$.
+\end{proposition}
+
+As promised, by instantiating $\Xi$ to be empty, we also get coherence for $[\blank]_{\Ty}$.
 
@@
 
-We conclude this extension with $\Pi$-types by noting that single term substitution
-\begin{alignat*}{3}
-  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
-  \left< t \right>    & \defeq (\idS , t),
-\end{alignat*}
-used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
-can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
-Similarly, the structural rule for substitution on $\mathsf{app}$ can be derived
+We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also ``ordinary'' application
+$t \mathop{\$} u \defeq [ (\idS , t) ]_{\Tm} (\mathsf{app}\,t)$ of type $\Tm\,\Gamma\,([(\idS , t)] B)$ can be defined without any transport.
+This is in contrast to the pure QII situation, where application is defined as $[ (\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}}) ]_{\Tm} (\mathsf{app}\,t)$.
+%
+%\begin{alignat*}{3}
+%  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
+%  \left< t \right>    & \defeq (\idS , t),
+%\end{alignat*}
+%used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
+%can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
+Similarly, the structural rule for substitution on $\mathsf{app}$
 \begin{align*}
@@
 \end{align*}
-without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.
+can be derived without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.
 
@@
 
+\section{Omitted proofs}
+
+\Fredrik{Do properly with restate-thm etc}
+The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
+\begin{proposition}[Coherence, I]
+  \danger
+  For the following equality constructors $p : \sigma = \tau$
+  \[
+    \mathsf{idr} : \sigma;\idS = \sigma, \quad
+    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
+    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
+  \]
+  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
+  hold strictly for any telescope $\Xi$ and any type $A$ and
+  \[
+    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
+    \tau   \upuparrows \Xi
+  \]
+  holds propositionally for any telescope $\Xi$.
+\end{proposition}
+\begin{proof}
+  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
+  Take $\mathsf{idr}$ as an example, we can see 
+  \begin{align*}
+    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
+                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
+                                   & = \sigma \upuparrows \Xi
+  \end{align*}
+  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.
+
+  For other two equality constructors, the arguments follow similarly.
+\end{proof}
+The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
+\begin{proposition}[Coherence, II]
+  \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
+  For the following equality constructors $p : \sigma = \tau$
+  \begin{alignat*}{5}
+    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
+    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
+  \end{alignat*}
+   the identities
+  \danger
+  \begin{align*}
+  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
+                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
+  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
+%    &
+%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
+  \end{align*}
+  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
+\end{proposition}
+\begin{proof}
+  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.
+
+  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$ and the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
+  \begin{enumerate}
+    \item Consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$ by induction on $A$.
+      \begin{description}
+        \item[$A$ is $\UU$:] By definition
+          \[
+            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
+          \]
+          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
+          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
+          \LT{Justify the use of congruence rule for QIIT, e.g. $\transfib{\Ty}{p}{\UU_{\Gamma}} = \UU_{\Delta}$ for $p : \Gamma = \Delta$}
+
+        \item[$A$ is $\El\,u$:]
+          Similar to \cref{prop:coherence-2}, we use \cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.
+
+        \item[$A$ is $\Pi\,B\,C$:]
+          Note that one of the induction hypothesises is
+          \[
+            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
+          \]
+          for \emph{arbitrary} $\Xi$, so we can apply the induction hypothesis to $B$ and $C$ below:
+          \begin{align*}
+            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
+            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
+            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
+            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by induction hypothesises} \\
+            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
+            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
+          \end{align*}
+      \end{description}
+    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
+      \begin{description}
+        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
+        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
+          \begin{align*}
+            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
+                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
+                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
+          \end{align*}
+      \end{description}
+    \item By induction on $\Xi$, consider the following cases for $\pi_1(\sigma, t) \upuparrows \Xi = \sigma \upuparrows\,\Xi$.
+      \begin{description}
+        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
+        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
+          \begin{align*}
+            \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
+                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
+                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
+                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
+                                                  & \judgeq \sigma \upuparrows (\Xi , A)
+          \end{align*}
+      \end{description}
+  \end{enumerate}
+
+\end{proof}
+
 \section{Formal definitions in Agda}

### 2025-02-11 tex/main.tex
Draft Section 6

@@
 \LT{Compare some formal proofs with pen-and-paper proofs here}
-\section{Discussion}
-\subsection{Related work}
-\paragraph*{Formalisation of type theory in type theory}
-\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
-\cite{Altenkirch2017}
-\cite{Munoz1998}
-\cite{Dybjer1996,Castellan2021}
-\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
-
-Substitution calculus is closely related to \emph{categories with families}~\cite{Dybjer1996}.
-\paragraph*{Quotient inductive-recursive definition of single substitution calculus}
-\cite{Kaposi2023,Kaposi2024a}
-\paragraph*{Schemata of inductive types}
-\cite{Kaposi2019}
-\LT{How does the general schema in \cite{Kaposi2019} model equality constructors?
+\section{Conclusions, and related and future work}
+
+We have used quotient inductive-recursive types and overlapping
+patterns to represent type theory in type theory. By defining type
+substitution recursively with overlapping patterns, we can avoid most
+transports that show up in terms in other formalisations of type
+theory in type theory. In other words, we have shown that so-called
+`transport hell' can be minimised when formalising type theory in type
+theory this way, without hiding any details in the formalisation.
+
+%\subsection{Related work}
+%\paragraph*{Formalisation of type theory in type theory}
+
+The idea of representing type theory in type theory in a well typed manner goes back to Dybjer~\cite{Dybjer1996} (see also Clairambault and Dybjer~\cite{Clairambault2014} and Castellan, Clairambault and Dybjer~\cite{Castellan2021} for recent developments).
+Dybjer's notion of \emph{category with families} can be seen as a semantic structure corresponding closely to the syntax of type theory, and many of the formalisations of type theory in type theory can be understood as capturing the initial category with families via inductive or quotient-inductive types.
+The equations involved are very similar to the equations described in Martin-L\"of's substitution calculus~\cite{Martin-Lof1992}.
+
+Like us, Danielsson~\cite{Danielsson2006} defines type substitution by a recursive function, while keeping term substitution a constructor.
+However because he does not use quotients, he also has an explicit term constructor for transporting terms along an inductively defined type of conversion rules, which explicitly contains congruence rules.
+We are able to remove a lot of the boilerplate by using actual equality constructors instead.
+Chapman~\cite{Chapman2009} similarly uses explicit term substitutions and explicit transports along conversion rules, but also uses explicit type substitition in favour of more straightforward foundations.
+McBride~\cite{McBride2010} offers another solution which uses an inductive-recursive type to reuse the equality of the metatheory rather than using explicit transports.
+This works well if the object theory and meta theory are similar, but does not allow to represent equalities that do not hold in the metatheory.
+
+Altenkirch and Kaposi~\cite{Altenkirch2016a} were the first to exploit QIITs to simplify the treatment of computation rules in representations of type theory in type theory.
+We build heavily on their ideas, but by using Danielsson's~\cite{Danielsson2006} idea of defining type substitution recursively, we are able to remove many explicit transports from the definition, with the result that it is closer to an informal exposition on paper.
+Such a benefit was also suggested by Kaposi's~\cite{Kaposi2023} work, where the simply typed lambda calculus was used as an example.
+Altenkirch and Kaposi has built on the QIIT representation with great success to derive non-trivial results such as normalisation by evaluation~\cite{Altenkirch2017}, and investigate the metatheory of QIITs themselves, together with Kov\'acs~\cite{Kaposi2019}.
+
+%\cite{Munoz1998}
+
+%\paragraph*{Quotient inductive-recursive definition of single substitution calculus}
+
+\Fredrik{As far as I can tell, ``Type theory in type theory with single substitutions''~\cite{Kaposi2024a} is just a standard QIIT --- so maybe no need to discuss?}
+%\cite{Kaposi2023,Kaposi2024a}
+
+%\paragraph*{Schemata of inductive types}
+%\cite{Kaposi2019}
+%\LT{How does the general schema in \cite{Kaposi2019} model equality constructors?
 extensionally?
@@
 
-\subsection{Future work}
-\LT{Remember that it is for further developments.}
-\paragraph*{Type theory in type theory}
-\cite{Pujet2022,Pujet2024,Pujet2022a}
-\LT{OTT is a concrete instance of a type theory with UIP and function extensionality.}
-
-\paragraph*{Semantics and design for definitions by overlapping patterns}
-\LT{Overlapping patters are discussed by Cocks \cite{Cockx2014}, but we would like to integrate it with the schema of QIIRT.}
-\LT{\cite{Eremondi2025} has recently hinted a possible direction for overlapping patterns using sheaf theory.}
-
-\paragraph*{A schema of quotient inductive-inductive-recursive types in observation type theory}
-\LT{What a general schema should have?  inductive types should be allowed as the codomain of a recursion.}
-
-\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.
-It will be nice to have OTT equipped with QII(R)T on a par with HIIT (or Cubical Agda).}
+%\subsection{Future work}
+
+In the future, we hope to settle the foundational status of the
+combination of quotient types, inductive-inductive types, and the kind
+of generalised inductive-recursive types we have considered here,
+including overlapping patterns. In fact, our original motivation for
+embarking on this work was seeking examples that could give evidence
+that such a combination could be useful.
+%
+Various schemas for inductive types have been developed in recent
+years~\cite{NordvallForsberg2017,Kaposi2018,Kovacs2023}, and it seems
+promising to see if they can be extended to also handle quotient
+inductive-inductive-recursive types. Overlapping patterns are
+discussed by Cockx~\cite{Cockx2014}, but we would like to integrate it
+with such a schema. Eremondi and Kammar~\cite{Eremondi2025} has
+recently given a promising possible direction for overlapping patterns
+using sheaf theory.
+
+%\paragraph*{Type theory in type theory}
+
+%\paragraph*{Semantics and design for definitions by overlapping patterns}
+
+%\paragraph*{A schema of quotient inductive-inductive-recursive types in observation type theory}
+%\LT{What a general schema should have?  inductive types should be allowed as the codomain of a recursion.}
+
+%\LT{OTT is a concrete instance of a type theory with UIP and function extensionality.}
+%\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.
+%It will be nice to have OTT equipped with QII(R)T on a par with HIIT (or Cubical Agda).}
+%\cite{Pujet2022,Pujet2024,Pujet2022a}
 

### 2025-02-12 tex/main.tex
Some points discussed during the meeting and revisions

@@
 
-A typical workaround in practice (as adopted by Kaposi et al.~\cite{Kaposi2019}) is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
-%
-As the host theory is assumed to have the uniqueness of identity proof and function extensionality, this workaround is anticipatively justified by the fact~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
-%
-However, this workaround is computationally ill-behaved, as confluence does not hold any more.
-\LT[noinline]{Do they actually justify this workaround anywhere?}
-%
-Any method that reduces the friction introduced by the use of equality constructors without breaking computational properties seem worth exploring.
+A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
+%
+As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by Hofmann's result~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
+%
+However, this workaround is computationally ill-behaved, as confluence does not hold in general.
+Also, the presentation often differs from the actual definition of type theory in intensional type theory.
+%
+Any method that reduces the friction introduced by the use of equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
 %
@@
     \end{itemize}
+   with their elimination principles and the standard model, 
    using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
-  \item Comparison with other definitions using quotient inductive-inductive types.
+  \item Comparison with definitions using quotient inductive-inductive types.
 \end{itemize}
@@
 \subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-Functions are typically defined informally using pattern matching for convenience and readability.
-Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
-This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
-For example, the following two definitions are essentially the same
-
+\begin{figure} 
 \begin{minipage}[t]{.5\textwidth}
@@
   0              & +' m         && \defeq m  \\
-  (\suc\;n)      & +' 0         && \defeq \suc\,n \\
+  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
   (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
@@
 \end{minipage}
-\vskip\belowdisplayskip\noindent
-after elaboration in \Agda: each clause of $\blank +' \blank$ is strict (hence the symbol $\defeq$ is used) while clauses of $\blank+\blank$ are not necessarily strict by definition.
-This semantics for function definitions affects how definitions reduce and therefore how identities are proved.
-
-Cockx explores an alternative semantics for pattern matching~\cite{Cockx2014}, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
-Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry of $\blank + \blank$ to be proved commutative directly by induction on $n$ and $m$.
-Thus, when a function $f$ is defined with overlapping patterns, we use the notation $f\;\overrightarrow{p_i} \reduce t_i$ to emphasize its reduction behaviour.
+\caption{Two seemingly different definitions of addition}
+\label{fig:additions}
+\end{figure}
+Functions are typically defined informally using pattern matching for convenience and readability.
+Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
+This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
+For example, two definitions of addition in \cref{fig:additions} are the same
+after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$ and makes the fourth clause useless.
+Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
+This semantics for function definitions decides how definitions reduce and therefore how identities are proved.
+
+Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
+Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
+Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
 
 A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
-Cockx shows~\cite[Proposition~2]{Cockx2014} that if a definition of $f$ is
+Cockx shows that if a definition of $f$ is
 \begin{enumerate}
-  \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
+  \item \emph{structurally recursive}~\cite{Abel2002}, i.e.\ arguments of recursive calls are structurally smaller, and
   \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, terms $\sigma t_1$ and $\sigma t_2$ have the same normal form,
@@
 then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
-We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ a \emph{local peak} informally.
-
-\subsection{Overlapping patterns via adding rewrite rules}
+We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.
+
+\subparagraph*{Implementing overlapping patterns via rewrite rules}
 Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
@@
 \end{proposition}
-\begin{proof}
-  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
-\end{proof}
+%\begin{proof}
+%  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
+%\end{proof}
 \begin{proposition}[Local confluence] \label{prop:local-confluence-1}
@@
 As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
-\LT[noinline]{Should we mention any general condition for coherence?}
-\Fredrik[noinline]{Is there one?}
-\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
+%\LT[noinline]{Should we mention any general condition for coherence?}
+%\Fredrik[noinline]{Is there one?}
+%\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
 \begin{proposition}[Coherence]\label{prop:coherence-1}
@@
 \begin{remark}\label{re:coherence-proof}
-  Note that in the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three identities follow from \eqref{eq:def-type-subst-5} by induction on $A$.
-  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how congruence rules for type substitution is defined.
+  In the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three follow from \eqref{eq:def-type-subst-5} by induction on $A$.
+  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how type substitution is defined.
 \end{remark}
-
-\LT{We may wonder if the last three identities can be made strict.}
 
@@
 \end{proposition}
-\begin{proof}
-  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
-\end{proof}
+%\begin{proof}
+%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
+%\end{proof}
 \begin{proposition}[Local confluence] \label{prop:local-confluence-2}
@@
 \end{alignat}
-where $\sigma^+ \defeq (\pi_1\idS ; \sigma, \pi_2\idS)$.
+where $\sigma^+ \defeq (\pi_1\idS ; \sigma), \pi_2\idS$.
 Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
@@
   For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
-  $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
+  $\sigma \uparrow A$ is propositionally equal to $\sigma^+$. 
 \end{proposition}
@@
 
-The type of telescopes over a ``base'' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.
+The type of telescopes over a `base' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.
 
@@
 \begin{alignat*}{5}
-  & [ \idS   ]_{\Tel}       \,\Xi && \reduce \Xi \\
-  & [ \sigma ; \tau ]_{\Tel}\,\Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
+  & [ \idS   ]_{\Tel}       \,&& \Xi && \reduce \Xi \\
+  & [ \sigma ; \tau ]_{\Tel}\,&& \Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \idS          \upuparrows \Xi && \reduce \idS \\
-  & (\sigma;\tau) \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
+  & \idS          && \upuparrows \Xi && \reduce \idS \\
+  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
 \end{alignat*}
@@
 
-We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also ``ordinary'' application
+We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also `ordinary' application
 $t \mathop{\$} u \defeq [ (\idS , t) ]_{\Tm} (\mathsf{app}\,t)$ of type $\Tm\,\Gamma\,([(\idS , t)] B)$ can be defined without any transport.
@@
 }
-With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
-Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).
+With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
+Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart (\cref{subsec:equivalence}).
 
@@
   [\blank]_{\Ty}^M\blank & : \implicit{\dots} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
-  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
+  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
+  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to
+    \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)
 \end{alignat*}
@@
 
-Interestingly, the key difference from its QII counterpart is that substitution operations $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}$ are functions rather than constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
+Interestingly, the key difference from its QII counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
 %type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
@@
 \end{equation}
-as the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
+where the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
 In the QIIR definition, on the contrary, the interpreted type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$, which is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
@@
 \begin{remark} \label{re:method-recursion-IIR}
-Note again that our definition is not an instance of Dybjer and Setzer's schema of inductive-recursion where simultaneously defined functions do not require corresponding interpretation as they target meta-theoretic types instead of inductive types being defined.
+Note again that our definition is not an instance of Dybjer and Setzer's schema of inductive-recursion where simultaneously defined functions do not require corresponding interpretation, as they target meta-theoretic types instead of inductive types being defined.
 \end{remark}
@@
 
-In the QII counterpart, none of identities hold strictly, since their indices always involve equality constructors.
+In the QII counterpart, none of identities hold strictly: their indices always involve equality constructors.
 E.g., the constructor $[\idS]_{\Ty}$ gives rise to
@@
 and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
-Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs one step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport to show.
+Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.
 
@@
 
+In general, we can translate the QIIR definition of type theory to its QII definition
+
+\LT{Mention the NbE for SC here}
 \section{Formal implementation} \label{sec:formal-implementation}
-\LT{Mention the NbE for SC here}
 \LT{Compare some formal proofs with pen-and-paper proofs here}
@@
 transports that show up in terms in other formalisations of type
-theory in type theory. In other words, we have shown that so-called
+theory in (intensional) type theory. In other words, we have shown that so-called
 `transport hell' can be minimised when formalising type theory in type
@@
 
-%\cite{Munoz1998}
-
-%\paragraph*{Quotient inductive-recursive definition of single substitution calculus}
-
-\Fredrik{As far as I can tell, ``Type theory in type theory with single substitutions''~\cite{Kaposi2024a} is just a standard QIIT --- so maybe no need to discuss?}
-%\cite{Kaposi2023,Kaposi2024a}
-
-%\paragraph*{Schemata of inductive types}
-%\cite{Kaposi2019}
-%\LT{How does the general schema in \cite{Kaposi2019} model equality constructors?
-extensionally?
-Can their schemata manifest the difference of our QIIR definition and the QII definition in \cite{Altenkirch2016a}?}
-
-%\subsection{Future work}
-
 In the future, we hope to settle the foundational status of the
@@
 Various schemas for inductive types have been developed in recent
-years~\cite{NordvallForsberg2017,Kaposi2018,Kovacs2023}, and it seems
+years~\cite{Ghani2017,Kaposi2018,Kovacs2023}, and it seems
 promising to see if they can be extended to also handle quotient
@@
 
-%\paragraph*{Type theory in type theory}
-
-%\paragraph*{Semantics and design for definitions by overlapping patterns}
-
-%\paragraph*{A schema of quotient inductive-inductive-recursive types in observation type theory}
-%\LT{What a general schema should have?  inductive types should be allowed as the codomain of a recursion.}
-
-%\LT{OTT is a concrete instance of a type theory with UIP and function extensionality.}
-%\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.
-%It will be nice to have OTT equipped with QII(R)T on a par with HIIT (or Cubical Agda).}
-%\cite{Pujet2022,Pujet2024,Pujet2022a}
-
-\IfFileExists{./reference.bib}{\bibliography{reference}}{\bibliography{ref}}
-%\bibliography{ref}
+%\IfFileExists{./reference.bib}{\bibliography{reference}}{\bibliography{ref}}
+\bibliography{ref}
 

### 2025-02-12 tex/main.tex
Appendix on complete definitions
Make the definitions in Appendix consistent with the definitions in Agda

@@
 \begin{alignat*}{3}
-  \Pi     &: \implicit{\Gamma}            && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
-  \lambda &: \implicit{\Gamma, A, B}      && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
-  \mathsf{app} &: \implicit{\Gamma, A, B} && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
+  \Pi     &: \implicit{\Gamma}           \; && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
+  \lambda &: \implicit{\Gamma, A, B}     \; && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
+  \mathsf{app} &: \implicit{\Gamma, A, B}\; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  \Pi\beta           & : \implicit{\Gamma, A, B, t} && \mathsf{app}\,(\lambda\,t) && =^{\Tm\,\Gamma,A} && t \\
-  \Pi\eta            & : \implicit{\Gamma, A, B, t} && t                        && =^{\Tm\,\Gamma,A} && \lambda(\mathsf{app}\,t)
+  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
+  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                          && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
 \end{alignat*}
@@
 \begin{alignat}{5}
-  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
-  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t} && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
+  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
+  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
 \end{alignat}
@@
                      & [ \sigma ]_{\Ty}\,(\Pi\,A\,B) && \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \label{eq:def-type-subst-9} \\
-  \mathsf{[]\lambda} & : [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]_{\Tm}\,t) \nonumber
+  \mathsf{[]\lambda} :{} & [\sigma]\, (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]\,\,t) \nonumber
 \end{alignat}
@@
 \begin{align*}
-  [\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
+  [\sigma \uparrow A ]\,(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]\,(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
                                               & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
@@
 \section{Complete definitions}
+We give the QIIR definition for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
+
+\paragraph*{Substitution calculus with a base type $\UU$}
 \begin{alignat*}{3}
@@
   \Sub      & : && \Ctx \to \Ctx \to \Set \\
-  \Tm       & : && (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set \\
+  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set \\
+  [\blank]_{\Ty}\blank & : \implicit{\Gamma, \Delta}\; && \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
   \emptyctx & : && \Ctx \\
-  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx \\
-  [\blank]\blank & : \implicit{\Gamma, \Delta} && \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
-  \emptysub & : \implicit{\Gamma} && \Sub\,\Gamma\,\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,([ \sigma ]_{\Ty} A) \to \Sub\,\Gamma,(\Delta, A) \\
-  \idS & : \implicit{\Gamma} && \Sub\,\Gamma\,\Gamma \\
-  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
-  \pi_1 & : \implicit{\Gamma, \Delta, A} && \Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
-  \pi_2 & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \pi_1\,\sigma ]\, A) \\
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]\, A) \\
-  \UU     & : \implicit{\Gamma} && \Ty\, \Gamma \\
-  \El     & : \implicit{\Gamma} && \Tm\,\Gamma\,U \to \Ty\,\Gamma \\
-  \Pi     & : \implicit{\Gamma} && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
-\end{alignat*}
-
-\begin{alignat*}{5}
-[ \idS ]_{\Ty}\,A             & \reduce A \\
-[ \sigma ; \tau ]_{\Ty}\,A    & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
-[ \pi_1(\sigma, t) ]_{\Ty}\,A & \reduce [\sigma]_\Ty\,A \\
-[ \pi_1(\sigma; \tau) ]_{\Ty}\,A & \reduce [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
-[ \sigma ]_{\Ty}\,\UU           & \reduce \UU \\
-[ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm}{u}) \\
-[ \sigma ]_{\Ty}\,(\Pi\,A\,B) & \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
+  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx \\
+  \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;([ \sigma ]_{\Ty}\,A) \to \Sub\;\Gamma,(\Delta, A) \\
+  \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
+  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
+  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\;\sigma ]_{\Ty}\, A) \\
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\, A) \\
+  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
+\end{alignat*}
+\begin{alignat*}{5}
+  & [ \idS ]_{\Ty}\,               && A           && \reduce A \\
+  & [ \sigma ; \tau ]_{\Ty}\,      && A           && \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
+  & [ \pi_1(\sigma, t) ]_{\Ty}\,   && A           && \reduce [\sigma]_\Ty\,A \\
+  & [ \pi_1(\sigma; \tau) ]_{\Ty}\,&& A           && \reduce [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
+  & [ \sigma ]_{\Ty}\,             && \UU         && \reduce \UU
+\end{alignat*}
+\begin{alignat*}{5}
+  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
+  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
+  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
+  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
+  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  [\idS]t         & : \implicit{\Gamma, A, t}\; && {[\idS]\;t}          && =^{\Tm\;\Gamma\;A} && t \\
+  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {[\sigma ; \tau]\;t} && =^{\Tm\;\Gamma\;[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
+\end{alignat*}
+
+\paragraph*{Universe of small types}
+\begin{alignat*}{5}
+  [\blank]_{\Tm}\blank            & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\,A) \\ 
+ [ \idS ]_{\Tm}\,t                & \reduce t \\
+ [ \sigma ; \tau ]_{\Tm}\,t       & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
+ [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \reduce [\sigma]_\Tm\,t \\
+ [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
+ [ \sigma ]_{\Tm}\,t              & \reduce [ \sigma ]\,t, \quad \text{otherwise}
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \El       & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \nameless & : [ \sigma ]_{\Ty}\,  && (\El\, u) \reduce \El\,([\sigma]_{\Tm}{u})
+\end{alignat*}
+
+\paragraph*{$\Pi$-types}
+\begin{alignat*}{5}
+  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
+  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
+\end{alignat*}
+\begin{alignat*}{5}
+  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
 \idS                \uparrow A  & \reduce \idS \\
@@
 \pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
-\sigma              \uparrow A & \reduce (\pi_1 \idS ; \sigma , \pi_2 \idS) & \text{otherwiese} \\
- [ \idS ]_{\Tm}\,t                & \reduce t \\
- [ \sigma ; \tau ]_{\Tm}\,t       & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
- [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \reduce [\sigma]_\Tm\,t \\
- [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
- [ \sigma ]_{\Tm}\,t              & \reduce [ \sigma ]\,t & \text{otherwise}
-\end{alignat*}
-
-\begin{alignat*}{5}
-  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
-  \emptyctx\eta   & : \implicit{\Gamma, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
-  \pi\eta         & : \implicit{\Gamma, A, \sigma} && \sigma                   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : \implicit{\Gamma, A, t} && {[\,\idS\,]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
-  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\,\sigma ; \tau\,]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\,[ \tau ]\,t} \\
+\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwiese}
+\end{alignat*}
+\begin{alignat*}{5}
+  \Pi             & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
+  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
+  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
+\end{alignat*}
+\begin{alignat*}{5}
+  \nameless       &: && [ \sigma ]_{\Ty}\, (\Pi\,A\,B) && {} \reduce {} && \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
+  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& [ \sigma ] \,(\lambda\,t) && {} =^{\Tm\;\Gamma\;([\sigma]_{\Ty}(\Pi\,A\,B)} {} && \lambda ([\sigma \uparrow A ]\,t)\\
+  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\;\Gamma\;A} && t \\
+  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
+\end{alignat*}
+
+\paragraph*{Extensional identity types}
+
+\begin{alignat*}{5}
+  \idtype[\blank]{\blank}{\blank} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
+  \nameless & : && [ \sigma ]_{\Ty}\, \idtype[a]{t}{u} \reduce \idtype[{[\sigma]_{\Tm}\,a}]{[\sigma]_{\Tm}\,t}{[\sigma]_{\Tm}\,u} \\
+  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
 \end{alignat*}

### 2025-02-13 tex/main.tex
Draft S4.3

@@
 
-We work in intensional Martin-L\"of type theory with uniqueness of identity proofs.
+We work in intensional Martin-L\"of type theory with uniqueness of identity proofs (\emph{UIP} for short).
 In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
@@
 %
-Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
+Since we assume UIP globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
 %
@@
 \begin{figure} 
-\begin{minipage}[t]{.5\textwidth}
+\begin{minipage}[t]{.48\textwidth}
   \begin{alignat*}{5}
@@
   (\suc\;n)      & +' 0         && \defeq \suc\;n \\
-  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
+  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
 \end{alignat*}
@@
 Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
-This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
+This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
 For example, two definitions of addition in \cref{fig:additions} are the same
-after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$ and makes the fourth clause useless.
+after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
 Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
-This semantics for function definitions decides how definitions reduce and therefore how identities are proved.
+%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.
 
@@
 Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
-Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
+Under this semantics, each clause of $+$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
 
@@
 \begin{alignat*}{5}
-  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
+  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
 \end{alignat*}
@@
 \[
-  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {[\idS]\,t} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
+  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {[\,\idS\,]\,t} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
   \qquad\qquad
-  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {[\sigma ; \tau]\,t} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{[ \sigma ]\,[ \tau ]\,t}
-\]
-equivalent to the following (homogeneous) identities
-\[
-  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
-  \quad\text{and}\quad
-  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
-\]
-respectively.
-Constructors introduced so far complete the definition of substitution calculus.
+  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {[\,\sigma ; \tau\,]\,t} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{[\, \sigma\,]\,[\,\tau\,]\,t}
+\]
+%equivalent to the following (homogeneous) identities
+%\[
+%  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
+%  \quad\text{and}\quad
+%  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
+%\]
+%respectively.
+Constructors introduced so far complete the inductive definition of substitution calculus.
 
@@
   \]
-  by a back-of-the-envelope calculation
+  and is checked by a back-of-the-envelope calculation
   \begin{equation} \label{eq:pi2-comp-proof}
@@
 
-  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
+  To better illustrate the annoyance, note that \eqref{eq:pi2-comp-real-proof} can be proved by following equations:
   \begin{alignat*}{3}
@@
 %As a result, neither  are strict.
-We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to
-\[
-[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
-[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
-[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
-\]
-where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
+We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to
+\[
+[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
+[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
+[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
+\]
+Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
 %
@@
 \subsection{Adding a universe} \label{subsec:SC+U}
-In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
+Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
 For a QII definition \cite{Altenkirch2016a}, this extension adds the following constructors
@@
   We show this statement by induction on $\sigma$.
-Since we are proving a propositional equality, and we are assuming uniqueness of identity proofs, all equality constructors are automatically respected, so we only need to worry about the point constructors.
+Since we are proving a propositional equality, and we are assuming UIP, all equality constructors are automatically respected, so we only need to worry about the point constructors.
   \begin{enumerate}
@@
 
-In this section, we extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
+We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
 For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
@@
   \item Discuss the Boolean type.
-  \item Discuss the type of natural numbers.
+%  \item Discuss the type of natural numbers.
 \end{enumerate}
@@
 \section{The elimination principle and models} \label{sec:models}
-\LT{%
-\begin{enumerate}
-  \item Show that two definitions are equivalent (ongoing).
-  \item Show that methods are equivalent (ongoing).
-\end{enumerate}
-}
+
 With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
@@
 \begin{remark} \label{re:method-recursion-IIR}
-Note again that our definition is not an instance of Dybjer and Setzer's schema of inductive-recursion where simultaneously defined functions do not require corresponding interpretation, as they target meta-theoretic types instead of inductive types being defined.
+Note that a simultaneously defined function does not need a corresponding interpretation for Dybjer and Setzer's inductive-recursive types, as their framework targets meta-theoretic types rather than the inductive types being defined.
 \end{remark}
@@
 Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.
+A workaround to avoid constant transports is to define the notion of recursors independent of eliminators and show that each recursor gives rise to an eliminator.
 
@@
 \subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
-The additional rules we have used to enable further reduction are derivable in type theory, and they can be justified by the same standard model with simpler proofs.
-
-In general, we can translate the QIIR definition of type theory to its QII definition
-
+We have used additional rules to enable further reduction which are derivable in the QII definition.
+Indeed, the QIIR definition of type theory is equivalent to its QII definition, as we can translate every constructor to its corresponding constructor and vice versa.
+For comparison, we add a subscript to constructors and types in the QIIR and QII definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIR definition to its QII counterpart and the converse, respectively.
+
+The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QII counterpart as well as `smart constructors'.
+Additional equality constructors are all derivable in the QII definition.
+
+On the other hand, the converse $(\blank)^{\prec}$ has some choices.
+As in the QII definition there is only one term substitution $[\blank]_{\I}\blank$ and lifting $(\blank)^+_{\I}$, we may choose to translate $[\sigma]_{\I} \,t$ to the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ to carry out some computation or the constructor form $[\sigma^{\prec}]_{\IR}\,t^{\prec}$ to stay verbatim.
+Either choice is possible, but we need to use \cref{prop:correctness-1,prop:correctness-3} to switch from one to the other whenever suitable.
+For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $[ \sigma ]\,t = [\sigma]_{\Tm}\,t$ if the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ is chosen for the translation.
+UIP is also needed,\footnote{%
+We can further eliminate the use of propositional UIP by identifying function clauses with the reflexivity identity proof using additional rewrite rules.}
+since the method $\mathsf{concat}^M_{\I}$ has a transport along $[;]_{\Ty}^M$ but $\mathsf{concat}_{\IR}$ does not.
+
+Going back and forth from one definition to the other gives us the isomorphism:
+\begin{theorem}
+  $\Ctx_{\IR}$ is isomorphic to $\Ctx_{\I}$.
+  In particular, $(\blank)^{\prec} : \Ctx_{\I} \to \Ctx_{\IR}$ and $(\blank)^{\succ} : \Ctx_{\IR} \to \Ctx{\I}$ are inverses to each other. 
+\end{theorem}
+
+\section{Formal implementation} \label{sec:formal-implementation}
 \LT{Mention the NbE for SC here}
-\section{Formal implementation} \label{sec:formal-implementation}
 \LT{Compare some formal proofs with pen-and-paper proofs here}
@@
 using sheaf theory.
+\LT{Perhaps also mention that the notion of models might be interesting, but it is out of the scope of this paper.}
 

### 2025-02-13 tex/main.tex
Section 3.5

@@
 \subsection{Adding other type formers} \label{subsec:SC+U+Pi+more}
-\LT{%
-\begin{enumerate}
-  \item Discuss the extensional identity type.
-  \item Discuss Coquand universes.
-  \item Discuss the unit type.
-  \item Discuss the Boolean type.
-%  \item Discuss the type of natural numbers.
-\end{enumerate}
-}
-\Fredrik{I think we should keep this brief.
-I started doing the natural numbers in Agda, but there were some problems with the rewrite $[\sigma] \mathsf{Nat} = \mathsf{Nat}$ not firing in the type of the eliminator...}
+
+We can add extensional identity types, a hierarchy of Coquand universes~\cite{Coquand2018}, and a unit type with an eta law, all without any additional difficulty.
+In particular, all these additions can be done without introducing any transports in the syntax.
+However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
+We introduce the type $\bool$ itself, it's two constructors $\btrue$ and $\bfalse$, and its elimination principle $\belim$:
+%
+\begin{alignat*}{3}
+  \bool     &: \implicit{\Gamma}         \;  && \Ty\;\Gamma \\
+  \btrue, \bfalse &: \implicit{\Gamma}   \;  && \Tm\;\Gamma\;\bool \\
+  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;([ \instSub{\bfalse} ]_{\Ty} P) \to \Tm\;\Gamma\;([ \instSub{\btrue} ]_{\Ty} P) \\
+  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;([ \instSub{b} ]_{\Ty} P)
+\end{alignat*}
+%
+where we have written $\instSub{t} = (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ for the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place.
+We can add the expected computation rules
+%
+\begin{alignat*}{5}
+  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;[\instSub{\bfalse}]_{\Ty} P} && f \\
+  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;[\instSub{\btrue}]_{\Ty} P} && t
+\end{alignat*}
+%
+and the structural rule for type substitution as another straightforward clause
+\[
+  \;[ \sigma ]_{\Ty}\,\bool \reduce \bool \enspace ,
+\]
+but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following:
+%
+\begin{alignat*}{5}
+  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \\ & \quad [ \sigma ] (\belim\,P\,f\,t\,b) =^{\Tm\,\Gamma\,([ \sigma; \instSub{b} ]\,P)}_{\alert{p(b)}} \belim\,([ \sigma^+ ] P)\,&&(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{[ \sigma ]\,f}}}) \\&&&(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{[ \sigma ]\,t}}})\\&&&([\sigma ]\,b)
+\end{alignat*}
+where the proof $p(x) : [ \instSub{[ \sigma ]_{\Ty}\,x} ; \sigma^+ ]\;P = [ \sigma ; \instSub{x} ]_{\Ty}\;P$ is needed to line up the types.
+Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems.
+Hence, we do not know how to get rid of these transports.
+
+%\LT{%
+%\begin{enumerate}
+%  \item Discuss the extensional identity type.
+%  \item Discuss Coquand universes.
+%  \item Discuss the unit type.
+%  \item Discuss the Boolean type.
+%  % \item Discuss the type of natural numbers.
+%\end{enumerate}
+%}
 

### 2025-02-13 tex/main.tex
attempt to tighten Section 4

@@
 
-With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
-Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart (\cref{subsec:equivalence}).
+We show how to eliminate out of our QIIRT definition of type theory (\cref{subsec:eliminator}).
+In particular, we show how the elimination principle can be used to construct a standard model (\cref{subsec:std-model}) and compare our QIIRT definition with its QIIT counterpart (\cref{subsec:equivalence}).
 
@@
 
-To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
+To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify \emph{motive} type families
 \begin{align*}
@@
 \end{align*}
-called the \emph{motivates} as the target types for the elimination principle:
+to act as the target types for the elimination principle:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, e.g.,\ 
+Each constructor $c$ should be interpreted by a corresponding \emph{method} $c^M$ indexed by it, e.g.,
 \begin{alignat*}{3}
-  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
-  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
-  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
-  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
-  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
+%  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
+  \blank,_{\Ctx}^M\blank & : \implicit{\Gamma, A}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) %\\
+%  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
+%  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
+%  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
 %  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
-  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
-\end{alignat*}
-and simultaneously defined functions are also interpreted by corresponding functions 
+%  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
+\end{alignat*}
+and simultaneously defined functions are also interpreted by corresponding functions, e.g.,
 \begin{alignat*}{3}
-  [\blank]_{\Ty}^M\blank & : \implicit{\dots} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
-  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
-  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to
-    \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)
+  [\blank]_{\Ty}^M\blank & : \implicit{\Gamma, \Gamma^M,\Delta, \Delta^M, \sigma, A} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) %\\
+%  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) %\\
+%  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to
+%    \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
-  [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
-  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && t^M
-\end{alignat*}
-called \emph{methods}.
-For brevity, we omit the rest of methods and refer to the Agda implementation for them.
-\LT[noinline]{Add a link.}
-We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$.
-
-\begin{definition}
-  An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
-  A \emph{recursor} is an eliminator whereas motives are constants over syntaxes.
-\end{definition}
+  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
+  %[;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
+  %[\idS]_{\Tm}^M & : \implicit{\Gamma^M, t, t^M}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M
+\end{alignat*}
+%
+The elimination principle says that the functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given.
+When the motive families are constant, we call the principle a \emph{recursion principle} instead.
+We refer to \cref{appendix:full-def} and our Agda development for the full definition of the elimination principle.
+%We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$. 
+%%
+%
+%\begin{definition}
+%  An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
+%  A \emph{recursor} is an eliminator whereas motives are constants over syntaxes.
+%\end{definition}
+%
 
 \begin{remark}
-To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
-However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
-That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
+  To state the methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory --- for example, we have
+  \[
+    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\;[\idS^M]^M_{\Tm}\;t^M  =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
+  \]
+  since identities occurring in a method %(e.g.\ $[\idS]_{\Ty}^M$)
+  are propositional just like equality constructors in its QIIT definition.
+  However, in the elimination principle for the QIIT definition, the situation is even worse, as we in general need yet another layer of transports for the methods to typecheck over the QIIT constructors.
+
+  % : the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
+%That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
 \end{remark}
 
-Given an eliminator, the actual computation for the elimination principle is given by
-\begin{alignat*}{5}
-  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
-  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
-  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
-  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
-                            & \vdots
-\end{alignat*}
-interpreting each constructor using the corresponding method inductively.
-
-Interestingly, the key difference from its QII counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
+The elimination principle computes by interpreting each constructor using the corresponding method inductively, e.g.,
+\begin{alignat*}{5}
+%  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
+  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) %\\
+%  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
+%  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
+%                            & \vdots
+\end{alignat*}
+%
+Interestingly, a key difference compared to its QIIT counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
 %type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
 \begin{proposition}\label{prop:provable-function-clauses}
-  Given the motives and the methods, we have
+  Given the motives and the methods, we can prove
 \begin{alignat*}{5}
   \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
-  \nameless & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
-  \nameless & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
-\end{alignat*}
-\end{proposition}
-
-These identities are needed to interpret constructors whereas strict identities are used in syntax.
-Consider eliminating an extension $\sigma, t$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
-\[
-  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M_{\Sub} \; \alert{t^M}.
-\]
-In the QII definition, $\elim_{\Tm}\,t$ is exactly $t^M$ and has the type
-\begin{equation} \label{eq:elim-subext}
-  \Tm^M\;\Gamma^M\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t\judgeq
-  \Tm^M\;\Gamma^M\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t
-\end{equation}
-where the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
-In the QIIR definition, on the contrary, the interpreted type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$, which is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
-\[
-  (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
-\]
-transporting $\elim_{\Tm}\,t$ along the identity proof $\elim_{\Ty}[]$.
+  \elim_{\Tm}[] & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
+  \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
+\end{alignat*}
+\end{proposition}
+
+These identities are needed for the computation rules where strict identities are used in syntax.
+For example, for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, we have 
+\[
+  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
+\]
+where the transport is needed since $\elim_{\Tm}\;t : \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t$ but $\blank ,_{\Sub}^M \blank$ expects its second argument to have type $\Tm^M\;(\elim_{\Ctx}\,\Gamma)\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t$.
+
+% Consider eliminating a substitution extension $\sigma, t$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
+% \[
+%   \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M_{\Sub} \; \alert{t^M}.
+% \]
+% In the QIIT definition, $\elim_{\Tm}\,t$ is exactly $t^M$ and has the type
+% \begin{equation} \label{eq:elim-subext}
+%   \Tm^M\;\Gamma^M\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t\judgeq
+%   \Tm^M\;\Gamma^M\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t
+% \end{equation}
+% where the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
+% In the QIIRT definition, on the contrary, the interpreted type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$, which is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
+% \[
+%   (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
+% \]
+% transporting $\elim_{\Tm}\,t$ along the identity proof $\elim_{\Ty}[]$.
 
@@
 \subsection{The standard model} \label{subsec:std-model}
-The standard model, justifying the soundness of our definition, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
+The standard model, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
 \begin{alignat*}{5}
@@
 \end{alignat*}
-with the additional methods to its QII counterpart:
-\begin{align*}
-  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
-                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
-  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
-                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
-\end{align*}
-where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
-
-In the QII counterpart, none of identities hold strictly: their indices always involve equality constructors.
-E.g., the constructor $[\idS]_{\Ty}$ gives rise to
-$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
-and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
-Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.
-A workaround to avoid constant transports is to define the notion of recursors independent of eliminators and show that each recursor gives rise to an eliminator.
-
-In contrast, in our QIIR definition, many identities just hold strictly.
-Consider
-\begin{alignat*}{5}
-  [\idS]^M_{\Ty}                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
-  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
-\end{alignat*}
-where $[\idS]^M_{\Ty}$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
-Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$.
-By definition, it is a strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq (\lam{\gamma}\gamma)$ by the $\eta$-law for $\Sigma$-types.
+%with the additional methods to its QII counterpart:
+%\begin{align*}
+%  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
+%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
+%  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
+%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
+%\end{align*}
+%where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
+%
+We once again refer to our Agda formalisation for details.
+We remark that because our lack of transports in the syntax, many of the identities in the standard model simply hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is often needed for reasoning about constant transports.
+
+%In the QIIT counterpart standard model, none of identities hold strictly, since their indices always involve equality constructors.
+%E.g., the constructor $[\idS]_{\Ty}$ gives rise to
+%$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
+%and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
+%Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.
+%A workaround to avoid constant transports is to define the notion of recursors independent of eliminators and show that each recursor gives rise to an eliminator. 
+%
+%In contrast, in our QIIR definition, many identities just hold strictly.
+%Consider
+%\begin{alignat*}{5}
+%  [\idS]^M_{\Ty}                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
+%  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
+%\end{alignat*}
+%where $[\idS]^M_{\Ty}$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
+%Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$.
+%By definition, it is a strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq (\lam{\gamma}\gamma)$ by the $\eta$-law for $\Sigma$-types.
 
 \subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
-We have used additional rules to enable further reduction which are derivable in the QII definition.
-Indeed, the QIIR definition of type theory is equivalent to its QII definition, as we can translate every constructor to its corresponding constructor and vice versa.
-For comparison, we add a subscript to constructors and types in the QIIR and QII definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIR definition to its QII counterpart and the converse, respectively.
-
-The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QII counterpart as well as `smart constructors'.
-Additional equality constructors are all derivable in the QII definition.
+We have used additional rules to enable further reduction which are derivable in the QIIT definition.
+Indeed, the QIIRT definition of type theory is equivalent to its QIIT definition, as we can translate every constructor to its corresponding constructor and vice versa.
+For comparison, we add a subscript to constructors and types in the QIIRT and QIIT definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIRT definition to its QIIT counterpart and the converse, respectively.
+
+The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart as well as `smart constructors'.
+Additional equality constructors are all derivable in the QIIT definition.
 
 On the other hand, the converse $(\blank)^{\prec}$ has some choices.
-As in the QII definition there is only one term substitution $[\blank]_{\I}\blank$ and lifting $(\blank)^+_{\I}$, we may choose to translate $[\sigma]_{\I} \,t$ to the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ to carry out some computation or the constructor form $[\sigma^{\prec}]_{\IR}\,t^{\prec}$ to stay verbatim.
+Since in the QIIT definition there is only one term substitution $[\blank]_{\I}\blank$ and lifting $(\blank)^+_{\I}$, we may choose to translate $[\sigma]_{\I} \,t$ to the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ to carry out some computation, or to the constructor form $[\sigma^{\prec}]_{\IR}\,t^{\prec}$ to stay verbatim.
 Either choice is possible, but we need to use \cref{prop:correctness-1,prop:correctness-3} to switch from one to the other whenever suitable.
@@
 UIP is also needed,\footnote{%
-We can further eliminate the use of propositional UIP by identifying function clauses with the reflexivity identity proof using additional rewrite rules.}
+  We can further eliminate the use of propositional UIP by identifying function clauses with the reflexivity identity proof using additional rewrite rules.
+\Fredrik{I would be surprised if this was equivalent to UIP. I would suggest to delete the remark.}}
 since the method $\mathsf{concat}^M_{\I}$ has a transport along $[;]_{\Ty}^M$ but $\mathsf{concat}_{\IR}$ does not.
@@
 \end{theorem}
+\Fredrik{Why only on contexts? If that's the limit of the proof at the moment, I think it would be better to state just the translations, and maybe have a running text remark after the theorem.}
 
@@
 
-\section{Complete definitions}
+\section{Complete definitions}\label{appendix:full-def}
 We give the QIIR definition for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
@@
   \nameless & : && [ \sigma ]_{\Ty}\, \idtype[a]{t}{u} \reduce \idtype[{[\sigma]_{\Tm}\,a}]{[\sigma]_{\Tm}\,t}{[\sigma]_{\Tm}\,u} \\
+  \mathsf{refl} & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\idtype[a]{t}{t}) \\
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
 \end{alignat*}
+
+\paragraph*{Elimination principle}
+
+\Fredrik{TODO}
 

### 2025-02-14 tex/main.tex
Draft S5 and revise S4.3

@@
 \newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
+\lstset{
+    basicstyle=\large\ttfamily,
+    breaklines=true,
+    breakatwhitespace=true,
+    literate=%
+      {≡}{$\equiv$}1
+      {Γ}{$\Gamma$}1
+      {Δ}{$\Delta$}1 
+      {→}{$\to$}1
+      {≅}{$\cong$}1
+}
 
@@
 %
-As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by Hofmann's result~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
+As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
 %
@@
 
-In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
+In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s for short).
 %
@@
 We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
-Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags to introduce additional rewrite rules.
+Nevertheless, Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
 Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
 
-\paragraph*{Contributions}
+\subparagraph*{Contributions}
+This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory.
+In particular, we give
 \begin{itemize}
-  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
-    In particular, we give in a type theory:
-    \begin{itemize}
-      \item a definition of parallel substitution calculus;
-      \item a definition of type theory with a universe and $\Pi$-types,
+  \item definitions of parallel substitution calculus, type theories with a universe and $\Pi$-types,
+  \item with their elimination principles and the standard model;
+  \item a comparison with these definitions using quotient inductive-inductive types.
 \end{itemize}
-   with their elimination principles and the standard model, 
-   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
-  \item Comparison with definitions using quotient inductive-inductive types.
-\end{itemize}
-
-\subsection{Plan of the paper}
+
+\subparagraph*{Plan of the paper}
+In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
+Then, we will develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
+We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.
 
@@
 
-We work in intensional Martin-L\"of type theory with uniqueness of identity proofs (\emph{UIP} for short).
+We work in intensional Martin-L\"of type theory with uniqueness of identity proofs.
 In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
@@
 %
-We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which is not given explicitly, but instead can be figured out from context.
+We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
 We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.
@@
 We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the `dependent' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$.
-One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
-%
-%Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
-%As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
-
-We have formally implemented our definitions and proofs in \Agda, using the options \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
-Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}.
-Some of our statements are meta-theoretical and thus not checked by \Agda.
-Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types.
+One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is the operation to transport values between the type family $P$ along the identity type.
+%
+
+We have implemented our definitions and proofs in \Agda; see \cref{sec:formal-implementation} for details.
+However, some statements are metatheoretical and not checked in \Agda, while others are only partially checked by \Agda, as proof obligations are given manually. %for postulated equality constructors.
 We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
@@
 %
-Since we assume UIP globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
+Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
 %
@@
 
-\subparagraph*{Implementing overlapping patterns via rewrite rules}
-Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns in \Agda, we declare each function clause an additional rewrite rules \cite{Cockx2020,Cockx2021} (via the option \texttt{-{}-rewriting}) and use the local confluence checker (via \texttt{-{}-local-confluence-check}).
-This is indeed one of the main use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
-The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.
-
 \section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
@@
 where $\Gamma$ and $\Delta$ are kept implicit.
-\Fredrik{Could we change this to $A[\sigma]$ instead?}
-\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
-\LT{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}
+\Fredrik[noinline]{Could we change this to $A[\sigma]$ instead?}
+\LT[noinline]{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
+\LT[noinline]{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}
 
@@
 
-One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
+One possibility is to define the substitution calculus as a QIIRT, where type substitution is defined by recursion, and hence computes.
 %Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
@@
 \end{alignat}
-We first check that this definition of type substitution is confluent.\footnote{%
-\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
-This option is turned on globally in our formal implementation.}
-
+We first check that this definition of type substitution is confluent.
 \begin{proposition}[Termination]
@@
 We revisit \cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
-\begin{example}
+\begin{example} \label{ex:pi-comp-qiirt}
   The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
@@
   These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
-\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
-\LT{How about deferring introducing heterogeneous equality until \cref{sec:formal-implementation}?}
 \end{proof}
@@
 However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
-We introduce the type $\bool$ itself, it's two constructors $\btrue$ and $\bfalse$, and its elimination principle $\belim$:
+We introduce the type $\bool$ itself, its two constructors $\btrue$ and $\bfalse$, and its elimination principle $\belim$:
 %
@@
 \begin{alignat*}{5}
-  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;[\instSub{\bfalse}]_{\Ty} P} && f \\
-  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;[\instSub{\btrue}]_{\Ty} P} && t
+  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;[\instSub{\bfalse}]_{\Ty} P} {} && f \\
+  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;[\instSub{\btrue}]_{\Ty} P} {} && t
 \end{alignat*}
@@
 to act as the target types for the elimination principle:
-\begin{alignat*}{5}
+\begin{alignat*}{10}
   \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma &
-  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
+  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\; && (\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;&&\sigma \\
   \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \quad
-               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
+               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\; && (\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;&&t.
 \end{alignat*}
@@
 The standard model, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
-\begin{alignat*}{5}
-  \Ctx^M & \defeq \lam{\nameless} \; && \Set & \qquad \Sub^M & \defeq \lam{\Gamma^M\,\Delta^M\,\nameless}\; && \Gamma^M \to \Delta^M \\
-  \Ty^M  & \defeq \lam{\Gamma^M\, \nameless} && \Gamma^M \to \Set & 
-  \Tm^M  & \defeq \lam{\Gamma^M\,A^M\,\nameless} && (\gamma : \Gamma^M) \to A^M\,\gamma
-\end{alignat*}
+\begin{align*}
+  \Ctx^M\;\nameless          & \defeq \Set              & \Sub^M\;\Gamma^M\;\Delta^M\;\nameless & \defeq \Gamma^M \to \Delta^M \\
+  \Ty^M\;\Gamma^M\;\nameless & \defeq \Gamma^M \to \Set & \Tm^M\;\Gamma^M\;A^M\;\nameless & \defeq (\gamma : \Gamma^M) \to A^M\;\gamma
+\end{align*}
 %with the additional methods to its QII counterpart:
@@
 
-The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart as well as `smart constructors'.
+The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart.
 Additional equality constructors are all derivable in the QIIT definition.
@@
 For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $[ \sigma ]\,t = [\sigma]_{\Tm}\,t$ if the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ is chosen for the translation.
-UIP is also needed,\footnote{%
-  We can further eliminate the use of propositional UIP by identifying function clauses with the reflexivity identity proof using additional rewrite rules.
-\Fredrik{I would be surprised if this was equivalent to UIP. I would suggest to delete the remark.}}
-since the method $\mathsf{concat}^M_{\I}$ has a transport along $[;]_{\Ty}^M$ but $\mathsf{concat}_{\IR}$ does not.
-
-Going back and forth from one definition to the other gives us the isomorphism:
+
+Going back and forth from one definition to the other gives us the equivalences
 \begin{theorem}
-  $\Ctx_{\IR}$ is isomorphic to $\Ctx_{\I}$.
-  In particular, $(\blank)^{\prec} : \Ctx_{\I} \to \Ctx_{\IR}$ and $(\blank)^{\succ} : \Ctx_{\IR} \to \Ctx{\I}$ are inverses to each other. 
+  Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 
+  \begin{itemize}
+    \item $\eqv{\Ctx_{\IR}}{\Ctx_{\I}}$, 
+    \item $\eqv{\left((\Gamma : \Ctx_{\IR})\to\Ty_{\IR}\;\Gamma\right)}{\left((\Gamma : \Ctx_{\I})\to\Ty_{\I}\;\Gamma\right)}$,
+    \item $\eqv{\left((\Gamma\;\Delta : \Ctx_{\IR}) \to \Sub_{\IR}\;\Gamma\;\Delta\right)}%
+               {\left((\Gamma\;\Delta: \Ctx_{\I}) \to \Sub_{\I}\;\Gamma\;\Delta\right)}$, and 
+    \item $\eqv{\left((\Gamma : \Ctx_{\IR}) \to (A : \Ty_{\IR}\;\Gamma) \to \Tm_{\IR}\;\Gamma\;A\right)}
+               {\left((\Gamma : \Ctx_{\I}) \to (A : \Ty_{\I}\;\Gamma) \to \Tm_{\I}\;\Gamma\;A\right)}$.
+  \end{itemize}
 \end{theorem}
-\Fredrik{Why only on contexts? If that's the limit of the proof at the moment, I think it would be better to state just the translations, and maybe have a running text remark after the theorem.}
+%\Fredrik{Why only on contexts? If that's the limit of the proof at the moment, I think it would be better to state just the translations, and maybe have a running text remark after the theorem.}
 
 \section{Formal implementation} \label{sec:formal-implementation}
+
+We have a formal implementation in \Agda, using the options \texttt{-{}-with-K} for the uniqueness of identity proofs, \texttt{-{}-exact-split} to ensure that all function clauses are definitional, and \texttt{-{}-rewriting} with \texttt{-{}-local-confluence-check} for implementing definitions by overlapping patterns.
+Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}, as \textsc{Cubical Agda} does not support user-defined rewrite rules.
+
+\subparagraph*{Implementing overlapping patterns via rewrite rules}
+%The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
+To emulate a function by overlapping patterns in \Agda, we declare each function clause a rewrite rule \cite{Cockx2020,Cockx2021} and another rewrite to identify the clause with the reflexivity proof and also turn on the option \texttt{-{}-local-confluence-check} to check local confluence.
+For example, type substitution $[\blank]_{\Ty}$ up to the first clause $[\idS]_{\Ty}\,A \reduce A$ is defined as
+\begin{lstlisting}
+postulate
+  [_]_      : Sub Γ Δ → Ty Δ → Ty Γ
+  [id]      : [ idS ] A ≡ A
+  {-# REWRITE [id] #-}
+  [id]≡refl : [id] {A = A} ≡ refl
+  {-# REWRITE [id]=refl #-}
+\end{lstlisting}
+where \lstinline|[id]≡refl| is used to ensure that \lstinline|[id]| need not be referred.
+
+Defining by overlapping patterns is indeed one of the use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
+The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.
+
+\subparagraph*{Heterogeneous identity}
+In our implementation, some statements are proved differently from our presentation. 
+We have sometimes used John Major equality~\cite{McBride1999} $x \cong y$  for $A, B : \Set$, and $x : A$ and $y : B$, where dependent congruence rule or dependent identity type are needed.
+
+For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\mathit{apd}_f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
+However, as $[\pi_1(\sigma; \tau)]_{\Ty}\,A$ and $[\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))]_{\Ty}\,A$ are strictly equal, the resulting identity is indeed homogeneous.
+Hence, we use the heterogeneous congruence $\mathit{hap}_f\colon (p : x \simeq y) \to f\,x\simeq f\,y$ (by \lstinline|hcong|) and switch back to the ordinary identity type $f\,x = f\, y$ (by \lstinline|≅-to-≡|) as their types match.
+
+%As shown by McBride, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
+
 \LT{Mention the NbE for SC here}
-\LT{Compare some formal proofs with pen-and-paper proofs here}
-\section{Conclusions, and related and future work}
+
+\section{Conclusions, and related and future work}\label{sec:conclusion}
 

### 2025-02-14 tex/main.tex
Minor revisions before adding definitions in the appendix

@@
 \subparagraph*{Plan of the paper}
-In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
-Then, we will develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
+In the rest of the paper, we introduce the metalanguage in \cref{sec:meta}.
+Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
 We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.
@@
 %
-We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
-%
-We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
-We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.
+We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using \Agda notation $(x : A) \to B\,x$.
+%
+We are further inspired by \Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
+We write $\Set_i$ for the hierarchy of universes with the level index $i$ suppressed.
 
@@
 
-An ordinary inductive type $A$, such as for example the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, by providing an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$.
-For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\mathsf{suc}\,n)$, where the latter might refer to the already computed value $f(n)$.
+An ordinary inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, using an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
+For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
 %
@@
 \]
-with $T(\pi(a, b)) = (x : T(a)) \to T(b(x))$ only makes sense because we already have access to the decoding function $T$.
+with $T(\pi(a, b)) = (x : T\,a) \to T(b\, x)$ only makes sense because we already have access to the decoding function $T$.
 %
@@
 %
-However, note that the recursive function $F : A \to D$ is not defined using this elimination rule, but part of the definition of the type itself.
+However, note that the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
 %
@@
 
-Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
+Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
 %
@@
 
-A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a family of types $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively; indeed, inductive-inductive definitions only make sense when $D = \Set$.
+A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
+%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
 The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
 
-We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list.
+We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list.
 We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
@@
 \subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-\begin{figure} 
+\begin{figure}[b]
 \begin{minipage}[t]{.48\textwidth}
@@
 Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
-Under this semantics, each clause of $+$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
+Under this semantics, each clause of $+$ is strictly equal to its right-hand side, also allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
 
@@
 
-We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
+We begin with a recall of the definition of substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
 We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning, and develop a type theory with a universe (\cref{subsec:SC+U}) of small types, $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
@@
 
-We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
+We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, where the context of its codomain is extended with the domain type.
 For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
@@
 \end{alignat*}
-Each constructor $c$ should be interpreted by a corresponding \emph{method} $c^M$ indexed by it, e.g.,
+Each constructor $\mathsf{c}$ should be interpreted by a corresponding \emph{method} $\mathsf{c}^M$ indexed by it, e.g.,
 \begin{alignat*}{3}
@@
 The elimination principle says that the functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given.
+\LT[noinline]{says what?}
 When the motive families are constant, we call the principle a \emph{recursion principle} instead.
-We refer to \cref{appendix:full-def} and our Agda development for the full definition of the elimination principle.
+We refer to \cref{appendix:full-def} and our \Agda development for the full definition of the elimination principle.
 %We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$. 
@@
 
-For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\mathit{apd}_f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
+For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\apdfunc f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
 However, as $[\pi_1(\sigma; \tau)]_{\Ty}\,A$ and $[\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))]_{\Ty}\,A$ are strictly equal, the resulting identity is indeed homogeneous.
@@
 using sheaf theory.
-\LT{Perhaps also mention that the notion of models might be interesting, but it is out of the scope of this paper.}
+%\LT{Perhaps also mention that the notion of models might be interesting, but it is out of the scope of this paper.}
 

### 2025-02-14 tex/main.tex
fix typo

@@
 %
-The elimination principle says that the functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given.
-\LT[noinline]{says what?}
+The elimination principle says that there are functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given.
 When the motive families are constant, we call the principle a \emph{recursion principle} instead.

### 2025-02-14 tex/main.tex
Add points discussed during the meeting

@@
 We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
-Nevertheless, Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
+Nevertheless, \Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
 Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
@@
 \subsection{Inductive-recursive types}
+\LT[noinline]{Shorten this section}
+\LT[noinline]{the recursion part of IR is not given by the elimination rule}
 
@@
 We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.
+
+\LT[noinline]{remark that recursions in IR are not defined by elimination rules}
 
@@
 %
-We once again refer to our Agda formalisation for details.
+We once again refer to our \Agda formalisation for details.
 We remark that because our lack of transports in the syntax, many of the identities in the standard model simply hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is often needed for reasoning about constant transports.
@@
 
-\LT{Mention the NbE for SC here}
+\LT[noinline]{Mention the NbE for SC here?}
 
@@
 
-\Fredrik{TODO}
+\begin{align*}
+  \Ctx^M & : \Ctx \to \Set &
+  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
+  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set &
+  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
+\end{align*}
+
+\begin{alignat*}{10}
+  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
+  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  [\blank]_{\Ty}^M\blank & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
+  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
+  \idS^M          & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
+  \blank;^M\blank & : \implicit{\dots}\; && (\sigma^M : \Sub^M\;\Gamma\;\Delta^M\;\sigma) \to (\tau^M : \Sub^M\;\Delta^M\;\Theta^M\;\tau) \to \Sub^M\;\Gamma^M\;\Delta^M\;(\sigma ; \tau) \\
+  \pi_1^M         & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma \to \Sub^M\;\Gamma^M\;\Delta^M\;(\pi_1\sigma) \\
+  \pi_2^M         & : \implicit{\dots}\; && (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;([ \pi_1^M\,\sigma^M ]^M_{\Ty}\,A^M)\;(\pi_2\,\sigma) \\
+  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t) \\
+  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
+  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
+  \UU^M & : \implicit{\dots}\; && \Ty^M\;\Gamma^M\;\UU \\
+  \El^M & : \implicit{\dots}\; && \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u) \\
+  \Pi^M & : \implicit{\dots}\; && (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
+  \lambda^M & : \implicit{\dots}\; && \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;t \to \Tm^M\;\Gamma^M\;(\Pi^M\;A^M\;B^M)\;(\lambda\,t) \\
+  \mathsf{app}^M & : \implicit{\dots}\; && \Tm^M\;\Gamma^M\;(\Pi^M\,A^M\,B^M)\;t \to \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;(\mathsf{app}\,t) \\
+\end{alignat*}
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
+  [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
+  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M
+\end{alignat*}
+\LT{To be finished}
 
@@
 
-\section{Formal definitions in Agda}
 \end{document}

### 2025-02-17 tex/main.tex
move many proofs to the appendix

@@
 We first check that this definition of type substitution is confluent.
-\begin{proposition}[Termination]
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
-  \danger
-\end{proposition}
-%\begin{proof}
-%  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
-%\end{proof}
-\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
-  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+%
+\begin{proposition}\label{prop:local-confluence-1}
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
   \danger
@@
 \begin{proof}
+  Termination is clear, since all recursive calls are on structurally smaller arguments~\cite{Abel2002}.
   To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
@@
 %\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
-\begin{proposition}[Coherence]\label{prop:coherence-1}
-  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
+\begin{proposition}\label{prop:coherence-1}
+  For every equality constructor $p : \sigma = \tau$, we have $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$.
   \danger
@@
 \end{align}
-to retain the local confluence for type substitution:
-\begin{proposition}[Termination]
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
-  \danger
-\end{proposition}
+to retain the local confluence for type substitution. The proof is similar to the proof of \cref{prop:local-confluence-1}, and can be found in \cref{appendix:proofs}.
+%\begin{proposition}[Termination]
+%  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+%  \danger
+%\end{proposition}
 %\begin{proof}
@@
 %\end{proof}
-\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
-  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+\begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
+  \qed
   \danger
-\end{proposition}
-\begin{proof}
-  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
-  \[
-    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
-  \]
-  which reduce to the following terms
-  \[
-    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
-    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
-  \]
-  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
-  Therefore, type substitution is locally confluent.
-\end{proof}
+\end{restatable}
 
@@
 
-\begin{proposition}\label{prop:correctness-1}
-  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
-\end{proposition}
-\begin{proof}
-  We show this statement by induction on $\sigma$.
-Since we are proving a propositional equality, and we are assuming UIP, all equality constructors are automatically respected, so we only need to worry about the point constructors.
-  \begin{enumerate}
-    \item For the identity substitution $\idS$, we have $[\idS]t = t \judgeq [\idS]_{\Tm} t$ by definition.
-    \item For a composite substitution $\sigma; \tau$, we have
-      \begin{align*}
-        [\sigma;\tau]t & = [\sigma]\,([\tau] t)            && \text{by $[;]t$} \\
-%                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
-                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
-                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
-      \end{align*}
-    \item For $\pi_1(\sigma, t)$, we have
-      \begin{align*}
-        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
-                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
-                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
-      \end{align*}
-    %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
-    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ strictly.
-  \end{enumerate}
-  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
-\end{proof}
+\begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
+  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$. \qed
+\end{restatable}
 
 Finally, we have to show that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent.
-\begin{proposition}[Coherence] \label{prop:coherence-2}
-  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
+\begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
+  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the identities
   \danger
@@
   \]
-  hold for any type $A$ and any term $t$.
-\end{proposition}
-\begin{proof}
-  By \cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
-  \begin{align*}
-    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
-  \end{align*}
-  while other cases hold strictly.
-  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
-  However, all cases follow easily from \cref{prop:correctness-1}.
-  For example, consider the first case:
-  \begin{align*}
-    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by \eqref{eq:def-type-subst-7}} \\
-                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
-                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
-                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
-                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by \eqref{eq:def-type-subst-7} }
-  \end{align*}
-  The remaining two cases are omitted.
-
-  As for term substitution $[\blank]_{\Tm}\blank$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
-\end{proof}
+  hold for any type $A$ and any term $t$. \qed
+\end{restatable}
 
@@
 
-\begin{proposition}[Termination]
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
+\begin{restatable}{proposition}{propConfluenceThree}\label{prop:local-confluence-3}
+  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
+  \qed
     \danger
-\end{proposition}
-\begin{proposition}[Local confluence]\label{prop:local-confluence-3}
-  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
-  \danger
-\end{proposition}
-\begin{proof}
-  In addition to previous cases in \cref{prop:local-confluence-1,prop:local-confluence-2}, consider
-  \begin{align*}
-    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
-  \end{align*}
-  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
-  \begin{align*}
-    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
-    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
-  \end{align*}
-  respectively.
-  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
-\end{proof}
+\end{restatable}
 
@@
 
-\begin{proposition} \label{prop:correctness-2}
+\begin{restatable}{proposition}{propCorrectnessTwo}\label{prop:correctness-2}
     For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
     $\sigma \uparrow A$ is propositionally equal to $\sigma^+$.
-\end{proposition}
-\begin{proof}
-  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
-  Every case follows from equational reasoning straightforwardly.
-  For example, if $\sigma$ is a composite $\sigma;\tau$, we have
-  \begin{align*}
-    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
-                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
-                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
-  \end{align*}
-  Other cases follow similarly.
-\end{proof}
+    \qed
+\end{restatable}
 
@@
 Finally, we show that these simultaneously defined functions are coherent.
-\begin{proposition}[Coherence]
-  For every constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
+\begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
+    For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
     \danger
@@
     for any type $A$ and any term $t$.
-\end{proposition}
-\begin{proof}
-  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
-\end{proof}
+    \qed
+\end{restatable}
 
@@
 Of course, we also have to prove coherence for the newly introduced recursive functions to verify that they are well defined.
-\begin{proposition}[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
+
+\begin{restatable}%[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
+{proposition}{coherenceTel}
+  For every equality constructor $p : \sigma = \tau$ and telescope $\Xi$, we have
+  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$. \qed
   \danger
-  For any equality constructor $p : \sigma = \tau$, we have
-  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$
-  for any telescope $\Xi$.
-\end{proposition}
+\end{restatable}
 
@@
 
-\begin{corollary}[Coherence]
-  The identity $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ holds for every equality constructor $p : \sigma = \tau$.
+\begin{corollary}
+  We have $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
+  \qed
   \danger
@@
 %type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
-\begin{proposition}\label{prop:provable-function-clauses}
+\begin{restatable}{proposition}{propProvableFunc}\label{prop:provable-function-clauses}
   Given the motives and the methods, we can prove
@@
 \end{alignat*}
-\end{proposition}
+\end{restatable}
 
@@
 Going back and forth from one definition to the other gives us the equivalences
-\begin{theorem}
+\begin{restatable}{theorem}{equivalence}
   Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 
@@
   \end{itemize}
-\end{theorem}
-%\Fredrik{Why only on contexts? If that's the limit of the proof at the moment, I think it would be better to state just the translations, and maybe have a running text remark after the theorem.}
+\end{restatable}
 
@@
 
-\section{Omitted proofs}
-
-\Fredrik{Do properly with restate-thm etc}
-The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
-\begin{proposition}[Coherence, I]
-  \danger
-  For the following equality constructors $p : \sigma = \tau$
+\section{Omitted proofs}\label{appendix:proofs}
+
+\propConfluenceTwo*
+\begin{proof}
+  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
+  \[
+    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
+  \]
+  which reduce to the following terms
+  \[
+    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
+    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
+  \]
+  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
+  Therefore, type substitution is locally confluent.
+\end{proof}
+
+\propCorrectness*
+\begin{proof}
+  We show this statement by induction on $\sigma$.
+Since we are proving a propositional equality, and we are assuming UIP, all equality constructors are automatically respected, so we only need to worry about the point constructors.
+  \begin{enumerate}
+    \item For the identity substitution $\idS$, we have $[\idS]t = t \judgeq [\idS]_{\Tm} t$ by definition.
+    \item For a composite substitution $\sigma; \tau$, we have
+      \begin{align*}
+        [\sigma;\tau]t & = [\sigma]\,([\tau] t)            && \text{by $[;]t$} \\
+%                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
+                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
+      \end{align*}
+    \item For $\pi_1(\sigma, t)$, we have
+      \begin{align*}
+        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
+                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
+      \end{align*}
+    %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
+    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ strictly.
+  \end{enumerate}
+  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
+\end{proof}
+
+\propCoherence*
+\begin{proof}
+  By \cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
+  \begin{align*}
+    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
+                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+  \end{align*}
+  while other cases hold strictly.
+  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
+  However, all cases follow easily from \cref{prop:correctness-1}.
+  For example, consider the first case:
+  \begin{align*}
+    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by \eqref{eq:def-type-subst-7}} \\
+                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
+                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
+                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
+                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by \eqref{eq:def-type-subst-7} }
+  \end{align*}
+  The remaining two cases are omitted.
+
+  As for term substitution $[\blank]_{\Tm}\blank$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
+\end{proof}
+
+\propConfluenceThree*
+\begin{proof}
+  In addition to previous cases in \cref{prop:local-confluence-1,prop:local-confluence-2}, consider
+  \begin{align*}
+    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
+  \end{align*}
+  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
+  \begin{align*}
+    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
+    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
+  \end{align*}
+  respectively.
+  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+\end{proof}
+
+\propCorrectnessTwo*
+\begin{proof}
+  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
+  Every case follows from equational reasoning straightforwardly.
+  For example, if $\sigma$ is a composite $\sigma;\tau$, we have
+  \begin{align*}
+    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
+                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
+                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
+  \end{align*}
+  Other cases follow similarly.
+\end{proof}
+
+\propCoherenceUparrow*
+\begin{proof}
+  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
+\end{proof}
+
+\coherenceTel*
+\begin{proof}
+  \Fredrik{TODO: spell out all cases here}
+  The proposition is easy to prove for the following equality
+  constructors, since the lifting and substitution operations compute
+  independently for these cases:
   \[
@@
   \]
-  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
-  hold strictly for any telescope $\Xi$ and any type $A$ and
-  \[
-    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
-    \tau   \upuparrows \Xi
-  \]
-  holds propositionally for any telescope $\Xi$.
-\end{proposition}
-\begin{proof}
-  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
   Take $\mathsf{idr}$ as an example, we can see 
@@
   For other two equality constructors, the arguments follow similarly.
-\end{proof}
+
 The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
-\begin{proposition}[Coherence, II]
   \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
@@
    the identities
-  \danger
   \begin{align*}
@@
   hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
-\end{proposition}
-\begin{proof}
   We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.
@@
   \end{enumerate}
-
 \end{proof}
 
+\propProvableFunc*
+\begin{proof}
+  TODO
+\end{proof}
+
+\equivalence*
+\begin{proof}
+  TODO
+\end{proof}
+
 \end{document}

### 2025-02-17 tex/main.tex
equations up to term substitution

@@
 
-We have a formal implementation in \Agda, using the options \texttt{-{}-with-K} for the uniqueness of identity proofs, \texttt{-{}-exact-split} to ensure that all function clauses are definitional, and \texttt{-{}-rewriting} with \texttt{-{}-local-confluence-check} for implementing definitions by overlapping patterns.
+We have a formal implementation in \Agda, using the options \texttt{-{}-with-K} for the uniqueness of identity proofs, \texttt{-{}-exact-split} to ensure that all function clauses are definitional, and \texttt{-{}-rewriting} with \texttt{-{}-local-confluence-check} \cite{Cockx2020,Cockx2021} to implement definitions by overlapping patterns.
 Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}, as \textsc{Cubical Agda} does not support user-defined rewrite rules.
@@
 %The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns in \Agda, we declare each function clause a rewrite rule \cite{Cockx2020,Cockx2021} and another rewrite to identify the clause with the reflexivity proof and also turn on the option \texttt{-{}-local-confluence-check} to check local confluence.
-For example, type substitution $[\blank]_{\Ty}$ up to the first clause $[\idS]_{\Ty}\,A \reduce A$ is defined as
+To emulate a function by overlapping patterns, we declare each function clause a rewrite and another rewrite to identify the clause with the reflexivity proof.
+E.g.,\ $[\blank]_{\Ty}$ up to the first clause $[\idS]_{\Ty}\,A \reduce A$ is defined as
 \begin{lstlisting}
@@
 where \lstinline|[id]≡refl| is used to ensure that \lstinline|[id]| need not be referred.
+We also turn on the option \texttt{-{}-local-confluence-check} to check local confluence.
 
@@
 \begin{alignat*}{5}
-  \idtype[\blank]{\blank}{\blank} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
+  \mathsf{Id} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
   \nameless & : && [ \sigma ]_{\Ty}\, \idtype[a]{t}{u} \reduce \idtype[{[\sigma]_{\Tm}\,a}]{[\sigma]_{\Tm}\,t}{[\sigma]_{\Tm}\,u} \\
-  \mathsf{refl} & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\idtype[a]{t}{t}) \\
+  \mathsf{refl} & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\mathsf{Id}\;a\;t\;t) \\
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
@@
 
+\subparagraph*{Motives}
 \begin{align*}
@@
 
+\subparagraph*{Methods}
+Note that implicitly quantified variables which would be otherwise declared inside $\implicit{\dots}$ are omitted below.
+Each variable $\mathsf{c}^M$ with the superscript $(\blank)^M$ is always indexed by $\mathsf{c}$ in addition to other indices.
+
 \begin{alignat*}{10}
-  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
-  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
-  [\blank]_{\Ty}^M\blank & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
-  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
-  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
-  \idS^M          & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
-  \blank;^M\blank & : \implicit{\dots}\; && (\sigma^M : \Sub^M\;\Gamma\;\Delta^M\;\sigma) \to (\tau^M : \Sub^M\;\Delta^M\;\Theta^M\;\tau) \to \Sub^M\;\Gamma^M\;\Delta^M\;(\sigma ; \tau) \\
-  \pi_1^M         & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma \to \Sub^M\;\Gamma^M\;\Delta^M\;(\pi_1\sigma) \\
-  \pi_2^M         & : \implicit{\dots}\; && (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;([ \pi_1^M\,\sigma^M ]^M_{\Ty}\,A^M)\;(\pi_2\,\sigma) \\
-  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t) \\
-  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
-  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
-  \UU^M & : \implicit{\dots}\; && \Ty^M\;\Gamma^M\;\UU \\
-  \El^M & : \implicit{\dots}\; && \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u) \\
-  \Pi^M & : \implicit{\dots}\; && (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
-  \lambda^M & : \implicit{\dots}\; && \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;t \to \Tm^M\;\Gamma^M\;(\Pi^M\;A^M\;B^M)\;(\lambda\,t) \\
-  \mathsf{app}^M & : \implicit{\dots}\; && \Tm^M\;\Gamma^M\;(\Pi^M\,A^M\,B^M)\;t \to \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;(\mathsf{app}\,t) \\
-\end{alignat*}
-\begin{alignat*}{5}
-  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
-  [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
-  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M
+  \emptyctx_{\Ctx}^M     & : \Ctx^M\;\emptyctx \\
+  \blank,_{\Ctx}^M\blank & : (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
+  [\blank]_{\Ty}^M\blank & : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  \emptysub_{\Sub}^M     & : \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
+  \blank,_{\Sub}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
+  \idS^M                 & : \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
+  \blank;^M\blank        & : (\sigma^M : \Sub^M\;\Gamma\;\Delta^M\;\sigma) \to (\tau^M : \Sub^M\;\Delta^M\;\Theta^M\;\tau) \to \Sub^M\;\Gamma^M\;\Delta^M\;(\sigma ; \tau) \\
+  \pi_1^M                & : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma \to \Sub^M\;\Gamma^M\;\Delta^M\;(\pi_1\sigma) \\
+  \pi_2^M                & : (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;([ \pi_1^M\,\sigma^M ]^M_{\Ty}\;A^M)\;(\pi_2\,\sigma) \\
+  [\blank]^M\blank       & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t) \\
+  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
+  [\blank]_{\Tm}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
+  \UU^M                  & : \Ty^M\;\Gamma^M\;\UU \\
+  \El^M                  & : \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u) \\
+  \Pi^M                  & : (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
+  \lambda^M              & : \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;t \to \Tm^M\;\Gamma^M\;(\Pi^M\;A^M\;B^M)\;(\lambda\,t) \\
+  \mathsf{app}^M         & : \Tm^M\;\Gamma^M\;(\Pi^M\,A^M\,B^M)\;t \to \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;(\mathsf{app}\,t) \\
+  \mathsf{Id}^M          & : (a^M : \Tm^M\;\Gamma^M\;\UU^M\;a) \to \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;t \to \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;u \to \Ty^M\;\Gamma^M\;(\mathsf{Id}\;a\;t\;u)  \\
+  \mathsf{refl}^M        & : (t^M : \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;t) \to \Tm^M\;\Gamma^M\;(\mathsf{Id}^M\;a^M\;t^M\;t^M)\;(\mathsf{refl}\;t) 
+\end{alignat*}
+
+\begin{alignat*}{5}
+  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M ([ \sigma^M ]^M\;A^M]))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
+  (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;\pi_2\idS}{([;]^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
+\end{alignat*}
+
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M      & : [ \idS^M ]^M_{\Ty}                             && \;A^M =^{\Ty^M\,\Gamma^M}\; && A^M \\
+  [;]_{\Ty}^M         & : [ \sigma^M ;^M \tau^M ]^M_{\Ty}                && \;A^M =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
+  [\pi_1,]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ,^M_{\Sub} t^M) ]^M_{\Ty} && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\,A^M \\
+  [\pi_1;]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ;^M \tau^M) ]^M_{\Ty}     && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\left([ \pi_1^M\,\tau^M]^M_{\Ty}\,A^M\right)
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \idS\mathord{\uparrow}^M      & : \idS^M              && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\idS}_{\alert{[\idS]^M_{\Ty}}} {} && \idS^M \\
+  ;\mathord{\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
+  \pi_1,\mathord{\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
+  \pi_1;\mathord{\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
+  \emptysub^M_{\Sub} \mathord{\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
+  ,\mathord{\uparrow}^M & : (\sigma^M ,_{\Sub}^M t^M)  && \uparrow^M A^M = (\sigma^M ,_{\Sub}^M t^M)^{+M} \\
+  \pi_1\idS\mathord{\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
+  \pi_1\pi_1\mathord{\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
+\end{alignat*}
+
+\begin{alignat*}{5}
+  [\idS]_{\Tm}^M      & : [\idS^M]^M_{\Tm}                               && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M \\
+  [;]^M_{\Tm}         & : [ \sigma^M ;^M \tau^M ]^M_{\Tm}                && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[;]^M_{\Ty}}} && [ \sigma^M ]^M\; \left([ \tau^M ]^M_{\Tm}\,t^M\right) \\
+  [\pi_1,]^M_{\Tm}    & : [\pi_1^M (\sigma^M ,_{\Sub}^M t^M) ]^M_{\Tm}   && \;u^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;u^M \\
+  [\pi_1;]^M_{\Tm}    & : [\pi_1^M (\sigma^M ;^M \tau^M) ]^M_{\Tm}       && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1;]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;\left([\pi_1^M\,\tau^M]^M_{\Tm}\;t^M\right) \\
+  [\emptysub]^M_{\Tm} & : [\emptysub^M_{\Sub}]^M_{\Tm}                   && \;t^M = %^{\Tm^M\,\Gamma^M\;([\emptysub_{\Sub}^M]^M_{\Ty}\,A^M)\,([\emptysub]_{\Tm}\,t)}
+                      && [ \emptysub^M_{\Sub} ]^M \;t^M \\
+  [,]_{\Tm}^M         & : [ \sigma^M ,_{\Sub}^M t^M ]^M_{\Tm}            && \;u^M = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)} &&
+                      [ \sigma^M ,_{\Sub}^M t^M ]^M \;u \\
+  [\pi_1\idS]^M_{\Tm} & : [ \pi_1^M\,\idS^M ]^M_{\Tm}                    && \;t^M = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{} &&
+                      [ \pi_1^M\,\idS^M ]^M\;t^M \\
+  [\pi_1\pi_1]^M_{\Tm}& : [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M_{\Tm}       && \;t^M = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)} &&
+                      [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M\;t^M
 \end{alignat*}

### 2025-02-18 tex/main.tex
Finish methods

@@
 %
-Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
+Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define$[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
 \begin{alignat}{3}
-[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
+%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
 [ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
@@
 \section{Complete definitions}\label{appendix:full-def}
+
+\paragraph*{Quotient inductive-inductive-recursive types}
 We give the QIIR definition for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
 
-\paragraph*{Substitution calculus with a base type $\UU$}
+\subparagraph*{Substitution calculus with a base type $\UU$}
 \begin{alignat*}{3}
@@
   \Sub      & : && \Ctx \to \Ctx \to \Set \\
-  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set \\
+  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
+\end{alignat*}
+\begin{alignat*}{5}
+  \emptyctx & : && \Ctx \\
+  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
+\end{alignat*}
+\begin{alignat*}{5}
   [\blank]_{\Ty}\blank & : \implicit{\Gamma, \Delta}\; && \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
-  \emptyctx & : && \Ctx \\
-  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx \\
+  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
+\end{alignat*}
+\begin{alignat*}{5}
   \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
@@
   \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\;\sigma ]_{\Ty}\, A) \\
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\, A) \\
-  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
+  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\, A)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
-  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
   ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
@@
   \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  [\idS]t         & : \implicit{\Gamma, A, t}\; && {[\idS]\;t}          && =^{\Tm\;\Gamma\;A} && t \\
-  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {[\sigma ; \tau]\;t} && =^{\Tm\;\Gamma\;[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
-\end{alignat*}
-
-\paragraph*{Universe of small types}
+  [\idS]          & : \implicit{\Gamma, A, t}\; && {[\idS]\;t}          && =^{\Tm\;\Gamma\;A} && t \\
+  [;]             & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {[\sigma ; \tau]\;t} && =^{\Tm\;\Gamma\;[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
+\end{alignat*}
+
+\subparagraph*{Universe of small types}
 \begin{alignat*}{5}
@@
 \begin{alignat*}{5}
-  \nameless & : [ \sigma ]_{\Ty}\,  && (\El\, u) \reduce \El\,([\sigma]_{\Tm}{u})
-\end{alignat*}
-
-\paragraph*{$\Pi$-types}
+  [ \sigma ]_{\Ty}\,  && (\El\, u) \reduce \El\,([\sigma]_{\Tm}{u})
+\end{alignat*}
+
+\subparagraph*{$\Pi$-types}
 \begin{alignat*}{5}
@@
 \begin{alignat*}{5}
-  \nameless       &: && [ \sigma ]_{\Ty}\, (\Pi\,A\,B) && {} \reduce {} && \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
+  [ \sigma ]_{\Ty}\, (\Pi\,A\,B) \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
+\end{alignat*}
+\begin{alignat*}{5}
   []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& [ \sigma ] \,(\lambda\,t) && {} =^{\Tm\;\Gamma\;([\sigma]_{\Ty}(\Pi\,A\,B)} {} && \lambda ([\sigma \uparrow A ]\,t)\\
@@
 
-\paragraph*{Extensional identity types}
+\subparagraph*{Extensional identity types}
 
@@
   \mathsf{Id} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
-  \nameless & : && [ \sigma ]_{\Ty}\, \idtype[a]{t}{u} \reduce \idtype[{[\sigma]_{\Tm}\,a}]{[\sigma]_{\Tm}\,t}{[\sigma]_{\Tm}\,u} \\
+  \nameless & : && [ \sigma ]_{\Ty}\;(\mathsf{Id}\;a\;t\;u) \reduce \mathsf{Id}\;([\sigma]_{\Tm}\,a)\;([\sigma]_{\Tm}\,t)\;([\sigma]_{\Tm}\,u) \\
   \mathsf{refl}   & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\mathsf{Id}\;a\;t\;t) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& [\sigma]\,(\mathsf{refl}\;t) =^{\Tm\;\Gamma\;([\sigma](\mathsf{Id}\,a\,t\,t))} \mathsf{refl}([ \sigma ]_{\Tm}\,t) \\
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
@@
 
-\paragraph*{Elimination principle}
+\paragraph*{Elimination principles for type theories as QIIRTs}
 
@@
   \idS\mathord{\uparrow}^M      & : \idS^M              && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\idS}_{\alert{[\idS]^M_{\Ty}}} {} && \idS^M \\
-  ;\mathord{\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
-  \pi_1,\mathord{\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
-  \pi_1;\mathord{\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
-  \emptysub^M_{\Sub} \mathord{\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
+  \mathord{;\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
+  \mathord{\pi_1,\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
+  \mathord{\pi_1;\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
+  \mathord{\emptysub^M_{\Sub}\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
   ,\mathord{\uparrow}^M & : (\sigma^M ,_{\Sub}^M t^M)  && \uparrow^M A^M = (\sigma^M ,_{\Sub}^M t^M)^{+M} \\
-  \pi_1\idS\mathord{\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
-  \pi_1\pi_1\mathord{\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
+  \mathord{\pi_1\idS\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
+  \mathord{\pi_1\pi_1\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
 \end{alignat*}
@@
   [;]^M_{\Tm}         & : [ \sigma^M ;^M \tau^M ]^M_{\Tm}                && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[;]^M_{\Ty}}} && [ \sigma^M ]^M\; \left([ \tau^M ]^M_{\Tm}\,t^M\right) \\
-  [\pi_1,]^M_{\Tm}    & : [\pi_1^M (\sigma^M ,_{\Sub}^M t^M) ]^M_{\Tm}   && \;u^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;u^M \\
+  [\pi_1,]^M_{\Tm}    & : [\pi_1^M (\sigma^M ,_{\Sub}^M u^M) ]^M_{\Tm}   && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;t^M \\
   [\pi_1;]^M_{\Tm}    & : [\pi_1^M (\sigma^M ;^M \tau^M) ]^M_{\Tm}       && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1;]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;\left([\pi_1^M\,\tau^M]^M_{\Tm}\;t^M\right) \\
   [\emptysub]^M_{\Tm} & : [\emptysub^M_{\Sub}]^M_{\Tm}                   && \;t^M = %^{\Tm^M\,\Gamma^M\;([\emptysub_{\Sub}^M]^M_{\Ty}\,A^M)\,([\emptysub]_{\Tm}\,t)}
-                      && [ \emptysub^M_{\Sub} ]^M \;t^M \\
-  [,]_{\Tm}^M         & : [ \sigma^M ,_{\Sub}^M t^M ]^M_{\Tm}            && \;u^M = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)} &&
-                      [ \sigma^M ,_{\Sub}^M t^M ]^M \;u \\
-  [\pi_1\idS]^M_{\Tm} & : [ \pi_1^M\,\idS^M ]^M_{\Tm}                    && \;t^M = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{} &&
-                      [ \pi_1^M\,\idS^M ]^M\;t^M \\
-  [\pi_1\pi_1]^M_{\Tm}& : [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M_{\Tm}       && \;t^M = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)} &&
-                      [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M\;t^M
-\end{alignat*}
-\LT{To be finished}
+                      [ \emptysub^M_{\Sub} ]^M  && t^M \\
+  [,]_{\Tm}^M         & : [ \sigma^M ,_{\Sub}^M u^M ]^M_{\Tm}            && \;t^M = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)}
+                      [ \sigma^M ,_{\Sub}^M u^M ]^M && t^M \\
+  [\pi_1\idS]^M_{\Tm} & : [ \pi_1^M\,\idS^M ]^M_{\Tm}                    && \;t^M = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{}
+                      [ \pi_1^M\,\idS^M ]^M && t^M \\
+  [\pi_1\pi_1]^M_{\Tm}& : [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M_{\Tm}       && \;t^M = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)}
+                      [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M && t^M
+\end{alignat*}
+
+\begin{alignat*}{5}
+  ;\text{-}\textsf{idr}^M   & : \sigma^M ;^M \idS^M && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{idr}}} && \sigma^M \\
+  ;\text{-}\textsf{idl}^M   & : \idS^M ;^M \sigma^M && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{idl}}} && \sigma^M \\
+  ;\text{-}\textsf{assoc}^M & : (\sigma^M ;^M \tau^M) ;^M \gamma^M   && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{assoc}}} && \sigma^M ;^M (\tau^M ;^M \gamma^M) \\
+  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black}[\sigma^M]_{\Ty}^M\,t^M}} \\
+  \pi_1\beta^M              & : \pi_1^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\pi_1\beta}}      && \sigma^M \\
+  \pi_2\beta^M              & : \pi_2^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Tm^M\,\Gamma^M}_{\alert{[\pi_1,]^M, \pi_2}}       && t^M \\
+  \cdot\eta^M               & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,\emptyctx^M_{\Ctx}}_{\alert{\cdot\eta}}  && \emptysub_{\Sub}^M \\
+  \pi\eta^M                 & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,(\Delta^M ,^M_{\Ctx} A^M)}_{\alert{\pi\eta}} {} && \pi_1^M \sigma^M ,_{\Sub}^M \pi_2^M \sigma^M \\
+  [\idS]^M                  & : [ \idS ]^M\,t^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
+  [;]^M                     & : [ \sigma^M ;^M \tau^M ]^M\,t^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && [\sigma^M]^M\,([\tau^M]^M\,t^M) 
+\end{alignat*}
+
+\begin{alignat*}{5}
+  []\UU^M & : [ \sigma^M ]^M_{\Ty}\;\UU^M          && =^{\Ty^M\,\Gamma^M\,\UU}      {} && \UU^M \\
+  []\El^M & : [ \sigma^M ]^M_{\Ty}\;(\El^M\;u^M)   && =^{\Ty^M\,\Gamma^M\,(\El\,u)} {} && \El^M\,(\alert{\transfib{}{[]\UU^M}{\color{black}[ \sigma^M ]^M_{\Tm}\,u^M}}
+\end{alignat*}
+
+\begin{alignat*}{5}
+  []\Pi^M         & : [ \sigma^M ]^M_{\Ty} \;(\Pi^M\;A^M\;B^M)  && =^{\Ty^M\,\Gamma^M\,([\sigma]_{\Ty} (\Pi\,A\,B))} && \Pi^M\;([ \sigma^M ]^M_{\Ty}\;A^M)\;([ \sigma^M ]^M_{\Ty}\;B^M) \\
+  []\lambda^M     & : [ \sigma^M ]^M\;(\lambda^M\;t^M)          && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} && \lambda^M\;([ \sigma^M \uparrow^M A^M ]^M\;t^M) \\
+  \Pi\beta^M      & : \mathsf{app}^M\;(\lambda^M\;t^M)          && =^{\Tm^M\,(\Gamma^M,_{\Ctx}^M A^M)\,B^M}_{\alert{\Pi\beta}}      && t^M  \\
+  \Pi\eta^M       & : \lambda^M\;(\mathsf{app}^M\;t^M)          && =^{\Tm^M\,\Gamma^M\,(\Pi\,A^M\,B^M)}_{\alert{\Pi\eta}}           && t^M 
+\end{alignat*}
+
+\begin{alignat*}{5}
+  []\mathsf{Id}^M    :{} & [ \sigma^M ]^M_{\Ty} (\mathsf{Id}^M\;a^M\;t^M\;u^M) =^{\Ty^M\,\Gamma^M\,([\sigma]_{\Ty}(\mathsf{Id}\,a\,t\,u))} \\
+                      & \mathsf{Id}^M\; (\transfib{}{[]\UU^M}{[ \sigma^M ]^M_{\Tm}\,a^M})\;
+                  (\transfib{}{[]\El^M}{[ \sigma^M ]^M_{\Tm}\,t^M})\;
+                  (\transfib{}{[]\El^M}{[ \sigma^M ]^M_{\Tm}\,u^M}) \\
+  []\mathsf{refl}    :{} & [ \sigma^M ]^M\;(\mathsf{refl}^M\;t^M) =^{\Tm^M\;\Gamma^M}_{\alert{[]\mathsf{Id}^M, []\mathsf{refl}}} \\
+                      & \mathsf{refl}^M\;(\transfib{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,\nameless}{[]\El^M}{[\sigma^M]^M_{\Tm} \,t^M}) \\
+  \mathsf{reflect}^M :{} & \Tm^M\;\Gamma^M\;(\mathsf{Id}^M\;a^M\;t^M\;u^M)\;p \to t^M =^{\Tm^M\;\Gamma^M\;A^M}_{\alert{\mathsf{reflect}\;p}} u^M
+\end{alignat*}
 

### 2025-02-18 tex/main.tex
Organise methods
Adjust the layout to save some space

@@
 \usepackage[obeyFinal,textsize=footnotesize]{todonotes}
+\usepackage{float}
 \usepackage{manfnt}
@@
 \lstset{
-    basicstyle=\large\ttfamily,
     breaklines=true,
@@
 \subparagraph*{Plan of the paper}
-In the rest of the paper, we introduce the metalanguage in \cref{sec:meta}.
+In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
 Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
@@
 %
-Inductive-recursive types (IRT) were introduced by Dybjer~\cite{Dybjer2000}, and then extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
+Inductive-recursive types (IRT), introduced by Dybjer~\cite{Dybjer2000}, were extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
 The general schema allows one to define a type $A$ inductively, together with a recursive function $F : A \to D$, for some fixed (possibly large) type $D$.
@@
 with the following constructors:
+%\begin{figure}[H]
+\begin{minipage}{.35\textwidth}
 \begin{alignat*}{3}
@@
   \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
-  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
-  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
-  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
+  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A}
+\end{alignat*}
+\end{minipage}
+\begin{minipage}{.64\textwidth}
+\begin{alignat*}{3}
+  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
+  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
   \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
 \end{alignat*}
-
+\end{minipage}
+%\end{figure}
+\vskip\belowdisplayskip\noindent
 A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
 %; indeed, inductive-inductive definitions only make sense when $D = \Set$.
-The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
+The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}
 
@@
 \subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-\begin{figure}[b]
-\begin{minipage}[t]{.48\textwidth}
+Functions are typically defined informally using pattern matching for convenience and readability.
+Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
+This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
+For example, two definitions of addition below
+%\begin{figure}[H]%
+\begin{minipage}{.48\textwidth}
   \begin{alignat*}{5}
@@
 \end{minipage}
-\begin{minipage}[t]{.5\textwidth}
+\begin{minipage}{.5\textwidth}
 \begin{alignat*}{5}
@@
 \end{minipage}
-\caption{Two seemingly different definitions of addition}
-\label{fig:additions}
-\end{figure}
-Functions are typically defined informally using pattern matching for convenience and readability.
-Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
-This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
-For example, two definitions of addition in \cref{fig:additions} are the same
-after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
+\vskip\belowdisplayskip\noindent
+%\caption{Two seemingly different definitions of addition}
+%\label{fig:additions}
+%\end{figure}
+are the same after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
 Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
@@
 
-Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types in context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from context $\Gamma$ to context $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms in context $\Gamma$ of type $A$.
-This amounts to the following type families, all mutually inductively defined (hence inductive-inductive):
+Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types in context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from context $\Gamma$ to context $\Delta$, and a type $\Tm\;\Gamma\;A$ of $A$-typed terms in $\Gamma$.
+This amounts to the following type families, mutually inductively defined:
 \begin{align*}
@@
 
-Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
+Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e.\ with $\Gamma$ and $\Delta$ being implicit type substitution as a constructor has type
 \begin{alignat*}{3}
@@
 \end{alignat*}
-where $\Gamma$ and $\Delta$ are kept implicit.
+
 \Fredrik[noinline]{Could we change this to $A[\sigma]$ instead?}
@@
 Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
-We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$, because the substitution calculus is supposed to be the initial \emph{category} with families:
+We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$
+%because the substitution calculus is supposed to be the initial \emph{category} with families:
 %Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
@@
 \end{alignat*}
-Note that $[\blank]\blank$ is overloaded for type and term substitution.
+%Note that $[\blank]\blank$ is overloaded for type and term substitution.
 
@@
 \end{minipage}\vskip\belowdisplayskip\noindent
-whose confluence property is easy to see.
+%whose confluence property is easy to see.
 
@@
 However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
-We introduce the type $\bool$ itself, its two constructors $\btrue$ and $\bfalse$, and its elimination principle $\belim$:
+We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$ and its elimination principle:
 %
 \begin{alignat*}{3}
-  \bool     &: \implicit{\Gamma}         \;  && \Ty\;\Gamma \\
-  \btrue, \bfalse &: \implicit{\Gamma}   \;  && \Tm\;\Gamma\;\bool \\
   \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;([ \instSub{\bfalse} ]_{\Ty} P) \to \Tm\;\Gamma\;([ \instSub{\btrue} ]_{\Ty} P) \\
@@
 %
-where we have written $\instSub{t} = (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ for the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place.
+where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ for the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place.
 We can add the expected computation rules
@@
 %
-and the structural rule for type substitution as another straightforward clause
-\[
-  \;[ \sigma ]_{\Ty}\,\bool \reduce \bool \enspace ,
-\]
+and the structural rule for type substitution as another straightforward clause $[ \sigma ]_{\Ty}\,\bool \reduce \bool$,
 but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following:
@@
 \begin{alignat*}{5}
-  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \\ & \quad [ \sigma ] (\belim\,P\,f\,t\,b) =^{\Tm\,\Gamma\,([ \sigma; \instSub{b} ]\,P)}_{\alert{p(b)}} \belim\,([ \sigma^+ ] P)\,&&(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{[ \sigma ]\,f}}}) \\&&&(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{[ \sigma ]\,t}}})\\&&&([\sigma ]\,b)
+  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; [ \sigma ] (\belim\,P\,f\,t\,b) \\
+           & =^{\Tm\,\Gamma\,([ \sigma; \instSub{b} ]\,P)}_{\alert{p(b)}} \belim\,([ \sigma^+ ] P)\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{[ \sigma ]\,f}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{[ \sigma ]\,t}}})\; ([\sigma ]\,b)
 \end{alignat*}
@@
 
-We have used quotient inductive-recursive types and overlapping
-patterns to represent type theory in type theory. By defining type
+We have used QIIRTs and overlapping
+patterns to represent type theory in (intensional) type theory. By defining type
 substitution recursively with overlapping patterns, we can avoid most
-transports that show up in terms in other formalisations of type
-theory in (intensional) type theory. In other words, we have shown that so-called
+coercions that appear in terms in other formalisations of type
+theory in type theory. In other words, we have shown that so-called
 `transport hell' can be minimised when formalising type theory in type
-theory this way, without hiding any details in the formalisation.
+theory this way, without hiding details in the formalisation.
 
@@
 
-The idea of representing type theory in type theory in a well typed manner goes back to Dybjer~\cite{Dybjer1996} (see also Clairambault and Dybjer~\cite{Clairambault2014} and Castellan, Clairambault and Dybjer~\cite{Castellan2021} for recent developments).
+The idea of representing type theory in type theory in a well typed manner goes back to Dybjer~\cite{Dybjer1996} (see also~\cite{Clairambault2014,Castellan2021}).
 Dybjer's notion of \emph{category with families} can be seen as a semantic structure corresponding closely to the syntax of type theory, and many of the formalisations of type theory in type theory can be understood as capturing the initial category with families via inductive or quotient-inductive types.
@@
 However because he does not use quotients, he also has an explicit term constructor for transporting terms along an inductively defined type of conversion rules, which explicitly contains congruence rules.
-We are able to remove a lot of the boilerplate by using actual equality constructors instead.
+We are able to remove a lot of the boilerplate by using equality constructors instead.
 Chapman~\cite{Chapman2009} similarly uses explicit term substitutions and explicit transports along conversion rules, but also uses explicit type substitition in favour of more straightforward foundations.
@@
 
-Altenkirch and Kaposi~\cite{Altenkirch2016a} were the first to exploit QIITs to simplify the treatment of computation rules in representations of type theory in type theory.
+Altenkirch and Kaposi~\cite{Altenkirch2016a} were the first to exploit QIITs to simplify the treatment of computation rules in representations of type theory.% in type theory.
 We build heavily on their ideas, but by using Danielsson's~\cite{Danielsson2006} idea of defining type substitution recursively, we are able to remove many explicit transports from the definition, with the result that it is closer to an informal exposition on paper.
-Such a benefit was also suggested by Kaposi's~\cite{Kaposi2023} work, where the simply typed lambda calculus was used as an example.
+Such a benefit was also suggested by Kaposi's~\cite{Kaposi2023} work, where simply typed lambda calculus was used as an example.
 Altenkirch and Kaposi has built on the QIIT representation with great success to derive non-trivial results such as normalisation by evaluation~\cite{Altenkirch2017}, and investigate the metatheory of QIITs themselves, together with Kov\'acs~\cite{Kaposi2019}.
@@
 
+\subparagraph*{Substitution calculus with a base type $\UU$}
 \begin{alignat*}{10}
@@
   [\blank]_{\Ty}^M\blank & : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  \UU^M                  & : \Ty^M\;\Gamma^M\;\UU \\
   \emptysub_{\Sub}^M     & : \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
@@
   \pi_2^M                & : (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;([ \pi_1^M\,\sigma^M ]^M_{\Ty}\;A^M)\;(\pi_2\,\sigma) \\
-  [\blank]^M\blank       & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t) \\
-  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
-  [\blank]_{\Tm}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
-  \UU^M                  & : \Ty^M\;\Gamma^M\;\UU \\
-  \El^M                  & : \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u) \\
-  \Pi^M                  & : (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
-  \lambda^M              & : \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;t \to \Tm^M\;\Gamma^M\;(\Pi^M\;A^M\;B^M)\;(\lambda\,t) \\
-  \mathsf{app}^M         & : \Tm^M\;\Gamma^M\;(\Pi^M\,A^M\,B^M)\;t \to \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;(\mathsf{app}\,t) \\
-  \mathsf{Id}^M          & : (a^M : \Tm^M\;\Gamma^M\;\UU^M\;a) \to \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;t \to \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;u \to \Ty^M\;\Gamma^M\;(\mathsf{Id}\;a\;t\;u)  \\
-  \mathsf{refl}^M        & : (t^M : \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;t) \to \Tm^M\;\Gamma^M\;(\mathsf{Id}^M\;a^M\;t^M\;t^M)\;(\mathsf{refl}\;t) 
-\end{alignat*}
-
-\begin{alignat*}{5}
-  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M ([ \sigma^M ]^M\;A^M]))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
-  (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;\pi_2\idS}{([;]^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
+  [\blank]^M\blank       & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t)
 \end{alignat*}
@@
   [\pi_1,]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ,^M_{\Sub} t^M) ]^M_{\Ty} && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\,A^M \\
-  [\pi_1;]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ;^M \tau^M) ]^M_{\Ty}     && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\left([ \pi_1^M\,\tau^M]^M_{\Ty}\,A^M\right)
-\end{alignat*}
-
-\begin{alignat*}{5}
-  \idS\mathord{\uparrow}^M      & : \idS^M              && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\idS}_{\alert{[\idS]^M_{\Ty}}} {} && \idS^M \\
-  \mathord{;\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
-  \mathord{\pi_1,\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
-  \mathord{\pi_1;\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
-  \mathord{\emptysub^M_{\Sub}\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
-  ,\mathord{\uparrow}^M & : (\sigma^M ,_{\Sub}^M t^M)  && \uparrow^M A^M = (\sigma^M ,_{\Sub}^M t^M)^{+M} \\
-  \mathord{\pi_1\idS\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
-  \mathord{\pi_1\pi_1\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
+  [\pi_1;]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ;^M \tau^M) ]^M_{\Ty}     && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\left([ \pi_1^M\,\tau^M]^M_{\Ty}\,A^M\right) \\
+  []\UU^M             & : [ \sigma^M ]^M_{\Ty}                           && \;\UU^M =^{\Ty^M\,\Gamma^M\,\UU}      {} && \UU^M
+\end{alignat*}
+
+\begin{alignat*}{5}
+  ;\text{-}\textsf{idr}^M   & : \sigma^M ;^M \idS^M && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{idr}}} && \sigma^M \\
+  ;\text{-}\textsf{idl}^M   & : \idS^M ;^M \sigma^M && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{idl}}} && \sigma^M \\
+  ;\text{-}\textsf{assoc}^M & : (\sigma^M ;^M \tau^M) ;^M \gamma^M   && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{assoc}}} && \sigma^M ;^M (\tau^M ;^M \gamma^M) \\
+  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black}[\sigma^M]_{\Ty}^M\,t^M}} \\
+  \pi_1\beta^M              & : \pi_1^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\pi_1\beta}}      && \sigma^M \\
+  \pi_2\beta^M              & : \pi_2^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Tm^M\,\Gamma^M}_{\alert{[\pi_1,]^M, \pi_2\beta}}       && t^M \\
+  \cdot\eta^M               & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,\emptyctx^M_{\Ctx}}_{\alert{\cdot\eta}}  && \emptysub_{\Sub}^M \\
+  \pi\eta^M                 & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,(\Delta^M ,^M_{\Ctx} A^M)}_{\alert{\pi\eta}} {} && \pi_1^M \sigma^M ,_{\Sub}^M \pi_2^M \sigma^M \\
+  [\idS]^M                  & : [ \idS ]^M\,t^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
+  [;]^M                     & : [ \sigma^M ;^M \tau^M ]^M\,t^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && [\sigma^M]^M\,([\tau^M]^M\,t^M) 
+\end{alignat*}
+
+\subparagraph*{Universe of small types}
+\begin{alignat*}{10}
+  [\blank]_{\Tm}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
+  \El^M                  & : \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  ;\text{-}\textsf{idr}^M   & : \sigma^M ;^M \idS^M && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{idr}}} && \sigma^M \\
-  ;\text{-}\textsf{idl}^M   & : \idS^M ;^M \sigma^M && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{idl}}} && \sigma^M \\
-  ;\text{-}\textsf{assoc}^M & : (\sigma^M ;^M \tau^M) ;^M \gamma^M   && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{assoc}}} && \sigma^M ;^M (\tau^M ;^M \gamma^M) \\
-  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black}[\sigma^M]_{\Ty}^M\,t^M}} \\
-  \pi_1\beta^M              & : \pi_1^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\pi_1\beta}}      && \sigma^M \\
-  \pi_2\beta^M              & : \pi_2^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Tm^M\,\Gamma^M}_{\alert{[\pi_1,]^M, \pi_2}}       && t^M \\
-  \cdot\eta^M               & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,\emptyctx^M_{\Ctx}}_{\alert{\cdot\eta}}  && \emptysub_{\Sub}^M \\
-  \pi\eta^M                 & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,(\Delta^M ,^M_{\Ctx} A^M)}_{\alert{\pi\eta}} {} && \pi_1^M \sigma^M ,_{\Sub}^M \pi_2^M \sigma^M \\
-  [\idS]^M                  & : [ \idS ]^M\,t^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
-  [;]^M                     & : [ \sigma^M ;^M \tau^M ]^M\,t^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && [\sigma^M]^M\,([\tau^M]^M\,t^M) 
-\end{alignat*}
-
-\begin{alignat*}{5}
-  []\UU^M & : [ \sigma^M ]^M_{\Ty}\;\UU^M          && =^{\Ty^M\,\Gamma^M\,\UU}      {} && \UU^M \\
   []\El^M & : [ \sigma^M ]^M_{\Ty}\;(\El^M\;u^M)   && =^{\Ty^M\,\Gamma^M\,(\El\,u)} {} && \El^M\,(\alert{\transfib{}{[]\UU^M}{\color{black}[ \sigma^M ]^M_{\Tm}\,u^M}}
+\end{alignat*}
+
+\subparagraph*{$\Pi$-types}
+\begin{alignat*}{5}
+  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M ([ \sigma^M ]^M\;A^M]))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
+  (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;\pi_2\idS}{([;]^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
+\end{alignat*}
+
+\begin{alignat*}{10}
+  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
+  \Pi^M                  & : (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
+  \lambda^M              & : \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;t \to \Tm^M\;\Gamma^M\;(\Pi^M\;A^M\;B^M)\;(\lambda\,t) \\
+  \mathsf{app}^M         & : \Tm^M\;\Gamma^M\;(\Pi^M\,A^M\,B^M)\;t \to \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;(\mathsf{app}\,t)
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \idS\mathord{\uparrow}^M      & : \idS^M              && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\idS}_{\alert{[\idS]^M_{\Ty}}} {} && \idS^M \\
+  \mathord{;\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} \\
+                                &&& (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
+  \mathord{\pi_1,\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
+  \mathord{\pi_1;\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} \\
+                                &&& (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
+  \mathord{\emptysub^M_{\Sub}\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
+  ,\mathord{\uparrow}^M & : (\sigma^M ,_{\Sub}^M t^M)  && \uparrow^M A^M = (\sigma^M ,_{\Sub}^M t^M)^{+M} \\
+  \mathord{\pi_1\idS\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
+  \mathord{\pi_1\pi_1\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
 \end{alignat*}
@@
   \Pi\eta^M       & : \lambda^M\;(\mathsf{app}^M\;t^M)          && =^{\Tm^M\,\Gamma^M\,(\Pi\,A^M\,B^M)}_{\alert{\Pi\eta}}           && t^M 
+\end{alignat*}
+
+\subparagraph*{Extensional identity types}
+\begin{alignat*}{10}
+  \mathsf{Id}^M          & : (a^M : \Tm^M\;\Gamma^M\;\UU^M\;a) \to \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;t \to \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;u \to \Ty^M\;\Gamma^M\;(\mathsf{Id}\;a\;t\;u)  \\
+  \mathsf{refl}^M        & : (t^M : \Tm^M\;\Gamma^M\;(\El^M\,a^M)\;t) \to \Tm^M\;\Gamma^M\;(\mathsf{Id}^M\;a^M\;t^M\;t^M)\;(\mathsf{refl}\;t) 
 \end{alignat*}

### 2025-02-18 tex/main.tex
Adjust the layout to save some space
[ sigma ] A replaced by A [ sigma ]

@@
 \begin{alignat*}{3}
-  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
-\end{alignat*}
-
-\Fredrik[noinline]{Could we change this to $A[\sigma]$ instead?}
-\LT[noinline]{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
-\LT[noinline]{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}
+  \sub{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma .
+\end{alignat*}
 
@@
   \qquad\qquad
-  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A).
-\]
-Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
-We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$
+  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}) \to \Sub\;\Gamma\;(\Delta, A).
+\]
+Note that type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
+We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$.
 %because the substitution calculus is supposed to be the initial \emph{category} with families:
-%Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
 \[
@@
   \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
-  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\, A)
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\sub{A}{\pi_1\,\sigma})
 \end{alignat*}
 respectively.
-Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
+Again, a type substitution $\sub{A}{\pi_1\,\sigma}$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
 Finally, we have term substitution, which given $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in $\Delta$ to terms in $\Gamma$, appropriately preserving the type, which is also substituted:
 \begin{alignat*}{3}
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\, A).
-\end{alignat*}
-%Note that $[\blank]\blank$ is overloaded for type and term substitution.
+  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
+\end{alignat*}
+%Note that $\sub{\blank}{\blank}$ is overloaded for type and term substitution.
 
@@
 \begin{alignat*}{3}
-  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && [ \idS ] \,A           && =^{\Ty\,\Gamma}\;A \\
-  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& [ \sigma ; \tau ]\,A  && =^{\Ty\,\Gamma}\;[ \sigma ]\,([ \tau ]\,A) \\
-  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma}\;\UU
+  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && \sub{A}{\idS}           && =^{\Ty\,\Gamma}\;A \\
+  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& \sub{A}{\sigma ; \tau}  && =^{\Ty\,\Gamma}\; \sub{(\sub{A}{\tau})}{\sigma} \\
+  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && \sub{\UU}{\sigma}        && =^{\Ty\,\Gamma}\;\UU
 \end{alignat*}
@@
   \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
 \end{alignat*}
@@
 Composition of substitutions $\blank;\blank$ acts like a list concatenation, so
-we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), [\sigma]\,t$.
-However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead.
-Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ which we can transport along to fix up the equation.
+we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), \sub{t}{\sigma}$.
+However $\sub{t}{\sigma}$ is of type $\Tm\;\Gamma\;(\sub{\sub{A}{\tau}}{\sigma})$, but the extension $(\sigma; \tau), \sub{t}{\sigma}$ requires $\sub{t}{\sigma}$ to have type $\Tm\;\Gamma\;(\sub{A}{\sigma; \tau})$ instead.
+Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon \sub{\sub{A}{\tau}}{\sigma} = \sub{A}{\sigma;\tau}$ which we can transport along to fix up the equation.
 
@@
   \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
-\end{alignat*}
-but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$.
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,(\sub{A}{\blank})}_{\alert{\pi_1\beta}} &&  t, 
+\end{alignat*}
+but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;(\sub{A}{\pi_1\,(\sigma, t)})$ instead of $\Tm\;\Gamma\;(\sub{A}{\sigma})$.
 %whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
@@
 \[
-  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {[\,\idS\,]\,t} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
+  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {\sub{t}{\idS}} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
   \qquad\qquad
-  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {[\,\sigma ; \tau\,]\,t} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{[\, \sigma\,]\,[\,\tau\,]\,t}
+  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {\sub{t}{\sigma ; \tau}} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{\sub{\sub{t}{\tau}}{\sigma}}
 \]
@@
 \begin{example}\label{ex:pi2-comp}
-  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$.
+  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $\sub{A}{\sigma;\tau}$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$.
   In short, the following identity apparently holds
   \[
-    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
+    \pi_2\,(\sigma ; \tau) = \sub{(\pi_2\,\tau)}{\sigma}
   \]
@@
     = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
-    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
-    = [\sigma] (\pi_2\,\tau).
+    = \pi_2\,(\sigma;\pi_1\,\tau, \sub{(\pi_2\,\tau)}{\sigma})
+    = \sub{(\pi_2\,\tau)}{\sigma}.
   \end{equation}
-  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
+  Yet, the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
   Hence the above identity does not even make sense, since their types do not match.
-  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
+  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} \sub{(\pi_2\,\tau)}{\sigma}$ for some $p: \sub{A}{\pi_1(\sigma; \tau)} = \sub{\sub{A}{\pi_1\tau}}{\sigma}$ or, equivalently
   \begin{equation}\label{eq:pi2-comp-real-proof}
-    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
+    (\sub{A}{\pi_1(\sigma; \tau)}, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} (\sub{\sub{A}{\pi_1\,\tau}}{\sigma}, \sub{(\pi_2\,\tau)}{\sigma})
   \end{equation}
@@
   \begin{alignat*}{3}
-         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
-    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
-    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
-    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
-    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
+         & (\sub{A}{\pi_1(\sigma; \tau)}                    &&, \pi_2\,(\sigma ; \tau)) \\
+    = {} & (\sub{A}{\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))} &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
+    = {} & (\sub{A}{\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}\sub{(\pi_2\;\tau)}{\sigma}})} &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}\sub{(\pi_2\;\tau)}{\sigma}})) \\
+    = {} & (\sub{A}{\sigma;\pi_1\,\tau} &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}\sub{(\pi_2\;\tau)}{\sigma}})) \nonumber \\
+    = {} & (\sub{\sub{A}{\pi_1\tau}}{\sigma} &&, \sub{(\pi_2\,\tau)}{\sigma}). \nonumber
   \end{alignat*}
@@
 \begin{alignat}{3}
-%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
-[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
-[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
-[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
+%\subTy{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \nonumber \\
+\subTy{\UU}{\sigma}  & = \UU                                   \label{eq:type-sub-at2-1} \\
+\subTy{A}{\idS}    & = A                                     \label{eq:type-sub-at2-2} \\
+\subTy{A}{\sigma ; \tau}    & = \subTy{(\subTy{A}{\tau})}{\sigma} \label{eq:type-sub-at2-3}
 \end{alignat}
@@
 \[
-[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
-[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
-[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
-\]
-Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
+\subTy{\UU}{\emptysub}         \judgeq \UU, \qquad
+\subTy{\UU}{\sigma, t}         \judgeq \UU, \qquad\text{and}\qquad 
+\subTy{\UU}{\pi_1\,\sigma}     \judgeq \UU,
+\]
+Thus in this case we have again lost the desired strict identity $\subTy{\UU}{\sigma} \equiv \UU$.
 %
@@
 {rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
-In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
+In addition, we can include other derived rules such as $\subTy{A}{\pi_1(\sigma, t)} = \subTy{A}{\sigma}$ as long as they are locally confluent and terminating.
 We end up with the following definition for type substitution:
 \begin{alignat}{3}
-[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
-[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
-[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
-[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
-[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
-[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
+\subTy{\blank}{\blank}            &: \implicit{\Gamma, \Delta} \; \Ty\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty\,\Gamma \nonumber \\
+\subTy{A}{\idS}                & \reduce A \label{eq:def-type-subst-1}\\
+\subTy{A}{\sigma ; \tau}       & \reduce \subTy{(\subTy{A}{\tau})}{\sigma} \label{eq:def-type-subst-2}\\
+\subTy{A}{\pi_1(\sigma, t)}    & \reduce \subTy{A}{\sigma} \label{eq:def-type-subst-3} \\
+\subTy{A}{\pi_1(\sigma; \tau)} & \reduce \subTy{(\subTy{A}{\pi_1\tau})}{\sigma} \label{eq:def-type-subst-4} \\
+\subTy{\UU}{\sigma}            & \reduce \UU                                   \label{eq:def-type-subst-5}
 \end{alignat}
@@
 \begin{proposition}\label{prop:local-confluence-1}
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
+  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
   \danger
@@
   \[
-    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
+    \subTy{\UU}{\idS} \qquad \subTy{\UU}{\sigma;\tau}, \qquad \subTy{\UU}{\pi_1(\sigma, t)}, \qquad\text{and}\qquad \subTy{\UU}{\pi_1(\sigma; \tau)}
   \]
   and show that each applicable clause does result in $\UU$.
-
-  For example, $[\sigma;\tau]_{\Ty}\;\UU \reduce [\sigma]_{\Ty}\;([\tau]_{\Ty}\;\UU) \reduce [\sigma]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $[\sigma;\tau]_{\Ty}\;\UU \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
+  %
+  For example, $\subTy{\UU}{\sigma;\tau} \reduce \subTy{(\subTy{\UU}{\tau})}{\sigma} \reduce \subTy{\UU}{\sigma} \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $\subTy{\UU}{\sigma;\tau} \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
   It is easy to see that remaining local peaks reduce to the same term.
@@
 \begin{alignat*}{5}
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t) && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), [ \sigma ]\;t \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t) && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
   \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
-  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;([ \tau ]\;t})
+  [\idS]t         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A} && t \\
+  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,\sub{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}}
 \end{alignat*}
@@
 \begin{example} \label{ex:pi-comp-qiirt}
-  The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
-  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is strictly equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
+  The equation $\pi_2\,(\sigma; \tau) = \sub{(\pi_2\,\tau)}{\sigma}$ can be formally stated without any transport.
+  The type of the term on the LHS is $\subTy{A}{\pi_1(\sigma;\tau)}$ which is strictly equal to the type $\subTy{(\subTy{A}{\pi_1\tau})}{\sigma}$ of the term on the RHS.
   Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are strictly equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
@@
 
-As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
+As we are defining a function on a quotient inductive type, we have to prove that $\subTy{\blank}{\blank}$ is coherent, i.e., that it respects all equality constructors.
 %\LT[noinline]{Should we mention any general condition for coherence?}
@@
 \begin{proposition}\label{prop:coherence-1}
-  For every equality constructor $p : \sigma = \tau$, we have $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$.
+  For every equality constructor $\cc : \sigma = \tau$, we have $\subTy{A}{\sigma} = \subTy{A}{\tau}$.
   \danger
@@
 \begin{proof}
-  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
-  \begin{align*}
-    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
-                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
-                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
-    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
-    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+  The equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ turn into definitional equalities for type substitution, so it remains to show that the following identities
+  \begin{align*}
+    \subTy{A}{\sigma;\idS}          & = \subTy{A}{\sigma},
+                                    & \subTy{A}{\idS;\sigma}          & = \subTy{A}{\sigma},
+                                     & \subTy{A}{(\sigma;\tau);\gamma} & = \subTy{A}{\sigma; (\tau; \gamma)}, \\
+    \subTy{A}{\pi_1(\sigma, t)}      & = \subTy{A}{\sigma}, \\
+    \subTy{A}{\sigma; (\tau, t)}    & = \subTy{A}{(\sigma;\tau), \sub{t}{\sigma}},
+                                     & \subTy{A}{\sigma}               & = \subTy{A}{\emptyctx},
+                                     & \subTy{A}{\sigma}               & = \subTy{A}{\pi_1\sigma, \pi_2\sigma}
   \end{align*}
   hold propositionally.
-  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.
-
-  The sole identity of the second row is clause \eqref{eq:def-type-subst-3} of the definition of $[\blank]_{\Ty}\blank$.
+  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $\subTy{(\subTy{(\subTy{A}{\gamma})}{\tau})}{\sigma}$ by \eqref{eq:def-type-subst-2}.
+
+  The sole identity of the second row is clause \eqref{eq:def-type-subst-3} of the definition of $\subTy{\blank}{\blank}$.
 
@@
   \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
-  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
+  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
 \end{alignat*}
 %as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
-The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.
-
+The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
+%
 In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
-Likewise, the equality constructor $[]\El$ may be replaced by 
-\begin{alignat*}{3}
-  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u})
+Likewise, the equality constructor $[]\El$ may be replaced by the reduction
+\begin{alignat*}{3}
+  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
 \end{alignat*}
@@
 \[
-  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u).
-\]
-cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.
-
-To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
-\begin{alignat*}{3}
-  [\blank]_{\Tm}\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
-[ \idS ]_{\Tm}\,t          & \reduce t \\
-[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
-[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
-[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
-[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
+  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
+\]
+cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
+
+To repair the local confluence, we add a simultaneously defined function $\subTm{\blank}{\blank}$ for term substitution apart from the explicit term substitution $\sub{\blank}{\blank}$ as follows
+\begin{alignat*}{3}
+  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
+\subTm{t}{\idS}          & \reduce t \\
+\subTm{t}{\sigma ; \tau} & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
+\subTm{t}{\pi_1(\sigma, s)} & \reduce \subTm{t}{\sigma} \\
+\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
+\subTm{t}{\sigma}        & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
 \end{alignat*}
 which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
-The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
-Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
+The remaining cases reduce to the explicit term substitution $\sub{t}{\sigma}$.
+Note that $\subTm{\blank}{\blank}$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
 \begin{remark}
-  The function $[\blank]_{\Tm}\blank$ can be understood as a \emph{smart constructor} for the explicit term substitution $[\blank]\blank$, a well-known idiom in functional programming, since the function $[\blank]_{\Tm}\blank$ (mostly) builds an explicit substitution but computes certain cases away.
+  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming, since the function $\subTm{\blank}{\blank}$ (mostly) builds an explicit substitution but computes certain cases away.
 \end{remark}
 
-Then, the equality constructor $[]\El$ is accordingly changed to 
+The equality constructor $[]\El$ is now accordingly replaced by the reduction
 \begin{align}
-  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
+  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\subTm{u}{\sigma}) \label{eq:def-type-subst-7}
 \end{align}
@@
 \begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
+  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
   \qed
@@
 
-We also have to show that the recursive substitution $[\blank]_{\Tm}\blank$ is propositionally equal to the explicit substitution $[\blank]\blank$, so they can be used together consistently.
+We also have to show that the recursive substitution $\subTm{\blank}{\blank}$ is propositionally equal to the explicit substitution $\sub{\blank}{\blank}$, so they can be used together consistently.
 
 \begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
-  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$. \qed
+  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$. \qed
 \end{restatable}
 
-Finally, we have to show that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent.
+Finally, we have to show that both $\subTy{\blank}{\blank}$ and $\subTm{\blank}{\blank}$ are coherent.
+%
 \begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
-  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the identities
+  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, the identities
   \danger
   \[
-    [\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A
+    \subTy{A}{\sigma} = \subTy{A}{\tau}
     \quad\text{and}\quad
-    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t
+    \subTm{t}{\sigma} = \subTm{t}{\tau}
   \]
@@
 \begin{alignat}{5}
-  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
-  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
+  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && \subTy{(\Pi\;A\;B)}{\sigma} && =^{\Ty\;\Gamma} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-8} \\
+  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && \subTm{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
 \end{alignat}
@@
 we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
-However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $[\blank]_{\Ty}\blank$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
-For example, $[\idS]_{\Ty}(\Pi\;A\;B)$ reduces to $\Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow [\idS]_{\Ty}(\Pi\;A\;B) \reduce \Pi\;A\;([\idS \uparrow A]_{\Ty}B)$ cannot be joined.
+However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $\subTy{\blank}{\blank}$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
+For example, $\subTy{(\Pi\;A\;B)}{\idS}$ reduces to $\Pi\;A\;(\subTy{B}{\idS \uparrow A})$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow \subTy{(\Pi\;A\;B)}{\idS} \reduce \Pi\;A\;(\subTy{B}{\idS \uparrow A})$ cannot be joined.
 
@@
 \begin{alignat}{3}
-  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, [\sigma]_{\Ty}\,A)\, (\Delta, A) \label{eq:type-of-lifting} \\
+  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, \subTy{A}{\sigma})\, (\Delta, A) \label{eq:type-of-lifting} \\
 \idS                \uparrow A & \reduce \idS \nonumber \\
@@
 \begin{alignat}{5}
-                     & [ \sigma ]_{\Ty}\,(\Pi\,A\,B) && \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \label{eq:def-type-subst-9} \\
-  \mathsf{[]\lambda} :{} & [\sigma]\, (\lambda\,t) && =^{\Tm\,\Gamma\,([\sigma]_{\Ty}(\Pi\,A\,B))} && \lambda\,([\sigma \uparrow A]\,\,t) \nonumber
+                     & \subTy{(\Pi\,A\,B)}{\sigma} && \reduce \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-9} \\
+  \mathsf{[]\lambda} :{} & \sub{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
 \end{alignat}
@@
 \begin{restatable}{proposition}{propConfluenceThree}\label{prop:local-confluence-3}
-  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
+  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
   \qed
@@
     \qquad\text{and}\qquad
-    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, [\sigma; \tau]_{\Ty}\,A)\;(\Theta, A)} \sigma^+ ; \tau^+.
+    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+.
   \]
@@
 
-As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $[\sigma]\,t = [\sigma]_{\Tm}\,t$ but the same argument still applies:
+As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ but the same argument still applies:
 \begin{proposition} \label{prop:correctness-3}
   For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$,
-  the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
+  the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
 \end{proposition}
@@
 \begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
-    For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
+    For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, we have
     \danger
     \[
-      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
+      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A
       \qquad\text{and}\qquad
-      [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
+      \subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau},
     \]
@@
 
-We also need to prove the coherence for type substitution, i.e., that $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
+We also need to prove the coherence for type substitution, i.e., that $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
 To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
 \begin{align*}
-  [\sigma]_{\Ty}\,(\Pi\;A\;B) & = \Pi\;([\sigma]_{\Ty}\,A)\;([\sigma\uparrow A]_{\Ty}\,B) \\
-  [\tau]_{\Ty}\,(\Pi\;A\;B) & = \Pi\;([\tau]_{\Ty}\,A)\;([\tau\uparrow A]_{\Ty}\,B)
-\end{align*}
-but the induction hypothesis does not give us $[\sigma\uparrow A]_{\Ty}\,B = [\tau\uparrow A]_{\Ty}\,B$.
+  \subTy{(\Pi\;A\;B)}{\sigma} & = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \\
+  \subTy{(\Pi\;A\;B)}{\tau} & = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
+\end{align*}
+but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
 Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
-Then, for each constructor $p : \sigma = \tau$, we show the identity
-$[\sigma \mathop{\upuparrows} \Xi ]_{\Ty}\,A = [\tau \mathop{\upuparrows} \Xi ]_{\Ty}\,A$
-for arbitrary $\Xi$, subsuming the special case $[\sigma ]_{\Ty}\,A = [\tau ]_{\Ty}\,A$ when $\Xi$ is empty.
+Then, for each constructor $\cc : \sigma = \tau$, we show the identity
+$\subTy{A}{\sigma \mathop{\upuparrows} \Xi} = \subTy{A}{\tau \mathop{\upuparrows} \Xi}$
+for arbitrary $\Xi$, subsuming the special case $\subTy{A}{\sigma} = \subTy{A}{\tau}$ when $\Xi$ is empty.
 
@@
 \blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
-\to \Sub\;(\Gamma \dplus ([\sigma]_{\Tel}\,\Xi))\;(\Delta \dplus \Xi)
-\]
-which requires us to also simultaneously define telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$.
+\to \Sub\;(\Gamma \dplus (\subTel{\Xi}{\sigma}))\;(\Delta \dplus \Xi)
+\]
+which requires us to also simultaneously define telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$.
 
@@
 \begin{alignat*}{5}
-  & [ \sigma ]_{\Tel}\,\emptytel && \reduce \emptytel \\
-  & [ \sigma ]_{\Tel}\,(\Xi, A)  && \reduce [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A
+  & \subTel{\emptytel}{\sigma} && \reduce \emptytel \\
+  & \subTel{(\Xi, A)}{\sigma}  && \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
 \end{alignat*}
@@
 
-To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows [\tau]_{\Tel} \Xi); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
-%
-Instead of resorting to \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
+To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows \subTel{\Xi}{\tau}); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus \subTel{\Xi}{\idS})\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
+%
+Instead of resorting to \transp, we also turn telescope substitution $\subTel{\blank}{\blank}$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
 
 \begin{minipage}{.5\textwidth}
-\begin{alignat*}{5}
-  & [ \idS   ]_{\Tel}       \,&& \Xi && \reduce \Xi \\
-  & [ \sigma ; \tau ]_{\Tel}\,&& \Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
+\begin{alignat*}{3}
+  & \subTel{\Xi}{\idS}       \Xi && \reduce \Xi \\
+  & \subTel{\Xi}{\sigma ; \tau} && \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
 \end{alignat*}
@@
   & \idS          && \upuparrows \Xi && \reduce \idS \\
-  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
+  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
 \end{alignat*}
@@
 {proposition}{coherenceTel}
-  For every equality constructor $p : \sigma = \tau$ and telescope $\Xi$, we have
-  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$. \qed
+  For every equality constructor $\cc : \sigma = \tau$ and telescope $\Xi$, we have
+  $\subTel{\Xi}{\sigma} = \subTel{\Xi}{\tau}$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\sigma})\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$. \qed
   \danger
@@
 
-As promised, by instantiating $\Xi$ to be empty, we also get coherence for $[\blank]_{\Ty}$.
+As promised, by instantiating $\Xi$ to be empty, we also get coherence for $\subTy{\blank}{\blank}$.
 
 \begin{corollary}
-  We have $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
+  We have $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
   \qed
@@
 We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also `ordinary' application
-$t \mathop{\$} u \defeq [ (\idS , t) ]_{\Tm} (\mathsf{app}\,t)$ of type $\Tm\,\Gamma\,([(\idS , t)] B)$ can be defined without any transport.
-This is in contrast to the pure QII situation, where application is defined as $[ (\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}}) ]_{\Tm} (\mathsf{app}\,t)$.
+$t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS , t)}$ of type $\Tm\,\Gamma\,(\sub{B}{(\idS , t)})$ can be defined without any transport.
+This is in contrast to the pure QII situation, where application is defined as $\subTm{(\mathsf{app}\,t)}{(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})}$.
 %
@@
 \begin{align*}
-  [\sigma \uparrow A ]\,(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]\,(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
-                                              & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
-                                              & = \mathsf{app}\,([\sigma]\,t) && \text{by $\Pi\eta^{-1}$.}
+  \sub{(\mathsf{app}\;t)}{\sigma \uparrow A} & = \mathsf{app}\,\lambda(\sub{(\mathsf{app}\;t)}{\sigma \uparrow A}) && \text{by $\Pi\beta^{-1}$} \\
+                                              & = \mathsf{app}\,\left(\sub{(\lambda\,(\mathsf{app}\,t))}{\sigma}\right) && \text{by $[]\lambda^{-1}$} \\
+                                              & = \mathsf{app}\,(\sub{t}{\sigma}) && \text{by $\Pi\eta^{-1}$.}
 \end{align*}
@@
 \begin{alignat*}{3}
-  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;([ \instSub{\bfalse} ]_{\Ty} P) \to \Tm\;\Gamma\;([ \instSub{\btrue} ]_{\Ty} P) \\
-  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;([ \instSub{b} ]_{\Ty} P)
+  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}}) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}}) \\
+  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{b}})
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;[\instSub{\bfalse}]_{\Ty} P} {} && f \\
-  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;[\instSub{\btrue}]_{\Ty} P} {} && t
-\end{alignat*}
-%
-and the structural rule for type substitution as another straightforward clause $[ \sigma ]_{\Ty}\,\bool \reduce \bool$,
+  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
+  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
+\end{alignat*}
+%
+and the structural rule for type substitution as another straightforward clause
+$\subTy{\bool}{\sigma} \reduce \bool$,
 but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following:
@@
 \begin{alignat*}{5}
-  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; [ \sigma ] (\belim\,P\,f\,t\,b) \\
-           & =^{\Tm\,\Gamma\,([ \sigma; \instSub{b} ]\,P)}_{\alert{p(b)}} \belim\,([ \sigma^+ ] P)\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{[ \sigma ]\,f}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{[ \sigma ]\,t}}})\; ([\sigma ]\,b)
-\end{alignat*}
-where the proof $p(x) : [ \instSub{[ \sigma ]_{\Ty}\,x} ; \sigma^+ ]\;P = [ \sigma ; \instSub{x} ]_{\Ty}\;P$ is needed to line up the types.
+  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
+           & =^{\Tm\,\Gamma\,(\sub{P}{\sigma; \instSub{b}})}_{\alert{p(b)}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
+\end{alignat*}
+where the proof $p(x) : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types.
 Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems.
@@
 \begin{alignat*}{3}
-  [\blank]_{\Ty}^M\blank & : \implicit{\Gamma, \Gamma^M,\Delta, \Delta^M, \sigma, A} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) %\\
+  \blank[\blank]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M,\Delta, \Delta^M, \sigma, A} \; && \Ty^M\;\Delta^M\;A \to \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Gamma^M\;(\subTy{A}{\sigma}) %\\
 %  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) %\\
@@
 \begin{alignat*}{5}
-  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
+  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && A^M [ \idS^M ]^M_{\Ty} && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
   %[;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
@@
   \[
-    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\;[\idS^M]^M_{\Tm}\;t^M  =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
+    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\;t^M [\idS^M]^M_{\Tm}  =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
   \]
@@
 %
-Interestingly, a key difference compared to its QIIT counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
+Interestingly, a key difference compared to its QIIT counterpart is that $\subTy{\blank}{\blank}$, $\subTm{\blank}{\blank}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
 %type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
@@
 \begin{alignat*}{5}
-  \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
-  \elim_{\Tm}[] & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
+  \elim_{\Ty}[] & : \elim_{\Ty}\,(\subTy{A}{\sigma}) && = && (\elim_{\Ty}\,A)[\elim_{\Sub}\,\sigma]^M_{\Ty} \\
+  \elim_{\Tm}[] & : \elim_{\Tm}\,(\subTm{t}{\sigma}) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Tm}\,t) [\elim_{\Sub}\,\sigma]^M_{\Tm} \\
   \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
@@
 These identities are needed for the computation rules where strict identities are used in syntax.
-For example, for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, we have 
+For example, for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;(\subTy{A}{\sigma})$, we have 
 \[
@@
 \]
-where the transport is needed since $\elim_{\Tm}\;t : \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t$ but $\blank ,_{\Sub}^M \blank$ expects its second argument to have type $\Tm^M\;(\elim_{\Ctx}\,\Gamma)\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t$.
+where the transport is needed since $\elim_{\Tm}\;t : \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\alert{\elim_{\Ty}\;(\subTy{A}{\sigma})})\;t$ but $\blank ,_{\Sub}^M \blank$ expects its second argument to have type $\Tm^M\;(\elim_{\Ctx}\,\Gamma)\;\left(\alert{(\elim_{\Ty}\,A)[\elim_{\Sub}\,\sigma]^M_{\Ty}}\right)\;t$.
 
@@
 On the other hand, the converse $(\blank)^{\prec}$ has some choices.
-Since in the QIIT definition there is only one term substitution $[\blank]_{\I}\blank$ and lifting $(\blank)^+_{\I}$, we may choose to translate $[\sigma]_{\I} \,t$ to the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ to carry out some computation, or to the constructor form $[\sigma^{\prec}]_{\IR}\,t^{\prec}$ to stay verbatim.
+Since in the QIIT definition there is only one term substitution $\blank [\blank]_{\I}$ and lifting $(\blank)^+_{\I}$, we may choose to translate $t [\sigma]_{\I}$ to the recursive form $t^{\prec} [\sigma^{\prec}]_{\Tm}$ to carry out some computation, or to the constructor form $t^{\prec} [\sigma^{\prec}]_{\IR}$ to stay verbatim.
 Either choice is possible, but we need to use \cref{prop:correctness-1,prop:correctness-3} to switch from one to the other whenever suitable.
-For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $[ \sigma ]\,t = [\sigma]_{\Tm}\,t$ if the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ is chosen for the translation.
+For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ if the recursive form $\subTm{t^{\prec}}{\sigma^{\prec}}$ is chosen for the translation.
 
@@
 %The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns, we declare each function clause a rewrite and another rewrite to identify the clause with the reflexivity proof.
-E.g.,\ $[\blank]_{\Ty}$ up to the first clause $[\idS]_{\Ty}\,A \reduce A$ is defined as
+To emulate a function by overlapping patterns, we declare each function clause a rewrite and another rewrite to identify the clause with the reflexivity proof
+For example, the first few clauses for type substitution (written $[\sigma]_{\Ty}\,A$ rather than $\subTy{A}{\sigma}$ in the formalisation) are defined as
 \begin{lstlisting}
@@
 For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\apdfunc f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
-However, as $[\pi_1(\sigma; \tau)]_{\Ty}\,A$ and $[\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))]_{\Ty}\,A$ are strictly equal, the resulting identity is indeed homogeneous.
+However, as $\subTy{A}{\pi_1(\sigma; \tau)}$ and $\subTy{A}{\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))}$ are strictly equal, the resulting identity is indeed homogeneous.
 Hence, we use the heterogeneous congruence $\mathit{hap}_f\colon (p : x \simeq y) \to f\,x\simeq f\,y$ (by \lstinline|hcong|) and switch back to the ordinary identity type $f\,x = f\, y$ (by \lstinline|≅-to-≡|) as their types match.
@@
 \begin{alignat*}{5}
-  [\blank]_{\Ty}\blank & : \implicit{\Gamma, \Delta}\; && \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
+\subTy{\blank}{\blank} & : \implicit{\Gamma, \Delta}\; && \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \\
   \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
@@
   \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;([ \sigma ]_{\Ty}\,A) \to \Sub\;\Gamma,(\Delta, A) \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \to \Sub\;\Gamma,(\Delta, A) \\
   \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
@@
   \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
-  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\;\sigma ]_{\Ty}\, A) \\
-  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\, A)
-\end{alignat*}
-\begin{alignat*}{5}
-  & [ \idS ]_{\Ty}\,               && A           && \reduce A \\
-  & [ \sigma ; \tau ]_{\Ty}\,      && A           && \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
-  & [ \pi_1(\sigma, t) ]_{\Ty}\,   && A           && \reduce [\sigma]_\Ty\,A \\
-  & [ \pi_1(\sigma; \tau) ]_{\Ty}\,&& A           && \reduce [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
-  & [ \sigma ]_{\Ty}\,             && \UU         && \reduce \UU
+  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\subTy{A}{\pi_1\;\sigma}) \\
+  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A} && \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma})
+\end{alignat*}
+\begin{alignat*}{3}
+  & \subTy{A}{\idS}           && \reduce A \\
+  & \subTy{A}{\sigma ; \tau}  && \reduce \subTy{(\subTy{A}{\tau})}{\sigma} \\
+  & \subTy{A}{\pi_1(\sigma, t)} && \reduce \subTy{A}{\sigma} \\
+  & \subTy{A}{\pi_1(\sigma; \tau)} && \reduce \subTy{(\subTy{A}{\pi_1\tau})}{\sigma} \\
+  & \subTy{\UU}{\sigma} && \reduce \UU
 \end{alignat*}
@@
   ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , \sub{t}{\sigma} \\
   \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma \\
@@
   \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  [\idS]          & : \implicit{\Gamma, A, t}\; && {[\idS]\;t}          && =^{\Tm\;\Gamma\;A} && t \\
-  [;]             & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {[\sigma ; \tau]\;t} && =^{\Tm\;\Gamma\;[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
+  [\idS]         & : \implicit{\Gamma, A, t}\; && {\sub{t}{\idS}}          && =^{\Tm\;\Gamma\;A} && t \\
+  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {\sub{t}{\sigma ; \tau}} && =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  [\blank]_{\Tm}\blank            & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\,A) \\ 
- [ \idS ]_{\Tm}\,t                & \reduce t \\
- [ \sigma ; \tau ]_{\Tm}\,t       & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
- [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \reduce [\sigma]_\Tm\,t \\
- [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
- [ \sigma ]_{\Tm}\,t              & \reduce [ \sigma ]\,t, \quad \text{otherwise}
-\end{alignat*}
-
+  \subTm{\blank}{\blank}            & : \implicit{\Gamma, \Delta, A}\; \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to  \Tm\;\Gamma\;(\subTy{A}{\sigma}) \\
+ \subTm{t}{\idS}                & \reduce t \\
+ \subTm{t}{\sigma ; \tau}       & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
+ \subTm{t}{\pi_1(\sigma, s)}    & \reduce \subTm{t}{\sigma} \\
+ \subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
+ \subTm{t}{\sigma}              & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
+\end{alignat*}
+%
 \begin{alignat*}{5}
@@
 \end{alignat*}
-
-\begin{alignat*}{5}
-  [ \sigma ]_{\Ty}\,  && (\El\, u) \reduce \El\,([\sigma]_{\Tm}{u})
+%
+\begin{alignat*}{3}
+  \subTy{(\El\, u)}{\sigma} \reduce \El\,(\subTm{u}{\sigma})
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
+  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
   \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
@@
 \begin{alignat*}{5}
-  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
+  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
 \idS                \uparrow A  & \reduce \idS \\
-\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
+\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \sub{A}{\tau}) ; (\tau \uparrow A) \\
 \pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
+\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\sub{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
 \sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwiese}
@@
 \end{alignat*}
-\begin{alignat*}{5}
-  [ \sigma ]_{\Ty}\, (\Pi\,A\,B) \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
-\end{alignat*}
-\begin{alignat*}{5}
-  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& [ \sigma ] \,(\lambda\,t) && {} =^{\Tm\;\Gamma\;([\sigma]_{\Ty}(\Pi\,A\,B)} {} && \lambda ([\sigma \uparrow A ]\,t)\\
+\begin{alignat*}{3}
+  & \subTy{(\Pi\,A\,B)}{\sigma}       && {} \reduce {} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A})
+\end{alignat*}
+\begin{alignat*}{5}
+  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\;\Gamma\;(\subTy{(\Pi\,A\,B)}{\sigma}} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
   \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\;\Gamma\;A} && t \\
@@
 \begin{alignat*}{5}
-  \mathsf{Id} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
-  \nameless & : && [ \sigma ]_{\Ty}\;(\mathsf{Id}\;a\;t\;u) \reduce \mathsf{Id}\;([\sigma]_{\Tm}\,a)\;([\sigma]_{\Tm}\,t)\;([\sigma]_{\Tm}\,u) \\
-  \mathsf{refl}   & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\mathsf{Id}\;a\;t\;t) \\
-  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& [\sigma]\,(\mathsf{refl}\;t) =^{\Tm\;\Gamma\;([\sigma](\mathsf{Id}\,a\,t\,t))} \mathsf{refl}([ \sigma ]_{\Tm}\,t) \\
+  \idtype[\blank]{\blank}{\blank} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
+  & \subTy{(\idtype[a]{t}{u})}{\sigma} && \reduce \idtype[{\subTm{a}{\sigma}}]{\subTm{t}{\sigma}}{\subTm{u}{\sigma}} \\
+  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\idtype[a]{t}{t}) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \subTm{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;([\sigma](\mathsf{Id}\,a\,t\,t))} \mathsf{refl}(\subTm{t}{\sigma}) \\
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
@@
   \[
-    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
+    \subTy{(\El\,u)}{\idS}, \qquad \subTy{(\El\,u)}{\sigma;\tau}, \qquad \subTy{(\El\,u)}{\pi_1(\sigma, t)}, \qquad\text{and}\qquad \subTy{(\El\,u)}{\pi_1(\sigma; \tau)}
   \]
@@
   \[
-    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
-    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
-  \]
-  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
+    \El\,u, \qquad \El(\subTm{(\subTm{u}{\tau})}{\sigma}, \qquad \El(\subTm{u}{\sigma}), \qquad\text{and}\qquad
+    \El(\subTm{(\subTm{u}{\pi_1\tau})}{\sigma})
+  \]
+  respectively, no matter which function clause of $\subTy{\blank}{\blank}$ is used.
   Therefore, type substitution is locally confluent.
@@
   \begin{enumerate}
-    \item For the identity substitution $\idS$, we have $[\idS]t = t \judgeq [\idS]_{\Tm} t$ by definition.
+    \item For the identity substitution $\idS$, we have $\sub{t}{\idS} = t \judgeq \subTm{t}{\idS}$ by definition.
     \item For a composite substitution $\sigma; \tau$, we have
       \begin{align*}
-        [\sigma;\tau]t & = [\sigma]\,([\tau] t)            && \text{by $[;]t$} \\
+        \sub{t}{\sigma;\tau} & = \sub{(\sub{t}{\tau})}{\sigma}            && \text{by $[;]t$} \\
 %                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
-                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
-                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
-      \end{align*}
-    \item For $\pi_1(\sigma, t)$, we have
-      \begin{align*}
-        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
-                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
-                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
+                       & = \subTm{(\subTm{t}{\tau})}{\sigma} && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
+                       & \judgeq \subTm{t}{\sigma;\tau} && \text{by definition.}
+      \end{align*}
+    \item For $\pi_1(\sigma, s)$, we have
+      \begin{align*}
+        \sub{t}{\pi_1(\sigma, s)} & = \sub{t}{\sigma}               && \text{by $\pi_1\beta$} \\
+                            & = \subTm{t}{\sigma}       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
+                            & \judgeq \subTm{t}{\pi_1(\sigma, s)} && \text{by definition.}
       \end{align*}
     %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
-    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ strictly.
+    \item For the remaining cases, $\subTm{t}{\sigma}$ is equal to $\sub{t}{\sigma}$ strictly.
   \end{enumerate}
-  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
+  Hence, we conclude that $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
 \end{proof}
@@
   \begin{align*}
-    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
-                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
+    \subTy{A}{\sigma; (\tau, t)}    & = \subTy{A}{(\sigma;\tau), \sub{t}{\sigma}},
+                                     & \subTy{A}{\sigma}               & = \subTy{A}{\emptyctx},
+                                     & \subTy{A}{\sigma}               & = \subTy{A}{\pi_1\sigma, \pi_2\sigma}
   \end{align*}
@@
   \begin{align*}
-    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by \eqref{eq:def-type-subst-7}} \\
-                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
-                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
-                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
-                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by \eqref{eq:def-type-subst-7} }
+    \subTy{(\El\,u)}{\sigma; (\tau, t)} & \judgeq \El(\subTm{u}{\sigma; (\tau, t)}) && \text{by \eqref{eq:def-type-subst-7}} \\
+                                      & = \El( \sub{u}{\sigma; (\tau, t)})            && \text{by \cref{prop:correctness-1}} \\
+                                      & = \El( \sub{u}{\sigma; \tau, \sub{t}{\sigma}})     && \text{by $\mathsf{concat}$} \\
+                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by \cref{prop:correctness-1}} \\
+                                      & \judgeq \subTy{\left(\El\,u\right)}{\sigma; \tau, \sub{t}{\sigma}} && \text{by \eqref{eq:def-type-subst-7} }
   \end{align*}
@@
 
-  As for term substitution $[\blank]_{\Tm}\blank$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
+  As for term substitution $\subTm{\blank}{\blank}$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
 \end{proof}
@@
   \begin{align*}
-    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
-  \end{align*}
-  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
-  \begin{align*}
-    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
-    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
+    \subTy{(\Pi\;A\;B)}{\idS} && \subTy{(\Pi\;A\;B)}{\sigma;\tau} && \subTy{(\Pi\;A\;B)}{\pi_1(\sigma, t)} && \subTy{(\Pi\;A\;B)}{\pi_1(\sigma; \tau)}.
+  \end{align*}
+  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $\subTy{\blank}{\blank}$:
+  \begin{align*}
+    & \Pi\;A\;B                && \Pi\;(\subTy{(\subTy{A}{\tau})}{\sigma})\;(\subTy{\subTy{B}{\tau \uparrow A}}{\sigma \uparrow \sub{A}{\tau}}) \\
+    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;(\subTy{(\subTy{A}{\pi_1\tau})}{\sigma})\;(\subTy{\subTy{B}{\sigma \uparrow (\subTy{A}{\pi_1\tau})}}{\pi_1 \tau \uparrow A})
   \end{align*}
   respectively.
-  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
+  Hence, the type substitution $\subTy{\blank}{\blank}$ is locally confluent.
 \end{proof}
@@
   \begin{align*}
-    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
+    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow \sub{A}{\tau});(\tau \uparrow A) && \text{by definition} \\
                              & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
@@
   \begin{align*}
-    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
+    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows \subTel{\Xi}{\idS}) ; (\idS \upuparrows \Xi) \\
                                    & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
@@
   \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
-  For the following equality constructors $p : \sigma = \tau$
-  \begin{alignat*}{5}
-    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
+  For the following equality constructors $\cc : \sigma = \tau$
+  \begin{alignat*}{5}
+    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), \sub{t}{\sigma} \\
     \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
@@
   \begin{align*}
-  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
-                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
-  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
+  \subTel{\Xi}{\sigma} & = \subTel{\Xi}{\tau}
+                         & \subTy{A}{\sigma \upuparrows \Xi} & =^{\lam{\gamma}\Ty\;(\Gamma \dplus \subTel{\Xi}{\gamma})}_{\alert{\cc}} \subTy{A}{\tau \upuparrows \Xi} \\
+  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus \subTel{\Xi}{\gamma})\;(\Delta\dplus\Xi)}_{\alert{\cc}} \tau \upuparrows \Xi
 %    &
@@
 
-  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$ and the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
+  First, we prove the coherence for $\subTy{A}{\blank}$ by induction on $A$ and the coherence for the other two identities $\subTel{\Xi}{\blank}$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
   \begin{enumerate}
-    \item Consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$ by induction on $A$.
+    \item Consider the following cases for $\subTy{A}{\sigma\upuparrows\Xi} = \subTy{A}{\tau \upuparrows \xi}$ by induction on $A$.
       \begin{description}
@@
           \[
-            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
-          \]
-          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
-          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
+            \subTy{\UU_{\Delta\dplus \Xi}}{\pi_1(\sigma, t) \upuparrows \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad \subTy{\UU_{\Delta\dplus \Xi}}{\sigma\upuparrows \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
+          \]
+          By assumption, $\subTel{\Xi}{\pi_1(\sigma, t)} = \subTel{\Xi}{\sigma}$ and by congruence $\Gamma \dplus \subTel{\Xi}{\pi_1(\sigma, t)} = \Gamma \dplus \subTel{\Xi}{\sigma}$ holds.
+          Then, $\UU_{\Gamma \dplus \subTel{\Xi}{\pi_1(\sigma, t)}} \simeq \UU_{\Gamma \dplus \subTel{\Xi}{\sigma}}$ holds by congruence for $\UU$.
           \LT{Justify the use of congruence rule for QIIT, e.g. $\transfib{\Ty}{p}{\UU_{\Gamma}} = \UU_{\Delta}$ for $p : \Gamma = \Delta$}
@@
         \item[$A$ is $\El\,u$:]
-          Similar to \cref{prop:coherence-2}, we use \cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.
+          Similar to \cref{prop:coherence-2}, we use \cref{prop:correctness-3} and the congruence law for $\sub{u}{\blank}$ to derive the desired identity.
 
@@
           \[
-            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
+            \subTy{C}{\pi_1(\sigma, t) \upuparrows \Xi} = \subTy{C}{\sigma \upuparrows \Xi}
           \]
@@
           \begin{align*}
-            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
-            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
-            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
-            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by induction hypothesises} \\
-            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
-            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
+            & \subTy{(\Pi\,B\,C)}{\pi_1(\sigma, t) \upuparrows \Xi} \\
+            \judgeq{} & \Pi\,(\subTy{B}{\pi_1(\sigma, t) \upuparrows \Xi})\,(\subTy{C}{\pi_1(\sigma, t) \upuparrows \Xi \uparrow B}) && \text{by the definition of $\subTy{(\Pi\,B\,C)}{\blank}$} \\
+            \judgeq{} & \Pi\,(\subTy{B}{\pi_1(\sigma, t) \upuparrows \Xi})\,(\sub{C}{\pi_1(\sigma, t) \upuparrows (\Xi, B)}) && \text{by the definition of $\upuparrows$} \\
+            ={}       & \Pi\,(\subTy{B}{\sigma \upuparrows \Xi})\,(\sub{C}{\sigma \upuparrows (\Xi, B)}) && \text{by induction hypothesises} \\
+            \judgeq{} & \Pi\,(\subTy{B}{\sigma \upuparrows \Xi})\,(\sub{C}{(\sigma \upuparrows \Xi) \uparrow B}) && \text{by the definition of $\upuparrows$} \\
+            \judgeq{} & \subTy{(\Pi\,B\,C)}{\sigma\upuparrows \Xi} && \text{by the definition of $\subTy{(\Pi\,B\,C)}{\blank}$.}
           \end{align*}
       \end{description}
-    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
+    \item By induction on $\Xi$, consider following cases for $\subTel{\Xi}{\pi_1(\sigma, t)} = \subTel{\Xi}{\sigma}$.
       \begin{description}
-        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
+        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $\subTel{\emptytel}{\pi_1(\sigma, t)} \judgeq \emptytel \judgeq \subTel{\emptytel}{\sigma}$.
         \item[$\Xi$ is the telescope extension $\Xi, A$:]
           \begin{align*}
-            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
-                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
-                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
+            \subTel{(\Xi, A)}{\pi_1(\sigma, t)} & \judgeq \subTel{\Xi}{\pi_1(\sigma, t)}, \subTy{A}{\pi_1(\sigma, t) \upuparrows \Xi} && \text{by definition} \\
+                                                & = \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi} \\
+                                                & \judgeq \subTel{(\Xi, A)}{\sigma} && \text{by definition}
           \end{align*}
@@
                                                   & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
-                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
+                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus\subTel{\Xi}{\gamma}, \subTy{A}{\gamma \upuparrows \Xi})\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
                                                   & = (\sigma \upuparrows \Xi) \uparrow A \\

### 2025-02-18 tex/main.tex
fix remaining `A[sigma]`

@@
   \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\idtype[a]{t}{t}) \\
-  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \subTm{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;([\sigma](\mathsf{Id}\,a\,t\,t))} \mathsf{refl}(\subTm{t}{\sigma}) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \subTm{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\sub{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\subTm{t}{\sigma}) \\
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
@@
   \blank,_{\Ctx}^M\blank & : (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
-  [\blank]_{\Ty}^M\blank & : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
+  \blank[\blank]_{\Ty}^M & : \Ty^M\;\Delta^M\;A \to \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Gamma^M\;(\subTy{A}{\sigma}) \\
   \UU^M                  & : \Ty^M\;\Gamma^M\;\UU \\
   \emptysub_{\Sub}^M     & : \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
-  \blank,_{\Sub}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
+  \blank,_{\Sub}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [\sigma^M]^M_{\Ty})\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
   \idS^M                 & : \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
@@
   \pi_1^M                & : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma \to \Sub^M\;\Gamma^M\;\Delta^M\;(\pi_1\sigma) \\
-  \pi_2^M                & : (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;([ \pi_1^M\,\sigma^M ]^M_{\Ty}\;A^M)\;(\pi_2\,\sigma) \\
-  [\blank]^M\blank       & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t)
-\end{alignat*}
-
-\begin{alignat*}{5}
-  [\idS]_{\Ty}^M      & : [ \idS^M ]^M_{\Ty}                             && \;A^M =^{\Ty^M\,\Gamma^M}\; && A^M \\
-  [;]_{\Ty}^M         & : [ \sigma^M ;^M \tau^M ]^M_{\Ty}                && \;A^M =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
-  [\pi_1,]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ,^M_{\Sub} t^M) ]^M_{\Ty} && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\,A^M \\
-  [\pi_1;]^M_{\Ty}    & : [ \pi_1^M\,(\sigma^M ;^M \tau^M) ]^M_{\Ty}     && \;A^M =^{\Ty^M\,\Gamma^M} \; && [\sigma^M]^M_{\Ty}\left([ \pi_1^M\,\tau^M]^M_{\Ty}\,A^M\right) \\
-  []\UU^M             & : [ \sigma^M ]^M_{\Ty}                           && \;\UU^M =^{\Ty^M\,\Gamma^M\,\UU}      {} && \UU^M
+  \pi_2^M                & : (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [ \pi_1^M\,\sigma^M ]^M_{\Ty})\;(\pi_2\,\sigma) \\
+  \blank[\blank]^M       & : \Tm^M\;\Delta^M\;A^M\;t \to (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [\sigma^M]_{\Ty}^M)\;(\sub{t}{\sigma} )
+\end{alignat*}
+
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M      & : A^M [ \idS^M ]^M_{\Ty}                             &&  =^{\Ty^M\,\Gamma^M}\; && A^M \\
+  [;]_{\Ty}^M         & : A^M [ \sigma^M ;^M \tau^M ]^M_{\Ty}                && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,(A^M [\tau^M]^M_{\Ty}) \\
+  [\pi_1,]^M_{\Ty}    & : A^M [ \pi_1^M\,(\sigma^M ,^M_{\Sub} t^M) ]^M_{\Ty} && =^{\Ty^M\,\Gamma^M} \; && A^M [\sigma^M]^M_{\Ty} \\
+  [\pi_1;]^M_{\Ty}    & : A^M [ \pi_1^M\,(\sigma^M ;^M \tau^M) ]^M_{\Ty}     && =^{\Ty^M\,\Gamma^M} \; && \left(A^M [ \pi_1^M\,\tau^M]^M_{\Ty}\right) [\sigma^M]^M_{\Ty} \\
+  []\UU^M             & : \UU^M [ \sigma^M ]^M_{\Ty}                           &&  =^{\Ty^M\,\Gamma^M\,\UU}      {} && \UU^M
 \end{alignat*}
@@
   ;\text{-}\textsf{assoc}^M & : (\sigma^M ;^M \tau^M) ;^M \gamma^M   && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{assoc}}} && \sigma^M ;^M (\tau^M ;^M \gamma^M) \\
-  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black}[\sigma^M]_{\Ty}^M\,t^M}} \\
+  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black}t^M [\sigma^M]_{\Ty}^M}} \\
   \pi_1\beta^M              & : \pi_1^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\pi_1\beta}}      && \sigma^M \\
@@
   \pi\eta^M                 & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,(\Delta^M ,^M_{\Ctx} A^M)}_{\alert{\pi\eta}} {} && \pi_1^M \sigma^M ,_{\Sub}^M \pi_2^M \sigma^M \\
-  [\idS]^M                  & : [ \idS ]^M\,t^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
-  [;]^M                     & : [ \sigma^M ;^M \tau^M ]^M\,t^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && [\sigma^M]^M\,([\tau^M]^M\,t^M) 
+  [\idS]^M                  & : t^M [ \idS ]^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
+  [;]^M                     & : t^M [ \sigma^M ;^M \tau^M ]^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && (t^M [\tau^M]^M) [\sigma^M]^M
 \end{alignat*}
@@
 \begin{alignat*}{10}
-  [\blank]_{\Tm}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
+  \blank[\blank]_{\Tm}^M & : \Tm^M\;\Delta^M\;A^M\;t \to (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [\sigma^M]_{\Ty}^M)\,(\subTm{t}{\sigma}) \\
   \El^M                  & : \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u)
@@
 \begin{alignat*}{5}
-  [\idS]_{\Tm}^M      & : [\idS^M]^M_{\Tm}                               && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M \\
-  [;]^M_{\Tm}         & : [ \sigma^M ;^M \tau^M ]^M_{\Tm}                && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[;]^M_{\Ty}}} && [ \sigma^M ]^M\; \left([ \tau^M ]^M_{\Tm}\,t^M\right) \\
-  [\pi_1,]^M_{\Tm}    & : [\pi_1^M (\sigma^M ,_{\Sub}^M u^M) ]^M_{\Tm}   && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;t^M \\
-  [\pi_1;]^M_{\Tm}    & : [\pi_1^M (\sigma^M ;^M \tau^M) ]^M_{\Tm}       && \;t^M =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1;]^M_{\Ty}}} && [\sigma^M]_{\Tm}^M\;\left([\pi_1^M\,\tau^M]^M_{\Tm}\;t^M\right) \\
-  [\emptysub]^M_{\Tm} & : [\emptysub^M_{\Sub}]^M_{\Tm}                   && \;t^M = %^{\Tm^M\,\Gamma^M\;([\emptysub_{\Sub}^M]^M_{\Ty}\,A^M)\,([\emptysub]_{\Tm}\,t)}
-                      [ \emptysub^M_{\Sub} ]^M  && t^M \\
-  [,]_{\Tm}^M         & : [ \sigma^M ,_{\Sub}^M u^M ]^M_{\Tm}            && \;t^M = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)}
-                      [ \sigma^M ,_{\Sub}^M u^M ]^M && t^M \\
-  [\pi_1\idS]^M_{\Tm} & : [ \pi_1^M\,\idS^M ]^M_{\Tm}                    && \;t^M = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{}
-                      [ \pi_1^M\,\idS^M ]^M && t^M \\
-  [\pi_1\pi_1]^M_{\Tm}& : [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M_{\Tm}       && \;t^M = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)}
-                      [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M && t^M
-\end{alignat*}
-
-\begin{alignat*}{5}
-  []\El^M & : [ \sigma^M ]^M_{\Ty}\;(\El^M\;u^M)   && =^{\Ty^M\,\Gamma^M\,(\El\,u)} {} && \El^M\,(\alert{\transfib{}{[]\UU^M}{\color{black}[ \sigma^M ]^M_{\Tm}\,u^M}}
+  [\idS]_{\Tm}^M      & : t^M [\idS^M]^M_{\Tm}                               && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M \\
+  [;]^M_{\Tm}         & : t^M [ \sigma^M ;^M \tau^M ]^M_{\Tm}                && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[;]^M_{\Ty}}} && \left(t^M [ \tau^M ]^M_{\Tm}\right) [ \sigma^M ]^M \\
+  [\pi_1,]^M_{\Tm}    & : t^M [\pi_1^M (\sigma^M ,_{\Sub}^M u^M) ]^M_{\Tm}   && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && t^M [\sigma^M]_{\Tm}^M \\
+  [\pi_1;]^M_{\Tm}    & : t^M [\pi_1^M (\sigma^M ;^M \tau^M) ]^M_{\Tm}       && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1;]^M_{\Ty}}} && \left(t^M [\pi_1^M\,\tau^M]^M_{\Tm}\right) [\sigma^M]_{\Tm}^M \\
+  [\emptysub]^M_{\Tm} & : t^M [\emptysub^M_{\Sub}]^M_{\Tm}                   && = %^{\Tm^M\,\Gamma^M\;([\emptysub_{\Sub}^M]^M_{\Ty}\,A^M)\,([\emptysub]_{\Tm}\,t)}
+                      t^M [ \emptysub^M_{\Sub} ]^M  &&  \\
+  [,]_{\Tm}^M         & : t^M [ \sigma^M ,_{\Sub}^M u^M ]^M_{\Tm}            && = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)}
+                      t^M [ \sigma^M ,_{\Sub}^M u^M ]^M && \\
+  [\pi_1\idS]^M_{\Tm} & : t^M [ \pi_1^M\,\idS^M ]^M_{\Tm}                    && = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{}
+                      t^M [ \pi_1^M\,\idS^M ]^M && \\
+  [\pi_1\pi_1]^M_{\Tm}& : t^M [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M_{\Tm}       && = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)}
+                      t^M [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M &&
+\end{alignat*}
+
+\begin{alignat*}{5}
+  []\El^M & : [ \sigma^M ]^M_{\Ty}\;(\El^M\;u^M)   && =^{\Ty^M\,\Gamma^M\,(\El\,u)} {} && \El^M\,(\alert{\transfib{}{[]\UU^M}{\color{black}u^M [ \sigma^M ]^M_{\Tm}}}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M ([ \sigma^M ]^M\;A^M]))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
+  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M (A^M [ \sigma^M ]^M))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
   (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;\pi_2\idS}{([;]^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
@@
 \begin{alignat*}{10}
-  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
+  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, A^M [\sigma^M]_{\Ty}^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
   \Pi^M                  & : (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
@@
   \mathord{;\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} \\
-                                &&& (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
+                                &&& (\sigma^M \uparrow^M A^M [ \tau^M ]^M_{\Ty} ) ;^M (\tau^M \uparrow^M A^M)  \\
   \mathord{\pi_1,\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
   \mathord{\pi_1;\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} \\
-                                &&& (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
+                                &&& (\sigma^M \uparrow^M  A^M [\pi_1^M\tau^M]^M_{\Ty});^M (\pi_1^M\tau^M \uparrow^M A^M) \\
   \mathord{\emptysub^M_{\Sub}\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
@@
 \begin{alignat*}{5}
-  []\Pi^M         & : [ \sigma^M ]^M_{\Ty} \;(\Pi^M\;A^M\;B^M)  && =^{\Ty^M\,\Gamma^M\,([\sigma]_{\Ty} (\Pi\,A\,B))} && \Pi^M\;([ \sigma^M ]^M_{\Ty}\;A^M)\;([ \sigma^M ]^M_{\Ty}\;B^M) \\
-  []\lambda^M     & : [ \sigma^M ]^M\;(\lambda^M\;t^M)          && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} && \lambda^M\;([ \sigma^M \uparrow^M A^M ]^M\;t^M) \\
+  []\Pi^M         & : (\Pi^M\;A^M\;B^M) [ \sigma^M ]^M_{\Ty}  && =^{\Ty^M\,\Gamma^M\,(\subTy{(\Pi\,A\,B)}{\sigma})} && \Pi^M\;(A^M [ \sigma^M ]^M_{\Ty})\;(B^M [ \sigma^M ]^M_{\Ty}) \\
+  []\lambda^M     & : (\lambda^M\;t^M) [ \sigma^M ]^M          && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} && \lambda^M\;(t^M [ \sigma^M \uparrow^M A^M ]^M) \\
   \Pi\beta^M      & : \mathsf{app}^M\;(\lambda^M\;t^M)          && =^{\Tm^M\,(\Gamma^M,_{\Ctx}^M A^M)\,B^M}_{\alert{\Pi\beta}}      && t^M  \\
@@
 \begin{alignat*}{5}
-  []\mathsf{Id}^M    :{} & [ \sigma^M ]^M_{\Ty} (\mathsf{Id}^M\;a^M\;t^M\;u^M) =^{\Ty^M\,\Gamma^M\,([\sigma]_{\Ty}(\mathsf{Id}\,a\,t\,u))} \\
-                      & \mathsf{Id}^M\; (\transfib{}{[]\UU^M}{[ \sigma^M ]^M_{\Tm}\,a^M})\;
-                  (\transfib{}{[]\El^M}{[ \sigma^M ]^M_{\Tm}\,t^M})\;
-                  (\transfib{}{[]\El^M}{[ \sigma^M ]^M_{\Tm}\,u^M}) \\
-  []\mathsf{refl}    :{} & [ \sigma^M ]^M\;(\mathsf{refl}^M\;t^M) =^{\Tm^M\;\Gamma^M}_{\alert{[]\mathsf{Id}^M, []\mathsf{refl}}} \\
-                      & \mathsf{refl}^M\;(\transfib{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,\nameless}{[]\El^M}{[\sigma^M]^M_{\Tm} \,t^M}) \\
+  []\mathsf{Id}^M    :{} & (\mathsf{Id}^M\;a^M\;t^M\;u^M) [ \sigma^M ]^M_{\Ty} =^{\Ty^M\,\Gamma^M\,(\subTy{(\mathsf{Id}\,a\,t\,u)}{\sigma})} \\
+                      & \mathsf{Id}^M\; (\transfib{}{[]\UU^M}{a^M [ \sigma^M ]^M_{\Tm}})\;
+                  (\transfib{}{[]\El^M}{t^M [ \sigma^M ]^M_{\Tm}})\;
+                  (\transfib{}{[]\El^M}{u^M [ \sigma^M ]^M_{\Tm}}) \\
+  []\mathsf{refl}    :{} & (\mathsf{refl}^M\;t^M) [ \sigma^M ]^M =^{\Tm^M\;\Gamma^M}_{\alert{[]\mathsf{Id}^M, []\mathsf{refl}}} \\
+                      & \mathsf{refl}^M\;(\transfib{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,\nameless}{[]\El^M}{t^M [\sigma^M]^M_{\Tm}}) \\
   \mathsf{reflect}^M :{} & \Tm^M\;\Gamma^M\;(\mathsf{Id}^M\;a^M\;t^M\;u^M)\;p \to t^M =^{\Tm^M\;\Gamma^M\;A^M}_{\alert{\mathsf{reflect}\;p}} u^M

### 2025-02-19 tex/main.tex
Shorten S2.1 and address the point in S2.3

@@
 %
-For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
+For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
 %
@@
 
-In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s for short).
+In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s, for short).
 %
@@
 \subsection{Inductive-recursive types}
-\LT[noinline]{Shorten this section}
-\LT[noinline]{the recursion part of IR is not given by the elimination rule}
-
-An ordinary inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, using an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
-For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
-%
-The type is first defined inductively by listing its constructors, and then afterwards recursive functions can be defined out of it.
-
-In contrast, an inductive-recursive type is defined inductively \emph{simultaneously} with a recursively defined function out of it.
-The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents.
-Crucially, the code for dependent function types
+An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined, using an elimination rule which says that $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
+%For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
+%
+The type is first defined inductively by listing its constructors, and then recursive functions can be defined out of it.
+
+In contrast, an inductive-recursive type(IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F$ out of $A$ to some fixed (possibly large) type $D$.
+The prototypical example is Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents.
+Crucially, the code
 \[
   \pi : (a : U) \to (T(a) \to U) \to U
-\]
-with $T(\pi(a, b)) = (x : T\,a) \to T(b\, x)$ only makes sense because we already have access to the decoding function $T$.
-%
-Inductive-recursive types (IRT), introduced by Dybjer~\cite{Dybjer2000}, were extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
-The general schema allows one to define a type $A$ inductively, together with a recursive function $F : A \to D$, for some fixed (possibly large) type $D$.
-
-An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, again reflecting that $A$ is inductively defined.
-%
-However, note that the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
-%
-In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must always target a fixed pre-existing type $D$.
-%
-In this work, we will consider an extension \emph{generalised} IRTs, where also $F : A \to A$ is allowed.
-%
-Such IRTs were previously used by Danielsson~\cite{Danielsson2006} to represent type theory in type theory, and are discussed by Kaposi~\cite{Kaposi2023}.
-%
-Their general metatheory would be interesting future work.
+  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T(b\, x)
+\]
+for dependent function types makes sense only because we have access to the function $T$.
+%
+
+An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, but the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
+%
+In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must target a pre-existing type $D$.
+%
+In this work, we will consider an extension \emph{generalised} IRTs, where $F : A \to A$ is also allowed.
+%
+Such IRTs are accepted in \Agda and were previously used~\cite{Danielsson2006} to represent type theory in type theory.
+%
+%Their general metatheory would be interesting future work.
 
@@
 %
-Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
-%
-For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
-with the following constructors:
-%\begin{figure}[H]
+%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
+%
+For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT in \cref{fig:QIT-join-list}.
+\begin{figure}
 \begin{minipage}{.35\textwidth}
@@
 \end{minipage}
-%\end{figure}
-\vskip\belowdisplayskip\noindent
-A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
+\caption{The quotient inductive type of join lists}
+\label{fig:QIT-join-list}
+\end{figure}
+%\vskip\belowdisplayskip\noindent
+A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
+% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
 %; indeed, inductive-inductive definitions only make sense when $D = \Set$.
@@
 For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
+%In this work, we will consider quotient inductive-inductive-recursive types, generalising QIITs and QITs.
 
@@
 \subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-Functions are typically defined informally using pattern matching for convenience and readability.
-Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
-This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
+In a proof assistant, functions are typically defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
+This elaboration offers the conventional notation, while making computational behaviour opaque.
+% and favouring asymmetric definitions.
 For example, two definitions of addition below
-%\begin{figure}[H]%
+%\begin{figure}[b]%
 \begin{minipage}{.48\textwidth}
@@
 are the same after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
-Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
+Each clause of $+'$ is strict while clauses of $+$ are not necessarily strict.
 %This semantics for function definitions decides how definitions reduce and therefore how identities are proved.
@@
 
-\LT[noinline]{remark that recursions in IR are not defined by elimination rules}
+In this work, we will consider inductive types defined simultaneously with recursive functions using overlapping patterns.
+Note that in the schema of IRTs these simultaneously defined functions are already not given by using elimination rules.
 
@@
 %\end{alignat*}
-Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
-  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
+Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer, since the codomain of the recursive function is the inductively defined typed itself.
 %Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
-%This definition fails to remove any transport in the definition of substitution calculus.
-%
 Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define$[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
@@
 We build heavily on their ideas, but by using Danielsson's~\cite{Danielsson2006} idea of defining type substitution recursively, we are able to remove many explicit transports from the definition, with the result that it is closer to an informal exposition on paper.
-Such a benefit was also suggested by Kaposi's~\cite{Kaposi2023} work, where simply typed lambda calculus was used as an example.
+Such a benefit was also suggested by Kaposi's~\cite{Kaposi2023} work, where the simply typed $\lambda$-calculus was used as an example.
 Altenkirch and Kaposi has built on the QIIT representation with great success to derive non-trivial results such as normalisation by evaluation~\cite{Altenkirch2017}, and investigate the metatheory of QIITs themselves, together with Kov\'acs~\cite{Kaposi2019}.
@@
 \paragraph*{Quotient inductive-inductive-recursive types}
-We give the QIIR definition for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
+We give the QIIRTs for type theories with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
 

### 2025-02-19 tex/main.tex
comment out het eq

@@
 
-\subparagraph*{Heterogeneous identity}
-In our implementation, some statements are proved differently from our presentation. 
-We have sometimes used John Major equality~\cite{McBride1999} $x \cong y$  for $A, B : \Set$, and $x : A$ and $y : B$, where dependent congruence rule or dependent identity type are needed.
-
-For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\apdfunc f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
-However, as $\subTy{A}{\pi_1(\sigma; \tau)}$ and $\subTy{A}{\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))}$ are strictly equal, the resulting identity is indeed homogeneous.
-Hence, we use the heterogeneous congruence $\mathit{hap}_f\colon (p : x \simeq y) \to f\,x\simeq f\,y$ (by \lstinline|hcong|) and switch back to the ordinary identity type $f\,x = f\, y$ (by \lstinline|≅-to-≡|) as their types match.
-
-%As shown by McBride, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
-
-\LT[noinline]{Mention the NbE for SC here?}
+%\subparagraph*{Heterogeneous identity}
+%In our implementation, some statements are proved differently from our presentation. 
+%We have sometimes used John Major equality~\cite{McBride1999} $x \cong y$  for $A, B : \Set$, and $x : A$ and $y : B$, where dependent congruence rule or dependent identity type are needed.
+%
+%For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\apdfunc f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
+%However, as $\subTy{A}{\pi_1(\sigma; \tau)}$ and $\subTy{A}{\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))}$ are strictly equal, the resulting identity is indeed homogeneous.
+%Hence, we use the heterogeneous congruence $\mathit{hap}_f\colon (p : x \simeq y) \to f\,x\simeq f\,y$ (by \lstinline|hcong|) and switch back to the ordinary identity type $f\,x = f\, y$ (by \lstinline|≅-to-≡|) as their types match.
+% 
+%%As shown by McBride, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.
+%
+%\LT[noinline]{Mention the NbE for SC here?}
 

### 2025-02-19 tex/main.tex
Some polishing Section 3

@@
   Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
-  \qed
   \danger
@@
 \begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
-  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$. \qed
+  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
 \end{restatable}
@@
   \]
-  hold for any type $A$ and any term $t$. \qed
+  hold for any type $A$ and any term $t$.
 \end{restatable}
@@
   Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
-  \qed
     \danger
@@
 
-Similar to term substitution, we can show that the lifted substitution $\sigma \uparrow A$ is propositionally equal to $\sigma^+$, based on two basic properties\footnote{%
-  The reader is invited to prove them using the QII definition with transports.}
-about the lifting.
+The lifting $\sigma^+$ interacts well with identities and composition.
+For comparision, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
+%
 \begin{lemma} \label{lem:lifting}
-  The following statements hold:
-  \[
-    \idS^+ =^{\Sub\;(\Gamma, A)\;(\Gamma, A)} \idS
-    \qquad\text{and}\qquad
-    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+.
-  \]
+  For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
 \end{lemma}
 
+Using this lemma, we can show that the lifted substitution $\sigma \uparrow A$ agrees propositionally with its `default' reduction $\sigma^+$.
+
 \begin{restatable}{proposition}{propCorrectnessTwo}\label{prop:correctness-2}
-    For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
-    $\sigma \uparrow A$ is propositionally equal to $\sigma^+$.
-    \qed
+    For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, we have
+    $\sigma \uparrow A = \sigma^+$.
 \end{restatable}
@@
 \begin{proposition} \label{prop:correctness-3}
-  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$,
-  the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
+  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
 \end{proposition}
@@
     for any type $A$ and any term $t$.
-    \qed
 \end{restatable}
 
-\subsubsection{Multiple lifting and the coherence of type substitution}
+\subsubsection*{Multiple lifting and the coherence of type substitution}
 
@@
 To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
-\begin{align*}
-  \subTy{(\Pi\;A\;B)}{\sigma} & = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \\
-  \subTy{(\Pi\;A\;B)}{\tau} & = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
-\end{align*}
+\[
+  \subTy{(\Pi\;A\;B)}{\sigma} = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \qquad
+  \subTy{(\Pi\;A\;B)}{\tau} = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
+\]
 but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
@@
   For every equality constructor $\cc : \sigma = \tau$ and telescope $\Xi$, we have
-  $\subTel{\Xi}{\sigma} = \subTel{\Xi}{\tau}$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\sigma})\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$. \qed
+  $\subTel{\Xi}{\sigma} = \subTel{\Xi}{\tau}$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\sigma})\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$.
   \danger
@@
   We have $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
-  \qed
   \danger

### 2025-02-19 tex/main.tex
polish introduction (saves some space, but no effect at the end yet)

@@
 \title{Type theories as quotient inductive-recursive types}
-\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
-\author{Fredrik Nordvall Forsberg}{Department of Computer and Information Sciences, University of Strathclyde, UK \and \url{https://fredriknf.com}}{fredrik.nordvall-forsberg@strath.ac.uk}{https://orcid.org/0000-0001-6157-9288}{Supported by the Engineering and Physical Sciences Research Council [EP/Y000455/1].}
-\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
+\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{The National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
+\author{Fredrik Nordvall Forsberg}{Department of Computer and Information Sciences, University of Strathclyde, UK \and \url{https://fredriknf.com}}{fredrik.nordvall-forsberg@strath.ac.uk}{https://orcid.org/0000-0001-6157-9288}{The Engineering and Physical Sciences Research Council [EP/Y000455/1].}
+\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{The National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
 \authorrunning{L.-T.~Chen, F.~Nordvall Forsberg, and T.-C.~Tsai}
@@
 \ccsdesc[500]{Theory of computation~Type theory}
-\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
+\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families, type theory in type theory.}
 \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
@@
 
-\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/L-TChen/TTasQIIRT}}%optional, e.g.
-related research data, source code, ...
-hosted on a repository like zenodo, figshare, GitHub, ...
+\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}./}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
 %\supplementdetails[linktext={opt.
@@
 %
-In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas, etc.
-%
-Doing this effectively seems to require several sophisticated features of the host type theory, most notably advanced inductive definitions.
+In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas.
+%
+Doing this effectively seems to require sophisticated features of the host type theory, most notably advanced inductive definitions.
 %
@@
 %
-This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.
-
-Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to equality constructors.
-%
-Yet, there is a general phenomenon, called `transport hell', when working with quotient inductive-inductive types: to make use of equalities in later constructors and when using the eliminator for the type, the equalities need to be transported across.
-%
+This is not just a matter of pure strength, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.
+
+Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types} (QIITs)~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equality constructors can be used to represent equations such as $\beta$- and $\eta$-equality.
+%
+Yet, there is a general phenomenon, called `transport hell', when working with QIITs: to make use of equalities in later constructors and when using eliminators, equalities need to be transported across,
+leading to intricate higher-dimensional equations between equations.
 
@@
 
-In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s, for short).
-%
-The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
-%
-Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
-%
-This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
-%
-As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
+In this paper, we show that the development can be streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s).
+%
+The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and precursors by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
+%
+Our development also makes use of another feature not usually found in standard presentations of type theory: pattern matching definitions with overlapping patterns.
+%
+This is helpful to avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees~\cite{Cockx2017}.
+%
+As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with QIITs alone).
 %
@@
 %
-We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
-Nevertheless, \Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
+We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are, so far, not yet well understood or foundationally well motivated.
+Nevertheless, the proof assistant \Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
 Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.
@@
 \subparagraph*{Contributions}
-This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory.
-In particular, we give
+This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory. Our contributions are:
 \begin{itemize}
-  \item definitions of parallel substitution calculus, type theories with a universe and $\Pi$-types,
-  \item with their elimination principles and the standard model;
-  \item a comparison with these definitions using quotient inductive-inductive types.
+  \item We give definitions of the parallel substitution calculus and type theories with a universe and $\Pi$-types as a QIIRT (\cref{sec:QIIRTs}).
+  \item We show how elimination principles for reasoning about the type theory can be derived, and use it to construct its standard model (\cref{sec:models}).
+  \item We compare our definition to definitions using quotient inductive-inductive types, and prove that the two definitions are actually equivalent (\cref{thm:equivalence}).
 \end{itemize}
 
-\subparagraph*{Plan of the paper}
-In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
-Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
-We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.
+%\subparagraph*{Plan of the paper}
+%In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
+%Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, follo%wed by their elimination principles and usage in \cref{sec:models}.
+%We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.
 
@@
 Going back and forth from one definition to the other gives us the equivalences
-\begin{restatable}{theorem}{equivalence}
+\begin{restatable}{theorem}{equivalence}\label{thm:equivalence}
   Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 

### 2025-02-19 tex/main.tex
Polish Section 2 (currently 15.8 pages)

@@
 
-\section{Metatheory and formalisation} \label{sec:meta}
+\section{Notation and metatheory} \label{sec:meta}
 
@@
 We are further inspired by \Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
-We write $\Set_i$ for the hierarchy of universes with the level index $i$ suppressed.
-
-We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the `dependent' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$.
-One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is the operation to transport values between the type family $P$ along the identity type.
+%We write $\Set_i$ for the hierarchy of universes with the level index $i$ suppressed.
+We write $\Set$ for the universe of types (suppressing to mention a hierarchy of universes $\Set_i$).
+
+We write $x \jdeq y$ for strict (definitional) equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the `dependent' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$.
+One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is the operation to transport values in the type family $P$ along an equality proof.
 %
@@
 We have implemented our definitions and proofs in \Agda; see \cref{sec:formal-implementation} for details.
-However, some statements are metatheoretical and not checked in \Agda, while others are only partially checked by \Agda, as proof obligations are given manually. %for postulated equality constructors.
+However, some statements are metatheoretical and thus not checked in \Agda.
+%, while others are only partially checked by \Agda, as proof obligations are given manually.
+%for postulated equality constructors.
 We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
@@
 \subsection{Inductive-recursive types}
-An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined, using an elimination rule which says that $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
+An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined using an elimination rule which says that $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
 %For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
@@
 
-In contrast, an inductive-recursive type(IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F$ out of $A$ to some fixed (possibly large) type $D$.
+In contrast, an inductive-recursive type (IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F : A \to D$ to some fixed (possibly large) type $D$.
 The prototypical example is Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents.
@@
 %
-In this work, we will consider an extension \emph{generalised} IRTs, where $F : A \to A$ is also allowed.
+In this work, we will consider an extension of \emph{generalised} IRTs, where $F : A \to A$ is also allowed.
 %
@@
 %; indeed, inductive-inductive definitions only make sense when $D = \Set$.
-The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}
-
-We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list.
-We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
+The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
+
+We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list.
+We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$ from \cref{fig:QIT-join-list}, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
 \[
@@
 \subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
-In a proof assistant, functions are typically defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
-This elaboration offers the conventional notation, while making computational behaviour opaque.
+In a proof assistant such as \Agda, functions are defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following a \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
+%This elaboration offers conventional notation, while making computational behaviour opaque.
 % and favouring asymmetric definitions.
-For example, two definitions of addition below
+For example, the two definitions of addition $\blank  + \blank$, $\blank  +' \blank $
+%
 %\begin{figure}[b]%
@@
   \begin{alignat*}{5}
-    \blank  & + \blank     && : \N \to \N \to \N \\
+%    \blank  & + \blank     && : \N \to \N \to \N \\
     0       &  + m         && = m  \\
@@
 \begin{alignat*}{5}
-  \blank         & +'\blank     && : \N \to \N \to \N \\
+%  \blank         & +'\blank     && : \N \to \N \to \N \\
   0              & +' m         && \defeq m  \\
@@
 Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
-Under this semantics, each clause of $+$ is strictly equal to its right-hand side, also allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
-
-A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
+Under this semantics, each clause of $+$ is strictly equal to its right-hand side. %, also allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
+
+A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any overlapping clauses must eventually compute to the same term, i.e.\ the definition must be confluent.
 Cockx shows that if a definition of $f$ is
@@
   \item \emph{structurally recursive}~\cite{Abel2002}, i.e.\ arguments of recursive calls are structurally smaller, and
-  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, terms $\sigma t_1$ and $\sigma t_2$ have the same normal form,
+  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, the terms $\sigma t_1$ and $\sigma t_2$ have the same normal form,
 \end{enumerate}
-then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
+then $f$ is confluent --- essentially a variant of Newman's Lemma~\cite{Newman1942}.
 We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.

### 2025-02-19 tex/main.tex
polish 3.1 and 3.1 (currently 15.5 pages)

@@
 
-We begin with a recall of the definition of substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
+We begin by recalling the definition of the substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
 We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning, and develop a type theory with a universe (\cref{subsec:SC+U}) of small types, $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).
@@
 
-Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e.\ with $\Gamma$ and $\Delta$ being implicit type substitution as a constructor has type
+A substitution $\sigma : \Sub\;\Gamma\;\Delta$ acts on types $A : \Ty\;\Delta$ over $\Delta$ to ``pull them back'' to types $\sub{A}{\sigma} : \Ty\;\Gamma$ over $\Gamma$.
+That is, the action of type substitution can be seen as a constructor
 \begin{alignat*}{3}
@@
 \end{alignat*}
-
+%
 Since $\Ty$ is inductively defined, we add a base case $\UU : \implicit{\Gamma} \Ty\,\Gamma$ to ensure that $\Ty\;\Gamma$ is not empty.
@@
 
-Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as (snoc-)lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
+Substitutions from $\Gamma$ to $\Delta = \emptyctx, A_1, \ldots, A_n$ can be understood intuitively as lists of terms $t_i : A_i$ in context $\Gamma$ for each type $A_i$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which substitutes the most recently bound variable by the term~$t$:
 \[
@@
 \]
-Note that type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
+Note that the type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
 We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$.
@@
 \]
-These need to satisfy certain laws, which we introduce later as equality constructors.
-Context extension is characterised by its projections, which picks out the rest of the context and the most recent variable
+These need to satisfy laws, which we will introduce later as equality constructors.
+Context extension is characterised by its projections, which forgets and picks out the most recently substituted term respectively:
 %Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
@@
 \end{alignat*}
-respectively.
+%
 Again, a type substitution $\sub{A}{\pi_1\,\sigma}$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
@@
 The structural rules for type substitution are stipulated by following equality constructors:
-\begin{alignat*}{3}
-  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && \sub{A}{\idS}           && =^{\Ty\,\Gamma}\;A \\
-  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& \sub{A}{\sigma ; \tau}  && =^{\Ty\,\Gamma}\; \sub{(\sub{A}{\tau})}{\sigma} \\
-  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && \sub{\UU}{\sigma}        && =^{\Ty\,\Gamma}\;\UU
-\end{alignat*}
+\begin{align*}
+  [\idS]_{\Ty} &: \implicit{\Gamma, A} \; \sub{A}{\idS} =^{\Ty\,\Gamma} A
+  \qquad\qquad
+  [;]_{\Ty}  : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\; \sub{A}{\sigma ; \tau} =^{\Ty\,\Gamma} \sub{(\sub{A}{\tau})}{\sigma}
+  \\
+  []\UU &: \implicit{\Gamma, \Delta, \sigma} \; \sub{\UU}{\sigma} =^{\Ty\,\Gamma} \UU
+\end{align*}
+%
+%\begin{alignat*}{3}
+%  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && \sub{A}{\idS}           && =^{\Ty\,\Gamma}\;A \\
+%  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& \sub{A}{\sigma ; \tau}  && =^{\Ty\,\Gamma}\; \sub{(\sub{A}{\tau})}{\sigma} \\
+%  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && \sub{\UU}{\sigma}        && =^{\Ty\,\Gamma}\;\UU
+%\end{alignat*}
 
@@
 \]
-
-The laws for substitution composition are stipulated using equality constructors, where, for the first time, we see the need to adjust the types involved with a $\transp$ in order for both sides to have the same type.
-\begin{alignat*}{5}
-  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
-  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
-\end{alignat*}
-%
-Composition of substitutions $\blank;\blank$ acts like a list concatenation, so
+%
+The laws for substitution composition are stated using equality constructors, where, for the first time, we see the use of $\transp$ for both sides of the equation to have the same type.
+\begin{align*}
+  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} {\sigma ; \idS_{\Delta}} =^{\Sub\,\Gamma\,\Delta} {\sigma}
+                                                                                    \qquad\qquad
+  \mathsf{idl}    : \implicit{\Gamma, \Delta, \sigma} {\idS_{\Gamma} ; \sigma} =^{\Sub\,\Gamma\,\Delta} {\sigma} \\
+  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; (\sigma ; \tau) ; \gamma =^{\Sub\,\Gamma\,\Theta} \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; \sigma ; (\tau , t) =^{\Sub\,\Gamma\,(\Theta, A)} \; (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
+\end{align*}
+%\begin{alignat*}{5}
+%  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+%  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
+%  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
+%  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; &%& \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
+%\end{alignat*}
+%
+Composition of substitutions $\blank;\blank$ acts like list concatenation, so
 we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), \sub{t}{\sigma}$.
@@
 
-The first and second projections of $(\sigma, t)$ should obviously be $\sigma$ and $t$ respectively
-\begin{alignat*}{5}
-  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,(\sub{A}{\blank})}_{\alert{\pi_1\beta}} &&  t, 
-\end{alignat*}
+We would expect the first and second projections of $(\sigma, t)$ to be $\sigma$ and $t$ respectively,
+\[
+  \pi_1\beta : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; \pi_1(\sigma , t)        =^{\Sub\,\Gamma,\Delta} \sigma
+  \qquad
+  \pi_2\beta : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; \pi_2(\sigma , t) =^{\Tm\,\Gamma\,(\sub{A}{\blank})}_{\alert{\pi_1\beta}} t,
+\]
+%\begin{alignat*}{5}
+%  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1%(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
+%  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2%(\sigma , t)        && =^{\Tm\,\Gamma\,(\sub{A}{\blank})}_{\alert{\pi_1\beta}} &&  t, 
+%\end{alignat*}
 but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;(\sub{A}{\pi_1\,(\sigma, t)})$ instead of $\Tm\;\Gamma\;(\sub{A}{\sigma})$.
@@
 %respectively.
-Constructors introduced so far complete the inductive definition of substitution calculus.
-
-The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
+Constructors introduced so far complete the inductive definition of the substitution calculus.
+
+The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts such as the one below.
 \begin{example}\label{ex:pi2-comp}
@@
 %\end{alignat*}
-Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer, since the codomain of the recursive function is the inductively defined typed itself.
+Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer1999}, since the codomain of the recursive function is the inductively defined typed itself.
 %Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
@@
 %\subTy{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \nonumber \\
-\subTy{\UU}{\sigma}  & = \UU                                   \label{eq:type-sub-at2-1} \\
-\subTy{A}{\idS}    & = A                                     \label{eq:type-sub-at2-2} \\
-\subTy{A}{\sigma ; \tau}    & = \subTy{(\subTy{A}{\tau})}{\sigma} \label{eq:type-sub-at2-3}
+\subTy{\UU}{\sigma}  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
+\subTy{A}{\idS}    & \defeq A                                     \label{eq:type-sub-at2-2} \\
+\subTy{A}{\sigma ; \tau}    & \defeq \subTy{(\subTy{A}{\tau})}{\sigma} \label{eq:type-sub-at2-3}
 \end{alignat}
@@
 
-To ensure that type substitution is strict for all %\emph
-{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
+To ensure that type substitution is strict for all clauses,
+we turn to a definition by overlapping patterns %(\cref{sec:meta:overlapping})
+instead. % to make each clause strict.
 In addition, we can include other derived rules such as $\subTy{A}{\pi_1(\sigma, t)} = \subTy{A}{\sigma}$ as long as they are locally confluent and terminating.
-We end up with the following definition for type substitution:
+We end up with the following definition for type substitution $\subTy{\blank}{\blank} : \implicit{\Gamma, \Delta} \; \Ty\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty\,\Gamma$:
 \begin{alignat}{3}
-\subTy{\blank}{\blank}            &: \implicit{\Gamma, \Delta} \; \Ty\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty\,\Gamma \nonumber \\
+% \subTy{\blank}{\blank}            &: \implicit{\Gamma, \Delta} \; \Ty\,\Delta \to \Sub\,\Gamma\,\Delta \to \Ty\,\Gamma \nonumber \\
 \subTy{A}{\idS}                & \reduce A \label{eq:def-type-subst-1}\\
@@
   Termination is clear, since all recursive calls are on structurally smaller arguments~\cite{Abel2002}.
-  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
+  For local confluence, we consider all local peaks, i.e., each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
   \[
@@
 
-As we are defining a function on a quotient inductive type, we have to prove that $\subTy{\blank}{\blank}$ is coherent, i.e., that it respects all equality constructors.
+However, we still have a little bit more work to do before we can declare victory: as we are defining a function on a quotient inductive type, we have to prove that $\subTy{\blank}{\blank}$ is coherent, i.e., that it respects all equality constructors.
 %\LT[noinline]{Should we mention any general condition for coherence?}
@@
   hold propositionally.
-  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $\subTy{(\subTy{(\subTy{A}{\gamma})}{\tau})}{\sigma}$ by \eqref{eq:def-type-subst-2}.
-
-  The sole identity of the second row is clause \eqref{eq:def-type-subst-3} of the definition of $\subTy{\blank}{\blank}$.
-
+  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while both sides of the third reduce to $\subTy{(\subTy{(\subTy{A}{\gamma})}{\tau})}{\sigma}$ by \eqref{eq:def-type-subst-2}.
+  %
+  The second row is clause \eqref{eq:def-type-subst-3} of the definition of $\subTy{\blank}{\blank}$.
+  %
   We show the remaining cases by induction on $A$.
-  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of the identities with $A \defeq \UU$ reduce to $\UU$.
+  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ and each of the identities with $A \judgeq \UU$ reduce to $\UU$.
 \end{proof}
@@
   In the above proof the first four identities hold strictly by \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4}, while the last three follow from \eqref{eq:def-type-subst-5} by induction on $A$.
-  As we extend substitution calculus with other type formers, the first four identities will always hold but the last three depend on how type substitution is defined.
+  As we add other type formers, the first four identities will always hold, but the last three will depend on how type substitution is defined.
 \end{remark}
@@
 cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
-
+%
 To repair the local confluence, we add a simultaneously defined function $\subTm{\blank}{\blank}$ for term substitution apart from the explicit term substitution $\sub{\blank}{\blank}$ as follows

### 2025-02-19 tex/main.tex
Polish 3.4 (currently 15.1 pages)

@@
 Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
-For a QII definition \cite{Altenkirch2016a}, this extension adds the following constructors
+In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
 \begin{alignat*}{3}
@@
 \end{alignat*}
-Yet, this naive change breaks the local confluence!
-The problem is that there is now additional interaction between types and terms. Observe that the local peak
-\[
+Yet, this naive change breaks local confluence!
+The problem is that there is now additional interaction between types and terms.
+% Observe that
+The local peak
+\(
   \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
-\]
+\)
 cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
 %
-To repair the local confluence, we add a simultaneously defined function $\subTm{\blank}{\blank}$ for term substitution apart from the explicit term substitution $\sub{\blank}{\blank}$ as follows
-\begin{alignat*}{3}
-  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
+To repair local confluence, we add a simultaneously defined term substitution function $\subTm{\blank}{\blank} : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma})$, different from the constructor $\sub{\blank}{\blank}$, with the following reductions:
+\begin{alignat*}{3}
+%  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
 \subTm{t}{\idS}          & \reduce t \\
@@
 \end{alignat*}
-which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
-The remaining cases reduce to the explicit term substitution $\sub{t}{\sigma}$.
-Note that $\subTm{\blank}{\blank}$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
+This reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from types to terms.
+The remaining ``default'' cases reduce to the constructor $\sub{t}{\sigma}$.
+Note that $\subTm{\blank}{\blank}$ can be defined with the standard first-match pattern matching semantics, since there are no overlapping patterns.
 \begin{remark}
-  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming, since the function $\subTm{\blank}{\blank}$ (mostly) builds an explicit substitution but computes certain cases away.
+  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming: the function $\subTm{\blank}{\blank}$ computes certain cases away, but mostly reduces to the non-smart constructor $\sub{\blank}{\blank}$.
 \end{remark}
@@
 \begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
+  \danger
   Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
-  \danger
 \end{restatable}
@@
 \begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
-  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
+  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
 \end{restatable}
@@
   \danger
-  \[
-    \subTy{A}{\sigma} = \subTy{A}{\tau}
-    \quad\text{and}\quad
-    \subTm{t}{\sigma} = \subTm{t}{\tau}
-  \]
+    $\subTy{A}{\sigma} = \subTy{A}{\tau}$
+    and
+    $\subTm{t}{\sigma} = \subTm{t}{\tau}$
   hold for any type $A$ and any term $t$.
@@
 \begin{remark}
-We have seen that a naive extension with a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
-In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
+We have seen that a naive addition of a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with existing clauses.
+In general, every term that occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this propagation can be achieved by turning a constructor involving a substitution into a recursive function.
 \end{remark}
@@
 
-We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, where the context of its codomain is extended with the domain type.
-For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
+We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types. % --- the context of the codomain is extended with the domain type.
+For a QIIT definition~\cite{Altenkirch2016a}, this amounts to adding the following constructors for the type formation rule and the term introduction/elimination rules
 \begin{alignat*}{3}
@@
 the computation rules
-\begin{alignat*}{5}
-  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
-  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                          && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
-\end{alignat*}
+\[
+  \Pi\beta           : \implicit{\Gamma, A, B, t}\; \mathsf{app}\,(\lambda\,t) =^{\Tm\;\Gamma\;A} t
+  \qquad\qquad
+  \Pi\eta            : \implicit{\Gamma, A, B, t}\; t                         =^{\Tm\;\Gamma\;A} \lambda(\mathsf{app}\,t)
+\]
+%\begin{alignat*}{5}
+%  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
+%  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                      %    && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
+%\end{alignat*}
 and the structural rules for substitution
@@
 we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
-However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $\subTy{\blank}{\blank}$ breaks the local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
+However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $\subTy{\blank}{\blank}$ breaks local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
 For example, $\subTy{(\Pi\;A\;B)}{\idS}$ reduces to $\Pi\;A\;(\subTy{B}{\idS \uparrow A})$ where $\idS \uparrow A$ is propositionally but not strictly equal to $\idS$, so the local peak $\Pi\;A\;B \Leftarrow \subTy{(\Pi\;A\;B)}{\idS} \reduce \Pi\;A\;(\subTy{B}{\idS \uparrow A})$ cannot be joined.
@@
 where $\sigma^+ \defeq (\pi_1\idS ; \sigma), \pi_2\idS$.
-Then, we add another clause for type substitution and an equality constructor for term substitution without transport:
+Then, we add another clause for type substitution
+\begin{equation*}
+  \subTy{(\Pi\,A\,B)}{\sigma} \reduce \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) %\label{eq:def-type-subst-9}
+\end{equation*}
+and an equality constructor for term substitution without transport:
 \begin{alignat}{5}
-                     & \subTy{(\Pi\,A\,B)}{\sigma} && \reduce \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-9} \\
   \mathsf{[]\lambda} :{} & \sub{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
@@
 
-As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ but the same argument still applies:
+As we are working with a different type theory from \cref{subsec:SC+U}, we  have to again prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$, but the same argument still applies:
 \begin{proposition} \label{prop:correctness-3}
@@
 \begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
-    For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, we have
-    \danger
-    \[
-      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A
-      \qquad\text{and}\qquad
-      \subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau},
-    \]
-    for any type $A$ and any term $t$.
+  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$ and  any type $A$ and term $t$,   \danger we have
+  $\sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A$
+      and
+      $\subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau}$.
     \end{restatable}
@@
 
-We also need to prove the coherence for type substitution, i.e., that $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
+We also need to prove coherence of type substitution, i.e., for every equality constructor $\cc : \sigma = \tau$, we have $\subTy{A}{\sigma} = \subTy{A}{\tau}$ .
 To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
@@
 but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
-Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
-Then, for each constructor $\cc : \sigma = \tau$, we show the identity
+Instead, we define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ of $\sigma$ by a telescope $\Xi = A_1, \dots, A_n$, given by multiple applications of lifting $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
+For each constructor $\cc : \sigma = \tau$, we then show %the identity
 $\subTy{A}{\sigma \mathop{\upuparrows} \Xi} = \subTy{A}{\tau \mathop{\upuparrows} \Xi}$
@@
 
-The type of telescopes over a `base' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.
+The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx$ as follows.
 
@@
 \begin{alignat*}{5}
-  \Tel               & :                   && \Ctx \to \Set                                     \\
+  %\Tel               & :                   && \Ctx \to \Set                                     \\
   \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
@@
 \begin{alignat*}{5}
-  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
+  %\blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
   \Gamma & \dplus \emptytel && \reduce \Gamma \\
@@
 \begin{alignat*}{3}
-  & \subTel{\Xi}{\idS}       \Xi && \reduce \Xi \\
+  & \subTel{\Xi}{\idS}          && \reduce \Xi \\
   & \subTel{\Xi}{\sigma ; \tau} && \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
@@
 
-We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also `ordinary' application
+Finally, we note that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also `ordinary' application
 $t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS , t)}$ of type $\Tm\,\Gamma\,(\sub{B}{(\idS , t)})$ can be defined without any transport.
-This is in contrast to the pure QII situation, where application is defined as $\subTm{(\mathsf{app}\,t)}{(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})}$.
+This is in contrast to the QIIT situation, where ordinary application is defined as $t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})}$.
 %
@@
 \end{align*}
-can be derived without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.
+can be derived without transport (cf.\ Altenkirch and Kaposi~\cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} is now strict.
 

### 2025-02-19 tex/main.tex
fit within the page limit

@@
 Similarly, the structural rule for substitution on $\mathsf{app}$
-\begin{align*}
-  \sub{(\mathsf{app}\;t)}{\sigma \uparrow A} & = \mathsf{app}\,\lambda(\sub{(\mathsf{app}\;t)}{\sigma \uparrow A}) && \text{by $\Pi\beta^{-1}$} \\
-                                              & = \mathsf{app}\,\left(\sub{(\lambda\,(\mathsf{app}\,t))}{\sigma}\right) && \text{by $[]\lambda^{-1}$} \\
-                                              & = \mathsf{app}\,(\sub{t}{\sigma}) && \text{by $\Pi\eta^{-1}$.}
-\end{align*}
+\[
+  \sub{(\mathsf{app}\;t)}{\sigma \uparrow A} \overset{\Pi\beta^{-1}}{=}
+  \mathsf{app}\,\lambda(\sub{(\mathsf{app}\;t)}{\sigma \uparrow A}) \overset{[]\lambda^{-1}}{=}
+  \mathsf{app}\,\left(\sub{(\lambda\,(\mathsf{app}\,t))}{\sigma}\right) \overset{\Pi\eta^{-1}}{=}
+  \mathsf{app}\,(\sub{t}{\sigma})
+\]
+%\begin{align*}
+%  \sub{(\mathsf{app}\;t)}{\sigma \uparrow A} & = \mathsf{app}\,\lambda(\sub{(\mathsf{app}\;t)}{\sigma \uparrow A}) && \text{by $\Pi\beta^{-1}$} \\
+%                                              & = \mathsf{app}\,\left(\sub{(\lambda\,(\mathsf{app}\,t))}{\sigma}\right) && \text{by $[]\lambda^{-1}$} \\
+%                                              & = \mathsf{app}\,(\sub{t}{\sigma}) && \text{by $\Pi\eta^{-1}$.}
+%\end{align*}
 can be derived without transport (cf.\ Altenkirch and Kaposi~\cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} is now strict.
@@
 %
-where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ for the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place.
+where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ is the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place.
 We can add the expected computation rules
@@
 %
-and the structural rule for type substitution as another straightforward clause
+as constructors, and the structural rule for type substitution as another straightforward clause
 $\subTy{\bool}{\sigma} \reduce \bool$,
-but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following:
+but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following
 %
@@
 
-We show how to eliminate out of our QIIRT definition of type theory (\cref{subsec:eliminator}).
-In particular, we show how the elimination principle can be used to construct a standard model (\cref{subsec:std-model}) and compare our QIIRT definition with its QIIT counterpart (\cref{subsec:equivalence}).
+We show how to eliminate out of our QIIRT definition of type theory,
+use the eliminator to construct a ``standard'' model,
+and compare our QIIRT definition with its QIIT counterpart.
 
@@
 
-The elimination principle computes by interpreting each constructor using the corresponding method inductively, e.g.,
+The elimination principle computes by recursively interpreting each constructor using the corresponding method. For example, we have
 \begin{alignat*}{5}
 %  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
-  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) %\\
+  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) . && %\\
 %  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
@@
 \begin{remark} \label{re:method-recursion-IIR}
-Note that a simultaneously defined function does not need a corresponding interpretation for Dybjer and Setzer's inductive-recursive types, as their framework targets meta-theoretic types rather than the inductive types being defined.
+Note that a simultaneously defined function does not need a corresponding interpretation for Dybjer and Setzer's inductive-recursive types~\cite{Dybjer1999}, as their framework targets meta-theoretic types rather than the inductive types being defined.
 \end{remark}
@@
 %
-We once again refer to our \Agda formalisation for details.
-We remark that because our lack of transports in the syntax, many of the identities in the standard model simply hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is often needed for reasoning about constant transports.
+See our \Agda formalisation for details.
+Because of the lack of transports in the syntax, many identities in the standard model hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is needed for reasoning about constant transports.
 
@@
 \subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
-We have used additional rules to enable further reduction which are derivable in the QIIT definition.
+Our additional rules, enabling further reduction, are derivable in the QIIT definition.
 Indeed, the QIIRT definition of type theory is equivalent to its QIIT definition, as we can translate every constructor to its corresponding constructor and vice versa.
 For comparison, we add a subscript to constructors and types in the QIIRT and QIIT definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIRT definition to its QIIT counterpart and the converse, respectively.
-
-The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart.
-Additional equality constructors are all derivable in the QIIT definition.
+%
+The definition of $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart.
+The additional equality constructors are all derivable in the QIIT definition.
 
@@
 For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ if the recursive form $\subTm{t^{\prec}}{\sigma^{\prec}}$ is chosen for the translation.
-
-Going back and forth from one definition to the other gives us the equivalences
+%
+% Going back and forth from one definition to the other gives us the equivalences:
+%The equivalence can be summed up as follows:
+%
 \begin{restatable}{theorem}{equivalence}\label{thm:equivalence}
@@
 
-We have a formal implementation in \Agda, using the options \texttt{-{}-with-K} for the uniqueness of identity proofs, \texttt{-{}-exact-split} to ensure that all function clauses are definitional, and \texttt{-{}-rewriting} with \texttt{-{}-local-confluence-check} \cite{Cockx2020,Cockx2021} to implement definitions by overlapping patterns.
-Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}, as \textsc{Cubical Agda} does not support user-defined rewrite rules.
+Our formal \Agda implementation uses the options \texttt{-{}-with-K} (for uniqueness of identity proofs), \texttt{-{}-exact-split} (to ensure that function clauses are definitional), and \texttt{-{}-rewriting} with \texttt{-{}-local-confluence-check} \cite{Cockx2020,Cockx2021} (to implement definitions by overlapping patterns).
+Quotient inductive types are simulated by postulating equations for equality constructors~\cite{Licata2011}. %We do not use \textsc{Cubical Agda}, since it does not support user-defined rewrite rules.
 
@@
 %The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns, we declare each function clause a rewrite and another rewrite to identify the clause with the reflexivity proof
+To emulate a function by overlapping patterns, we declare each function clause a rewrite, together with another rewrite to identify the witness as the reflexivity proof $\mathsf{refl}$.
 For example, the first few clauses for type substitution (written $[\sigma]_{\Ty}\,A$ rather than $\subTy{A}{\sigma}$ in the formalisation) are defined as
@@
 \end{lstlisting}
-where \lstinline|[id]≡refl| is used to ensure that \lstinline|[id]| need not be referred.
+where \lstinline|[id]≡refl| ensures that \lstinline|[id]| disappears in transports.
 We also turn on the option \texttt{-{}-local-confluence-check}  to check local confluence.
-
-Defining by overlapping patterns is indeed one of the use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
-The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.
+%
+Overlapping patterns is indeed one of the use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}.
+The caveat, though, is that the termination checking is not imposed, so we need to check termination informally.
 
@@
 We have used QIIRTs and overlapping
-patterns to represent type theory in (intensional) type theory. By defining type
+patterns to represent type theory in intensional type theory. By defining type
 substitution recursively with overlapping patterns, we can avoid most
 coercions that appear in terms in other formalisations of type
-theory in type theory. In other words, we have shown that so-called
+theory in type theory. That is, we have shown that so-called
 `transport hell' can be minimised when formalising type theory in type
@@
 
-Altenkirch and Kaposi~\cite{Altenkirch2016a} were the first to exploit QIITs to simplify the treatment of computation rules in representations of type theory.% in type theory.
+Altenkirch and Kaposi~\cite{Altenkirch2016a} were the first to exploit QIITs to simplify the treatment of computation rules in representations of type theory. % in type theory.
 We build heavily on their ideas, but by using Danielsson's~\cite{Danielsson2006} idea of defining type substitution recursively, we are able to remove many explicit transports from the definition, with the result that it is closer to an informal exposition on paper.
-Such a benefit was also suggested by Kaposi's~\cite{Kaposi2023} work, where the simply typed $\lambda$-calculus was used as an example.
-Altenkirch and Kaposi has built on the QIIT representation with great success to derive non-trivial results such as normalisation by evaluation~\cite{Altenkirch2017}, and investigate the metatheory of QIITs themselves, together with Kov\'acs~\cite{Kaposi2019}.
+Such a benefit was also suggested by Kaposi~\cite{Kaposi2023}, who used the simply typed $\lambda$-calculus as an example.
+Altenkirch and Kaposi have built on the QIIT representation with great success to derive non-trivial results such as normalisation by evaluation~\cite{Altenkirch2017}, and used it to investigate the metatheory of QIITs themselves, together with Kov\'acs~\cite{Kaposi2019}.
 

### 2025-02-20 tex/main.tex
Polish and fix the notation for substitution up to Section A

@@
 Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional
+\acknowledgements{We'd like to thank Hsiang-Shang Ko for useful discussion.}%optional
 
@@
 %
-In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas.
+In the spirit of type theory, we would like to represent well typed terms only --- this has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas.
 %
@@
 %
-For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
+For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions are used for dependent type theory~\cite{Danielsson2006,Chapman2009}.
 %
@@
 
-A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
+A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules~\cite{Cockx2021,Cockx2020} after an inductive definition.
 %
@@
 However, this workaround is computationally ill-behaved, as confluence does not hold in general.
-Also, the presentation often differs from the actual definition of type theory in intensional type theory.
-%
-Any method that reduces the friction introduced by the use of equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
+Also, the presentation often differs from the actual definition in type theory.
+%
+Any method that reduces the friction introduced by equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
 %
@@
 For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT in \cref{fig:QIT-join-list}.
-\begin{figure}
+\begin{figure}[b]
 \begin{minipage}{.35\textwidth}
 \begin{alignat*}{3}
-  \Jnil                    & : \implicit{A}      && \JList{A} \\
-  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
-  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A}
+  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
+  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
+  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
 \end{alignat*}
@@
 % and favouring asymmetric definitions.
-For example, the two definitions of addition $\blank  + \blank$, $\blank  +' \blank $
+For example, the two definitions of addition $+$ and $+'$
 %
@@
 
-A substitution $\sigma : \Sub\;\Gamma\;\Delta$ acts on types $A : \Ty\;\Delta$ over $\Delta$ to ``pull them back'' to types $\sub{A}{\sigma} : \Ty\;\Gamma$ over $\Gamma$.
+A substitution $\sigma : \Sub\;\Gamma\;\Delta$ acts on types $A : \Ty\;\Delta$ over $\Delta$ to `pull them back' to types $\sub{A}{\sigma} : \Ty\;\Gamma$ over $\Gamma$.
 That is, the action of type substitution can be seen as a constructor
@@
 \begin{alignat*}{3}
-  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
+  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;\Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
 \end{alignat*}
@@
 The structural rules for type substitution are stipulated by following equality constructors:
-\begin{align*}
-  [\idS]_{\Ty} &: \implicit{\Gamma, A} \; \sub{A}{\idS} =^{\Ty\,\Gamma} A
+\begin{alignat*}{9}
+  [\idS]_{\Ty} &: \implicit{\Gamma, A} \; && \sub{A}{\idS} && =^{\Ty\,\Gamma} A
   \qquad\qquad
@@
   \\
-  []\UU &: \implicit{\Gamma, \Delta, \sigma} \; \sub{\UU}{\sigma} =^{\Ty\,\Gamma} \UU
-\end{align*}
+  []\UU &: \implicit{\Gamma, \Delta, \sigma} \; && \sub{\UU}{\sigma} && =^{\Ty\,\Gamma} \UU
+\end{alignat*}
 %
@@
 \begin{example}\label{ex:pi2-comp}
-  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $\sub{A}{\sigma;\tau}$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$.
-  In short, the following identity apparently holds
+  Given any substitution $\sigma : \Sub\;\Gamma\; \Delta$ and a non-empty substitution $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, the first term of the composite $(\sigma; \tau)$ should intuitively be equal to the first term of $\tau$ after applying the substitution $\sigma$, i.e.\ the identity
   \[
@@
   \]
-  and is checked by a back-of-the-envelope calculation
+  apparently holds and can checked by a back-of-the-envelope calculation
   \begin{equation} \label{eq:pi2-comp-proof}
@@
 %Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
-Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define$[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
+Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define $[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
 \begin{alignat}{3}
 %\subTy{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \nonumber \\
-\subTy{\UU}{\sigma}  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
-\subTy{A}{\idS}    & \defeq A                                     \label{eq:type-sub-at2-2} \\
-\subTy{A}{\sigma ; \tau}    & \defeq \subTy{(\subTy{A}{\tau})}{\sigma} \label{eq:type-sub-at2-3}
+\subTy{\UU}{\sigma}         & \defid \UU                                   \label{eq:type-sub-at2-1} \\
+\subTy{A}{\idS}             & \defid A                                     \label{eq:type-sub-at2-2} \\
+\subTy{A}{\sigma ; \tau}    & \defid \subTy{\subTy{A}{\tau}}{\sigma} \label{eq:type-sub-at2-3}
 \end{alignat}
@@
 \subTy{A}{\idS}                & \reduce A \label{eq:def-type-subst-1}\\
-\subTy{A}{\sigma ; \tau}       & \reduce \subTy{(\subTy{A}{\tau})}{\sigma} \label{eq:def-type-subst-2}\\
+\subTy{A}{\sigma ; \tau}       & \reduce \subTy{\subTy{A}{\tau}}{\sigma} \label{eq:def-type-subst-2}\\
 \subTy{A}{\pi_1(\sigma, t)}    & \reduce \subTy{A}{\sigma} \label{eq:def-type-subst-3} \\
-\subTy{A}{\pi_1(\sigma; \tau)} & \reduce \subTy{(\subTy{A}{\pi_1\tau})}{\sigma} \label{eq:def-type-subst-4} \\
+\subTy{A}{\pi_1(\sigma; \tau)} & \reduce \subTy{\subTy{A}{\pi_1\tau}}{\sigma} \label{eq:def-type-subst-4} \\
 \subTy{\UU}{\sigma}            & \reduce \UU                                   \label{eq:def-type-subst-5}
@@
   Termination is clear, since all recursive calls are on structurally smaller arguments~\cite{Abel2002}.
-  For local confluence, we consider all local peaks, i.e., each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
+  For local confluence, we consider each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
   \[
@@
   \]
-  and show that each applicable clause does result in $\UU$.
+  and show that each local peak results in $\UU$ using any applicable rules.
   %
   For example, $\subTy{\UU}{\sigma;\tau} \reduce \subTy{(\subTy{\UU}{\tau})}{\sigma} \reduce \subTy{\UU}{\sigma} \reduce \UU$ using \eqref{eq:def-type-subst-2} as the first reduction and $\subTy{\UU}{\sigma;\tau} \reduce \UU$ using \eqref{eq:def-type-subst-5} as the first reduction.
-  It is easy to see that remaining local peaks reduce to the same term.
+  It is easy to see that other local peaks reduce to the same term.
 \end{proof}
@@
 \begin{alignat*}{5}
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t) && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A} && t \\
-  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,\sub{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}}
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) {} && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)          {} && =^{\Tm\,\Gamma\, A} &&  t \\
+  [\idS]t         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A}     {} && t \\
+  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}  && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,\sub{A}{\sigma ; \tau}} {} && {\sub{\sub{t}{\tau}}{\sigma}}
 \end{alignat*}
@@
 \begin{example} \label{ex:pi-comp-qiirt}
-  The equation $\pi_2\,(\sigma; \tau) = \sub{(\pi_2\,\tau)}{\sigma}$ can be formally stated without any transport.
-  The type of the term on the LHS is $\subTy{A}{\pi_1(\sigma;\tau)}$ which is strictly equal to the type $\subTy{(\subTy{A}{\pi_1\tau})}{\sigma}$ of the term on the RHS.
+  Now, the equation $\pi_2\,(\sigma; \tau) = \sub{(\pi_2\,\tau)}{\sigma}$ can be formally stated without any transport: the type of the term on the left is $\subTy{A}{\pi_1(\sigma;\tau)}$ which is strictly equal to the type $\subTy{(\subTy{A}{\pi_1\tau})}{\sigma}$ of the term on the right.
   Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are strictly equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
@@
 %
+
 To repair local confluence, we add a simultaneously defined term substitution function $\subTm{\blank}{\blank} : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma})$, different from the constructor $\sub{\blank}{\blank}$, with the following reductions:
@@
 This reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from types to terms.
-The remaining ``default'' cases reduce to the constructor $\sub{t}{\sigma}$.
+The remaining `default' cases reduce to the constructor $\sub{t}{\sigma}$.
 Note that $\subTm{\blank}{\blank}$ can be defined with the standard first-match pattern matching semantics, since there are no overlapping patterns.
@@
 \begin{alignat}{5}
-  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && \subTy{(\Pi\;A\;B)}{\sigma} && =^{\Ty\;\Gamma} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-8} \\
-  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && \subTm{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
+  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && \sub{(\Pi\;A\;B)}{\sigma} && =^{\Ty\;\Gamma} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) \label{eq:def-type-subst-8} \\
+  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && \sub{(\lambda\,t)}{\sigma} && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,(\sub{t}{\sigma \uparrow A}) \nonumber
 \end{alignat}
@@
 \idS                \uparrow A & \reduce \idS \nonumber \\
-(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
+(\sigma ; \tau)     \uparrow A & \reduce (\sigma \uparrow \subTy{\tau}{A} ) ; (\tau \uparrow A) \nonumber \\
 \pi_1(\sigma, t)    \uparrow A & \reduce \sigma \uparrow A \nonumber \\
-\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
+\pi_1(\sigma; \tau) \uparrow A & \reduce (\sigma \uparrow \subTy{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A)\nonumber \\
 \sigma              \uparrow A & \reduce \sigma^+ , \quad \text{otherwise} \nonumber
@@
 \begin{equation*}
-  \subTy{(\Pi\,A\,B)}{\sigma} \reduce \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A}) %\label{eq:def-type-subst-9}
+  \subTy{(\Pi\,A\,B)}{\sigma} \reduce \Pi\,(\subTy{A}{\sigma})\,(\subTy{B}{\sigma\uparrow A}) %\label{eq:def-type-subst-9}
 \end{equation*}
@@
 but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
-Instead, we define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ of $\sigma$ by a telescope $\Xi = A_1, \dots, A_n$, given by multiple applications of lifting $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
+Instead, we define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, given by multiple applications of lifting $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
 For each constructor $\cc : \sigma = \tau$, we then show %the identity
@@
 
-The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx$ as follows.
+The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) \to \Tel\;\Gamma \to \Ctx$ as follows.
 
@@
   %\Tel               & :                   && \Ctx \to \Set                                     \\
-  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
-  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
+  \emptytel          & : \implicit{\Gamma}\; && \Tel\;\Gamma \\
+  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
 \end{alignat*}
@@
 \]
-which requires us to also simultaneously define telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$.
+which asks for telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$ defined mutually:
 
@@
 \begin{alignat*}{5}
-  & \subTel{\emptytel}{\sigma} && \reduce \emptytel \\
-  & \subTel{(\Xi, A)}{\sigma}  && \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
+  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \\
+  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
 \end{alignat*}
@@
 %
-Instead of resorting to \transp, we also turn telescope substitution $\subTel{\blank}{\blank}$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
+Instead of resorting to \transp, we can also turn telescope substitution $\subTel{\blank}{\blank}$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
 
@@
 \begin{alignat*}{3}
-  & \subTel{\Xi}{\idS}          && \reduce \Xi \\
-  & \subTel{\Xi}{\sigma ; \tau} && \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
+  \subTel{\Xi}{\idS}          & \reduce \Xi \\
+  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
 \end{alignat*}
@@
   For every equality constructor $\cc : \sigma = \tau$ and telescope $\Xi$, we have
-  $\subTel{\Xi}{\sigma} = \subTel{\Xi}{\tau}$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\sigma})\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$.
   \danger
+  \begin{align*}
+    \subTel{\Xi}{\sigma} & = \subTel{\Xi}{\tau}
+                         && \sigma \upuparrows \Xi =^{\lam{\gamma}\Sub\;(\Gamma\dplus \subTel{\Xi}{\gamma})\;(\Delta\dplus\Xi)}_{\alert{\mathsf{c}}} \tau \upuparrows \Xi \\
+    \subTy{A}{\sigma \upuparrows \Xi} & =^{\lam{\gamma}\Ty\;(\Gamma \dplus \subTel{\Xi}{\gamma})}_{\alert{\mathsf{c}}} \subTy{A}{\tau \upuparrows \Xi}.
+  \end{align*}
 \end{restatable}
@@
 %\end{align*}
-can be derived without transport (cf.\ Altenkirch and Kaposi~\cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} is now strict.
+can be derived without transport, since \eqref{eq:def-type-subst-8} is now strict.
 
@@
 We show how to eliminate out of our QIIRT definition of type theory,
-use the eliminator to construct a ``standard'' model,
+use the eliminator to construct a `standard' model,
 and compare our QIIRT definition with its QIIT counterpart.
@@
 \begin{alignat*}{5}
-  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && A^M [ \idS^M ]^M_{\Ty} && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
+  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && \subTyM{A^M}{\idS^M} && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
   %[;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
@@
   \[
-    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\;t^M [\idS^M]^M_{\Tm}  =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
+    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\subTmM{t^M}{\idS^M} =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
   \]
@@
 \begin{alignat*}{5}
-  \elim_{\Ty}[] & : \elim_{\Ty}\,(\subTy{A}{\sigma}) && = && (\elim_{\Ty}\,A)[\elim_{\Sub}\,\sigma]^M_{\Ty} \\
-  \elim_{\Tm}[] & : \elim_{\Tm}\,(\subTm{t}{\sigma}) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Tm}\,t) [\elim_{\Sub}\,\sigma]^M_{\Tm} \\
+  \elim_{\Ty}[] & : \elim_{\Ty}\,(\subTy{A}{\sigma}) && = && \subTyM{(\elim_{\Ty}\,A)}{\elim_{\Sub}\,\sigma} \\
+  \elim_{\Tm}[] & : \elim_{\Tm}\,(\subTm{t}{\sigma}) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && \subTmM{(\elim_{\Tm}\,t)}{\elim_{\Sub}\,\sigma} \\
   \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
@@
 \]
-where the transport is needed since $\elim_{\Tm}\;t : \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\alert{\elim_{\Ty}\;(\subTy{A}{\sigma})})\;t$ but $\blank ,_{\Sub}^M \blank$ expects its second argument to have type $\Tm^M\;(\elim_{\Ctx}\,\Gamma)\;\left(\alert{(\elim_{\Ty}\,A)[\elim_{\Sub}\,\sigma]^M_{\Ty}}\right)\;t$.
+where the transport is needed since $\elim_{\Tm}\;t : \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\alert{\elim_{\Ty}\;(\subTy{A}{\sigma})})\;t$ but $\blank ,_{\Sub}^M \blank$ expects its second argument to have type $\Tm^M\;(\elim_{\Ctx}\,\Gamma)\;\left(\alert{\subTyM{(\elim_{\Ty}\,A)}{\elim_{\Sub}\,\sigma}}\right)\;t$.
 
@@
 \begin{remark} \label{re:method-recursion-IIR}
-Note that a simultaneously defined function does not need a corresponding interpretation for Dybjer and Setzer's inductive-recursive types~\cite{Dybjer1999}, as their framework targets meta-theoretic types rather than the inductive types being defined.
+Note that a simultaneously defined function does not need a corresponding interpretation for Dybjer and Setzer's inductive-recursive types~\cite{Dybjer1999}, as their framework targets pre-existing types rather than the inductive types being defined.
 \end{remark}
@@
 Our additional rules, enabling further reduction, are derivable in the QIIT definition.
-Indeed, the QIIRT definition of type theory is equivalent to its QIIT definition, as we can translate every constructor to its corresponding constructor and vice versa.
+Indeed, the QIIRT definition of type theory is equivalent to its QIIT definition, as we can translate each constructor to its corresponding constructor and vice versa.
 For comparison, we add a subscript to constructors and types in the QIIRT and QIIT definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIRT definition to its QIIT counterpart and the converse, respectively.
@@
 On the other hand, the converse $(\blank)^{\prec}$ has some choices.
-Since in the QIIT definition there is only one term substitution $\blank [\blank]_{\I}$ and lifting $(\blank)^+_{\I}$, we may choose to translate $t [\sigma]_{\I}$ to the recursive form $t^{\prec} [\sigma^{\prec}]_{\Tm}$ to carry out some computation, or to the constructor form $t^{\prec} [\sigma^{\prec}]_{\IR}$ to stay verbatim.
+Since in the QIIT definition there is only one term substitution $\sub{\blank}{\blank}_{\I}$ and lifting $(\blank)^+_{\I}$, we may choose to translate $\sub{t}{\sigma}_{\I}$ to the recursive form $\subTm{t^{\prec}}{\sigma^{\prec}}$ to carry out some computation, or to the constructor form $\sub{t^{\prec}}{\sigma^{\prec}}_{\IR}$ to stay verbatim.
 Either choice is possible, but we need to use \cref{prop:correctness-1,prop:correctness-3} to switch from one to the other whenever suitable.
-For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ if the recursive form $\subTm{t^{\prec}}{\sigma^{\prec}}$ is chosen for the translation.
+For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ if the recursive form $\subTm{t^{\prec}}{\sigma^{\prec}}$ is chosen for $(\sub{t}{\sigma}_{\I})^{\prec}$.
 %
@@
 Like us, Danielsson~\cite{Danielsson2006} defines type substitution by a recursive function, while keeping term substitution a constructor.
-However because he does not use quotients, he also has an explicit term constructor for transporting terms along an inductively defined type of conversion rules, which explicitly contains congruence rules.
+However he does not use quotients, so he also has an explicit term constructor for transporting terms along an inductively defined type of conversion rules, which explicitly contains congruence rules.
 We are able to remove a lot of the boilerplate by using equality constructors instead.
@@
 with such a schema. Eremondi and Kammar~\cite{Eremondi2025} has
-recently given a promising possible direction for overlapping patterns
+recently given a promising direction for overlapping patterns
 using sheaf theory.
@@
 \idS                \uparrow A  & \reduce \idS \\
-\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \sub{A}{\tau}) ; (\tau \uparrow A) \\
+\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\tau}) ; (\tau \uparrow A) \\
 \pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\sub{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
+\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
 \sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwiese}
@@
 \begin{alignat*}{3}
-  & \subTy{(\Pi\,A\,B)}{\sigma}       && {} \reduce {} && \Pi\,(\sub{A}{\sigma})\,(\sub{B}{\sigma\uparrow A})
-\end{alignat*}
-\begin{alignat*}{5}
-  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\;\Gamma\;(\subTy{(\Pi\,A\,B)}{\sigma}} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
+  & \subTy{(\Pi\,A\,B)}{\sigma}       && {} \reduce {} && \Pi\,(\subTy{A}{\sigma})\,(\subTy{B}{\sigma\uparrow A})
+\end{alignat*}
+\begin{alignat*}{5}
+  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\;\Gamma\;(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
   \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\;\Gamma\;A} && t \\
@@
 \begin{alignat*}{5}
-  \idtype[\blank]{\blank}{\blank} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
-  & \subTy{(\idtype[a]{t}{u})}{\sigma} && \reduce \idtype[{\subTm{a}{\sigma}}]{\subTm{t}{\sigma}}{\subTm{u}{\sigma}} \\
-  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\idtype[a]{t}{t}) \\
-  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \subTm{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\sub{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\subTm{t}{\sigma}) \\
-  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
+  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a))(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma
+\end{alignat*}
+\begin{alignat*}{5}
+  \subTy{(\mathsf{Id}\;a\;t\;u)}{\sigma} \reduce \mathsf{Id}\,(\subTm{a}{\sigma})\,(\subTm{t}{\sigma})\,(\subTm{u}{\sigma})
+\end{alignat*}
+\begin{alignat*}{5}
+  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a)) \to \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,t) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\sub{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\sub{t}{\sigma}) \\
+  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,u) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
 \end{alignat*}
@@
   \blank,_{\Ctx}^M\blank    & : (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
-  \blank[\blank]_{\Ty}^M & : \Ty^M\;\Delta^M\;A \to \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Gamma^M\;(\subTy{A}{\sigma}) \\
+  \subTyM{\blank}{\blank} & : \Ty^M\;\Delta^M\;A \to \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Gamma^M\;(\subTy{A}{\sigma}) \\
   \UU^M                  & : \Ty^M\;\Gamma^M\;\UU \\
   \emptysub_{\Sub}^M     & : \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
-  \blank,_{\Sub}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [\sigma^M]^M_{\Ty})\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
+  \blank,_{\Sub}^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(\subTyM{A^M}{\sigma^M})\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
   \idS^M                 & : \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
@@
   \pi_1^M                & : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma \to \Sub^M\;\Gamma^M\;\Delta^M\;(\pi_1\sigma) \\
-  \pi_2^M                & : (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [ \pi_1^M\,\sigma^M ]^M_{\Ty})\;(\pi_2\,\sigma) \\
-  \blank[\blank]^M       & : \Tm^M\;\Delta^M\;A^M\;t \to (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [\sigma^M]_{\Ty}^M)\;(\sub{t}{\sigma} )
-\end{alignat*}
-
-\begin{alignat*}{5}
-  [\idS]_{\Ty}^M      & : A^M [ \idS^M ]^M_{\Ty}                             &&  =^{\Ty^M\,\Gamma^M}\; && A^M \\
-  [;]_{\Ty}^M         & : A^M [ \sigma^M ;^M \tau^M ]^M_{\Ty}                && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,(A^M [\tau^M]^M_{\Ty}) \\
-  [\pi_1,]^M_{\Ty}    & : A^M [ \pi_1^M\,(\sigma^M ,^M_{\Sub} t^M) ]^M_{\Ty} && =^{\Ty^M\,\Gamma^M} \; && A^M [\sigma^M]^M_{\Ty} \\
-  [\pi_1;]^M_{\Ty}    & : A^M [ \pi_1^M\,(\sigma^M ;^M \tau^M) ]^M_{\Ty}     && =^{\Ty^M\,\Gamma^M} \; && \left(A^M [ \pi_1^M\,\tau^M]^M_{\Ty}\right) [\sigma^M]^M_{\Ty} \\
-  []\UU^M             & : \UU^M [ \sigma^M ]^M_{\Ty}                           &&  =^{\Ty^M\,\Gamma^M\,\UU}      {} && \UU^M
+  \pi_2^M                & : (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;(\subTyM{A^M}{\pi_1^M\,\sigma^M})\;(\pi_2\,\sigma) \\
+  \subM{\blank}{\blank}  & : \Tm^M\;\Delta^M\;A^M\;t \to (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(\subTyM{A^M}{\sigma^M})\;(\sub{t}{\sigma} )
+\end{alignat*}
+
+\begin{alignat*}{5}
+  [\idS]_{\Ty}^M      & : \subTyM{A^M}{\idS^M}                             && = A^M \\
+  [;]_{\Ty}^M         & : \subTyM{A^M}{\sigma^M ;^M \tau^M}                && = \subTyM{(\subTyM{A^M}{\tau^M})}{\sigma^M} \\
+  [\pi_1,]^M_{\Ty}    & : \subTyM{A^M}{\pi_1^M\,(\sigma^M ,^M_{\Sub} t^M)} && = \subTyM{A^M}{\sigma^M} \\
+  [\pi_1;]^M_{\Ty}    & : \subTyM{A^M}{\pi_1^M\,(\sigma^M ;^M \tau^M)}     && = \left(\subTyM{A^M}{\pi_1^M\,\tau^M}\right) [\sigma^M]^M_{\Ty} \\
+  []\UU^M             & : \subTyM{\UU^M}{\sigma^M}                         && = \UU^M
 \end{alignat*}
@@
   ;\text{-}\textsf{assoc}^M & : (\sigma^M ;^M \tau^M) ;^M \gamma^M   && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{assoc}}} && \sigma^M ;^M (\tau^M ;^M \gamma^M) \\
-  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black}t^M [\sigma^M]_{\Ty}^M}} \\
+  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black} \subM{t^M}{\sigma^M}}} \\
   \pi_1\beta^M              & : \pi_1^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\pi_1\beta}}      && \sigma^M \\
@@
 \begin{alignat*}{10}
-  \blank[\blank]_{\Tm}^M & : \Tm^M\;\Delta^M\;A^M\;t \to (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(A^M [\sigma^M]_{\Ty}^M)\,(\subTm{t}{\sigma}) \\
+  \blank[\blank]_{\Tm}^M & : \Tm^M\;\Delta^M\;A^M\;t \to (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;(\subTyM{A^M}{\sigma^M})\,(\subTm{t}{\sigma}) \\
   \El^M                  & : \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u)
@@
 \begin{alignat*}{5}
-  [\idS]_{\Tm}^M      & : t^M [\idS^M]^M_{\Tm}                               && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M \\
-  [;]^M_{\Tm}         & : t^M [ \sigma^M ;^M \tau^M ]^M_{\Tm}                && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[;]^M_{\Ty}}} && \left(t^M [ \tau^M ]^M_{\Tm}\right) [ \sigma^M ]^M \\
-  [\pi_1,]^M_{\Tm}    & : t^M [\pi_1^M (\sigma^M ,_{\Sub}^M u^M) ]^M_{\Tm}   && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && t^M [\sigma^M]_{\Tm}^M \\
-  [\pi_1;]^M_{\Tm}    & : t^M [\pi_1^M (\sigma^M ;^M \tau^M) ]^M_{\Tm}       && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1;]^M_{\Ty}}} && \left(t^M [\pi_1^M\,\tau^M]^M_{\Tm}\right) [\sigma^M]_{\Tm}^M \\
-  [\emptysub]^M_{\Tm} & : t^M [\emptysub^M_{\Sub}]^M_{\Tm}                   && = %^{\Tm^M\,\Gamma^M\;([\emptysub_{\Sub}^M]^M_{\Ty}\,A^M)\,([\emptysub]_{\Tm}\,t)}
-                      t^M [ \emptysub^M_{\Sub} ]^M  &&  \\
-  [,]_{\Tm}^M         & : t^M [ \sigma^M ,_{\Sub}^M u^M ]^M_{\Tm}            && = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)}
-                      t^M [ \sigma^M ,_{\Sub}^M u^M ]^M && \\
-  [\pi_1\idS]^M_{\Tm} & : t^M [ \pi_1^M\,\idS^M ]^M_{\Tm}                    && = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{}
-                      t^M [ \pi_1^M\,\idS^M ]^M && \\
-  [\pi_1\pi_1]^M_{\Tm}& : t^M [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M_{\Tm}       && = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)}
-                      t^M [ \pi_1^M\,(\pi_1^M\,\sigma^M) ]^M &&
-\end{alignat*}
-
-\begin{alignat*}{5}
-  []\El^M & : [ \sigma^M ]^M_{\Ty}\;(\El^M\;u^M)   && =^{\Ty^M\,\Gamma^M\,(\El\,u)} {} && \El^M\,(\alert{\transfib{}{[]\UU^M}{\color{black}u^M [ \sigma^M ]^M_{\Tm}}}
+  [\idS]_{\Tm}^M      & : \subTmM{t^M}{\idS^M}                             && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\idS]^M_{\Ty}}}{} &&  t^M \\
+  [;]^M_{\Tm}         & : \subTmM{t^M}{\sigma^M ;^M \tau^M}                && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[;]^M_{\Ty}}} {} && \subTmM{\left(\subTmM{t^M}{\tau^M}\right)}{\sigma^M}  \\
+  [\pi_1,]^M_{\Tm}    & : \subTmM{t^M}{\pi_1^M (\sigma^M ,_{\Sub}^M u^M)}   && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1,]^M_{\Ty}}} && \subTmM{t^M}{\sigma^M} \\
+  [\pi_1;]^M_{\Tm}    & : \subTmM{t^M}{\pi_1^M (\sigma^M ;^M \tau^M)}      && =^{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,t}_{\alert{[\pi_1;]^M_{\Ty}}} && \subTmM{\left(\subTmM{t^M}{\pi_1^M\,\tau^M}\right)}{\sigma^M} \\
+  [\emptysub]^M_{\Tm} & : \subTmM{t^M}{\emptysub^M_{\Sub}}                 && = %^{\Tm^M\,\Gamma^M\;([\emptysub_{\Sub}^M]^M_{\Ty}\,A^M)\,([\emptysub]_{\Tm}\,t)}
+    \subTmM{t^M}{\emptysub^M_{\Sub}}  &&  \\
+    [,]_{\Tm}^M         & : \subTmM{t^M}{\sigma^M ,_{\Sub}^M u^M}          && = %^{\Tm^M\,\Gamma^M\;([\sigma^M ,_{\Sub}^M t ]^M_{\Ty}\,A^M)\,([\sigma , t]_{\Tm}\,t)}
+    \subTmM{t^M}{\sigma^M ,_{\Sub}^M u^M} && \\
+    [\pi_1\idS]^M_{\Tm} & : \subTmM{t^M}{\pi_1^M\,\idS^M}                    && = %^{\Tm^M\,(\Gamma^M ,^M_{\Ctx} A^M)\,([\pi_1^M\,\idS^M ]^M_{\Ty})\,A^M)\;([\pi_1\idS]_{\Tm}\,t)}{}
+    \subTmM{t^M}{\pi_1^M\,\idS^M} && \\
+                      [\pi_1\pi_1]^M_{\Tm}& : \subTmM{t^M}{\pi_1^M\,(\pi_1^M\,\sigma^M)}       && = %^{\Tm^M\,\Gamma^M\,([\pi_1^M\,(\pi_1^1\,\sigma^M)]^M_{\Ty}\,A)\;([\pi_1(\pi_1\,\sigma)]_{\Tm}\,t)}
+                      \subTmM{t^M}{\pi_1^M\,(\pi_1^M\,\sigma^M)} &&
+\end{alignat*}
+
+\begin{alignat*}{5}
+  []\El^M & : \subTyM{(\El^M\;u^M)}{\sigma^M} && =^{\Ty^M\,\Gamma^M\,(\El\,u)} {} && \El^M\,(\alert{\transfib{}{[]\UU^M}{\color{black}\subTmM{u^M}{\sigma^M}}}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M (A^M [ \sigma^M ]^M))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
+  (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M (\subTyM{A^M}{\sigma^M}))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
   (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;\pi_2\idS}{([;]^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
@@
 \begin{alignat*}{10}
-  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, A^M [\sigma^M]_{\Ty}^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
+  \blank\uparrow^M\blank & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M,_{\Ctx}^M (\subTyM{A^M}{\sigma^M}))\;(\Delta^M,_{\Ctx}^M A^M)\;(\sigma \uparrow A) \\
   \Pi^M                  & : (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
@@
   \mathord{;\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} \\
-                                &&& (\sigma^M \uparrow^M A^M [ \tau^M ]^M_{\Ty} ) ;^M (\tau^M \uparrow^M A^M)  \\
+                                &&& (\sigma^M \uparrow^M \subTyM{A^M}{\tau^M}) ;^M (\tau^M \uparrow^M A^M)  \\
   \mathord{\pi_1,\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
   \mathord{\pi_1;\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} \\
-                                &&& (\sigma^M \uparrow^M  A^M [\pi_1^M\tau^M]^M_{\Ty});^M (\pi_1^M\tau^M \uparrow^M A^M) \\
+                                &&& (\sigma^M \uparrow^M  \subTyM{A^M}{\pi_1^M\tau^M})\,;^M\, (\pi_1^M\tau^M \uparrow^M A^M) \\
   \mathord{\emptysub^M_{\Sub}\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
@@
 \begin{alignat*}{5}
-  []\Pi^M         & : (\Pi^M\;A^M\;B^M) [ \sigma^M ]^M_{\Ty}  && =^{\Ty^M\,\Gamma^M\,(\subTy{(\Pi\,A\,B)}{\sigma})} && \Pi^M\;(A^M [ \sigma^M ]^M_{\Ty})\;(B^M [ \sigma^M ]^M_{\Ty}) \\
-  []\lambda^M     & : (\lambda^M\;t^M) [ \sigma^M ]^M          && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} && \lambda^M\;(t^M [ \sigma^M \uparrow^M A^M ]^M) \\
+  []\Pi^M         & : \subTyM{(\Pi^M\;A^M\;B^M)}{\sigma^M}  && =^{\Ty^M\,\Gamma^M\,\left(\subTy{(\Pi\,A\,B)}{\sigma}\right)} {} && \Pi^M\;(\subTyM{A^M}{\sigma^M})\;(\subTyM{B^M}{\sigma^M}) \\
+  []\lambda^M     & : \subM{(\lambda^M\;t^M)}{\sigma^M}     && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} && \lambda^M\;(\subM{t^M}{\sigma^M \uparrow^M A^M}) \\
   \Pi\beta^M      & : \mathsf{app}^M\;(\lambda^M\;t^M)      && =^{\Tm^M\,(\Gamma^M,_{\Ctx}^M A^M)\,B^M}_{\alert{\Pi\beta}}      && t^M  \\
@@
 \begin{alignat*}{5}
-  []\mathsf{Id}^M    :{} & (\mathsf{Id}^M\;a^M\;t^M\;u^M) [ \sigma^M ]^M_{\Ty} =^{\Ty^M\,\Gamma^M\,(\subTy{(\mathsf{Id}\,a\,t\,u)}{\sigma})} \\
-                      & \mathsf{Id}^M\; (\transfib{}{[]\UU^M}{a^M [ \sigma^M ]^M_{\Tm}})\;
-                  (\transfib{}{[]\El^M}{t^M [ \sigma^M ]^M_{\Tm}})\;
-                  (\transfib{}{[]\El^M}{u^M [ \sigma^M ]^M_{\Tm}}) \\
-  []\mathsf{refl}    :{} & (\mathsf{refl}^M\;t^M) [ \sigma^M ]^M =^{\Tm^M\;\Gamma^M}_{\alert{[]\mathsf{Id}^M, []\mathsf{refl}}} \\
-                      & \mathsf{refl}^M\;(\transfib{\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,\nameless}{[]\El^M}{t^M [\sigma^M]^M_{\Tm}}) \\
+  []\mathsf{Id}^M    :{} & \subTyM{(\mathsf{Id}^M\;a^M\;t^M\;u^M)}{\sigma^M} =^{\Ty^M\,\Gamma^M\,(\subTy{(\mathsf{Id}\,a\,t\,u)}{\sigma})} \\
+                         & \mathsf{Id}^M\; (\transfib{}{[]\UU^M}{\subTmM{a^M}{\sigma^M}})\;
+                         (\transfib{}{[]\El^M}{\subTmM{t^M}{\sigma^M}})\;
+                         (\transfib{}{[]\El^M}{\subTmM{u^M}{\sigma^M}}) \\
+  []\mathsf{refl}    :{} & \subTm{(\mathsf{refl}^M\;t^M)}{\sigma^M} =^{\Tm^M\;\Gamma^M}_{\alert{[]\mathsf{Id}^M, []\mathsf{refl}}} {} \mathsf{refl}^M\;(\transfib{%\lam{A^M}\Tm^M\,\Gamma^M\,A^M\,\nameless
+                         }{[]\El^M}{\subTmM{t^M}{\sigma^M]^M}}) \\
   \mathsf{reflect}^M :{} & \Tm^M\;\Gamma^M\;(\mathsf{Id}^M\;a^M\;t^M\;u^M)\;p \to t^M =^{\Tm^M\;\Gamma^M\;A^M}_{\alert{\mathsf{reflect}\;p}} u^M

### 2025-02-20 tex/main.tex
Fix typos and minor revisions

@@
 
-\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}./}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
 %\supplementdetails[linktext={opt.
@@
 
-In this paper, we show that the development can be streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s).
+In this paper, we show that the development can be streamlined by combining quotient inductive-inductive types and inductive-recursive types, called \emph{quotient inductive-inductive-recursive types} (QIIRTs).
 %
@@
   \item We show how elimination principles for reasoning about the type theory can be derived, and use it to construct its standard model (\cref{sec:models}).
-  \item We compare our definition to definitions using quotient inductive-inductive types, and prove that the two definitions are actually equivalent (\cref{thm:equivalence}).
+  \item We compare our definition to definitions using quotient inductive-inductive types, and prove that the two definitions are actually equivalent (\cref{thm:equivalence}) in type theory.
 \end{itemize}
@@
 \subsection{Inductive-recursive types}
-An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined using an elimination rule which says that $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$.
+An inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined using an elimination rule which says that $f$ is defined by its values $f(\cc_i\,x)$ on all of the constructors $\cc_i$ of~$A$.
 %For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
@@
 \[
-  \pi : (a : U) \to (T(a) \to U) \to U
-  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T(b\, x)
+  \pi : (a : U) \to (T\,a \to U) \to U
+  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T\,(b\, x)
 \]
@@
 
-An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, but the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
+An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P\,x$ by giving values $f(\cc_i\,\overrightarrow{x})$ on all of the constructors $\cc_i$ of $A$, but the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
 %
@@
 %
-For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT in \cref{fig:QIT-join-list}.
-\begin{figure}[b]
-\begin{minipage}{.35\textwidth}
+For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT below.
+%\begin{figure}[b]
+\begin{minipage}{.4\textwidth}
 \begin{alignat*}{3}
@@
 \end{minipage}
-\begin{minipage}{.64\textwidth}
+\begin{minipage}{.58\textwidth}
 \begin{alignat*}{3}
@@
 \end{minipage}
-\caption{The quotient inductive type of join lists}
-\label{fig:QIT-join-list}
-\end{figure}
-%\vskip\belowdisplayskip\noindent
+%\caption{The quotient inductive type of join lists}
+%\label{fig:QIT-join-list}
+%\end{figure}
+\vskip\belowdisplayskip\noindent
 A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
@@
 
-We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list.
-We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$ from \cref{fig:QIT-join-list}, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
-\[
-  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'
-\]
-which states that join lists are equal as soon as their flattenings are equal.
+We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list and replaces equality constructors $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
+\[
+  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'.
+\]
+This constructor states that join lists are equal as soon as their flattenings are equal.
 For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
@@
 then $f$ is confluent --- essentially a variant of Newman's Lemma~\cite{Newman1942}.
-We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.
+We call a diagram $\sigma t_2 \Leftarrow f\,(\sigma\overrightarrow{p_1}) \Rightarrow \sigma t_1$ used in the analysis of local confluence a \emph{local peak}.
 
@@
   \]
-  apparently holds and can checked by a back-of-the-envelope calculation
+  apparently holds and can be checked by a back-of-the-envelope calculation
   \begin{equation} \label{eq:pi2-comp-proof}
@@
   \end{equation}
-  Yet, the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
-  Hence the above identity does not even make sense, since their types do not match.
+  Yet, the above identity does not even make sense, since their types do not match --- the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
   Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} \sub{(\pi_2\,\tau)}{\sigma}$ for some $p: \sub{A}{\pi_1(\sigma; \tau)} = \sub{\sub{A}{\pi_1\tau}}{\sigma}$ or, equivalently
@@
   as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
-  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.
+  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right-hand side.
 
@@
 %Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
-Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define $[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
+Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define $\subTy{\blank}{\blank} : \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma$ recursively as follows:
 \begin{alignat}{3}
@@
 \begin{alignat*}{5}
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) {} && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)          {} && =^{\Tm\,\Gamma\, A} &&  t \\
-  [\idS]t         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A}     {} && t \\
-  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}  && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,\sub{A}{\sigma ; \tau}} {} && {\sub{\sub{t}{\tau}}{\sigma}}
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) {} && =^{\Sub\,\Gamma\,(\Theta, A)} {} &&  (\sigma ; \tau), \sub{t}{\sigma} \\
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)          {} && =^{\Tm\,\Gamma\, (\subTy{A}{\sigma})} {} &&  t \\
+  [\idS]         & : \implicit{\Gamma, A, t} && {\sub{t}{\idS}}          && =^{\Tm\,\Gamma\,A}      {} && t \\
+  [;]            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}  && {\sub{t}{\sigma ; \tau}} && =^{\Tm\,\Gamma\,(\subTy{A}{\sigma ; \tau})} {} && {\sub{\sub{t}{\tau}}{\sigma}}
 \end{alignat*}
@@
   \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
-  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
+  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Ty\;\Gamma} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
 \end{alignat*}
@@
 \subTm{t}{\idS}          & \reduce t \\
-\subTm{t}{\sigma ; \tau} & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
+\subTm{t}{\sigma ; \tau} & \reduce \subTm{\subTm{t}{\tau}}{\sigma} \\
 \subTm{t}{\pi_1(\sigma, s)} & \reduce \subTm{t}{\sigma} \\
-\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
+\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{\subTm{t}{\pi_1\tau}}{\sigma} \\
 \subTm{t}{\sigma}        & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
@@
 \end{alignat}
-where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.
+where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_{\Ty}}{\color{black}\pi_2\idS}})$.
 
 For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
-we do not need any transport since $[;]_T$ and $[]\Pi$ are strict.
+we do not need any transport since $[;]_{\Ty}$ and $[]\Pi$ are strict.
 However, once again, adding \eqref{eq:def-type-subst-8} directly to the type substitution $\subTy{\blank}{\blank}$ breaks local confluence, because the reduction gets stuck upon the lifting $\sigma \uparrow A$.
@@
   For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$ and  any type $A$ and term $t$,   \danger we have
-  $\sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A$
+  $\sigma \uparrow A =^{\Sub\;(\Gamma, \subTy{A}{\blank})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A$
       and
-      $\subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau}$.
+      $\subTm{t}{\sigma} =^{\Tm\;\Gamma\;(\subTy{A}{\blank})}_{\alert{\cc}} \subTm{t}{\tau}$.
     \end{restatable}
@@
 
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.6\textwidth}
 \begin{alignat*}{5}
@@
 \end{minipage}
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.4\textwidth}
 \begin{alignat*}{5}
@@
 
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{.6\textwidth}
 \begin{alignat*}{5}
@@
 \end{minipage}
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{.4\textwidth}
 \begin{alignat*}{5}
@@
     \subTel{\Xi}{\sigma} & = \subTel{\Xi}{\tau}
-                         && \sigma \upuparrows \Xi =^{\lam{\gamma}\Sub\;(\Gamma\dplus \subTel{\Xi}{\gamma})\;(\Delta\dplus\Xi)}_{\alert{\mathsf{c}}} \tau \upuparrows \Xi \\
-    \subTy{A}{\sigma \upuparrows \Xi} & =^{\lam{\gamma}\Ty\;(\Gamma \dplus \subTel{\Xi}{\gamma})}_{\alert{\mathsf{c}}} \subTy{A}{\tau \upuparrows \Xi}.
+                         && \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\blank})\;(\Delta\dplus\Xi)}_{\alert{\cc}} \tau \upuparrows \Xi \\
+    \subTy{A}{\sigma \upuparrows \Xi} & =^{\Ty\;(\Gamma \dplus \subTel{\Xi}{\blank})}_{\alert{\cc}} \subTy{A}{\tau \upuparrows \Xi}.
   \end{align*}
@@
 $t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS , t)}$ of type $\Tm\,\Gamma\,(\sub{B}{(\idS , t)})$ can be defined without any transport.
-This is in contrast to the QIIT situation, where ordinary application is defined as $t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})}$.
+This is in contrast to the QIIT situation, where ordinary application is defined as $t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS, \alert{\transfib{\Tm\;\Gamma}{[\idS]_{\Tm}^{-1}}{\color{black}t}})}$.
 %
@@
 %
-\begin{alignat*}{5}
-  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
-  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
-\end{alignat*}
-%
-as constructors, and the structural rule for type substitution as another straightforward clause
+%\begin{alignat*}{5}
+%  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
+%  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
+%\end{alignat*}
+%
+as constructors, e.g.\ $ \bool\beta_{\bfalse} : \implicit{\Gamma, P, f, t}\; \belim\,P\,f\,t\,\bfalse = f$, and the structural rule for type substitution as another straightforward clause
 $\subTy{\bool}{\sigma} \reduce \bool$,
@@
   []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
-           & =^{\Tm\,\Gamma\,(\sub{P}{\sigma; \instSub{b}})}_{\alert{p(b)}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
+           & =^{\Tm\,\Gamma}_{\alert{p(b)}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
 \end{alignat*}
@@
 \end{alignat*}
-Each constructor $\mathsf{c}$ should be interpreted by a corresponding \emph{method} $\mathsf{c}^M$ indexed by it, e.g.,
+Each constructor $\cc$ should be interpreted by a corresponding \emph{method} $\cc^M$ indexed by it, e.g.,
 \begin{alignat*}{3}
@@
 \begin{alignat*}{5}
-  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && \subTyM{A^M}{\idS^M} && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
+  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && \subTyM{A^M}{\idS^M} && =^{\Ty^M\,\Gamma^M\,A}\;&& A^M %\\
   %[;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
@@
 \begin{remark}
-  To state the methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory --- for example, we have
-  \[
-    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\subTmM{t^M}{\idS^M} =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
+  To state the methods, we cannot avoid using transports in identities that would need transports in its QIIT definition of a type theory --- for example, we have
+  \[
+    [\idS]_{\Tm}^M : \implicit{\Gamma, \Gamma^M, t, t^M} \;\subTmM{t^M}{\idS^M} =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
   \]
@@
 years~\cite{Ghani2017,Kaposi2018,Kovacs2023}, and it seems
-promising to see if they can be extended to also handle quotient
-inductive-inductive-recursive types. Overlapping patterns are
+promising to see if they can be extended to also handle QIIRTs.
+Overlapping patterns are
 discussed by Cockx~\cite{Cockx2014}, but we would like to integrate it
@@
   \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
-  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \to \Sub\;\Gamma,(\Delta, A) \\
+  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \to \Sub\;\Gamma\,(\Delta, A) \\
   \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
@@
   & \subTy{A}{\idS}           && \reduce A \\
-  & \subTy{A}{\sigma ; \tau}  && \reduce \subTy{(\subTy{A}{\tau})}{\sigma} \\
+  & \subTy{A}{\sigma ; \tau}  && \reduce \subTy{\subTy{A}{\tau}}{\sigma} \\
   & \subTy{A}{\pi_1(\sigma, t)} && \reduce \subTy{A}{\sigma} \\
-  & \subTy{A}{\pi_1(\sigma; \tau)} && \reduce \subTy{(\subTy{A}{\pi_1\tau})}{\sigma} \\
+  & \subTy{A}{\pi_1(\sigma; \tau)} && \reduce \subTy{\subTy{A}{\pi_1\tau}}{\sigma} \\
   & \subTy{\UU}{\sigma} && \reduce \UU
@@
  \subTm{t}{\idS}                & \reduce t \\
- \subTm{t}{\sigma ; \tau}       & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
+ \subTm{t}{\sigma ; \tau}       & \reduce \subTm{\subTm{t}{\tau}}{\sigma} \\
  \subTm{t}{\pi_1(\sigma, s)}    & \reduce \subTm{t}{\sigma} \\
- \subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
+ \subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{\subTm{t}{\pi_1\tau}}{\sigma} \\
  \subTm{t}{\sigma}              & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
@@
 \pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
-\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwiese}
+\pi_1(\sigma; \tau) \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
+\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwise}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\;\Gamma\;(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
-  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\;\Gamma\;A} && t \\
-  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
+  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
+  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\,(\Gamma, A)\,B} && t \\
+  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t)
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a))(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma
+  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
 \end{alignat*}
@@
   \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a)) \to \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,t) \\
-  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\sub{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\sub{t}{\sigma}) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\sub{t}{\sigma}) \\
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,u) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
@@
 Note that implicitly quantified variables which would be otherwise declared inside $\implicit{\dots}$ are omitted below.
-Each variable $\mathsf{c}^M$ with the superscript $(\blank)^M$ is always indexed by $\mathsf{c}$ in addition to other indices.
+Each variable $\cc^M$ with the superscript $(\blank)^M$ is always indexed by $\cc$ in addition to other indices.
 
@@
   [\idS]_{\Ty}^M      & : \subTyM{A^M}{\idS^M}                             && = A^M \\
-  [;]_{\Ty}^M         & : \subTyM{A^M}{\sigma^M ;^M \tau^M}                && = \subTyM{(\subTyM{A^M}{\tau^M})}{\sigma^M} \\
+  [;]_{\Ty}^M         & : \subTyM{A^M}{\sigma^M ;^M \tau^M}                && = \subTyM{\subTyM{A^M}{\tau^M}}{\sigma^M} \\
   [\pi_1,]^M_{\Ty}    & : \subTyM{A^M}{\pi_1^M\,(\sigma^M ,^M_{\Sub} t^M)} && = \subTyM{A^M}{\sigma^M} \\
-  [\pi_1;]^M_{\Ty}    & : \subTyM{A^M}{\pi_1^M\,(\sigma^M ;^M \tau^M)}     && = \left(\subTyM{A^M}{\pi_1^M\,\tau^M}\right) [\sigma^M]^M_{\Ty} \\
+  [\pi_1;]^M_{\Ty}    & : \subTyM{A^M}{\pi_1^M\,(\sigma^M ;^M \tau^M)}     && = \subTyM{\subTyM{A^M}{\pi_1^M\,\tau^M}}{\sigma^M} \\
   []\UU^M             & : \subTyM{\UU^M}{\sigma^M}                         && = \UU^M
@@
   ;\text{-}\textsf{assoc}^M & : (\sigma^M ;^M \tau^M) ;^M \gamma^M   && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{;\text{-}\mathsf{assoc}}} && \sigma^M ;^M (\tau^M ;^M \gamma^M) \\
-  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M)^{-1}}{\color{black} \subM{t^M}{\sigma^M}}} \\
+  \mathsf{concat}^M         & : \sigma^M ;^M (\tau^M ,_{\Sub}^M t^M) && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\mathsf{concat}}} && (\sigma^M ;^M \tau^M) ,_{\Sub}^M \alert{\transfib{}{([;]^M_{\Ty})^{-1}}{\color{black} \subM{t^M}{\sigma^M}}} \\
   \pi_1\beta^M              & : \pi_1^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Sub^M\,\Gamma^M\,\Delta^M}_{\alert{\pi_1\beta}}      && \sigma^M \\
-  \pi_2\beta^M              & : \pi_2^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Tm^M\,\Gamma^M}_{\alert{[\pi_1,]^M, \pi_2\beta}}       && t^M \\
+  \pi_2\beta^M              & : \pi_2^M (\sigma^M ,^M_{\Sub} t^M)    && =^{\Tm^M\,\Gamma^M}_{\alert{[\pi_1,]^M_{\Ty}, \pi_2\beta}}       && t^M \\
   \cdot\eta^M               & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,\emptyctx^M_{\Ctx}}_{\alert{\cdot\eta}}  && \emptysub_{\Sub}^M \\
@@
   [\idS]^M                  & : t^M [ \idS ]^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
-  [;]^M                     & : t^M [ \sigma^M ;^M \tau^M ]^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && (t^M [\tau^M]^M) [\sigma^M]^M
+  [;]^M                     & : t^M [ \sigma^M ;^M \tau^M ]^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && \subM{\subM{t^M}{\tau^M}}{\sigma^M}
 \end{alignat*}
@@
   (\blank)^{+M} & : (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Sub^M\;(\Gamma^M ,_{\Ctx}^M (\subTyM{A^M}{\sigma^M}))\;(\Delta^M ,_{\Ctx}^M A^M)\;(\sigma^+) \\
-  (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;\pi_2\idS}{([;]^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
+  (\sigma^M)^{+M} & \defeq (\pi_1^M \idS^M ;^M \sigma^M) ,_{\Sub}^M \alert{\transfib{\Tm^M\;\Gamma^M\;\blank\;\pi_2\idS}{([;]_{\Ty}^M)^{-1}}{\color{black}\pi_2^M \idS^M}}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  []\Pi^M         & : \subTyM{(\Pi^M\;A^M\;B^M)}{\sigma^M}  && =^{\Ty^M\,\Gamma^M\,\left(\subTy{(\Pi\,A\,B)}{\sigma}\right)} {} && \Pi^M\;(\subTyM{A^M}{\sigma^M})\;(\subTyM{B^M}{\sigma^M}) \\
-  []\lambda^M     & : \subM{(\lambda^M\;t^M)}{\sigma^M}     && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} && \lambda^M\;(\subM{t^M}{\sigma^M \uparrow^M A^M}) \\
-  \Pi\beta^M      & : \mathsf{app}^M\;(\lambda^M\;t^M)      && =^{\Tm^M\,(\Gamma^M,_{\Ctx}^M A^M)\,B^M}_{\alert{\Pi\beta}}      && t^M  \\
-  \Pi\eta^M       & : \lambda^M\;(\mathsf{app}^M\;t^M)      && =^{\Tm^M\,\Gamma^M\,(\Pi\,A^M\,B^M)}_{\alert{\Pi\eta}}           && t^M 
+  []\Pi^M         & : \subTyM{(\Pi^M\;A^M\;B^M)}{\sigma^M}  && =^{\Ty^M\,\Gamma^M\,\left(\subTy{(\Pi\,A\,B)}{\sigma}\right)} {} && \Pi^M\;(\subTyM{A^M}{\sigma^M})\;(\subTyM{B^M}{\sigma^M \uparrow^M A^M}) \\
+  []\lambda^M     & : \subM{(\lambda^M\;t^M)}{\sigma^M}     && =^{\Tm^M\,\Gamma^M}_{\alert{[]\Pi^M, []\lambda}} {} && \lambda^M\;(\subM{t^M}{\sigma^M \uparrow^M A^M}) \\
+  \Pi\beta^M      & : \mathsf{app}^M\;(\lambda^M\;t^M)      && =^{\Tm^M\,(\Gamma^M,_{\Ctx}^M A^M)\,B^M}_{\alert{\Pi\beta}}   {} && t^M  \\
+  \Pi\eta^M       & : t^M                                   && =^{\Tm^M\,\Gamma^M\,(\Pi\,A^M\,B^M)}_{\alert{\Pi\eta}}        {} && \lambda^M\;(\mathsf{app}^M\;t^M)
 \end{alignat*}
@@
   \[
-    \El\,u, \qquad \El(\subTm{(\subTm{u}{\tau})}{\sigma}, \qquad \El(\subTm{u}{\sigma}), \qquad\text{and}\qquad
-    \El(\subTm{(\subTm{u}{\pi_1\tau})}{\sigma})
+    \El\,u, \qquad \El\,(\subTm{\subTm{u}{\tau}}{\sigma}), \qquad \El\,(\subTm{u}{\sigma}), \qquad\text{and}\qquad
+    \El\,(\subTm{\subTm{u}{\pi_1\tau}}{\sigma})
   \]
@@
 \begin{proof}
-  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
+  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily.
 \end{proof}
@@
           Then, $\UU_{\Gamma \dplus \subTel{\Xi}{\pi_1(\sigma, t)}} \simeq \UU_{\Gamma \dplus \subTel{\Xi}{\sigma}}$ holds by congruence for $\UU$.
-          \LT{Justify the use of congruence rule for QIIT, e.g. $\transfib{\Ty}{p}{\UU_{\Gamma}} = \UU_{\Delta}$ for $p : \Gamma = \Delta$}
 

### 2025-02-20 tex/main.tex
Some more appendix proof details

@@
 %
-\begin{lemma} \label{lem:lifting}
+\begin{restatable}{lemma}{lemIdentities}\label{lem:lifting}
   For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
-\end{lemma}
+\end{restatable}
 
@@
 As we are working with a different type theory from \cref{subsec:SC+U}, we  have to again prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$, but the same argument still applies:
-\begin{proposition} \label{prop:correctness-3}
+\begin{restatable}{proposition}{propCorrectnessThree}\label{prop:correctness-3}
   For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
-\end{proposition}
+\end{restatable}
 
@@
 \begin{proof}
+  Termination is once again clear, since all recursion is structural.
   In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms
@@
 \begin{proof}
-  By \cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
+  For term substitution, the claim follows by \cref{prop:correctness-1}: if
+  $\cc : \sigma = \tau$ then
+  \begin{align*}
+    \subTm{t}{\sigma} & = \sub{t}{\sigma} && \text{by \cref{prop:correctness-1}} \\
+                      & = \sub{t}{\tau} && \text{by congruence and $\cc$} \\
+                      & = \subTm{t}{\tau} && \text{by \cref{prop:correctness-1}}
+  \end{align*}
+  For type substitution, by \cref{re:coherence-proof}, we only have to consider the following cases
   \begin{align*}
@@
     \subTy{(\El\,u)}{\sigma; (\tau, t)} & \judgeq \El(\subTm{u}{\sigma; (\tau, t)}) && \text{by \eqref{eq:def-type-subst-7}} \\
-                                      & = \El( \sub{u}{\sigma; (\tau, t)})            && \text{by \cref{prop:correctness-1}} \\
-                                      & = \El( \sub{u}{\sigma; \tau, \sub{t}{\sigma}})     && \text{by $\mathsf{concat}$} \\
-                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by \cref{prop:correctness-1}} \\
+%                                      & = \El( \sub{u}{\sigma; (\tau, t)})            && \text{by \cref{prop:correctness-1}} \\
+%                                      & = \El( \sub{u}{\sigma; \tau, \sub{t}{\sigma}})     && \text{by $\mathsf{concat}$} \\
+%                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by \cref{prop:correctness-1}} \\
+                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by coherence of $\subTm{\blank}{\blank}$} \\
                                       & \judgeq \subTy{\left(\El\,u\right)}{\sigma; \tau, \sub{t}{\sigma}} && \text{by \eqref{eq:def-type-subst-7} }
   \end{align*}
-  The remaining two cases are omitted.
-
-  As for term substitution $\subTm{\blank}{\blank}$, each case follows from \cref{prop:correctness-1} in conjunction with the congruence rule for the identity type.
+  The other two cases are dealt with in exactly the same way.
 \end{proof}
@@
   Hence, the type substitution $\subTy{\blank}{\blank}$ is locally confluent.
+\end{proof}
+
+\lemIdentities*
+\begin{proof}
+  We have
+  \[
+    \idS^+ \judgeq ((\pi_1\idS; \idS), \pi_2 \idS) = (\pi_1\idS, \pi_2 \idS) = \idS
+  \]
+  For commuting with composition, first note that $\pi_1(\sigma; \tau) = \sigma; \pi_1 \tau$ by $\mathsf{concat}$. In particular, we have $\sigma; \pi_1 \idS = \sigma^+; \pi_1\idS$. Now
+  \begin{align*}
+    (\sigma; \tau)^+ &= (\pi_1\idS; (\sigma; \tau), \pi_2 \idS) \\
+                     &= ((\pi_1\idS; \sigma); \tau, \pi_2 \idS) \\
+                     &= ((\sigma^+; \pi_1\idS); \tau, \sub{\pi_2 \idS}{\sigma^+}) \\
+                     &= (\sigma^+; (\pi_1\idS; \tau), \pi_2 \idS) \\
+                     &= \sigma^+; \tau^+
+  \end{align*}
+  as required.
 \end{proof}
@@
 
+\propCorrectnessThree*
+\begin{proof}
+  Again by induction on $\sigma$.
+\end{proof}
+
 \propCoherenceUparrow*
 \begin{proof}
-  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily.
+  These two identities follow from \cref{prop:correctness-2,prop:correctness-3}.
 \end{proof}
@@
 \begin{proof}
-  \Fredrik{TODO: spell out all cases here}
   The proposition is easy to prove for the following equality
@@
 The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
-  \LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
+  %\LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
   For the following equality constructors $\cc : \sigma = \tau$
@@
 \begin{proof}
-  TODO
+  These properties are proven mutually with the definition of the
+  eliminators. The property $\elim_{\Ty}[]$ is proven by induction on the type $A$, while $\elim_{\Tm}[]$ and $\elim_{\Sub}[]$ are proven by induction on the substitution.
 \end{proof}
@@
 \begin{proof}
-  TODO
+  See our \Agda formalisation for the exact definition of the functions back and forth, and how the roundtrips are identities.
 \end{proof}

### 2025-02-21 tex/main.tex
Tidy up the use of brackets

@@
   \qquad\qquad
-  [;]_{\Ty}  : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\; \sub{A}{\sigma ; \tau} =^{\Ty\,\Gamma} \sub{(\sub{A}{\tau})}{\sigma}
+  [;]_{\Ty}  : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\; \sub{A}{\sigma ; \tau} =^{\Ty\,\Gamma} \sub{\sub{A}{\tau}}{\sigma}
   \\
@@
   hold propositionally.
-  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while both sides of the third reduce to $\subTy{(\subTy{(\subTy{A}{\gamma})}{\tau})}{\sigma}$ by \eqref{eq:def-type-subst-2}.
+  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while both sides of the third reduce to $\subTy{\subTy{\subTy{A}{\gamma}}{\tau}}{\sigma}$ by \eqref{eq:def-type-subst-2}.
   %
@@
   \emptytel          & : \implicit{\Gamma}\; && \Tel\;\Gamma \\
-  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
+  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) \to \Ty\;(\Gamma \dplus \Xi) \to \Tel\;\Gamma
 \end{alignat*}
@@
 \blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
-\to \Sub\;(\Gamma \dplus (\subTel{\Xi}{\sigma}))\;(\Delta \dplus \Xi)
+\to \Sub\;(\Gamma \dplus \subTel{\Xi}{\sigma})\;(\Delta \dplus \Xi)
 \]
@@
   \subTel{\Xi}{\idS}          & \reduce \Xi \\
-  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{(\subTel{\Xi}{\tau})}{\sigma}
+  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma}
 \end{alignat*}
@@
 Finally, we note that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also `ordinary' application
-$t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS , t)}$ of type $\Tm\,\Gamma\,(\sub{B}{(\idS , t)})$ can be defined without any transport.
-This is in contrast to the QIIT situation, where ordinary application is defined as $t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{(\idS, \alert{\transfib{\Tm\;\Gamma}{[\idS]_{\Tm}^{-1}}{\color{black}t}})}$.
+$t \mathop{\$} u \defeq \subTm{(\mathsf{app}\,t)}{\idS , t}$ of type $\Tm\,\Gamma\,(\sub{B}{\idS , t})$ can be defined without any transport.
+This is in contrast to the QIIT situation, where ordinary application is defined as $t \mathop{\$} u \defeq \sub{(\mathsf{app}\,t)}{\idS, \alert{\transfib{\Tm\;\Gamma}{[\idS]_{\Ty}^{-1}}{\color{black}t}}}$.
 %
@@
   \[
-    [\idS]_{\Tm}^M : \implicit{\Gamma, \Gamma^M, t, t^M} \;\subTmM{t^M}{\idS^M} =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
+    [\idS]_{\Tm}^M : \implicit{\Gamma, \Gamma^M, t, t^M} \;\subTmM{t^M}{\idS^M} =^{\Tm^M\,\Gamma^M\,\blank\,t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
   \]
@@
   \elim_{\Ty}[] & : \elim_{\Ty}\,(\subTy{A}{\sigma}) && = && \subTyM{(\elim_{\Ty}\,A)}{\elim_{\Sub}\,\sigma} \\
-  \elim_{\Tm}[] & : \elim_{\Tm}\,(\subTm{t}{\sigma}) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && \subTmM{(\elim_{\Tm}\,t)}{\elim_{\Sub}\,\sigma} \\
-  \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
+  \elim_{\Tm}[] & : \elim_{\Tm}\,(\subTm{t}{\sigma}) && =^{\Tm^M\,\Gamma^M\,\blank\,t}_{\alert{\elim_{\Ty}[]}} {} && \subTmM{(\elim_{\Tm}\,t)}{\elim_{\Sub}\,\sigma} \\
+  \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\Sub^M\,(\Gamma^M,^M \blank)\,(\Delta^M, A^M)\,(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} {} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
 \end{alignat*}
@@
   \pi\eta^M                 & : \sigma^M                             && =^{\Sub^M\,\Gamma^M\,(\Delta^M ,^M_{\Ctx} A^M)}_{\alert{\pi\eta}} {} && \pi_1^M \sigma^M ,_{\Sub}^M \pi_2^M \sigma^M \\
-  [\idS]^M                  & : t^M [ \idS ]^M                      && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
-  [;]^M                     & : t^M [ \sigma^M ;^M \tau^M ]^M       && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && \subM{\subM{t^M}{\tau^M}}{\sigma^M}
+  [\idS]^M                  & : \subM{t^M}{\idS}                     && =^{\Tm^M\,\Gamma^M}_{\alert{[\idS]^M_{\Ty}, [\idS]}} && t^M \\
+  [;]^M                     & : \subM{t^M}{\sigma^M ;^M \tau^M}      && =^{\Tm^M\,\Gamma^M}_{\alert{[;]^M_{\Ty}, [;]}} && \subM{\subM{t^M}{\tau^M}}{\sigma^M}
 \end{alignat*}

### 2025-02-21 tex/main.tex
Fix "overfull"s

@@
 \usepackage{manfnt}
+\usepackage{mathpartir}
 \newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
@@
 %
-For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT below.
+For example, the type $\JList{A}$ of \emph{join lists} can be given below
 %\begin{figure}[b]
-\begin{minipage}{.4\textwidth}
-\begin{alignat*}{3}
-  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
-  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
-  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
-\end{alignat*}
-\end{minipage}
-\begin{minipage}{.58\textwidth}
-\begin{alignat*}{3}
-  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
-  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
-  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
-\end{alignat*}
-\end{minipage}
+\begin{mathpar}
+  \Jnil : \implicit{A}\; \JList{A} \and
+  \Jsing{\blank} : \implicit{A}\; A \to \JList{A} \and
+  \Jconcat{\blank}{\blank} : \implicit{A}\; \JList{A} \to \JList{A} \to \JList{A} \and
+  \mathsf{idr} : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \and
+  \mathsf{idl} : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \and
+  \mathsf{assoc} : \implicit{\ell_0, \ell_1, \ell_2}\; \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
+\end{mathpar}
+%\begin{minipage}{.6\textwidth}
+%\begin{alignat*}{3}
+%  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
+%  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
+%  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
+%\end{alignat*}
+%\end{minipage}
+%\begin{minipage}{.4\textwidth}
+%\begin{alignat*}{3}
+%\end{alignat*}
+%\end{minipage}
 %\caption{The quotient inductive type of join lists}
@@
 %\end{figure}
-\vskip\belowdisplayskip\noindent
+%\vskip\belowdisplayskip\noindent
 A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
@@
 %; indeed, inductive-inductive definitions only make sense when $D = \Set$.
-The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
+The metatheory of QIITs has been extensively studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
 
@@
 \begin{align*}
-  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} {\sigma ; \idS_{\Delta}} =^{\Sub\,\Gamma\,\Delta} {\sigma}
+  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; {\sigma ; \idS_{\Delta}} =^{\Sub\,\Gamma\,\Delta} {\sigma}
                                                                                     \qquad\qquad
-  \mathsf{idl}    : \implicit{\Gamma, \Delta, \sigma} {\idS_{\Gamma} ; \sigma} =^{\Sub\,\Gamma\,\Delta} {\sigma} \\
+  \mathsf{idl}    : \implicit{\Gamma, \Delta, \sigma}\; {\idS_{\Gamma} ; \sigma} =^{\Sub\,\Gamma\,\Delta} {\sigma} \\
   \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; (\sigma ; \tau) ; \gamma =^{\Sub\,\Gamma\,\Theta} \sigma ; (\tau ; \gamma) \\
@@
 The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) \to \Tel\;\Gamma \to \Ctx$ as follows.
-
-\begin{minipage}{.6\textwidth}
-\begin{alignat*}{5}
-  %\Tel               & :                   && \Ctx \to \Set                                     \\
-  \emptytel          & : \implicit{\Gamma}\; && \Tel\;\Gamma \\
-  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) \to \Ty\;(\Gamma \dplus \Xi) \to \Tel\;\Gamma
-\end{alignat*}
-\end{minipage}
-\begin{minipage}{.4\textwidth}
-\begin{alignat*}{5}
-  %\blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
-  \Gamma & \dplus \emptytel && \reduce \Gamma \\
-  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
-\end{alignat*}
-\end{minipage}
-\vskip\belowdisplayskip\noindent
+\begin{alignat*}{9}
+  \emptytel          & : \implicit{\Gamma}\; && \Tel\;\Gamma \qquad 
+                     && \Gamma && \dplus \emptytel && \reduce \Gamma \\
+  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) \to \Ty\;(\Gamma \dplus \Xi) \to \Tel\;\Gamma\qquad
+                     && \Gamma && \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
+\end{alignat*}
 Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
@@
 which asks for telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$ defined mutually:
-
-\begin{minipage}{.6\textwidth}
-\begin{alignat*}{5}
-  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \\
-  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
-\end{alignat*}
-\end{minipage}
-\begin{minipage}{.4\textwidth}
-\begin{alignat*}{5}
-  & \sigma \upuparrows \emptytel && \reduce \sigma \\
-  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
-\end{alignat*}
-\end{minipage}
-\vskip\belowdisplayskip
+\begin{alignat*}{9}
+  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \qquad
+                             && \sigma \upuparrows \emptytel && \reduce \sigma \\
+  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi} \qquad\qquad
+                             && \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
+\end{alignat*}
+%\begin{minipage}{.6\textwidth}
+%\begin{alignat*}{5}
+%  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \\
+%  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
+%\end{alignat*}
+%\end{minipage}
+%\begin{minipage}{.4\textwidth}
+%\begin{alignat*}{5}
+%  & \sigma \upuparrows \emptytel && \reduce \sigma \\
+%  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
+%\end{alignat*}
+%\end{minipage}
+%\vskip\belowdisplayskip
 
@@
 Instead of resorting to \transp, we can also turn telescope substitution $\subTel{\blank}{\blank}$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
-
-\begin{minipage}{.5\textwidth}
-\begin{alignat*}{3}
-  \subTel{\Xi}{\idS}          & \reduce \Xi \\
-  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma}
-\end{alignat*}
-\end{minipage}
-\begin{minipage}{.5\textwidth}
-\begin{alignat*}{5}
-  & \idS          && \upuparrows \Xi && \reduce \idS \\
-  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
-\end{alignat*}
-\end{minipage}\vskip\belowdisplayskip\noindent
+\begin{alignat*}{9}
+  \subTel{\Xi}{\idS}          & \reduce \Xi\qquad\qquad
+                              && \idS          && \upuparrows \Xi && \reduce \idS \\
+  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma} \qquad\qquad
+                              && (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
+\end{alignat*}
+%\begin{minipage}{.5\textwidth}
+%\begin{alignat*}{3}
+%  \subTel{\Xi}{\idS}          & \reduce \Xi \\
+%  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma}
+%\end{alignat*}
+%\end{minipage}
+%\begin{minipage}{.5\textwidth}
+%\begin{alignat*}{5}
+%  & \idS          && \upuparrows \Xi && \reduce \idS \\
+%  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
+%\end{alignat*}
+%\end{minipage}\vskip\belowdisplayskip\noindent
 %whose confluence property is easy to see.
@@
   []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
-           & =^{\Tm\,\Gamma}_{\alert{p(b)}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
-\end{alignat*}
-where the proof $p(x) : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types.
+           & =^{\Tm\,\Gamma}_{\alert{p\,b}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\bfalse}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\btrue}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
+\end{alignat*}
+where the proof $p\,x : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types.
 Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems.

### 2025-02-21 tex/main.tex
Polish proofs until Prop 3.16 in Appendix

@@
   \begin{align*}
-    \subTy{(\El\,u)}{\sigma; (\tau, t)} & \judgeq \El(\subTm{u}{\sigma; (\tau, t)}) && \text{by \eqref{eq:def-type-subst-7}} \\
+    \subTy{(\El\,u)}{\sigma; (\tau, t)} & \judgeq \El\,(\subTm{u}{\sigma; (\tau, t)}) && \text{by \eqref{eq:def-type-subst-7}} \\
 %                                      & = \El( \sub{u}{\sigma; (\tau, t)})            && \text{by \cref{prop:correctness-1}} \\
@@
 %                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by \cref{prop:correctness-1}} \\
-                                      & = \El( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by coherence of $\subTm{\blank}{\blank}$} \\
+                                      & = \El\,( \subTm{u}{\sigma; \tau, \sub{t}{\sigma}}) && \text{by coherence of $\subTm{\blank}{\blank}$} \\
                                       & \judgeq \subTy{\left(\El\,u\right)}{\sigma; \tau, \sub{t}{\sigma}} && \text{by \eqref{eq:def-type-subst-7} }
@@
 \begin{proof}
-  We have
-  \[
-    \idS^+ \judgeq ((\pi_1\idS; \idS), \pi_2 \idS) = (\pi_1\idS, \pi_2 \idS) = \idS
-  \]
-  For commuting with composition, first note that $\pi_1(\sigma; \tau) = \sigma; \pi_1 \tau$ by $\mathsf{concat}$. In particular, we have $\sigma; \pi_1 \idS = \sigma^+; \pi_1\idS$. Now
-  \begin{align*}
-    (\sigma; \tau)^+ &= (\pi_1\idS; (\sigma; \tau), \pi_2 \idS) \\
-                     &= ((\pi_1\idS; \sigma); \tau, \pi_2 \idS) \\
-                     &= ((\sigma^+; \pi_1\idS); \tau, \sub{\pi_2 \idS}{\sigma^+}) \\
-                     &= (\sigma^+; (\pi_1\idS; \tau), \pi_2 \idS) \\
-                     &= \sigma^+; \tau^+
+  By the identity law $\mathsf{idr}$ and the $\eta$-law $\pi\eta$, we have
+  \[
+    \idS^+ \judgeq ((\pi_1\idS; \idS), \pi_2 \idS) = (\pi_1\idS, \pi_2 \idS) = \idS.
+  \]
+  For commuting with composition, first note that $\pi_1(\sigma; \tau) = \sigma; \pi_1 \tau$ holds similarly to \eqref{eq:pi2-comp-proof} and, in particular, we have $\pi_1 \idS ; \sigma = \sigma^+; \pi_1\idS$.
+  Also note that $\pi_2\idS = \sub{\pi_2\idS}{\sigma^+}$ by \eqref{eq:pi2-comp-proof}.
+  Now
+  \begin{align*}
+    (\sigma; \tau)^+ & \judgeq (\pi_1\idS; (\sigma; \tau)), \pi_2 \idS && \text{by definition} \\
+                     & = ((\pi_1\idS; \sigma); \tau), \pi_2 \idS       && \text{by ;-$\mathsf{assoc}$}\\
+                     & = ((\sigma^+; \pi_1\idS); \tau), \sub{\pi_2 \idS}{\sigma^+} && \text{by $\pi_1\idS; \sigma = \sigma^+ ; \pi_1\idS$ and $\pi_2\idS = \sub{\pi_2\idS}{\sigma^+}$}\\
+                     & = (\sigma^+; (\pi_1\idS; \tau)) , \sub{\pi_2\idS}{\sigma^+}  && \text{by ;-$\mathsf{assoc}$} \\
+                     & = \sigma^+; ((\pi_1\idS ; \tau) , \pi_2\idS) && \text{by $\mathsf{concat}^{-1}$} \\
+                     & \judgeq \sigma^+; \tau^+ && \text{by definition}
   \end{align*}

### 2025-02-21 tex/main.tex
Polish the proof of Prop 3.16

@@
     \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
-    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma),
+    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma).
   \]
   Take $\mathsf{idr}$ as an example, we can see 
-  \begin{align*}
-    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows \subTel{\Xi}{\idS}) ; (\idS \upuparrows \Xi) \\
-                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
-                                   & = \sigma \upuparrows \Xi
-  \end{align*}
-  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.
-
+  \[
+    (\sigma; \idS) \upuparrows \Xi \judgeq (\sigma \upuparrows \subTel{\Xi}{\idS}) ; (\idS \upuparrows \Xi) \judgeq (\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi
+  \]
+  where the last identity follows from $\mathsf{idr}$.
   For other two equality constructors, the arguments follow similarly.
 
-The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
-  %\LT[noinline]{Josh Ko suggests us to follow the proof structure in our formal implementation.}
-  For the following equality constructors $\cc : \sigma = \tau$
+  However, for the following equality constructors
   \begin{alignat*}{5}
@@
   \end{alignat*}
-   the identities
+  the coherence proofs are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually, i.e.\ 
   \begin{align*}
@@
   First, we prove the coherence for $\subTy{A}{\blank}$ by induction on $A$ and the coherence for the other two identities $\subTel{\Xi}{\blank}$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
+  For convenience, we use heterogeneous identity $t \simeq u$ (also known as John Major equality~\cite{McBride1999}) instead of dependent identity.
   \begin{enumerate}
-    \item Consider the following cases for $\subTy{A}{\sigma\upuparrows\Xi} = \subTy{A}{\tau \upuparrows \xi}$ by induction on $A$.
+    \item Consider the following cases for $\subTy{A}{\sigma\upuparrows\Xi} = \subTy{A}{\tau \upuparrows \Xi}$ by induction on $A$.
       \begin{description}
@@
             \judgeq{} & \Pi\,(\subTy{B}{\pi_1(\sigma, t) \upuparrows \Xi})\,(\subTy{C}{\pi_1(\sigma, t) \upuparrows \Xi \uparrow B}) && \text{by the definition of $\subTy{(\Pi\,B\,C)}{\blank}$} \\
-            \judgeq{} & \Pi\,(\subTy{B}{\pi_1(\sigma, t) \upuparrows \Xi})\,(\sub{C}{\pi_1(\sigma, t) \upuparrows (\Xi, B)}) && \text{by the definition of $\upuparrows$} \\
-            ={}       & \Pi\,(\subTy{B}{\sigma \upuparrows \Xi})\,(\sub{C}{\sigma \upuparrows (\Xi, B)}) && \text{by induction hypothesises} \\
-            \judgeq{} & \Pi\,(\subTy{B}{\sigma \upuparrows \Xi})\,(\sub{C}{(\sigma \upuparrows \Xi) \uparrow B}) && \text{by the definition of $\upuparrows$} \\
-            \judgeq{} & \subTy{(\Pi\,B\,C)}{\sigma\upuparrows \Xi} && \text{by the definition of $\subTy{(\Pi\,B\,C)}{\blank}$.}
+            \judgeq{} & \Pi\,(\subTy{B}{\pi_1(\sigma, t) \upuparrows \Xi})\,(\subTy{C}{\pi_1(\sigma, t) \upuparrows (\Xi, B)}) && \text{by the definition of $\upuparrows$} \\
+            ={}       & \Pi\,(\subTy{B}{\sigma \upuparrows \Xi})\,(\subTy{C}{\sigma \upuparrows (\Xi, B)}) && \text{by induction hypothesises} \\
+            \judgeq{} & \Pi\,(\subTy{B}{\sigma \upuparrows \Xi})\,(\subTy{C}{(\sigma \upuparrows \Xi) \uparrow B}) && \text{by definition} \\
+            \judgeq{} & \subTy{(\Pi\,B\,C)}{\sigma\upuparrows \Xi} && \text{by definition.}
           \end{align*}
@@
             \subTel{(\Xi, A)}{\pi_1(\sigma, t)} & \judgeq \subTel{\Xi}{\pi_1(\sigma, t)}, \subTy{A}{\pi_1(\sigma, t) \upuparrows \Xi} && \text{by definition} \\
-                                                & = \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi} \\
+                                                & = \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi} && \text{by induction hypothesis} \\
                                                 & \judgeq \subTel{(\Xi, A)}{\sigma} && \text{by definition}
@@
       \begin{description}
-        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
+        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor $\pi_1\beta$.
         \item[$\Xi$ is the telescope extension $\Xi, A$:] 
@@
             \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
-                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
-                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus\subTel{\Xi}{\gamma}, \subTy{A}{\gamma \upuparrows \Xi})\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
-                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
+                                                  & = (\pi_1(\sigma, t) \upuparrows \Xi)^{+} && \text{by \cref{prop:correctness-2}} \\
+                                                  & = (\sigma\upuparrows \Xi)^{+} && \text{by induction hypothesis} \\
+                                                  & = (\sigma \upuparrows \Xi) \uparrow A \text{by \cref{prop:correctness-2}}\\
                                                   & \judgeq \sigma \upuparrows (\Xi , A)

### 2025-02-21 tex/main.tex
Group constructors by sort instead of by feature in Appendix

@@
 
-\paragraph*{Quotient inductive-inductive-recursive types}
-We give the QIIRTs for type theories with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
-
-\subparagraph*{Substitution calculus with a base type $\UU$}
+\paragraph*{Type theory as a quotient inductive-inductive-recursive type}
+We give the QIIRT for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
+
 \begin{alignat*}{3}
@@
 \begin{alignat*}{5}
-  \emptyctx & : && \Ctx \\
-  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
+  \emptyctx     & : \Ctx \\
+  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
 \end{alignat*}
@@
 \subTy{\blank}{\blank} & : \implicit{\Gamma, \Delta}\; && \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \\
-  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
-\end{alignat*}
+  \UU         & : \implicit{\Gamma}\; && \Ty\; \Gamma \\
+  \El         & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma \\
+  \Pi         & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
+  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
+\end{alignat*}
+
 \begin{alignat*}{5}
@@
   \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
-  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
+  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A} && \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \\
   \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\subTy{A}{\pi_1\;\sigma}) \\
-  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A} && \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma})
-\end{alignat*}
-\begin{alignat*}{3}
-  & \subTy{A}{\idS}           && \reduce A \\
-  & \subTy{A}{\sigma ; \tau}  && \reduce \subTy{\subTy{A}{\tau}}{\sigma} \\
-  & \subTy{A}{\pi_1(\sigma, t)} && \reduce \subTy{A}{\sigma} \\
-  & \subTy{A}{\pi_1(\sigma; \tau)} && \reduce \subTy{\subTy{A}{\pi_1\tau}}{\sigma} \\
-  & \subTy{\UU}{\sigma} && \reduce \UU
-\end{alignat*}
-\begin{alignat*}{5}
-  ;\text{-}\mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
-  ;\text{-}\mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
-  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
-  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , \sub{t}{\sigma} \\
-  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma \\
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
-  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
-  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
-  [\idS]         & : \implicit{\Gamma, A, t}\; && {\sub{t}{\idS}}          && =^{\Tm\;\Gamma\;A} && t \\
-  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {\sub{t}{\sigma ; \tau}} && =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}}
-\end{alignat*}
-
-\subparagraph*{Universe of small types}
+  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
+  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B \\
+  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a)) \to \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,t)
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
+  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
+\idS                \uparrow A  & \reduce \idS \\
+\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\tau}) ; (\tau \uparrow A) \\
+\pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
+\pi_1(\sigma; \tau) \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
+\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwise}
+\end{alignat*}
+
 \begin{alignat*}{5}
@@
 \end{alignat*}
-%
-\begin{alignat*}{5}
-  \El       & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma
-\end{alignat*}
-%
+
 \begin{alignat*}{3}
-  \subTy{(\El\, u)}{\sigma} \reduce \El\,(\subTm{u}{\sigma})
-\end{alignat*}
-
-\subparagraph*{$\Pi$-types}
-\begin{alignat*}{5}
-  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
-  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
-\end{alignat*}
-\begin{alignat*}{5}
-  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
-\idS                \uparrow A  & \reduce \idS \\
-\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\tau}) ; (\tau \uparrow A) \\
-\pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
-\pi_1(\sigma; \tau) \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
-\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwise}
-\end{alignat*}
-\begin{alignat*}{5}
-  \Pi             & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
-  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
-  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
-\end{alignat*}
-\begin{alignat*}{3}
-  & \subTy{(\Pi\,A\,B)}{\sigma}       && {} \reduce {} && \Pi\,(\subTy{A}{\sigma})\,(\subTy{B}{\sigma\uparrow A})
-\end{alignat*}
-\begin{alignat*}{5}
+  \subTy{A}{\idS}                & \reduce A \\
+  \subTy{A}{\sigma ; \tau}       & \reduce \subTy{\subTy{A}{\tau}}{\sigma} \\
+  \subTy{A}{\pi_1(\sigma, t)}    & \reduce \subTy{A}{\sigma} \\
+  \subTy{A}{\pi_1(\sigma; \tau)} & \reduce \subTy{\subTy{A}{\pi_1\tau}}{\sigma} \\
+  \subTy{\UU}{\sigma}            & \reduce \UU \\
+  \subTy{(\El\, u)}{\sigma}      & \reduce \El\,(\subTm{u}{\sigma}) \\
+  \subTy{(\Pi\,A\,B)}{\sigma}    & \reduce \Pi\,(\subTy{A}{\sigma})\,(\subTy{B}{\sigma\uparrow A}) \\
+  \subTy{(\mathsf{Id}\;a\;t\;u)}{\sigma} & \reduce \mathsf{Id}\,(\subTm{a}{\sigma})\,(\subTm{t}{\sigma})\,(\subTm{u}{\sigma})
+\end{alignat*}
+
+\begin{alignat*}{5}
+  ;\text{-}\mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
+  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
+  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , \sub{t}{\sigma} \\
+  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
+  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
+  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma
+\end{alignat*}
+
+\begin{alignat*}{5}
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
+  [\idS]         & : \implicit{\Gamma, A, t}\; && {\sub{t}{\idS}}          && =^{\Tm\;\Gamma\;A} && t \\
+  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {\sub{t}{\sigma ; \tau}} && =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}} \\
   []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
   \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\,(\Gamma, A)\,B} && t \\
-  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t)
-\end{alignat*}
-
-\subparagraph*{Extensional identity types}
-
-\begin{alignat*}{5}
-  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
-\end{alignat*}
-\begin{alignat*}{5}
-  \subTy{(\mathsf{Id}\;a\;t\;u)}{\sigma} \reduce \mathsf{Id}\,(\subTm{a}{\sigma})\,(\subTm{t}{\sigma})\,(\subTm{u}{\sigma})
-\end{alignat*}
-\begin{alignat*}{5}
-  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a)) \to \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,t) \\
-  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\sub{t}{\sigma}) \\
+  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} && =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} {} && \mathsf{refl}(\sub{t}{\sigma})
+\end{alignat*}
+
+\begin{alignat*}{5}
   \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,u) \to t =^{\Tm\;\Gamma\;(\El\,a)} u

### 2025-02-21 tex/main.tex
Number the section on the elimination rule in appendix

@@
 When the motive families are constant, we call the principle a \emph{recursion principle} instead.
-We refer to \cref{appendix:full-def} and our \Agda development for the full definition of the elimination principle.
+We refer to \cref{subsec:def:elimination} for the full definition of the elimination principle.
 %We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$. 
@@
 
-\paragraph*{Type theory as a quotient inductive-inductive-recursive type}
+\subsection{Type theory as a quotient inductive-inductive-recursive type}
 We give the QIIRT for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.
@@
 \begin{alignat*}{5}
-  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
-  [\idS]         & : \implicit{\Gamma, A, t}\; && {\sub{t}{\idS}}          && =^{\Tm\;\Gamma\;A} && t \\
-  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {\sub{t}{\sigma ; \tau}} && =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}} \\
-  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
-  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\,(\Gamma, A)\,B} && t \\
-  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t) \\
-  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} && =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} {} && \mathsf{refl}(\sub{t}{\sigma})
+  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\;  & \pi_2(\sigma , t)                & =^{\Tm\;\Gamma\; A} &&  t \\
+  [\idS]         & : \implicit{\Gamma, A, t}\;                           & {\sub{t}{\idS}}                  & =^{\Tm\;\Gamma\;A} && t \\
+  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; & {\sub{t}{\sigma ; \tau}}         & =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{\sub{t}{\tau}}{\sigma}} \\
+  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;          & \sub{(\lambda\,t)}{\sigma}       & =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
+  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;                          & \sub{(\mathsf{refl}\;t)}{\sigma} & =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} {} && \mathsf{refl}\,(\sub{t}{\sigma}) \\
+  \Pi\beta        & : \implicit{\Gamma, A, B, t}\;                       & \mathsf{app}\,(\lambda\,t)       & =^{\Tm\,(\Gamma, A)\,B} && t \\
+  \Pi\eta         & : \implicit{\Gamma, A, B, t}\;                       & t                                & =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t)
 \end{alignat*}
@@
 
-\paragraph*{Elimination principles for type theories as QIIRTs}
-
-\subparagraph*{Motives}
+\subsection{The elimination principle for type theory as QIIRT} \label{subsec:def:elimination}
+
+\paragraph*{Motives}
 \begin{align*}
@@
 
-\subparagraph*{Methods}
+\paragraph*{Methods}
 Note that implicitly quantified variables which would be otherwise declared inside $\implicit{\dots}$ are omitted below.

### 2025-02-21 tex/main.tex
coquand universes

@@
 \subsection{The standard model} \label{subsec:std-model}
-The standard model, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
+The standard model of the type theory in \cref{subsec:SC+U+Pi}, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
 \begin{align*}

### 2025-02-21 tex/main.tex
cosmetic changes
final (?) pass Sections 1, 2

@@
 Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-\acknowledgements{We'd like to thank Hsiang-Shang Ko for useful discussion.}%optional
+\acknowledgements{We thank Hsiang-Shang Ko for useful discussions.}%optional
 
@@
 %
-As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
+As the host theory is assumed to satisfy uniqueness of identity proofs and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
 %
@@
 In contrast, an inductive-recursive type (IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F : A \to D$ to some fixed (possibly large) type $D$.
-The prototypical example is Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents.
+The prototypical example is a Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents.
 Crucially, the code
@@
   \pi : (a : U) \to (T\,a \to U) \to U
-  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T\,(b\, x)
+  \quad\text{with decoding}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T\,(b\, x)
 \]
@@
 %
-For example, the type $\JList{A}$ of \emph{join lists} can be given below
+For example, the type $\JList{A}$ of \emph{join lists} can be given as follows:
 %\begin{figure}[b]
@@
 %; indeed, inductive-inductive definitions only make sense when $D = \Set$.
-The metatheory of QIITs has been extensively studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
-
-We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list and replaces equality constructors $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
+The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi, and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.
+
+We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list, and we replace the equality constructors $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
 \[
@@
 %  \blank         & +'\blank     && : \N \to \N \to \N \\
-  0              & +' m         && \defeq m  \\
-  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
-  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
+  0              & +' m         && = m  \\
+  (\suc\;n)      & +' 0         && = \suc\;n \\
+  (\suc\;n)      & +' (\suc\;m) && = \suc\;(n +' \suc\;m)
 \end{alignat*}
@@
   \UU         & : \implicit{\Gamma}\; && \Ty\; \Gamma \\
-  \El         & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma \\
+  \El         & : \implicit{\Gamma}\; && \Tm\;\Gamma\;\UU \to \Ty\;\Gamma \\
   \Pi         & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\

### 2025-02-21 tex/main.tex
another pass through Sections 3--6

@@
 \]
-Note that the type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
-We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$.
+Note that the type substitution $\sub{A}{\sigma}$ is needed in the type of $\blank,\blank$, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
+We also ask for identity substitutions and composition $\sigma;\tau$.
 %because the substitution calculus is supposed to be the initial \emph{category} with families:
@@
 \subsection{Adding a universe} \label{subsec:SC+U}
-Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
+Now we extend the substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
 In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
@@
 %
-In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
+In our QIIRT definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
 Likewise, the equality constructor $[]\El$ may be replaced by the reduction
@@
 The lifting $\sigma^+$ interacts well with identities and composition.
-For comparision, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
+For comparison, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
 %
@@
 However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
-We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$ and its elimination principle:
+We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$, and its elimination principle:
 %

### 2025-02-21 tex/main.tex
spell check

@@
 Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
-Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
+Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasise its reduction behaviour.
 Under this semantics, each clause of $+$ is strictly equal to its right-hand side. %, also allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.
@@
   \item \emph{structurally recursive}~\cite{Abel2002}, i.e.\ arguments of recursive calls are structurally smaller, and
-  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, the terms $\sigma t_1$ and $\sigma t_2$ have the same normal form,
+  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifier $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, the terms $\sigma t_1$ and $\sigma t_2$ have the same normal form,
 \end{enumerate}
@@
 
-For our QIIR definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
+For our QIIRT definition, %$\sigma\uparrow A$ and $[]\lambda$ do not need a transport since $[;]_T$ and $[]\Pi$ are strict.
 we do not need any transport since $[;]_{\Ty}$ and $[]\Pi$ are strict.
@@
 
-To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows \subTel{\Xi}{\tau}); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus \subTel{\Xi}{\idS})\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
+To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows \subTel{\Xi}{\tau}); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus \subTel{\Xi}{\idS})\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the right hand side is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
 %
@@
 We are able to remove a lot of the boilerplate by using equality constructors instead.
-Chapman~\cite{Chapman2009} similarly uses explicit term substitutions and explicit transports along conversion rules, but also uses explicit type substitition in favour of more straightforward foundations.
+Chapman~\cite{Chapman2009} similarly uses explicit term substitutions and explicit transports along conversion rules, but also uses explicit type substitution in favour of more straightforward foundations.
 McBride~\cite{McBride2010} offers another solution which uses an inductive-recursive type to reuse the equality of the metatheory rather than using explicit transports.

### 2025-02-26 tex/main.tex
Remove the referral to function clauses

@@
       {≅}{$\cong$}1
+      {σ}{$\sigma$}1
 }
@@
 %The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns, we declare each function clause a rewrite, together with another rewrite to identify the witness as the reflexivity proof $\mathsf{refl}$.
+To emulate a function by overlapping patterns, we declare each function clause a rewriting rule and turn on the option \texttt{-{}-local-confluence-check}  to check local confluence.
 For example, the first few clauses for type substitution (written $[\sigma]_{\Ty}\,A$ rather than $\subTy{A}{\sigma}$ in the formalisation) are defined as
@@
   [id] : [ idS ] A ≡ A
-  {-# REWRITE [id]      #-}
-  [id]≡refl : [id] {A = A} ≡ refl
-  {-# REWRITE [id]=refl #-}
+  []U  : [ σ ]   U ≡ U
+  {-# REWRITE [id] []U #-}
 \end{lstlisting}
-where \lstinline|[id]≡refl| ensures that \lstinline|[id]| disappears in transports.
-We also turn on the option \texttt{-{}-local-confluence-check}  to check local confluence.
+where \lstinline|A| and $\sigma$ are declared as genearlised variables.
 %

### 2025-02-26 tex/main.tex
no message

--- 2025-02-21 tex/main.tex
@@
 %The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns, we declare each function clause a rewrite, together with another rewrite to identify the witness as the reflexivity proof $\mathsf{refl}$.
+To emulate a function by overlapping patterns, we implement each function clause as a rewriting rule, together with another rule to identify the clause as the reflexivity proof $\mathsf{refl}$ (which ensures that the named function clause is merely synonymous with \lstinline|refl|).
 For example, the first few clauses for type substitution (written $[\sigma]_{\Ty}\,A$ rather than $\subTy{A}{\sigma}$ in the formalisation) are defined as
@@
   {-# REWRITE [id]      #-}
-  [id]≡refl : [id] {A = A} ≡ refl
+  [id]=refl : [id] {A = A} ≡ refl
   {-# REWRITE [id]=refl #-}
 \end{lstlisting}
-where \lstinline|[id]≡refl| ensures that \lstinline|[id]| disappears in transports.
-We also turn on the option \texttt{-{}-local-confluence-check}  to check local confluence.
+We also turn on the option \texttt{-{}-local-confluence-check} to check local confluence globally.
 %

### 2025-03-07 tex/main.tex
mathematical typo
Minor wording change

@@
 \begin{alignat}{3}
-  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, \subTy{A}{\sigma})\, (\Delta, A) \label{eq:type-of-lifting} \\
+  \blank\uparrow \blank & : \implicit{\Gamma, \Delta}\; (\sigma : \Sub\,\Gamma\,\Delta) (A : \Ty\,\Delta) \to \Sub\,(\Gamma, \subTy{A}{\sigma})\, (\Delta, A) \label{eq:type-of-lifting} \\
 \idS                \uparrow A & \reduce \idS \nonumber \\
@@
 %The mechanism of defining a function by overlapping patterns is only implemented in an outdated experimental fork of \Agda.
-To emulate a function by overlapping patterns, we implement each function clause as a rewriting rule, together with another rule to identify the clause as the reflexivity proof $\mathsf{refl}$ (which ensures that the named function clause is merely synonymous with \lstinline|refl|).
+To emulate a function by overlapping patterns, we implement each function clause as a rewriting rule, together with another rule to identify the clause as the reflexivity proof $\mathsf{refl}$ (which ensures that the named function clause is merely a synonym for \lstinline|refl|).
 For example, the first few clauses for type substitution (written $[\sigma]_{\Ty}\,A$ rather than $\subTy{A}{\sigma}$ in the formalisation) are defined as
@@
 \end{lstlisting}
-We also turn on the option \texttt{-{}-local-confluence-check} to check local confluence globally.
+We also turn on the option \texttt{-{}-local-confluence-check} globally to check local confluence.
 %

### 2025-05-20 tex/FSCD25/rebuttal.txt
Aftermath of the FSCD rejection

We thank the reviewers for their detailed and constructive feedback.

Reviewer 1:

>Lack of focus on metaproperties about object type theory

The interpretation into the standard model is described in Section 4.2, albeit
very briefly due to a lack of space. But it demonstrates that our approach can
remove noise in definitions and proofs, with the details available in the Agda
formalisation, as stated in Section 4.2.

>Practical effectiveness of the approach

We didn't manage to remove all transports in the case of inductive types, as we
acknowledge in Section 3.5. However, we still improve on the number of
transports needed for types and introduction forms.  We concentrate on
eliminating transports along type substitutions, which introduce overhead early
in the formalisation. Addressing this specific source of transports hopefully
serves as a use case of QIIRTs and a step towards a solution to the transport
hell. Our preliminary substitution calculus NbE experiment that the reviewer
mentions indicate that the QIIRT-based definition reduces the code size by
approximately 20% (from 229 to 182 LoC), for this simple case.

>It seems like [your GitHub repository] contains more, such as work-in-progress
>proof of NbE. Is there a reason to exclude that from your future work?

The current experiments are small in terms of code size, and thus we didn't
think it was worthwhile to mention them in the paper. But we of course hope to
turn them into larger projects at a later stage. Since the QIIRT definition is
(syntactically) isomorphic to the QIIT one, and the eliminator only introduces
additional derived equations, we expect existing meta-theoretical results based
on QIITs to transfer with minimal changes.

Reviewer 2:

>Do you have any (even vague) idea of how your generalisation of IRT would fare
>with any of the justification for induction-recursion?

It seems plausible that Dybjer and Setzer's set-theoretic model ([20] in our
paper) could be adapted to justify this generalisation. If so, it should also
be straightforward to extend this model to also include equality constructors,
since quotients are easy in set theory.

>Can you explain how you derive your notion of eliminator?
>In particular, what (if any) is the difference between that eliminator and
>that for the standard QIIT?
>And do you have a justification/intuition for why you require that your model
>satisfies only propositionally the rules which are definitional in the QIIRT?

The eliminator for a QIIRT is derived following the general QIIT schema, where
each function clause in the recursive part of the QIIRT definition corresponds
to an equality constructor.

Since we implement some constructors as recursive functions rather than
constructors, the associated computation rules become propositional equalities
in the eliminator (Proposition 4.2). These propositional equalities cannot be
stated as definitional equalities, since there is no way to talk about
definitional equality within type theory.

For the same reason, models cannot be required to satisfy definitional equalities for
these clauses.

>Can you explain why you are not able to have a notion of term-level
>substitution defined solely as a recursive function?
>Why you chose to have composition and other operation not be themselves
>defined? And why substitution is defined by induction on the structure of
>substitution, rather than on that of terms?

We do not define term-level substitution recursively, as doing so does not seem
to reduce the number of transports needed, beyond the "smart constructor"
[–]Tm– (Section 3.3). Defining more operations recursively would increase the
number of proof obligations such as verifying confluence.

The smart constructor is defined by induction on the structure of substitution,
because the clauses based on the substitution structure are exactly those needed
to align with the definitional behaviour of type substitution.

>Can you comment on the relation with research on explicit substitution
>calculi? In particular, do you view any fundamental obstruction to/reason not
>to adapt the complete rewriting system of Schäfer, Smolka and Tebbi, which
>seems to solve your issue in section 3.5, to a typed setting?

The work of Schäfer et al. employs a normalisation function into normal
expressions (see Section 6 and Theorem 32, op. cit.). In our setting, this
would correspond to replacing equality constructors with a normalisation
function, similar to the QIRT definition of join lists in Section 2.2 and the
work by Bense et al. Adapting their approach to a type theory with universes
and inductive types seems useful but likely requires a definition that departs
far from the typical CwF-based definition of type theory.

Reviewer 3:

>In your formalisation there is less transport hell, but is there any kind of
>new hell? E.g. "implicit argument hell" [...].

Indeed, some implicit arguments must be given explicitly in the QIIRT version.
Compared to the work saved from dealing with transports explicitly, we believe
it is a good deal, since inserting implicit arguments does not complicate the
equational theory.

### 2025-08-22 tex/CPP26/main.tex
Paper abstract (treated as planned goals)

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,10pt,review]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.

\usepackage{xspace}
\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Formalising type theory in a cubical language}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Liang-Ting Chen}
\affiliation{%
  \institution{Institute of Information Science, Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
\email{liangtingchen@as.edu.tw}

\author{Fredrik Nordvall Forsberg}
\affiliation{%
  \institution{Department of Computer and Information Sciences, University of Strathclyde}
  \city{Glasgow}
  \country{United Kingdom}}

\author{Tzu-Chun Tsai}
\affiliation{%
  \institution{Institute of Information Science, Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
%\affiliation{%
%  \institution{Institute for Logic, Language and Computation, University of Amsterdam}
%  \city{Amsterdam}
%  \country{Netherlands}}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
This paper presents a formalisation of a core type theory and its metatheory within \CA.
The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.

The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
Moreover, type theory and its elimination are defined with dependent identities without transports.
This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
  Your, Paper}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

%\received{20 February 2007}
%\received[revised]{12 March 2009}
%\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
%\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

### 2025-09-02 tex/CPP26/main.tex
[ tex ] outline

@@
-%%
-%% This is file `sample-sigplan.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-sigplan.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[sigplan,10pt,review]{acmart}
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
+\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% Rights management information.  This information is sent to you
@@
 \setcopyright{acmlicensed}
-\copyrightyear{2018}
-\acmYear{2018}
+\copyrightyear{2025}
+\acmYear{2025}
 \acmDOI{XXXXXXX.XXXXXXX}
@@
 
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
+%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{xspace}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
+%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%%
-%% end of the preamble, start of the body of the document source.
-
-\usepackage{xspace}
-\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
 \begin{document}
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Formalising type theory in a cubical language}
+\title{Can Natural Models Simplify The Metatheory of Type Theory in Cubical Agda?}
 
@@
 \begin{abstract}
-This paper presents a formalisation of a core type theory and its metatheory within \CA.
-The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
-We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.
 
-The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
-Moreover, type theory and its elimination are defined with dependent identities without transports.
-This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
+  We formalise type theory in type theory, using Awodey's notion of \emph{natural model} of type theory.
+  The initial natural model can be represented using quotient inductive-inductive-recursive types in the proof assistant \CA.
+Using natural models leads to fewer instances of so-called ``transport hell'' in the syntax of type theory and its elimination rules.
+To show that the approach is feasible, we formalise some meta-properties, including the standard % and logical predicate
+interpretation, normalisation by evaluation for typed terms, and strictification constructions.
+Since our formalisation is carried out using \CA's built-in support for quotient types, all our constructions compute.
+However, the use of transports along equations often reappeared when developing more sophisticated metatheory.
+Hence it is still a considerable struggle to develop the metatheory of type theory using welltyped terms with current proof assistant technology, and the effort is about the same using natural models or not.
+
+%   This paper presents a formalisation of a core type theory and its metatheory within \CA.
+% The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
+% We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.
+
+% The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
+% Moreover, type theory and its elimination are defined with dependent identities without transports.
+% This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
 \end{abstract}
@@
 \maketitle
+\bibliographystyle{ACM-Reference-Format}
 
-%%
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
-\end{acks}
+\section{Introduction}
+% FNF (Fri 5 Sep)
 
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-%\bibliography{sample-base}
+Transport hell = transports appear inside terms that you actually want to study, hence you need to do a lot of reasoning about transport rather than the actual objects of interest.
 
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
+\section{Setting and metatheory}
+% FNF (Sun 7 Sep)
 
-\section{Research Methods}
+Cubical Agda with UIP
 
-\subsection{Part One}
+QIIRTs
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
-malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
-sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
-vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
-lacinia dolor. Integer ultricies commodo sem nec semper.
+\section{Type theory as a natural model}
+% LTC (Sun 7 Sep)
 
-\subsection{Part Two}
+\subsection{Natural models}
 
-Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
-ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
-ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
-eros. Vivamus non purus placerat, scelerisque diam eu, cursus
-ante. Etiam aliquam tortor auctor efficitur mattis.
+\subsection{The Syntax of Type Theory as a Quotient Inductive-Inductive-Recursive Type}
+
+QIIRT def of the syntax
+
+\subsection{Recursion and Elimination Principles}
+
+\subsection{Strictification}
+
+\section{Metatheory}
+% LTC (Tue 9 Sep)
+
+\subsection{Standard Model}
+
+% Argue for relative consistency?
+
+\subsection{Normalisation by Evaluation}
+
+% (for SC)
+
+\section{Comparison with Other Approaches}
+% FNF (Tue 9 Sep)
+
+Compared to QIIT:
+
+\begin{itemize}
+\item Fewer transports in the syntax, but they tend to come back in concrete models
+\item Strictification orthogonal
+\end{itemize}
+
+Compared to untyped version:
+\begin{itemize}
+\item Untyped version might still be easiest to work with, with current proof assistant technology
+\end{itemize}
+
+\section{Discussion}
+% LTC (at first)
+
+Practical considerations (eg NBE computes, support for interleaved mutual definitions)
+
+Support from proof assistant (eg safety guarantees)
+
+Easier in OTT-based proof assistant
 
 \end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.
+
+%\begin{acks}
+%To Robert, for the bagels and explaining CMYK and color spaces.
+%\end{acks}

### 2025-09-03 tex/CPP26/main.tex
Assign CCS

@@
 \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
 \setcopyright{acmlicensed}
@@
 \acmDOI{XXXXXXX.XXXXXXX}
-%% These commands are for a PROCEEDINGS abstract or paper.
 \acmConference[Conference acronym 'XX]{Make sure to enter the correct
@@
 %%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[utf8]{inputenc}
+\usepackage{newunicodechar}
 \usepackage{xspace}
+\usepackage{xcolor}
+\usepackage{mathtools}
+\let\Bbbk\relax
+\usepackage{amsmath,amssymb,mathbbol}
+\let\Bbbk\relax
+\usepackage{bbold}
+
+%% Remove the following if there are no todo items.
+\setlength {\marginparwidth }{2cm}
+\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
    <concept>
-  <concept_id>00000000.0000000.0000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
        <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
        </concept>
@@
 
-\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc[500]{Theory of computation~Type theory}
 
-%%
 %% Keywords. The author(s) should pick words that accurately describe
 %% the work being presented. Separate the keywords with commas.
-\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
-  Your, Paper}
+\keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}
 %% A "teaser" image appears between the author and affiliation

### 2025-12-03 tex/CPP26/main.tex
Three remaining issues to resolve (termination, strictification, and the notion of models)

@@
-\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,screen,natbib,pbalance]{acmart}
+\settopmatter{printfolios=true,printccs=false,printacmref=false}
 \setcopyright{acmlicensed}
@@
 \acmDOI{XXXXXXX.XXXXXXX}
-\acmConference[Conference acronym 'XX]{Make sure to enter the correct
-  conference title from your rights confirmation email}{June 03--05,
-  2018}{Woodstock, NY}
+\acmConference[CPP '26]{Certified Programs and Proofs}{January 12--13, 2026}{Rennes, France}
 %%
@@
 \usepackage[utf8]{inputenc}
+\usepackage[UKenglish]{babel}
 \usepackage{newunicodechar}
@@
 \usepackage{xcolor}
+%\usepackage{natbib}
+\setcitestyle{numbers,sort&compress}
 \usepackage{mathtools}
@@
 \usepackage{amsmath,amssymb,mathbbol}
-\let\Bbbk\relax
 \usepackage{bbold}
+\usepackage[inline]{enumitem}
 
@@
 \usepackage[obeyFinal,textsize=footnotesize]{todonotes}
+\usepackage[capitalise]{cleveref}
+\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
+\newcommand{\FNF}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
+
+\usepackage{microtype}
+\usepackage{newunicodechar}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
+\newcommand{\CA}{\textrm{Cubical Agda}\xspace}
+\newcommand{\Agda}{\textrm{Agda}\xspace}
+\newcommand{\Set}{\mathbf{Set}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% lhs2tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\cons}[1]{\mathbf{#1}}
+\newcommand{\iden}{\mathit}
+
+\newcommand{\awa}[2]{\mathrlap{#2}\phantom{#1}} % as wide as
+
+\definecolor{addition}{RGB}{204,255,216}
+\definecolor{keyword}{RGB}{204,255,216}
+\definecolor{identifier}{RGB}{204,255,216}
+\newcommand{\highlight}[2]{\smash{\text{\colorbox{#1}{\kern-.1em\vphantom{\vrule height 1.2ex depth 0.1ex}\smash{\ensuremath{#2}}\kern-.1em}}}}
+
+\let\Bbbk\relax
+%% ODER: format ==         = "\mathrel{==}"
+%% ODER: format /=         = "\neq "
+%
+%
+\makeatletter
+\@ifundefined{lhs2tex.lhs2tex.sty.read}%
+  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
+   \newcommand\SkipToFmtEnd{}%
+   \newcommand\EndFmtInput{}%
+   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
+  }\SkipToFmtEnd
+
+\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
+\usepackage{amstext}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\DeclareFontFamily{OT1}{cmtex}{}
+\DeclareFontShape{OT1}{cmtex}{m}{n}
+  {<5><6><7><8>cmtex8
+   <9>cmtex9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
+\DeclareFontShape{OT1}{cmtex}{m}{it}
+  {<-> ssub * cmtt/m/it}{}
+\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+  {<5><6><7><8>cmtt8
+   <9>cmbtt9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
+\DeclareFontShape{OT1}{cmtex}{bx}{n}
+  {<-> ssub * cmtt/bx/n}{}
+\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
+
+\newcommand{\Sp}{\hskip.33334em\relax}
+
+\newcommand{\Conid}[1]{\mathit{#1}}
+\newcommand{\Varid}[1]{\mathit{#1}}
+\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
+\newcommand{\plus}{\mathbin{+\!\!\!+}}
+\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
+\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
+\newcommand{\sequ}{\mathbin{>\!\!\!>}}
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\usepackage{polytable}
+
+%mathindent has to be defined
+\@ifundefined{mathindent}%
+  {\newdimen\mathindent\mathindent\leftmargini}%
+  {}%
+
+\def\resethooks{%
+  \global\let\SaveRestoreHook\empty
+  \global\let\ColumnHook\empty}
+\newcommand*{\savecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
+\newcommand*{\restorecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
+\newcommand*{\aligncolumn}[2]%
+  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
+
+\resethooks
+
+\newcommand{\onelinecommentchars}{\quad-{}- }
+\newcommand{\commentbeginchars}{\enskip\{-}
+\newcommand{\commentendchars}{-\}\enskip}
+
+\newcommand{\visiblecomments}{%
+  \let\onelinecomment=\onelinecommentchars
+  \let\commentbegin=\commentbeginchars
+  \let\commentend=\commentendchars}
+
+\newcommand{\invisiblecomments}{%
+  \let\onelinecomment=\empty
+  \let\commentbegin=\empty
+  \let\commentend=\empty}
+
+\visiblecomments
+
+\newlength{\blanklineskip}
+\setlength{\blanklineskip}{0.66084ex}
+
+\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
+\let\hspre\empty
+\let\hspost\empty
+\newcommand{\NB}{\textbf{NB}}
+\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
+
+\EndFmtInput
+\makeatother
+%
+%
+%
+%
+%
+%
+% This package provides two environments suitable to take the place
+% of hscode, called "plainhscode" and "arrayhscode". 
+%
+% The plain environment surrounds each code block by vertical space,
+% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
+% similar to formulas. Note that if these dimensions are changed,
+% the spacing around displayed math formulas changes as well.
+% All code is indented using \leftskip.
+%
+% Changed 19.08.2004 to reflect changes in colorcode. Should work with
+% CodeGroup.sty.
+%
+\ReadOnlyOnce{polycode.fmt}%
+\makeatletter
+
+\newcommand{\hsnewpar}[1]%
+  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
+
+% can be used, for instance, to redefine the code size, by setting the
+% command to \small or something alike
+\newcommand{\hscodestyle}{}
+
+% The command \sethscode can be used to switch the code formatting
+% behaviour by mapping the hscode environment in the subst directive
+% to a new LaTeX environment.
+
+\newcommand{\sethscode}[1]%
+  {\expandafter\let\expandafter\hscode\csname #1\endcsname
+   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
+
+% "compatibility" mode restores the non-polycode.fmt layout.
+
+\newenvironment{compathscode}%
+  {\par\noindent
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed\)%
+   \par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\compaths}{\sethscode{compathscode}}
+
+% "plain" mode is the proposed default.
+% It should now work with \centering.
+% This required some changes. The old version
+% is still available for reference as oldplainhscode.
+
+\newenvironment{plainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newenvironment{oldplainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+% Here, we make plainhscode the default environment.
+
+\newcommand{\plainhs}{\sethscode{plainhscode}}
+\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
+\plainhs
+
+% The arrayhscode is like plain, but makes use of polytable's
+% parray environment which disallows page breaks in code blocks.
+
+\newenvironment{arrayhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\parray}%
+  {\endparray\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newcommand{\arrayhs}{\sethscode{arrayhscode}}
+
+% The mathhscode environment also makes use of polytable's parray 
+% environment. It is supposed to be used only inside math mode 
+% (I used it to typeset the type rules in my thesis).
+
+\newenvironment{mathhscode}%
+  {\parray}{\endparray}
+
+\newcommand{\mathhs}{\sethscode{mathhscode}}
+
+% texths is similar to mathhs, but works in text mode.
+
+\newenvironment{texthscode}%
+  {\(\parray}{\endparray\)}
+
+\newcommand{\texths}{\sethscode{texthscode}}
+
+% The framed environment places code in a framed box.
+
+\def\codeframewidth{\arrayrulewidth}
+\RequirePackage{calc}
+
+\newenvironment{framedhscode}%
+  {\parskip=\abovedisplayskip\par\noindent
+   \hscodestyle
+   \arrayrulewidth=\codeframewidth
+   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
+   \hline\framedhslinecorrect\\{-1.5ex}%
+   \let\endoflinesave=\\
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \framedhslinecorrect\endoflinesave{.5ex}\hline
+   \endtabular
+   \parskip=\belowdisplayskip\par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\framedhslinecorrect}[2]%
+  {#1[#2]}
+
+\newcommand{\framedhs}{\sethscode{framedhscode}}
+
+% The inlinehscode environment is an experimental environment
+% that can be used to typeset displayed code inline.
+
+\newenvironment{inlinehscode}%
+  {\(\def\column##1##2{}%
+   \let\>\undefined\let\<\undefined\let\\\undefined
+   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
+   \def\fromto##1##2##3{##3}%
+   \def\nextline{}}{\) }%
+
+\newcommand{\inlinehs}{\sethscode{inlinehscode}}
+
+% The joincode environment is a separate environment that
+% can be used to surround and thereby connect multiple code
+% blocks.
+
+\newenvironment{joincode}%
+  {\let\orighscode=\hscode
+   \let\origendhscode=\endhscode
+   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
+   %\let\SaveRestoreHook=\empty
+   %\let\ColumnHook=\empty
+   %\let\resethooks=\empty
+   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
+  {\origendhscode
+   \global\let\hscode=\orighscode
+   \global\let\endhscode=\origendhscode}%
+
+\makeatother
+\EndFmtInput
+%
+%
+\ReadOnlyOnce{agda.fmt}%
+
+\RequirePackage[T1]{fontenc}
+\RequirePackage[utf8]{inputenc}
+\RequirePackage{amsfonts}
+\RequirePackage{newunicodechar}
+
+\providecommand\mathbbm{\mathbb}
+
+% TODO: Define more of these ...
+\DeclareUnicodeCharacter{737}{\textsuperscript{l}}
+\DeclareUnicodeCharacter{8718}{\ensuremath{\blacksquare}}
+\DeclareUnicodeCharacter{8759}{::}
+\DeclareUnicodeCharacter{9669}{\ensuremath{\triangleleft}}
+\DeclareUnicodeCharacter{8799}{\ensuremath{\stackrel{\scriptscriptstyle ?}{=}}}
+\DeclareUnicodeCharacter{10214}{\ensuremath{\llbracket}}
+\DeclareUnicodeCharacter{10215}{\ensuremath{\rrbracket}}
+\DeclareUnicodeCharacter{27E6}{\ensuremath{\llbracket}}
+\DeclareUnicodeCharacter{27E7}{\ensuremath{\rrbracket}}
+\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}
+
+\DeclareUnicodeCharacter{2294}{\ensuremath{\sqcup}}
+\DeclareUnicodeCharacter{2080}{\ensuremath{_0}}
+\DeclareUnicodeCharacter{2081}{\ensuremath{_1}}
+\DeclareUnicodeCharacter{2082}{\ensuremath{_2}}
+\DeclareUnicodeCharacter{2083}{\ensuremath{_3}}
+
+\DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}}
+\DeclareUnicodeCharacter{2236}{:}
+\DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
+\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
+\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}
+\DeclareUnicodeCharacter{039B}{\ensuremath{\Lambda}}
+\DeclareUnicodeCharacter{03C0}{\ensuremath{\pi}}
+\DeclareUnicodeCharacter{03C3}{\ensuremath{\sigma}}
+\DeclareUnicodeCharacter{03C9}{\ensuremath{\omega}}
+\newunicodechar{β}{\ensuremath{\beta}}
+\newunicodechar{η}{\ensuremath{\eta}}
+\newunicodechar{Γ}{\ensuremath{\Gamma}}
+\newunicodechar{Δ}{\ensuremath{\Delta}}
+\newunicodechar{Θ}{\ensuremath{\Theta}}
+\newunicodechar{Ξ}{\ensuremath{\Xi}}
+\newunicodechar{Π}{\ensuremath{\Pi}}
+\newunicodechar{τ}{\ensuremath{\tau}}
+\newunicodechar{γ}{\ensuremath{\gamma}}
+\newunicodechar{δ}{\ensuremath{\delta}}
+\newunicodechar{∅}{\ensuremath{\emptyset}}
+\newunicodechar{≡}{\ensuremath{\equiv}}
+\newunicodechar{∘}{\ensuremath{\circ}}
+\newunicodechar{𝔹}{\ensuremath{\mathbb{B}}}
+\newunicodechar{𝕓}{\ensuremath{\mathbb{b}}}
+\newunicodechar{∎}{\ensuremath{\blacksquare}}
+\newunicodechar{∈}{\ensuremath{\in}}
+\newunicodechar{⋆}{\ensuremath{\star}}
+\newunicodechar{⌜}{\ensuremath{\ulcorner}}
+\newunicodechar{⌝}{\ensuremath{\urcorner}}
+\newunicodechar{ʸ}{\ensuremath{{}^y}}
+\newunicodechar{ⁿ}{\ensuremath{{}^n}}
+\newunicodechar{ᴾ}{\ensuremath{{}^P}}
+
+\DeclareUnicodeCharacter{2032}{\ensuremath{\prime}}
+\DeclareUnicodeCharacter{2113}{\ensuremath{\ell}}
+\DeclareUnicodeCharacter{2207}{\ensuremath{\nabla}}
+\DeclareUnicodeCharacter{220B}{\ensuremath{\ni}}
+\DeclareUnicodeCharacter{2264}{\ensuremath{\leq}}
+\DeclareUnicodeCharacter{21D2}{\ensuremath{\Rightarrow}}
+\DeclareUnicodeCharacter{22A2}{\ensuremath{\vdash}}
+\DeclareUnicodeCharacter{22A4}{\ensuremath{\top}}
+\DeclareUnicodeCharacter{22A5}{\ensuremath{\bot}}
+
+\DeclareUnicodeCharacter{1D57}{\ensuremath{^t}}
+\newunicodechar{∙}{\ensuremath{\cdot}}
+
+% TODO: This is in general not a good idea.
+\providecommand\textepsilon{$\epsilon$}
+\providecommand\textmu{$\mu$}
+
+%Actually, varsyms should not occur in Agda output.
+
+% TODO: Make this configurable. IMHO, italics doesn't work well
+% for Agda code.
+
+\renewcommand\Varid[1]{\mathord{\textsf{#1}}}
+\let\Conid\Varid
+\newcommand\Keyword[1]{\textsf{\textbf{#1}}}
+\EndFmtInput
+
+\renewcommand\Varid[1]{\mathord{\textsf{#1}}}
+%\let\Conid\Varid
+%\newcommand\Keyword[1]{\textsf{\textbf{#1}}}
+
+%% end of the preamble, start of the body of the document source.
 
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Can Natural Models Simplify The Metatheory of Type Theory in Cubical Agda?}
-
+\title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
+\subtitle{A Case Study in \CA}
 %%
@@
 \affiliation{%
-  \institution{Institute of Information Science, Academia Sinica}
+  \institution{Academia Sinica}
   \city{Taipei}
@@
 \affiliation{%
-  \institution{Department of Computer and Information Sciences, University of Strathclyde}
+  \institution{University of Strathclyde}
   \city{Glasgow}
   \country{United Kingdom}}
+\email{fredrik.nordvall-forsberg@strath.ac.uk}
 
@@
 %\affiliation{%
-%  \institution{Institute for Logic, Language and Computation, University of Amsterdam}
+%  \institution{University of Amsterdam}
 %  \city{Amsterdam}
@@
 
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
-
-  We formalise type theory in type theory, using Awodey's notion of \emph{natural model} of type theory.
-  The initial natural model can be represented using quotient inductive-inductive-recursive types in the proof assistant \CA.
-Using natural models leads to fewer instances of so-called ``transport hell'' in the syntax of type theory and its elimination rules.
-To show that the approach is feasible, we formalise some meta-properties, including the standard % and logical predicate
-interpretation, normalisation by evaluation for typed terms, and strictification constructions.
-Since our formalisation is carried out using \CA's built-in support for quotient types, all our constructions compute.
-However, the use of transports along equations often reappeared when developing more sophisticated metatheory.
-Hence it is still a considerable struggle to develop the metatheory of type theory using welltyped terms with current proof assistant technology, and the effort is about the same using natural models or not.
-
-%   This paper presents a formalisation of a core type theory and its metatheory within \CA.
-% The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
-% We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.
-
-% The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
-% Moreover, type theory and its elimination are defined with dependent identities without transports.
-% This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
+  We present an intrinsic representation of type theory in the proof assistant \CA, inspired by Awodey’s natural models of type theory.
+  The initial natural model is defined as quotient inductive-inductive-recursive types, leading us to a syntax accepted by \CA without using any transports, postulates, or custom rewrite rules.
+We formalise some meta-properties such as the standard model, normalisation by evaluation for typed terms, and strictification constructions.
+Since our formalisation is carried out using \CA's native support for quotient inductive types, all our constructions compute at a reasonable speed.
+When we try to develop more sophisticated metatheory, however, the `transport hell' problem reappears.
+Ultimately, it remains a considerable struggle to develop the metatheory of type theory using an intrinsic representation that lacks strict equations.
+The effort required is about the same whether or not the notion of natural model is used.
 \end{abstract}
 
-%%
 %% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
 %% Please copy and paste the code instead of the example below.
-%%
-\begin{CCSXML}
-<ccs2012>
+\begin{CCSXML}\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{ccs2012>}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
    <concept>
@@
 
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
 \keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}
@@
 
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
 \maketitle
@@
 \section{Introduction}
+
+%\LT[inline]{Clearly articulate what has been formalised and what has not, when appropriate? Ideally finish formalisation, otherwise scale back and discuss what has been done. (R1, R3)}
+%\LT[inline]{Alternative formulation? E.g., \url{https://web.archive.org/web/20241004151846/https://lists.chalmers.se/pipermail/agda/2019/011176.html} (R2)}
+
+%\listoftodos
+
 % FNF (Fri 5 Sep)
 
-Transport hell = transports appear inside terms that you actually want to study, hence you need to do a lot of reasoning about transport rather than the actual objects of interest.
+Internalising the syntax and semantics of type theory in type theory is a long-standing problem which stretches the limits of the theory~\cite{Dybjer1996,Danielsson2006,Chapman2009,McBride2010,Altenkirch2016a}.
+%
+There are both practical and theoretical reasons to pursue this problem.
+%
+On the practical side, an internal representation of type theory is needed for mechanised metatheory and metaprogramming.
+%
+On the theoretic side, if type theory is supposed to be a general constructive foundation of mathematics, then it should in particular be able to reason about its own syntax and semantics (up to inherent limitations due to G\"odel's incompleteness theorems).
+%
+In dependent type theory, types can depend on terms, which means that all of contexts, types and terms need to be defined simultaneously.
+%
+This is one reason why formalising type theory in type theory is hard.
+%
 
+Early approaches to formalising type theory, e.g.\ McKinna and Pollack~\cite{McKinna1999}, dealt with untyped terms that were later refined by a typing relation, setoid equality, or quotients, and thus had to prove a lot of congruence lemmas by hand~\cite{Danielsson2006,Chapman2009,Brunerie2019,Boer2020}.
+%
+A breakthrough was achieved by Altenkirch and Kaposi~\cite{Altenkirch2016a}, who showed that quotient inductive-inductive types (QIITs)~\cite{Altenkirch2018} can be employed to significantly simplify the internal representation of well typed terms (or, more precisely, derivations), since equality constructors can be used to represent equations such as $\beta$- and $\eta$-equality.
+%
+They took Dybjer's notion of a model of type theory in the form of a category with families~\cite{Dybjer1996}, and translated it into a QIIT definition.
+%
+In effect, this gives rise to the \emph{initial} category with families, with the elimination principles of the QIIT giving a unique morphism of categories with families to any other model.
+%
+This thus gives a both principled and practical way to formalise the syntax and equational theory of type theory in type theory; the feasibility of the approach was demonstrated by e.g.\ formalising normalisation by evaluation using this representation~\cite{Altenkirch2017}.
+
+At the time of publication of Altenkirch and Kaposi~\cite{Altenkirch2016a}, the proof assistant \Agda did not allow equality constructors in data type declarations, so a workaround known as `Licata's trick'~\cite{Licata2011} was used by postulating (equality) constructors and writing down the eliminator explicitly, which meant giving up features of the proof assistant such as dependent pattern matching.
+\CA, the cubical variant of \Agda~\cite{Vezzosi2021}, is now equipped with a native support for QIITs, so it is natural to ask if we can use this support to formalise the intrinsic representation of type theory without the trick or any other compromise.
+
+In this paper, we explore this question and see what the proof assistant provides and lacks to achieve this goal.
+We quickly find that their QIIT definitions break the strict positivity --- a syntactic restriction imposed by \CA to ensure consistency.
+Moreover, their definition is cumbersome to work with, since the type of later constructors or even equations often only make sense because of earlier equations.
+%
+In an intensional type theory, such as those implemented in proof assistants, this manifests itself in transport terms across equality proofs \emph{inside} other terms, and leads to the so-called `transport hell' --- rather than just reasoning about the terms you actually want to study, you now also have to do a lot of reasoning about transports themselves and their algebraic properties.
+%
+It turns out that we need an alternative way of representing type theory intrinsically without any transport hell, in order to make our formalisation of type theory more lightweight and accepted by \CA.
+
+%
+The framework of categories with families is only one of several (more or less) equivalent notions of models of type theory~\cite{Hofmann1997}, and we were wondering if any of the other notions might offer any advantages.
+%
+Bense~\cite{Bense2024} suggested that Awodey's notion of natural model~\cite{Awodey2018} might be a good candidate.
+%
+Indeed, in a natural model, the indexing of terms over their types $\mathsf{Tm}_{\Gamma} : \mathsf{Ty}(\Gamma) \to \mathsf{Set}$ (as in a category with families) is replaced by a `fibred' perspective where each term instead \emph{has} a type, as picked out by a function $\mathsf{tyOf} : \mathsf{Tm}(\Gamma) \to \mathsf{Ty}(\Gamma)$.
+%
+Terms and types are still indexed by contexts $\Gamma$, but since most `type mismatches' arise from equations between types, not equations between contexts (indeed many formulations of type theory does not even have a notion of context equality), this should mean that many uses of transports can be avoided.
+
+We test this hypothesis by formalising type theory in a form inspired by natural models.
+%
+\CA is particularly a good fit for such a project, because not only does it support QIITs, it also supports inductive-recursive types~\cite{Dybjer1999}, which are needed to simultaneously define the recursive $\mathsf{tyOf}$ function together with the inductively defined types $\mathsf{Tm}(\Gamma)$ and $\mathsf{Ty}(\Gamma)$.
+%
+Indeed, it could be the lack of support for inductive-recursive definitions in many proof assistants which has held back formalisation attempts based on natural models so far.
+
+While we manage to avoid transports occurring in its own syntax, the experiment is not an outright success.
+%
+Indeed, we found that when developing more sophisticated metatheory, such as when defining a logical predicate model, the use of transports along equations often reappeared.
+%
+Strictification~\cite{Donko2022,Kaposi2025}, a recent technique for strictifying equations, cannot be applied directly because of the lack of strict proposition in the metatheory.
+%
+Furthermore, we found that the use of natural models is less well supported in the Cubical Agda of today, compared to approaches based purely on QIITs.
+%
+This is because we are more reliant on the computational behaviour of the recursively defined $\mathsf{tyOf}$ function, and this behaviour is only available in `later' clauses, which leads to the need for hacks and tricks to work around this limitation.
+%
+We discuss proof assistant features and their helpfulness further towards the end of the paper, after presenting our formalisation.
+
+\paragraph{Contributions} We make the following contributions:
+\begin{itemize}
+\item We present an intrinsically well typed representation of the syntax of type theory, inspired by Awodey's natural models (\cref{sec:tt}).
+\item We derive elimination and recursion principles for the syntax (\cref{sec:tt:elim}), and show how the standard model and the term model are constructed (\cref{sec:standard-model}).
+\item We discuss strictification constructions on models, and show that they also apply to our notion of natural models (\cref{sec:strictify}).
+\item We develop normalisation by evaluation for substitution calculus~(\cref{sec:nbe}) as a proof of concept: our development is carried out in \CA, which has a computational implementation of QIITs and principles such as function extensionality, so the resulting normaliser computes, and can potentially be extracted as a verified program.
+\item We discuss pros and cons of our approach compared to other approaches, and which features of a proof assistant and its metatheory would make future formalisation more feasible (\cref{sec:discussion}).
+\end{itemize}
+
+%\LT{the idea of using natural model appears at least in 2024 \cite{Bense2024}, and it is a natural idea to formalise type theory in this way.}
 \section{Setting and metatheory}
@@
 
-Cubical Agda with UIP
+%\LT{Should we discharge this assumption in our formalisation, eg set-truncate where needed instead? (R1, R2)}
+Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0.
+%
+We explicitly set-truncate the types we define.
+Therefore, the term \emph{set} is used interchangeably with type.
+%
+For the sake of simplicity we occasionally postulate uniqueness of identity proofs (UIP) locally, which is of course inconsistent with univalence but in principle compatible with cubical type theory. 
+%
+We believe that the cubical type theory XTT~\cite{Sterling2022}, which enjoys definitional UIP without univalence, justifies this local assumption.
+%
 
-QIIRTs
+\CA implements cubical type theory, and one of the important concepts therein is the interval type \ensuremath{\Conid{I}} with two distinguished endpoints \ensuremath{\Varid{i0}} and \ensuremath{\Varid{i1}}.
+%
+Propositional equality \ensuremath{\Varid{x}\;\mathop\equiv\;\Varid{y}} in a type \ensuremath{\Conid{A}} is given by \emph{paths} in \ensuremath{\Conid{A}}, i.e., by a functions \ensuremath{\Conid{I}\;\Varid{→}\;\Conid{A}}.
+More generally, dependent paths \ensuremath{\Varid{p}\;\mathbin{:}\;\Conid{PathP}\;\Conid{P}\;\Varid{a}\;\Varid{b}} are dependent functions \ensuremath{\Varid{p}\;\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{I})\;\Varid{→}\;\Conid{P}\;\Varid{i}} sending \ensuremath{\Varid{i0}} to \ensuremath{\Varid{a}\;\mathbin{:}\;\Conid{P}\;\Varid{i0}} and \ensuremath{\Varid{i1}} to \ensuremath{\Varid{b}\;\mathbin{:}\;\Conid{P}\;\Varid{i1}}.
+Note that \ensuremath{\Conid{P}\;\mathbin{:}\;\Conid{I}\;\Varid{→}\;\Conid{Type}} itself is a path in the universe \ensuremath{\Conid{Type}}, hence a witness that \ensuremath{\Conid{P}\;\Varid{i0}\;\mathop\equiv\;\Conid{P}\;\Varid{i1}}, which the dependent path is \emph{over}, so paths are special cases of \ensuremath{\Conid{PathP}} with \ensuremath{\Conid{P}} being a constant path.
+%
+The constant path \ensuremath{\Varid{refl}} defined as \ensuremath{\Varid{λ}\;\anonymous \;\Varid{→}\;\Varid{a}} witnesses that equality is reflexive \ensuremath{\Varid{a}\;\mathop\equiv\;\Varid{a}}, and paths can be lifted to type families in the sense that there is a transport operation \ensuremath{\Varid{subst}\;\mathbin{:}\;(\Conid{P}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type})\;\Varid{→}\;\Varid{x}\;\mathop\equiv\;\Varid{y}\;\Varid{→}\;\Conid{P}\;\Varid{x}\;\Varid{→}\;\Conid{P}\;\Varid{y}}.
+%
+See the literature on cubical type theory for details~\cite{Vezzosi2021}.
+%
 
-\section{Type theory as a natural model}
-% LTC (Sun 7 Sep)
+\CA also allows paths to appear as the target of constructors in inductive definitions.
+That is, it implements higher inductive types~\cite{Lumsdaine2020}.
+%
+When defining a function \ensuremath{\Varid{f}\;\mathbin{:}\;\Conid{H}\;\Varid{→}\;\Conid{X}} by pattern matching out of a higher inductive type, \ensuremath{\Varid{f}} also needs to be defined on the path constructors: if \ensuremath{\Varid{e}\;\mathbin{:}\;\Varid{s}\;\mathop\equiv\;\Varid{t}} is a path from \ensuremath{\Varid{s}} to \ensuremath{\Varid{t}} in \ensuremath{\Conid{H}}, then \ensuremath{\Varid{f}\;\Varid{e}\;\mathbin{:}\;\Varid{f}\;\Varid{s}\;\mathop\equiv\;\Varid{f}\;\Varid{t}} should be a path from \ensuremath{\Varid{f}\;\Varid{s}} to \ensuremath{\Varid{f}\;\Varid{t}} in \ensuremath{\Conid{X}}.
+%
+Agda's support for simultaneous definitions allows us to define quotient inductive-inductive types (QIITs)~\cite{Altenkirch2018}, where a type \ensuremath{\Conid{A}\;\mathbin{:}\;\Conid{Type}} and a type family \ensuremath{\Conid{B}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type}} are defined inductively simultaneously.
+%
+Agda even allows us to define quotient inductive-inductive-recursive types (QIIRTs), where \ensuremath{\Conid{A}\;\mathbin{:}\;\Conid{Type}} and \ensuremath{\Conid{B}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type}} are defined inductively together with a recursive function \ensuremath{\Varid{f}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{C}}.
+%
+We will make use of this feature to define types, terms and the \ensuremath{\Varid{tyOf}} function from terms to types simultaneously.
 
-\subsection{Natural models}
+For the brevity of presentation, we have made arguments implicit for equality constructors, even though they are explicit in our formalisation.
+%
+Similarly, we are ignoring universe levels, but they are all present in the formalisation.
 
-\subsection{The Syntax of Type Theory as a Quotient Inductive-Inductive-Recursive Type}
+\section{Type theory as quotient inductive types} \label{sec:tt}
 
-QIIRT def of the syntax
+%\FNF{Discuss ``another approach is staying completely indexed and always using PathP instead of transports'' here, or in Discussion (R2)}
 
-\subsection{Recursion and Elimination Principles}
+In this section, we will explain why Altenkirch and Kaposi's representation~\cite{Altenkirch2016a} is hard to use in practice and rejected by \CA arising from transports in its definition.
+Then, we show how their representation can be transformed to a representation based on Awodey's natural models.
+This representation is accepted by \CA, since it is free from transports.
 
-\subsection{Strictification}
+%This section's aim is to exhibit that Altenkirch and Kaposi's representation, which contains the transport hells and violates the syntactic restriction imposed by \CA, can be transformed to a representation based on Awodey's natural model, which is free from transports and accepted by \CA.
+%Then, we will show how other type formers can be represented in this way.
+%In the reminder of this section, we will give its elimination principle and explain how these definitions are formalised in \CA.
 
-\section{Metatheory}
-% LTC (Tue 9 Sep)
+\subsection{Type theory as the initial CwF model} \label{sec:tt:cwf}
+We briefly recall the QIIT representation given by Altenkirch and Kaposi.
+Each judgement therein is defined as an inductive type, each typing rule as a constructor, and each equality between types, terms, and substitutions as an \emph{equality constructor}.
+The inhabitants of these types are valid derivations in type theory, because their validity is enforced by typing constraints.
+The four types of judgements in type theory are represented inductive-inductively %and indexed by their context and by their types for terms
+as
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Conid{Ctx}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\Conid{Sub}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;{}\<[24]%
+\>[24]{}\Varid{→}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{Ctx})\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\Conid{Ty}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\Conid{Tm}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;{}\<[24]%
+\>[24]{}\Varid{→}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+For example, an inhabitant of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}} represents a derivation for a term of type $A$ in context \ensuremath{\Conid{Γ}}.
+Rules are represented by constructors of these inductive types:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{∅}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ,\char95 }\;{}\<[10]%
+\>[10]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\_[\_]_{\text{T}}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\_[\_]_{\text{t}}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Δ}\;\Conid{A})\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{idS}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∘\char95 }\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Δ}\;\Conid{Θ}\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ}\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Θ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ,\char95 }\;{}\<[10]%
+\>[10]{}\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}))\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}[\circ]_{\text{T}}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{A}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{τ}\;\Varid{∘}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+ The constructor \ensuremath{\Varid{∅}} represents the empty context, and \ensuremath{\Conid{Γ}\;\Varid{,}\;\Conid{A}} a context extension, while \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} and \ensuremath{\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}} represent substituted types and terms, respectively.
+Further, \ensuremath{\Varid{idS}} is the identity substitution, \ensuremath{\Varid{\char95 ∘\char95 }} the constructor for substitution composition, and the second \ensuremath{\Varid{\char95 ,\char95 }} the constructor for extending a substitution \ensuremath{\Varid{σ}} with a term \ensuremath{\Varid{t}} of type \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} (making use of \Agda's support for overloaded constructor names).
+The equality constructor~\ensuremath{[\circ]_{\text{T}}} states that type substitution by \ensuremath{\Varid{τ}} followed by type substitution by \ensuremath{\Varid{σ}} is the same as a single substitution by the composition \ensuremath{\Varid{τ}\;\Varid{∘}\;\Varid{σ}}.
 
-\subsection{Standard Model}
+When formulating the corresponding rule for the interaction between \ensuremath{\Varid{\char95 ∘\char95 }} and \ensuremath{\Varid{\char95 ,\char95 }}, we encounter a type mismatch that needs to be resolved by a transport (highlighted in \highlight{addition}{\text{green}}): %, leading to the transport hell when reasoning with this equality:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{,∘}\;{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Varid{σ}\;\Varid{,}\;\Varid{t})\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\highlight{addition}{\Varid{subst}\;(\Conid{Tm}\;\Conid{Γ})\;[\circ]_{\text{T}}}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The reason is that the type of \ensuremath{\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}} is \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}} rather than the required \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}}.
+However, since \ensuremath{\Conid{Tm}} is an argument to \ensuremath{\Varid{subst}}, the use of transport violates a syntactic restriction of \Agda, namely its strict positivity check.
+In theory, transports are allowed in QIITs~\cite{Kaposi2019}, but it is not clear to us how this syntactic restriction should be relaxed for higher inductive types supported by \CA to take into account other cubical primitives (such as \ensuremath{\Varid{hcomp}}).
 
-% Argue for relative consistency?
+In other words, transport hell is not only an obstacle for reasoning, but also breaks strict positivity in \CA when arising in inductive definitions themselves.
+The situation becomes worse once additional type formers are introduced --- such as $\Pi$-types and the type \ensuremath{\Conid{El}} of elements~\cite{Altenkirch2016a} --- since each brings further instances of this problem.
+%
+Of course, one could bypass the strict positivity check, but doing so would undermine the general trustworthiness of formalisation.
+Another possibility is to fix the syntactic restriction for HIITs, but it is unclear what conditions should be.
+Therefore, we seek for an equivalent definition without transports first.
 
-\subsection{Normalisation by Evaluation}
+Another source of transports arises from equations over equations, but this can be avoided by using dependent paths.
+For example, the fact that the identity term substitution really acts as an identity is introduced as an equality constructor \ensuremath{[\mathsf{idS}]_{\text{t}}}, defined over the highlighted equality constructor \ensuremath{[\mathsf{idS}]_{\text{T}}\;\mathbin{:}\;\Conid{A}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} for the identity type substitution:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}[\mathsf{idS}]_{\text{t}}\;\mathbin{:}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}\;(\highlight{addition}{[\mathsf{idS}]_{\text{T}}\;\Varid{i}}))\;\Varid{t}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{t}}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Although equations over equations are in principle more manageable, it quickly leads us to \emph{equations over equations over yet more equations} in their elimination rules.
+Hence, it is still preferable to avoid them if possible.
 
-% (for SC)
+\subsection{The `Ford transformation' and index elimination} \label{sec:tt:terms-without-indices}
 
-\section{Comparison with Other Approaches}
-% FNF (Tue 9 Sep)
+To avoid transports in the definition itself, we note that the index \ensuremath{\Conid{A}} of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}} often needs an explicit proof for the typing constraint --- for example, that the term \ensuremath{\Varid{t}} in the substitution \ensuremath{(\Varid{σ}\;\Varid{,}\;\Varid{t})} has type \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} --- if this does not happen to hold strictly (i.e., up to judgemental equality), so enforcing this constraint in the index of \ensuremath{\Conid{Tm}} just shoots ourselves in the foot.
+Hence, we apply the `Ford transformation'~\cite{McBride1999} (`You can have any index you want, as long as it is equal to the specified one') to move the constraint on its index to its argument as an equality proof (highlighted below):
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{\char95 ,\char95 ∶[\char95 ]}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Conid{B})\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Conid{B}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The constructor \ensuremath{\Varid{,∘}} which had a transport in its type becomes
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{,∘}\;{}\<[5]%
+\>[5]{}\mathbin{:}\;\Varid{...}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{6}{}\<[11]%
+\>[11]{}\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{cong}\;\Varid{\char95 [}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\Varid{pt}\;\Varid{∙}\;[\circ]_{\text{T}}}\;\mskip1.5mu]){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where \ensuremath{\Varid{\char95 ∙\char95 }} is the transitivity of equality.
+Although transport is not needed this time, the use of \ensuremath{\Varid{cong}} and \ensuremath{\Varid{\char95 ∙\char95 }}
+still prevent the definition from being seen as strictly positive.
+Similar to the Ford transformation, this problem can be overcome by asking for another equality proof, highlighted below, as an argument:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{,∘}\;\mathbin{:}\;\Varid{...}\;\highlight{addition}{(\Varid{qt}\;\mathbin{:}\;\Conid{B}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{→}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ})\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{qt}}\;\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any information but only defers the proof obligation.
+%This redundant argument can be removed later when defining its eliminator (\Cref{sec:tt:elim}).
 
-Compared to QIIT:
+Once the Ford transformation has been applied, the index \ensuremath{\Conid{B}} in \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{B}} no longer plays the role of enforcing constraints.
+This opens the door to a simpler design: instead of carrying the index around, we can `Ford' all \ensuremath{\Conid{Tm}} constructors uniformly and remove the index entirely.
+To preserve the necessary typing information, we simultaneously introduce an auxiliary function \ensuremath{\Varid{tyOf}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}} that records it explicitly.
+In the end, the constructor \ensuremath{\Varid{,∘}} becomes
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{,∘}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Δ}\;\Conid{Θ})\;(\Varid{t}\;\mathbin{:}\;\highlight{addition}{\Conid{Tm}\;\Conid{Δ}})\;(\Varid{τ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{→}\;(\Varid{pt}\;\mathbin{:}\;\highlight{addition}{\Varid{tyOf}\;\Varid{t}}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{→}\;(\Varid{qt}\;\mathbin{:}\;\highlight{addition}{\Varid{tyOf}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}})}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\Varid{→}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ})\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\Varid{qt}\;\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
-\begin{itemize}
-\item Fewer transports in the syntax, but they tend to come back in concrete models
-\item Strictification orthogonal
-\end{itemize}
+As a side effect, this approach also removes the need for dependent paths in the definition.
+Two terms can now be compared even when it is not known in advance whether their types are equal.
+For instance, the equality constructor for the identity substitution becomes
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}[\mathsf{idS}]_{\text{t}}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\Varid{t}\;\mathop\equiv\;\Varid{t}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{t}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where the fact that \ensuremath{\Varid{t}} and \ensuremath{\Varid{t}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{t}}} share the same type follows from their term equality, rather than being a \emph{requirement}.
 
-Compared to untyped version:
-\begin{itemize}
-\item Untyped version might still be easiest to work with, with current proof assistant technology
-\end{itemize}
+\subsection{Representing substitution calculus using QIIRT}
+Building on the changes described in \Cref{sec:tt:terms-without-indices}, we now spell out our version of substitution calculus.
+The following types are defined simultaneously with a recursive function (changes compared to the QIIT version highlighted):
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Conid{Ctx}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\Conid{Sub}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;\Varid{→}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{Ctx})\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\Conid{Ty}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\Conid{Tm}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\highlight{addition}{(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;\Varid{→}\;\Conid{Set}}{}\<[E]%
+\\
+\>[B]{}\highlight{addition}{\Varid{tyOf}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Similarly to the QIIT representation, constructors are introduced for rules and equalities as follows, where we highlight constructors that are different from their QIIT counterpart:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{31}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{39}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{∅}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ,\char95 }\;{}\<[18]%
+\>[18]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\_[\_]_{\text{T}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\_[\_]_{\text{t}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Δ})\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{∅}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Varid{∅}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{\Varid{\char95 ,\char95 ∶[\char95 ]}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{idS}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∘\char95 }\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Δ}\;\Conid{Θ}\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ}\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Θ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π₁}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A})\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π₂}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A})\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{idS∘\char95 }\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Varid{idS}\;\Varid{∘}\;\Varid{σ}\;\mathop\equiv\;\Varid{σ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∘idS}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Varid{σ}\;\Varid{∘}\;\Varid{idS}\;\mathop\equiv\;\Varid{σ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{assocS}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;(\Varid{γ}\;\Varid{∘}\;\Varid{τ})\;\Varid{∘}\;\Varid{σ}\;\mathop\equiv\;\Varid{γ}\;\Varid{∘}\;(\Varid{τ}\;\Varid{∘}\;\Varid{σ}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}[\mathsf{idS}]_{\text{T}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{A}\;{}\<[23]%
+\>[23]{}\mathop\equiv\;{}\<[26]%
+\>[26]{}\Conid{A}\;{}\<[29]%
+\>[29]{}[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{[\mathsf{idS}]_{\text{t}}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Varid{t}\;{}\<[23]%
+\>[23]{}\mathop\equiv\;{}\<[26]%
+\>[26]{}\Varid{t}\;{}\<[29]%
+\>[29]{}[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{t}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}[\circ]_{\text{T}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{A}\;{}\<[23]%
+\>[23]{}[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;{}\<[31]%
+\>[31]{}[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;{}\<[39]%
+\>[39]{}\mathop\equiv\;\Conid{A}\;{}\<[44]%
+\>[44]{}[\kern-1.5pt\;\Varid{τ}\;\Varid{∘}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{[\circ]_{\text{t}}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Varid{t}\;{}\<[23]%
+\>[23]{}[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;{}\<[31]%
+\>[31]{}[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;{}\<[39]%
+\>[39]{}\mathop\equiv\;\Varid{t}\;{}\<[44]%
+\>[44]{}[\kern-1.5pt\;\Varid{τ}\;\Varid{∘}\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{\Varid{,∘}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\Varid{qt}\;\mskip1.5mu]){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We write \ensuremath{\Varid{wk}} for the `weakening' substitution \ensuremath{\Varid{wk}\;\mathrel{=}\;\Varid{π₁}\;\Varid{idS}\;\mathbin{:}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})\;\Conid{Γ}}, and \ensuremath{\Varid{vz}} for the `top variable' \ensuremath{\Varid{vz}\;\mathrel{=}\;\Varid{π₂}\;\Varid{idS}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})}.
+This works well except that we have to interleave the function clauses of \ensuremath{\Varid{tyOf}} with constructors.
+For example, we need define the function clause for \ensuremath{\Varid{π₂}\;\Varid{σ}} before the $\eta$-law for substitution:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{tyOf}\;(\Varid{π₂}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{Δ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{A}\mskip1.5mu\}\;\Varid{σ})\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{ηπ}\;\mathbin{:}\;\Varid{σ}\;\mathop\equiv\;(\Varid{π₁}\;\Varid{σ}\;\Varid{,}\;\Varid{π₂}\;\Varid{σ}\;\Varid{∶[}\;\Varid{refl}\;\mskip1.5mu]){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Otherwise, the proof obligation \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} on the right hand side of \ensuremath{\Varid{ηπ}} cannot be fulfilled by \ensuremath{\Varid{refl}}.
+We proceed with other equality constructors:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{η∅}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{σ}\;\mathop\equiv\;\Varid{∅S}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{βπ₁}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{π₁}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\mskip1.5mu])\;\mathop\equiv\;\Varid{σ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{βπ₂}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{...}\;\highlight{addition}{(\Varid{q}\;\mathbin{:}\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\mskip1.5mu])\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Varid{tyOf}\;\Varid{t})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{π₂}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\mskip1.5mu])\;\mathop\equiv\;\Varid{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Note that \ensuremath{\Varid{βπ₂}} has an additional derivable equality proof \ensuremath{\highlight{addition}{\Varid{q}}}.
+This argument is needed as the coherence condition for
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{tyOf}\;(\Varid{βπ₂}\;\Varid{...}\;\Varid{q}\;\Varid{i})\;{}\<[21]%
+\>[21]{}\mathrel{=}\;\Varid{q}\;\Varid{i}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+since again using any other function while defining inductive types breaks the strict positivity check.
+The remaining clauses are given as
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{tyOf}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;{}\<[18]%
+\>[18]{}\mathrel{=}\;(\Varid{tyOf}\;\Varid{t})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;([\mathsf{idS}]_{\text{t}}\;\Varid{i})\;{}\<[18]%
+\>[18]{}\mathrel{=}\;[\mathsf{idS}]_{\text{T}}\;\Varid{i}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;([\circ]_{\text{t}}\;\Varid{i})\;{}\<[18]%
+\>[18]{}\mathrel{=}\;[\circ]_{\text{T}}\;\Varid{i}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+This definition is accepted by \CA\footnote{At the time of writing, \CA does not support interleaved mutual definitions, but it can be equivalently defined using forward declarations.
+We will discuss this idiom in \Cref{sec:tt:mutual}.}
+without any warnings or errors.
+Although \ensuremath{\Conid{Tm}} is only indexed by \ensuremath{\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx}}, the function \ensuremath{\Varid{tyOf}} ensures that every \ensuremath{\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}} has a type.
+Hence, \ensuremath{\Conid{Tm}} only consists of valid derivations and is still an intrinsic representation of type theory.
 
-\section{Discussion}
-% LTC (at first)
+Replacing the index \ensuremath{\Conid{A}\;\mathbin{:}\;\Conid{Ty}} of \ensuremath{\Conid{Tm}} by a function \ensuremath{\Varid{tyOf}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}} aligns with Awodey's notion of natural model~\cite{Awodey2018} where the collections of terms and types are represented as presheaves $\mathsf{Tm}, \mathsf{Ty} \colon \mathbb{C} \to \Set$ over the category of contexts $\mathbb{C}$ and connected by a natural transformation $\mathsf{Tm} \to \mathsf{Ty}$ satisfying that each substitution into a non-empty context is equivalent to a pair of substitution and a term.
+That is, we have derived the intrinsic representation of type theory as a natural model using QIIRT in \CA.
+This situates our family of inductive types and their algebras within a well-studied categorical models for type theory.
 
-Practical considerations (eg NBE computes, support for interleaved mutual definitions)
+\subsection{\texorpdfstring{$\Pi$}{Pi}-types}
 
-Support from proof assistant (eg safety guarantees)
+We extend our object type theory with dependent function types.
+First we define the lifting of a substitution by a type as the following abbreviation:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{\char95 ↑\char95 }\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;\Varid{→}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\\
+\>[B]{}\Varid{\char95 ↑\char95 }\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\Varid{σ}\;\Conid{A}\;\mathrel{=}\;{}\<[16]%
+\>[16]{}\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\mskip1.5mu\}\;\Varid{idS,}\;{}\<[E]%
+\\
+\>[16]{}\Varid{π₂}\;(\Varid{idS}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\mskip1.5mu\})\;\Varid{∶[}\;\Varid{p}\;\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where \ensuremath{\Varid{p}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{π₂}\;\Varid{idS})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}}.
+We can use \ensuremath{[\circ]_{\text{T}}} to define \ensuremath{\Varid{p}}, as \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;(\Varid{idS}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\mskip1.5mu\}))} is equal to \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} by definition.\footnote{%
+Yet, as interleaving function clauses with inductive types is also not supported, the strict equality \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathrel{=}\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} is not available at this point for \ensuremath{\Varid{p}}.
+We use forward declarations to introduce the required equalities \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} and \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{idS})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} to be defined later as \ensuremath{\Varid{refl}} and \ensuremath{[\circ]_{\text{T}}}, see \Cref{sec:tt:mutual} for further details.}
 
-Easier in OTT-based proof assistant
+Other constructors are introduced following the `Ford transformation', with differences compared to the usual QIIT presentation highlighted:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{49}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Π}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;(\Conid{B}\;\mathbin{:}\;\Conid{Ty}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}))\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{\Varid{app}}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;(\Conid{B}\;\mathbin{:}\;\Conid{Ty}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}))\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;\Conid{B})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{abs}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}))\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Π[]}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;(\Conid{Π}\;\Conid{A}\;\Conid{B})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{Π}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;(\Conid{B}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑}\;\Conid{A}\;\kern-1.5pt]_{\text{T}}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{\Varid{abs[]}}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Varid{abs}\;\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\mathop\equiv\;\Varid{abs}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑}\;\Conid{A}\;\kern-1.5pt]_{\text{t}}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Πβ}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Varid{app}\;(\Varid{abs}\;\Varid{t})\;(\Varid{tyOf}\;\Varid{t})\;\highlight{addition}{\Varid{pt}}\;{}\<[49]%
+\>[49]{}\mathop\equiv\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Πη}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Varid{abs}\;(\Varid{app}\;\Varid{t}\;\Conid{B}\;\Varid{pt})\;{}\<[49]%
+\>[49]{}\mathop\equiv\;\Varid{t}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;(\Varid{app}\;\Varid{t}\;\Conid{B}\;\Varid{p})\;{}\<[23]%
+\>[23]{}\mathrel{=}\;\Conid{B}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Varid{abs}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;\Varid{t})\;{}\<[23]%
+\>[23]{}\mathrel{=}\;\Conid{Π}\;\Conid{A}\;(\Varid{tyOf}\;\Varid{t}){}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Varid{abs[]}\;\Varid{σ}\;\Varid{t}\;\Varid{i})\;{}\<[23]%
+\>[23]{}\mathrel{=}\;\Conid{Π[]}\;\Varid{σ}\;(\Varid{tyOf}\;\Varid{t})\;\Varid{i}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Conid{Πβ}\;\Varid{t}\;\Varid{pt}\;\Varid{i})\;{}\<[23]%
+\>[23]{}\mathrel{=}\;\Varid{tyOf}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Conid{Πη}\;\Varid{t}\;\Varid{pt}\;\Varid{i})\;{}\<[23]%
+\>[23]{}\mathrel{=}\;\Varid{pt}\;(\mathord{\sim}\;\Varid{i}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Apart from the extra clauses of \ensuremath{\Varid{tyOf}}, the main change happens in the constructor \ensuremath{\Varid{app}}.
+The constraint that \ensuremath{\Varid{t}} is of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} is enforced there, but every other constructor remains as almost the same as their QIIT definition.
+We have formulated application and abstraction as an isomorphism between terms of type \ensuremath{\Conid{B}} in context \ensuremath{\Conid{Γ}\;\Varid{,}\;\Conid{A}} and terms of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} in context \ensuremath{\Conid{Γ}}, but we can also derive ordinary application:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{\char95 ∶[\char95 ]\$\$\char95 ∶[\char95 ]}\;{}\<[15]%
+\>[15]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;\Conid{B}\;{}\<[E]%
+\\
+\>[15]{}\Varid{→}\;(\Varid{s}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Varid{tyOf}\;\Varid{s}\;\mathop\equiv\;\Conid{A}\;{}\<[E]%
+\\
+\>[15]{}\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\Varid{]\$\$}\;\Varid{s}\;\Varid{∶[}\;\Varid{q}\;\mskip1.5mu]\;\mathrel{=}\;\Varid{app}\;\Varid{t}\;\anonymous \;\Varid{p}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{s}\;\Varid{∶[}\;\Varid{q}\;\Varid{∙}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{t}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+As an example, we can write the identity function at type \ensuremath{\Conid{A}} as \ensuremath{\Varid{id}\;\Conid{A}\;\mathrel{=}\;\Varid{abs}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;(\Varid{π₂}\;\Varid{idS})}, and the identity function at type \ensuremath{\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\kern-1.5pt]_{\text{T}})} applied to the identity function at type \ensuremath{\Conid{A}} is represented by
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{idid}\;\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\Varid{idid}\;\Conid{A}\;\mathrel{=}\;\Varid{id}\;(\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\mskip1.5mu]))\;\Varid{∶[}\;\Varid{refl}\;\Varid{]\$\$}\;\Varid{id}\;\Conid{A}\;\Varid{∶[}\;\Varid{refl}\;\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+and indeed using the equality constructors, we have a proof of \ensuremath{\Varid{tyOf}\;(\Varid{idid}\;\Conid{A})\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\kern-1.5pt]_{\text{T}})} as expected.
+
+\subsection{The type of Booleans}
+
+To introduce the inductive type of Booleans, we need to specialise the substitution lifting.
+Let us see its constructors (with differences highlighted) and explain why a specialisation is needed.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{𝔹}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{𝔹[]}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{𝔹}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{𝔹}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{tt}\;\Varid{ff}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{tt[]}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Varid{tt}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\mathop\equiv\;\Varid{tt}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{ff[]}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Varid{ff}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\mathop\equiv\;\Varid{ff}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;\Varid{tt}\;{}\<[10]%
+\>[10]{}\mathrel{=}\;\Conid{𝔹}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;\Varid{ff}\;{}\<[10]%
+\>[10]{}\mathrel{=}\;\Conid{𝔹}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Varid{tt[]}\;\Varid{σ}\;\Varid{i})\;{}\<[18]%
+\>[18]{}\mathrel{=}\;\Conid{𝔹[]}\;\Varid{σ}\;\Varid{i}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Varid{ff[]}\;\Varid{σ}\;\Varid{i})\;{}\<[18]%
+\>[18]{}\mathrel{=}\;\Conid{𝔹[]}\;\Varid{σ}\;\Varid{i}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{elim𝔹}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;(\Conid{P}\;\mathbin{:}\;\Conid{Ty}\;(\Conid{Γ}\;\Varid{,}\;\Conid{𝔹}))\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[17]%
+\>[17]{}\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{tt}\;\Varid{∶[}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{u}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[17]%
+\>[17]{}\highlight{addition}{(\Varid{pu}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{u}\;\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{ff}\;\Varid{∶[}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[17]%
+\>[17]{}\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{𝔹}\;[\kern-1.5pt\;\Varid{idS}\;\mskip1.5mu])}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;(\Varid{elim𝔹}\;\Conid{P}\;\Varid{u}\;\Varid{t}\;\Varid{pu}\;\Varid{pt}\;\Varid{b}\;\Varid{pb})\;\mathrel{=}\;\Conid{P}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{b}\;\Varid{∶[}\;\Varid{pb}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The only thing missing from the above definition is the substitution rule for \ensuremath{\Varid{elim𝔹}}:
+applying the substitution \ensuremath{\Varid{σ}} to `\ensuremath{\Varid{elim𝔹}\;\Conid{P}\;\Varid{t}\;\Varid{pt}\;\Varid{u}\;\Varid{pu}\;\Varid{b}\;\Varid{pb}}' is equal to applying a lifted substitution  \ensuremath{\Varid{σ}\;\Varid{↑}\;\Conid{𝔹}} to \ensuremath{\Conid{P}} and \ensuremath{\Varid{σ}} to \ensuremath{\Varid{t}}, \ensuremath{\Varid{u}}, and \ensuremath{\Varid{b}}.
+However, \ensuremath{\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑}\;\Conid{𝔹}\;\kern-1.5pt]_{\text{T}}} gives us a type in the context \ensuremath{\Conid{Δ}\;\Varid{,}\;\Conid{𝔹}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} instead of \ensuremath{\Conid{Δ}\;\Varid{,}\;\Conid{𝔹}}, so we provide a lifting with a type \ensuremath{\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ}\;\Varid{→}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{𝔹})\;(\Conid{Δ}\;\Varid{,}\;\Conid{𝔹})} with a proof that \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{𝔹}\mskip1.5mu\}\;\Varid{idS})\;\mathop\equiv\;\Conid{𝔹}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}}.
+The proof, however, requires the transitivity of equalities, and \CA would see this as a strict positivity problem.
+Instead we introduce a \emph{superfluous} equality constructor \ensuremath{\Conid{𝔹[]₂}} to satisfy its proof obligation (highlighted):
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{𝔹[]₂}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Varid{tyOf}\;(\Varid{π₂}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{𝔹}\mskip1.5mu\}\;\Varid{idS})\;\mathop\equiv\;\Conid{𝔹}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{\char95 ↑𝔹}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{𝔹})\;(\Conid{Δ}\;\Varid{,}\;\Conid{𝔹}){}\<[E]%
+\\
+\>[B]{}\Varid{\char95 ↑𝔹}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{Δ}\mskip1.5mu\}\;\Varid{σ}\;\mathrel{=}\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{𝔹}\mskip1.5mu\}\;\Varid{idS}\;\Varid{,}\;\Varid{π₂}\;\Varid{idS}\;\Varid{∶[}\;\highlight{addition}{\Conid{𝔹[]₂}}\;\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Finally, we introduce the equality constructor for the interaction between \ensuremath{\Varid{elim𝔹}} and substitution:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{elim𝔹[]}\;\mathbin{:}\;\Varid{...}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{pt₂}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑𝔹}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{tt}\;\Varid{∶[}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{pu₂}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑𝔹}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{ff}\;\Varid{∶[}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{pb₂}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{𝔹}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}(\Varid{q}\;\mathbin{:}\;{}\<[11]%
+\>[11]{}\Conid{P}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{b}\;\Varid{∶[}\;\Varid{pb}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;{}\<[E]%
+\\
+\>[11]{}\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{wk}\;\Varid{,}\;\Varid{vz}\;\Varid{∶[}\;\Conid{𝔹[]₂}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\Varid{pb₂}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;(\Varid{elim𝔹}\;{}\<[15]%
+\>[15]{}\Conid{P}\;\Varid{t}\;\Varid{pt}\;\Varid{u}\;\Varid{pu}\;\Varid{b}\;\Varid{pb})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{6}{}\<[11]%
+\>[11]{}\mathop\equiv\;\Varid{elim𝔹}\;(\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑𝔹}\;\kern-1.5pt]_{\text{T}})\;{}\<[34]%
+\>[34]{}(\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\Varid{pt₂}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\Varid{pu₂}\;{}\<[E]%
+\\
+\>[34]{}(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\Varid{pb₂}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;(\Varid{elim𝔹[]}\;\Conid{P}\;\Varid{u}\;\Varid{t}\;\Varid{pu}\;\Varid{pt}\;\Varid{b}\;\Varid{pb}\;\Varid{pt₂}\;\Varid{pu₂}\;\Varid{pb₂}\;\Varid{q}\;\Varid{i})\;\mathrel{=}\;\Varid{q}\;\Varid{i}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Note again that we also defer the coherence proof of \ensuremath{\Varid{tyOf}} for \ensuremath{\Varid{elim𝔹[]}} by introducing another argument \ensuremath{\Varid{q}} in \ensuremath{\Varid{elim𝔹}} which can be removed when defining its elimination rule.
+
+\subsection{A Tarski universe} \label{sec:tt:univ}
+Using the same idiom described previously, a Tarski universe of types is introduced to our type theory in the same vein.
+First we need \ensuremath{\Conid{U}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ}} as the type of codes, and a type family \ensuremath{\Conid{El}} of elements for a given code (differences compared to the usual presentation highlighted):
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{U}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{U[]}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Conid{U}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{U}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{El}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;(\Varid{u}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pu}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{u}\;\mathop\equiv\;\Conid{U})}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{El[]}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\highlight{addition}{(\Varid{q}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;(\Conid{El}\;\Varid{u}\;\highlight{addition}{\Varid{pu}})\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{El}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{q}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+For the type \ensuremath{\Conid{𝔹}} of Boolean, its code \ensuremath{\Varid{𝕓}} is introduced with a type equality \ensuremath{\Conid{El𝕓}} such that the elements of \ensuremath{\Varid{𝕓}} are exactly \ensuremath{\Conid{𝔹}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{𝕓}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{𝕓[]}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Varid{𝕓}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\mathop\equiv\;\Varid{𝕓}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;\Varid{𝕓}\;{}\<[17]%
+\>[17]{}\mathrel{=}\;\Conid{U}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOf}\;(\Varid{𝕓[]}\;\Varid{σ}\;\Varid{i})\;{}\<[17]%
+\>[17]{}\mathrel{=}\;\Conid{U[]}\;\{\mskip1.5mu \Varid{σ}\;\mathrel{=}\;\Varid{σ}\mskip1.5mu\}\;\Varid{i}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{El𝕓}\;\mathbin{:}\;\Conid{El}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\Varid{𝕓}\;\Varid{refl}\;\mathop\equiv\;\Conid{𝔹}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+For the \ensuremath{\Conid{Π}}-type, we again need a specialised substitution lifting.
+This continues the pattern of introducing superfluous constructors to satisfy proof obligations (differences again highlighted).
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{El[]₂}\;\mathbin{:}\;(\Varid{u}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Δ})\;(\Varid{pu}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{u}\;\mathop\equiv\;\Conid{U})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\highlight{addition}{(\Varid{pu₂}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{tyOf}\;(\Varid{π₂}\;\{\mskip1.5mu \Conid{Γ}\;\Varid{,}\;\Conid{El}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pu₂}}\mskip1.5mu\}\;\Varid{idS})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathop\equiv\;\Conid{El}\;\Varid{u}\;\Varid{pu}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{\char95 ↑El}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\{\mskip1.5mu \Varid{u}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Δ}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\{\mskip1.5mu \Varid{pu}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{u}\;\mathop\equiv\;\Conid{U}\mskip1.5mu\}\;\highlight{addition}{\{\mskip1.5mu \Varid{pu'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{U}\mskip1.5mu\}}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\highlight{addition}{\Conid{El}\;(\Varid{u}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\Varid{pu'}})\;(\Conid{Δ}\;\Varid{,}\;\Conid{El}\;\Varid{u}\;\Varid{pu}){}\<[E]%
+\\
+\>[B]{}(\Varid{σ}\;\Varid{↑El})\;\{\mskip1.5mu \Varid{u}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{pu}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{pu'}\mskip1.5mu\}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\Varid{,}\;\Varid{π₂}\;\Varid{idS}\;\Varid{∶[}\;\highlight{addition}{\Conid{El[]₂}\;\Varid{u}\;\Varid{pu}\;\Varid{pu'}}\;\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Finally, we introduce the code \ensuremath{\Varid{π}} for \ensuremath{\Conid{Π}} and the type equality \ensuremath{\Conid{Elπ}} to complete our definition of type theory using QIIRT:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;{}\<[11]%
+\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π[]}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;{}\<[11]%
+\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[35]%
+\>[35]{}\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;{}\<[11]%
+\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;{}\<[35]%
+\>[35]{}\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;{}\<[11]%
+\>[11]{}\highlight{addition}{(\Varid{pa'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{a}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;{}\<[11]%
+\>[11]{}\highlight{addition}{(\Varid{pb'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑El}\;\kern-1.5pt]_{\text{t}})\;{}\<[43]%
+\>[43]{}\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;{}\<[11]%
+\>[11]{}(\Varid{π}\;\Varid{a}\;\Varid{pa}\;\Varid{b}\;\Varid{pb})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;{}\<[E]%
+\\
+\>[8]{}\mathop\equiv\;{}\<[11]%
+\>[11]{}\Varid{π}\;(\Varid{a}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pa'}}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑El}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pb'}}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;(\Varid{π}\;\anonymous \;\anonymous \;\anonymous \;\anonymous )\;\mathrel{=}\;\Conid{U}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Elπ}\;\mathbin{:}\;(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{El}\;(\Varid{π}\;\Varid{a}\;\Varid{pa}\;\Varid{b}\;\Varid{pb})\;\highlight{addition}{\Varid{refl}}\;\mathop\equiv\;\Conid{Π}\;(\Conid{El}\;\Varid{a}\;\Varid{pa})\;(\Conid{El}\;\Varid{b}\;\Varid{pb}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{tyOf}\;(\Varid{π[]}\;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\Varid{i})\;\mathrel{=}\;\Conid{U[]}\;\Varid{i}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+In the end, we emphasise that the introduction of superfluous equality proofs and constructors only makes sense because of set-truncation.
+These additional arguments are essentially unique and thus do not add any new laws to type theory, but merely serve as devices to meet the syntactic restriction of strict positivity in the current implementation of \CA.
+
+\subsection{Recursion and elimination principles} \label{sec:tt:elim}
+We turn to recursion and elimination principles for our syntax.
+Our QIIRT definition yields an \emph{initial model}.
+This means that for any other model (algebra) of our theory, there is a unique structure-preserving map from our syntax to that model.
+The recursion and elimination principles make this property concrete.
+Here, we only discuss the part for substitution calculus, since other type formers are addressed similarly.
+For the interested reader, see our formalisation.
+
+The signature for an algebra is packed in a record type~\ensuremath{\Conid{SC}} (short for Substitution Calculus).
+Inductive types and the function \ensuremath{\Varid{tyOf}} are interpreted as indexed types and a function between sets.
+Constructors of our syntax, except superfluous ones, correspond to function fields in this record, including equality constructors and clauses of \ensuremath{\Varid{tyOf}}.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{SC}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\Conid{Set}\;{}\<[19]%
+\>[19]{}\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Ctx}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Ty}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Ctx}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Tm}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Ctx}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Sub}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Ctx}\;\Varid{→}\;\Conid{Ctx}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{tyOf}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{∅}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\_,^{\mathsf{C}}\_\;{}\<[13]%
+\>[13]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\_[\_]_{\text{T}}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\_[\_]_{\text{t}}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Δ})\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{idS∘\char95 }\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Varid{idS}\;\Varid{∘}\;\Varid{σ}\;\mathop\equiv\;\Varid{σ}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{...}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{βπ₂}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Varid{π₂}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\mskip1.5mu])\;\mathop\equiv\;\Varid{t}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{...}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{tyOf[]}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Varid{tyOf}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;{}\<[36]%
+\>[36]{}\mathop\equiv\;(\Varid{tyOf}\;\Varid{t})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{tyOfπ₂}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Varid{tyOf}\;(\Varid{π₂}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;\Varid{σ})\;{}\<[36]%
+\>[36]{}\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+To distinguish syntactic constructors from the semantic methods in \ensuremath{\Conid{SC}}, we qualify the syntactic constructors with `\ensuremath{\Conid{S.}}' in the following discussion.
+
+Superfluous equality constructors, like \ensuremath{\Conid{S.tyOfπ₂idS}}, are not required as fields in the record.
+Instead, their semantic counterparts are defined within any given model using the other methods.
+For example,
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{tyOfπ₂idS}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{tyOf}\;(\Varid{π₂}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\mskip1.5mu\}\;\Varid{idS})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[3]{}\Varid{tyOfπ₂idS}\;\Varid{σ}\;\Conid{A}\;\mathrel{=}\;\Varid{tyOfπ₂}\;\Varid{idS}\;\Varid{∙}\;[\circ]_{\text{T}}\;\anonymous \;\anonymous \;\anonymous {}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+This simplifies the definition of models, as we only need to provide interpretations for the essential constructors.
+
+The recursion principle consists of a family of functions that map syntax to their semantic counterparts:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{recCtx}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{S.Ctx}\;{}\<[23]%
+\>[23]{}\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\Varid{recTy}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{S.Ty}\;\Conid{Γ}\;{}\<[23]%
+\>[23]{}\Varid{→}\;\Conid{Ty}\;(\Varid{recCtx}\;\Conid{Γ}){}\<[E]%
+\\
+\>[B]{}\Varid{recTm}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{S.Tm}\;\Conid{Γ}\;{}\<[23]%
+\>[23]{}\Varid{→}\;\Conid{Tm}\;(\Varid{recCtx}\;\Conid{Γ}){}\<[E]%
+\\
+\>[B]{}\Varid{recSub}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{S.Sub}\;\Conid{Γ}\;\Conid{Δ}\;{}\<[23]%
+\>[23]{}\Varid{→}\;\Conid{Sub}\;(\Varid{recCtx}\;\Conid{Γ})\;(\Varid{recCtx}\;\Conid{Δ}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We also need a function that translates proofs about syntactic equalities into semantic equalities:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{recTyOf}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{S.tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{B}\;\Varid{→}\;\Varid{tyOf}\;(\Varid{recTm}\;\Varid{t})\;\mathop\equiv\;\Varid{recTy}\;\Conid{B}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The definition of these functions proceeds by pattern matching on the syntactic structure.
+Each clause is an application of the corresponding method from the \ensuremath{\Conid{SC}} record:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{recCtx}\;\Conid{S.∅}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{∅}{}\<[E]%
+\\
+\>[B]{}\Varid{recCtx}\;(\Conid{Γ}\;\Conid{S.,}\;\Conid{A})\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{recCtx}\;\Conid{Γ}\;\Varid{,}^{\mathsf{C}}\;\Varid{recTy}\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\Varid{recSub}\;(\Varid{σ}\;\Conid{S.,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{recSub}\;\Varid{σ}\;\Varid{,}\;\Varid{recTm}\;\Varid{t}\;\Varid{∶[}\;\Varid{recTyOf}\;\Varid{t}\;\Varid{pt}\;\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The most interesting case is perhaps \ensuremath{\Varid{recTyOf}}, which handles the translation of syntactic equations.
+For a given syntactic equality proof \ensuremath{\Varid{p}\;\mathbin{:}\;\Conid{S.tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{B}}, we must construct a semantic equality proof.
+This is done by applying \ensuremath{\Varid{recTy}} to both sides of the syntactic equality to get \ensuremath{\Varid{recTy}\;(\Conid{S.tyOf}\;\Varid{t})\;\mathop\equiv\;\Varid{recTy}\;\Conid{B}}, and then using the semantic counterpart of the \ensuremath{\Varid{tyOf}} clause to derive \ensuremath{\Varid{tyOf}\;(\Varid{recTm}\;\Varid{t})\;\mathop\equiv\;\Varid{recTy}\;(\Conid{S.tyOf}\;\Varid{t})}.
+Taking \ensuremath{\Conid{S.π₂}} as an example, we have:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{recTyOf}\;\{\mskip1.5mu \Conid{B}\;\mathrel{=}\;\Conid{B}\mskip1.5mu\}\;(\Conid{S.π₂}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;\Varid{σ})\;\Varid{p}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{tyOf}\;(\Varid{recTm}\;(\Conid{S.π₂}\;\Varid{σ}))\;{}\<[33]%
+\>[33]{}\Varid{≡⟨⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{tyOf}\;(\Varid{π₂}\;(\Varid{recSub}\;\Varid{σ}))\;{}\<[33]%
+\>[33]{}\Varid{≡⟨}\;\Varid{tyOfπ₂}\;(\Varid{recSub}\;\Varid{σ})\;\Varid{⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}(\Varid{recTy}\;\Conid{A})\;[\kern-1.5pt\;\Varid{π₁}\;(\Varid{recSub}\;\Varid{σ})\;\kern-1.5pt]_{\text{T}}\;{}\<[33]%
+\>[33]{}\Varid{≡⟨⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{recTy}\;(\Conid{A}\;\Conid{S.[}\;\Conid{S.π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;{}\<[33]%
+\>[33]{}\Varid{≡⟨⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{recTy}\;(\Conid{S.tyOf}\;(\Conid{S.π₂}\;\Varid{σ}))\;{}\<[33]%
+\>[33]{}\Varid{≡⟨}\;\Varid{cong}\;\Varid{recTy}\;\Varid{p}\;\Varid{⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{recTy}\;\Conid{B}\;{}\<[33]%
+\>[33]{}\Varid{∎}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The coherence conditions for \ensuremath{\Varid{recTyOf}} over equality constructors are trivial because of set-truncation.
+
+For the elimination principle, we consider the notion of displayed algebras over an \ensuremath{\Conid{SC}}-algebra \ensuremath{\Conid{M}}, as a record type \ensuremath{\mathsf{SC}^{\bullet}} parametric in \ensuremath{\Conid{M}}, and later instantiate \ensuremath{\Conid{M}} to the term algebra, i.e.\ the syntax.
+Carriers of a displayed algebra as well as the semantics of \ensuremath{\Varid{tyOf}} are given below.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\mathsf{SC}^{\bullet}\;(\Conid{M}\;\mathbin{:}\;\Conid{SC})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{open}\;\Conid{SC}\;\Conid{M}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathsf{Ctx}^{\bullet}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\Conid{Ctx}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathsf{Ty}^{\bullet}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\mathsf{Ctx}^{\bullet}\;\Conid{Γ}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathsf{Tm}^{\bullet}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\mathsf{Ctx}^{\bullet}\;\Conid{Γ}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathsf{Sub}^{\bullet}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\mathsf{Ctx}^{\bullet}\;\Conid{Γ}\;\Varid{→}\;\mathsf{Ctx}^{\bullet}\;\Conid{Δ}\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathsf{tyOf}^{\bullet}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\mathsf{Tm}^{\bullet}\;{}\<[19]%
+\>[19]{}\Conid{Γ}^{\bullet}\;\Varid{t}\;\Varid{→}\;\mathsf{Ty}^{\bullet}\;\Conid{Γ}^{\bullet}\;(\Varid{tyOf}\;\Varid{t}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+As motives are indexed by their underlying model, we will have equations over equations of the underlying model.
+It is convenient to specialise dependent paths for them, e.g.,
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{\char95 ≡Tm[\char95 ]\char95 }\;\mathbin{:}\;\mathsf{Tm}^{\bullet}\;\Conid{Γ}^{\bullet}\;\Varid{t}\;\Varid{→}\;\Varid{t}\;\mathop\equiv\;\Varid{u}\;\Varid{→}\;\mathsf{Tm}^{\bullet}\;\Conid{Γ}^{\bullet}\;\Varid{u}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\Varid{\char95 ≡Tm[\char95 ]\char95 }\;\{\mskip1.5mu \Conid{Γ}^{\bullet}\;\mathrel{=}\;\Conid{Γ}^{\bullet}\mskip1.5mu\}\;\Varid{t}^{\bullet}\;\Varid{e}\;\Varid{u}^{\bullet}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\mathsf{Tm}^{\bullet}\;\Conid{Γ}^{\bullet}\;\highlight{addition}{(\Varid{e}\;\Varid{i})})\;\Varid{t}^{\bullet}\;\Varid{u}^{\bullet}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The signature for \ensuremath{\mathsf{SC}^{\bullet}}-algebras is similar to the one for \ensuremath{\Conid{SC}}-algebras.
+The difference here is that each displayed operation is indexed by their underlying operation and thus equations become equations over equations.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{53}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{∅}^{\bullet}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\mathsf{Ctx}^{\bullet}\;\Varid{∅}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\_,^{\bullet}\_\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\mathsf{Ctx}^{\bullet}\;\Conid{Γ}\;\Varid{→}\;\mathsf{Ty}^{\bullet}\;\Conid{Γ}^{\bullet}\;\Conid{A}\;\Varid{→}\;\mathsf{Ctx}^{\bullet}\;(\Conid{Γ}\;\Varid{,}^{\mathsf{C}}\;\Conid{A}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\_[\_]^{\bullet}_{\text{T}}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\mathsf{Ty}^{\bullet}\;\Conid{Δ}^{\bullet}\;\Conid{A}\;\Varid{→}\;\mathsf{Sub}^{\bullet}\;\Conid{Γ}^{\bullet}\;\Conid{Δ}^{\bullet}\;\Varid{σ}\;\Varid{→}\;\mathsf{Ty}^{\bullet}\;\Conid{Γ}^{\bullet}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\_[\_]^{\bullet}_{\text{t}}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\mathsf{Tm}^{\bullet}\;\Conid{Δ}^{\bullet}\;\Varid{t}\;\Varid{→}\;\mathsf{Sub}^{\bullet}\;\Conid{Γ}^{\bullet}\;\Conid{Δ}^{\bullet}\;\Varid{σ}\;\Varid{→}\;\mathsf{Tm}^{\bullet}\;\Conid{Γ}^{\bullet}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{tyOf[]}^{\bullet}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\mathsf{tyOf}^{\bullet}\;{}\<[23]%
+\>[23]{}(\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;\Varid{≡Ty[}\;\Varid{tyOf[]}\;\mskip1.5mu]\;(\mathsf{tyOf}^{\bullet}\;\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{T}}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{...}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}[\mathsf{idS}]_{\text{t}}^{\bullet}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Varid{t}^{\bullet}\;{}\<[38]%
+\>[38]{}\Varid{≡Tm[}\;[\mathsf{idS}]_{\text{t}}\;\mskip1.5mu]\;{}\<[53]%
+\>[53]{}\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{idS}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}[\circ]_{\text{t}}^{\bullet}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{τ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[38]%
+\>[38]{}\Varid{≡Tm[}\;[\circ]_{\text{t}}\;\mskip1.5mu]\;{}\<[53]%
+\>[53]{}\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{τ}^{\bullet}\;\Varid{∘}^{\bullet}\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Note that if \ensuremath{[\mathsf{idS}]_{\text{t}}} in its QIIT definition is formulated with a dependent path, the equality proof in the middle of \ensuremath{\Varid{\char95 ≡Tm[\char95 ]\char95 }} has to be a dependent path.
+As a result, we would have to specify two underlying equations as
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{t}^{\bullet}\;\Varid{≡Tm[}\;\highlight{addition}{[\mathsf{idS}]_{\text{T}}}\;\Varid{][}\;\highlight{addition}{[\mathsf{idS}]_{\text{t}}}\;\mskip1.5mu]\;\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{idS}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+and equational reasoning with them would involve three equations altogether.
+It is nice that we do not have to deal with this extra proof obligation in our formulation.
+
+The elimination principle is stated similarly to the recursion principle but indexed over the term algebra~(\Cref{sec:meta:term}),
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{31}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{elimCtx}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{S.Ctx})\;{}\<[31]%
+\>[31]{}\Varid{→}\;\mathsf{Ctx}^{\bullet}\;\Conid{Γ}{}\<[E]%
+\\
+\>[B]{}\Varid{elimTy}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{S.Ty}\;\Conid{Γ})\;{}\<[31]%
+\>[31]{}\Varid{→}\;\mathsf{Ty}^{\bullet}\;(\Varid{elimCtx}\;\Conid{Γ})\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\Varid{elimTm}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{S.Tm}\;\Conid{Γ})\;{}\<[31]%
+\>[31]{}\Varid{→}\;\mathsf{Tm}^{\bullet}\;(\Varid{elimCtx}\;\Conid{Γ})\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\Varid{elimSub}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{S.Sub}\;\Conid{Γ}\;\Conid{Δ})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\mathsf{Sub}^{\bullet}\;(\Varid{elimCtx}\;\Conid{Γ})\;(\Varid{elimCtx}\;\Conid{Δ})\;\Varid{σ}{}\<[E]%
+\\
+\>[B]{}\Varid{elimTyOf}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{S.Tm}\;\Conid{Γ})\;(\Varid{p}\;\mathbin{:}\;\Conid{S.tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{A})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;{}\<[6]%
+\>[6]{}\mathsf{tyOf}^{\bullet}\;(\Varid{elimTm}\;\Varid{t})\;\Varid{≡Ty[}\;\Varid{p}\;\mskip1.5mu]\;\Varid{elimTy}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+For the coherence conditions in the definition of the eliminators, we may need additional reasoning steps instead of just using the semantics equation, so we use transitivity of dependent paths:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\_\Varid{∙}_{\mathsf{P}}\_\;\mathbin{:}\;{}\<[9]%
+\>[9]{}\{\mskip1.5mu \Varid{x'}\;\mathbin{:}\;\Conid{B}\;\Varid{x}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{y'}\;\mathbin{:}\;\Conid{B}\;\Varid{y}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{z'}\;\mathbin{:}\;\Conid{B}\;\Varid{z}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\{\mskip1.5mu \Varid{p}\;\mathbin{:}\;\Varid{x}\;\mathop\equiv\;\Varid{y}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{q}\;\mathbin{:}\;\Varid{y}\;\mathop\equiv\;\Varid{z}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\Varid{p}\;\Varid{i}))\;\Varid{x'}\;\Varid{y'}\;\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\Varid{q}\;\Varid{i}))\;\Varid{y'}\;\Varid{z'}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\highlight{addition}{(\Varid{p}\;\Varid{∙}\;\Varid{q})}\;\Varid{i}))\;\Varid{x'}\;\Varid{z'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We also use set-truncation to identify the highlighted \ensuremath{\Varid{p}\;\Varid{∙}\;\Varid{q}} with the desired underlying equation in special-purpose equational reasoning combinators such as the following:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{beginSub[\char95 ]\char95 }\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\{\mskip1.5mu \Varid{p}\mskip1.5mu\}\;\Varid{q}\;\mathbin{:}\;\Varid{σ}\;\mathop\equiv\;\Varid{τ})\;\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{p}\;\mskip1.5mu]\;\Varid{τ}^{\bullet}\;\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{q}\;\mskip1.5mu]\;\Varid{τ}^{\bullet}{}\<[E]%
+\\
+\>[3]{}\Varid{beginSub[\char95 ]\char95 }\;\{\mskip1.5mu \Varid{σ}^{\bullet}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{τ}^{\bullet}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{p}\mskip1.5mu\}\;\Varid{q}\;\Varid{p}^{\bullet}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{subst}\;(\Varid{λ}\;\Varid{r}\;\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{r}\;\mskip1.5mu]\;\Varid{τ}^{\bullet})\;(\Conid{Sub-is-set}\;\Varid{p}\;\Varid{q})\;\Varid{p}^{\bullet}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+For example, the coherence proof for \ensuremath{\Varid{ηπ}} is given by
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{beginSub[}\;\Varid{ηπ}\;\mskip1.5mu]\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{elimSub}\;\Varid{σ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{≡Sub[}\;\Varid{ηπ}\;\Varid{]⟨}\;\Varid{ηπ∙}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π₁}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{,}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π₂}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{tyOfπ₂}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{≡Sub[}\;\Varid{refl}\;\Varid{]⟨}\;\Varid{cong}\;(\Varid{λ}\;\Varid{z}\;\Varid{→}\;\Varid{...}\;\Varid{,}\;\Varid{...}\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{z}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;(\Conid{Ty∙-is-set}\;{}\<[71]%
+\>[71]{}\anonymous \;\anonymous )\;\Varid{⟩}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{π₁}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{,}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{elimTm}\;(\Varid{π₂}\;\Varid{σ})\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{elimTyOf}\;(\Varid{π₂}\;\Varid{σ})\;\Varid{refl}\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{∎}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+with the steps \ensuremath{\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{p}\;\Varid{]⟨}\;\Varid{p}^{\bullet}\;\Varid{⟩}\;\Varid{τ}^{\bullet}} implemented using \ensuremath{\_\Varid{∙}_{\mathsf{P}}\_}.
+These steps give us an equation over \ensuremath{\Varid{ηπ}\;\Varid{∙}\;(\Varid{refl}\;\Varid{∙}\;\Varid{refl})}, and \ensuremath{\Varid{beginSub[}\;\Varid{ηπ}\;\Varid{]\char95 }} gives us an easy way to correct this to an equation over \ensuremath{\Varid{ηπ}} as desired instead.
+
+\FNF{Comment on the trick of including extra fields in the record type of models}
+
+\subsection{Practical workarounds for mutual definitions}  \label{sec:tt:mutual}
+
+So far, we have outlined how the recursion and elimination principles should be defined \emph{ideally}.
+In practice, however, limitations (and occasional mysterious bugs) of the proof assistant require us to adopt certain workarounds in order to implement the intended definitions.
+%This shows that the current design of \CA is not yet fully aligned with the theory of quotient inductive-inductive types~\cite{Kaposi2019}.
+
+\paragraph{Mutually interleaved QIITs}
+Constructors of QIITs currently can not be interleaved in \CA~\cite{Agdaissue2021}, even within an \ensuremath{\Keyword{interleaved}\;\Keyword{mutual}} block.
+The reason is that such a block is desugared into a collection of forward declarations for the \ensuremath{\Keyword{data}} types, rather than for the constructors.
+In principle, all constructors belonging to the same family of QIITs should be declared within the same context~\cite{Kaposi2019}.
+However, due to this desugaring, equality constructors may end up depending on other constructors that are not yet in scope.
+
+We work around this issue as follows:
+\begin{enumerate*}[label=(\roman*)]
+  \item we make forward declarations for the \emph{entire definition} of the QIITs, including the constructors;
+  \item we introduce each constructor but only refer to forward declarations when referring to other constructors;
+  \item we define the forward declarations by their corresponding constructors;
+  \item finally, we expose only the actual constructors, omitting the forward declarations.
+\end{enumerate*}
+The following snippet illustrates this approach:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{module}\;\Conid{S}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{data}\;\Conid{Ctx}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{∅}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Ctx}\mbox{\onelinecomment  note indentation, not a constructor!}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ,\char95 }\;{}\<[8]%
+\>[8]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{data}\;\Conid{Ctx}\;\Keyword{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{∅'}\;{}\<[11]%
+\>[11]{}\mathbin{:}\;\Conid{Ctx}\mbox{\onelinecomment  this is a constructor}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{\char95 ,'\char95 }\;{}\<[11]%
+\>[11]{}\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{∅}\;{}\<[11]%
+\>[11]{}\mathrel{=}\;\Varid{∅'}\mbox{\onelinecomment  make definition for forward declarations}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ,\char95 }\;{}\<[11]%
+\>[11]{}\mathrel{=}\;\Varid{\char95 ,'\char95 }{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\Keyword{open}\;\Conid{S}\;\Keyword{public}\;\mbox{\onelinecomment  expose actual constructors}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{hiding}\;(\Varid{∅};\Varid{\char95 ,\char95 };\Varid{...}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{renaming}\;(\Varid{∅'}\;\Varid{to}\;\Varid{∅};\Varid{\char95 ,'\char95 }\;\Varid{to}\;\Varid{\char95 ,\char95 };\Varid{...}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+This translation from QIITs in theory to their actual definitions in \CA should be sufficient to define mutually interleaved QIITs.
+Indeed, pedagogical presentations of type theory typically introduces one type former at a time, together with its formation, introduction, elimination, and equality rules (see, e.g.~\cite{Hofmann1997}), rather than presenting all type formers at once using a few large monolithic sets of rules.
+
+\paragraph{Mutual interleaved QIIRTs}
+Interleaving function clauses with inductive types is a different matter, since we cannot forward declare a function clause together with its computational behaviour.\footnote{Custom rewrite rules are not allowed in \CA.}
+However, because we have `Forded' the typing constraints into equality proofs, what we actually need at the point of introducing constructors is only the existence of such an equality proof, not its computational content.
+
+Our workaround is therefore as follows:
+\begin{enumerate*}[label=(\roman*)]
+  \item we declare the existence of the required equality proof before it is used,
+  \item we define \ensuremath{\Varid{tyOf}} only after all datatype declarations have been given, and
+  \item we provide the actual definition corresponding to the forward declaration.
+\end{enumerate*}
+For instance, the equality constructor \ensuremath{\Varid{ηπ}} asks for a proof of \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}}.
+In this case, we simply forward declare such a proof:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{tyOfπ₂}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\Varid{ηπ}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Varid{σ}\;\mathop\equiv\;(\Varid{π₁}\;\Varid{σ}\;\Varid{,}\;\Varid{π₂}\;\Varid{σ}\;\Varid{∶[}\;\Varid{tyOfπ₂}\;\mskip1.5mu]){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Then, once \ensuremath{\Varid{tyOf}} has been defined, we simply set \ensuremath{\Varid{tyOfπ₂}} to \ensuremath{\Varid{refl}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{tyOf}\;(\Varid{π₂'}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{Δ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{A}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{σ}\mskip1.5mu\})\;\mathrel{=}\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}{}\<[E]%
+\\
+\>[B]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOfπ₂}\;\mathrel{=}\;\Varid{refl}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+This translation is valid as long as the computational behaviour of the interleaved function clauses is not needed up to judgemental equality.
+
+\paragraph{Mutually-defined functions}
+%\LT[noinline]{(Low) Create an Agda issue?}
+Since the constructors of QII(R)Ts can be mutually interleaved, their recursion and elimination principles also need to be given in the same vein.
+However, \Agda does not allow us to interleave clauses of different functions directly.
+One workaround is to use forward declarations as a lifting of the entire clause and then perform the necessary coercions along the corresponding equality proofs by hand.
+
+Another possibility is to define a single family of functions indexed by tags.
+For instance, the recursion principle can be implemented by introducing a datatype \ensuremath{\Conid{Tag}} with one constructor for each motive:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Conid{Tag}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{ctx}\;\Varid{ty}\;\Varid{sub}\;\Varid{tm}\;\Varid{tyof}\;\mathbin{:}\;\Conid{Tag}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Then, we define the recursion principle uniformly as \ensuremath{\Varid{rec}}, with \ensuremath{\Varid{tyOfRec}} computing its type.
+Each actual function is introduced as a synonym for \ensuremath{\Varid{rec}} at the appropriate tag:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{tyOfRec}\;\mathbin{:}\;\Conid{Tag}\;{}\<[18]%
+\>[18]{}\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Varid{rec}\;\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tag})\;{}\<[18]%
+\>[18]{}\Varid{→}\;\Varid{tyOfRec}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\Varid{tyOfRec}\;\Varid{ctx}\;{}\<[15]%
+\>[15]{}\mathrel{=}\;\Conid{S.Ctx}\;\Varid{→}\;\Conid{Ctx}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{rec}\;\Varid{ctx}\;\Conid{S.∅}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{∅}{}\<[E]%
+\\
+\>[B]{}\Varid{rec}\;\Varid{ctx}\;(\Conid{Γ}\;\Conid{S.,}\;\Conid{A})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{rec}\;\Varid{ctx}\;\Conid{Γ}\;\Varid{,}^{\mathsf{C}}\;\Varid{rec}\;\Varid{ty}\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\Varid{...}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{recCtx}\;{}\<[10]%
+\>[10]{}\mathrel{=}\;\Varid{rec}\;\Varid{ctx}{}\<[E]%
+\\
+\>[B]{}\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+At the time of writing, however, this encoding cannot be fully carried out in \CA: some terms that should be strictly equal are not recognised as such during type checking.
+For example, in the following clause of \ensuremath{\Varid{rec}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{rec}\;\Varid{sub}\;(\Conid{S.\char95 ,\char95 ∶[\char95 ]}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{Δ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{A}\mskip1.5mu\}\;\Varid{σ}\;\Varid{t}\;\Varid{p})\;\mathrel{=}\;\Varid{\char95 ,\char95 ∶[\char95 ]}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\{\mskip1.5mu \Varid{rec}\;\Varid{ctx}\;\Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{rec}\;\Varid{ctx}\;\Conid{Δ}\mskip1.5mu\}\;\{\mskip1.5mu \Varid{rec}\;\Varid{ty}\;\Conid{A}\mskip1.5mu\}\;(\Varid{rec}\;\Varid{sub}\;\Varid{σ})\;(\Varid{rec}\;\Varid{tm}\;\Varid{t})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\highlight{addition}{\{\mskip1.5mu \mathbin{!}\;\Varid{rec}\;\Varid{tyof}\;\Varid{t}\;\Varid{p}\;\mathbin{!}\mskip1.5mu\}}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+the subterm in the hole is accepted by \Agda, but refining it results an error, as the terms \ensuremath{\Varid{rec}\;\Varid{ty}\;(\Conid{A}\;\Conid{S.[}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})} and \ensuremath{\Varid{rec}\;\Varid{ty}\;\Conid{A}\;[\kern-1.5pt\;\Varid{rec}\;\Varid{sub}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} are not recognised as equal --- even though the former was already defined to be the latter.
+
+In our formalisation, we fall back on forward declarations alone with coercions.
+We are still investigating the root cause of this behaviour, but it may point to a design flaw.
+
+\LT[inline]{TERMINATION checker? (R1, R3)}
+
+\section{Metatheory}\label{sec:meta}
+Having defined type theory as QIIRTs, we now turn to models of type theory as well as constructions of new models from existing ones.
+
+%We find that reasoning with this definition \emph{per se} in \CA is hard even with substitution calculus: the lack of strict equalities in type theory is known to cause the transport hell~\cite{Altenkirch2016a} and even worse it cannot be mitigated by using custom rewriting rules~\cite{Cockx2020,Cockx2021} in \CA.
+%Nevertheless, programming with type theory seems doable, as normalisation can be
+
+\subsection{Term model} \label{sec:meta:term}
+The term model is a self-interpretation of syntax, allowing displayed models to be instantiated over it and ensuring that the elimination rule computes.
+The definition is routine: each field is interpreted by the corresponding constructor, except that additional equality constraints (such as the one in \ensuremath{\Varid{βπ₂}} highlighted below) are replaced by actual proofs:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Term}\;\mathbin{:}\;\Conid{SC}{}\<[E]%
+\\
+\>[B]{}\Conid{Term}\;\mathrel{=}\;\Keyword{record}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\{\mskip1.5mu \Varid{∅}\;{}\<[13]%
+\>[13]{}\mathrel{=}\;\Conid{S.∅}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{tyOf[]}\;{}\<[13]%
+\>[13]{}\mathrel{=}\;\Varid{refl}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{...}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{βπ₂}\;{}\<[13]%
+\>[13]{}\mathrel{=}\;\Varid{λ}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{Δ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{A}\mskip1.5mu\}\;\Varid{σ}\;\Varid{t}\;\Varid{p}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{S.βπ₂}\;\Varid{σ}\;\Varid{t}\;\Varid{p}\;\highlight{addition}{(\Varid{cong}\;(\Conid{A}\;\Varid{[\char95 ]T})\;(\Conid{S.βπ₁}\;\Varid{σ}\;\Varid{t}\;\Varid{p})\;\Varid{∙}\;\Varid{sym}\;\Varid{p})}\mskip1.5mu\}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Other type formers are given similarly.
+The recursion and elimination principles justify that the term model is indeed the initial model.
+
+\subsection{Standard model} \label{sec:standard-model}
+
+In the standard model, contexts are interpreted as sets in \CA, types as sets indexed by a context~\ensuremath{\Conid{Γ}}, substitutions as functions between these sets, and terms as \emph{pairs} \ensuremath{(\Conid{A,}\;\Varid{t})} consisting of an interpreted type \ensuremath{\Conid{A}\;\mathbin{:}\;\Conid{Γ}\;\Varid{→}\;\Conid{Set}} together with a dependent function \ensuremath{\Varid{t}\;\mathbin{:}\;(\Varid{γ}\;\mathbin{:}\;\Conid{Γ})\;\Varid{→}\;\Conid{A}\;\Varid{γ}}.
+The interpretation of \ensuremath{\Varid{tyOf}} is simply the first component \ensuremath{\Conid{A}}.
+In other words, terms are interpreted as a type \ensuremath{\Conid{A}}, together with a `section' \ensuremath{\Varid{t}\;\mathbin{:}\;(\Varid{γ}\;\mathbin{:}\;\Conid{Γ})\;\Varid{→}\;\Conid{A}\;\Varid{γ}} of that type, as usual:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{std}\;\mathbin{:}\;\Conid{SC}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.Ctx}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.Ty}\;{}\<[12]%
+\>[12]{}\Conid{Γ}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{Γ}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.Sub}\;\Conid{Γ}\;\Conid{Δ}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{Γ}\;\Varid{→}\;\Conid{Δ}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.Tm}\;{}\<[12]%
+\>[12]{}\Conid{Γ}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{Σ[}\;\Conid{A}\;\Varid{∈}\;(\Conid{Γ}\;\Varid{→}\;\Conid{Set})\;\mskip1.5mu]\;((\Varid{γ}\;\mathbin{:}\;\Conid{Γ})\;\Varid{→}\;\Conid{A}\;\Varid{γ}){}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.tyOf}\;(\Conid{A}\;\Varid{,}\;\Varid{t})\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+The main construction of the model is the same as in the standard model of type theory using QIITs~\cite[Section~4]{Altenkirch2017}, except that the typing constraint \ensuremath{\Varid{p}} in \ensuremath{\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\mskip1.5mu]} is `Forded'.
+As a result, the value \ensuremath{\Varid{t}\;\Varid{γ}} below must be transported along \ensuremath{\Varid{p}}, as highlighted below:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{std}\;\Varid{.∅}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{Unit}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.\char95 ,C\char95 }\;\Conid{Γ}\;\Conid{A}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{Σ}\;\Conid{Γ}\;\Conid{A}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.\char95 [\char95 ]T}\;\Conid{A}\;\Varid{σ}\;\Varid{γ}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{A}\;(\Varid{σ}\;\Varid{γ}){}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.\char95 [\char95 ]t}\;(\Conid{A}\;\Varid{,}\;\Varid{t})\;\Varid{σ}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;(\Varid{λ}\;\Varid{γ}\;\Varid{→}\;\Conid{A}\;(\Varid{σ}\;\Varid{γ}))\;\Varid{,}\;(\Varid{λ}\;\Varid{γ}\;\Varid{→}\;\Varid{t}\;(\Varid{σ}\;\Varid{γ})){}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.tyOf[]}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Varid{refl}{}\<[E]%
+\\
+\>[3]{}\Varid{...}{}\<[E]%
+\\
+\>[3]{}\Varid{std}\;\Varid{.\char95 ,\char95 ∶[\char95 ]}\;\Varid{σ}\;(\Conid{A}\;\Varid{,}\;\Varid{t})\;\Varid{p}\;\Varid{γ}\;\mathrel{=}\;(\Varid{σ}\;\Varid{γ}\;\Varid{,}\;\highlight{addition}{\Varid{transport}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{p}\;\Varid{i}\;\Varid{γ})\;(\Varid{t}\;\Varid{γ})}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+To extend the standard model for the universe \ensuremath{\Conid{U}}, we define a Tarski universe of codes and its interpretation
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Conid{UU}\;\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Conid{T}\;\mathbin{:}\;\Conid{UU}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\Conid{UU}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{bool}\;\mathbin{:}\;\Conid{UU}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{pi}\;\mathbin{:}\;(\Varid{a}\;\mathbin{:}\;\Conid{UU})\;\Varid{→}\;(\Conid{T}\;\Varid{a}\;\Varid{→}\;\Conid{UU})\;\Varid{→}\;\Conid{UU}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Conid{T}\;\mathbin{:}\;\Conid{UU}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Conid{T}\;\Varid{bool}\;{}\<[13]%
+\>[13]{}\mathrel{=}\;\Conid{Bool}{}\<[E]%
+\\
+\>[B]{}\Conid{T}\;(\Varid{pi}\;\Varid{a}\;\Varid{b})\;{}\<[13]%
+\>[13]{}\mathrel{=}\;(\Varid{x}\;\mathbin{:}\;\Conid{T}\;\Varid{a})\;\Varid{→}\;\Conid{T}\;(\Varid{b}\;\Varid{x}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Each of the constructs in \Cref{sec:tt:univ} can now be interpreted:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{std}\;{}\<[6]%
+\>[6]{}\Varid{.U}\;{}\<[11]%
+\>[11]{}\anonymous \;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{UU}{}\<[E]%
+\\
+\>[B]{}\Varid{std}\;{}\<[6]%
+\>[6]{}\Varid{.U[]}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Varid{refl}{}\<[E]%
+\\
+\>[B]{}\Varid{std}\;{}\<[6]%
+\>[6]{}\Varid{.El}\;{}\<[11]%
+\>[11]{}(\Conid{A}\;\Varid{,}\;\Varid{u})\;\Varid{pu}\;\Varid{γ}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;\Conid{T}\;(\Varid{subst}\;(\Varid{λ}\;\Conid{A}\;\Varid{→}\;\Conid{A}\;\Varid{γ})\;\Varid{pu}\;(\Varid{u}\;\Varid{γ})){}\<[E]%
+\\
+\>[B]{}\Varid{std}\;{}\<[6]%
+\>[6]{}\Varid{.𝕓}\;{}\<[25]%
+\>[25]{}\mathrel{=}\;(\Varid{λ}\;\anonymous \;\Varid{→}\;\Conid{UU})\;\Varid{,}\;\Varid{λ}\;\anonymous \;\Varid{→}\;\Varid{bool}{}\<[E]%
+\\
+\>[B]{}\Varid{std}\;{}\<[6]%
+\>[6]{}\Varid{.π}\;(\Conid{A}\;\Varid{,}\;\Varid{a})\;\Varid{pa}\;(\Conid{B}\;\Varid{,}\;\Varid{b})\;\Varid{pb}\;\mathrel{=}\;(\Varid{λ}\;\anonymous \;\Varid{→}\;\Conid{UU})\;\Varid{,}\;\Varid{λ}\;\Varid{γ}\;\Varid{→}\;\Varid{pi}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(\Varid{transport}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{pa}\;\Varid{i}\;\Varid{γ})\;(\Varid{a}\;\Varid{γ}))\;{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(\Varid{λ}\;\Varid{a}\;\Varid{→}\;\Varid{transport}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{pb}\;\Varid{i}\;(\Varid{γ}\;\Varid{,}\;\Varid{a}))\;(\Varid{b}\;(\Varid{γ}\;\Varid{,}\;\Varid{a}))){}\<[E]%
+\\
+\>[B]{}\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Coherence conditions are then verified using standard properties of transport.
+We have formalised the standard model for type theory with $\Pi$-types, Booleans, and a Tarski universe, except the case for \ensuremath{\Varid{π[]}}.
+
+The main effort in the formalisation arises from the lack of \emph{regularity}~\cite{Sterling2022}: there is a path \ensuremath{\Varid{transportRefl}} between the transport along reflexivity and the identity but they are not strictly equal.
+For instance, the coherence condition for \ensuremath{\Conid{Π[]}} is given as
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{stdPi}\;\Varid{.Π[]}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{Δ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{A}\mskip1.5mu\}\;\Varid{σ}\;\Conid{B}\;\Varid{i}\;\Varid{γ}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}(\Varid{a}\;\mathbin{:}\;\Conid{A}\;(\Varid{σ}\;\Varid{γ}))\;\Varid{→}\;\Conid{B}\;(\Varid{σ}\;\Varid{γ}\;\Varid{,}\;\Varid{transportRefl³}\;\Varid{a})\;(\mathord{\sim}\;\Varid{i})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where \ensuremath{\Varid{transportRefl³}} amounts to using \ensuremath{\Varid{transportRefl}} three times.
+If regularity were available, this would collapse to the trivial reflexivity proof.
+%The case |π[]| above involves an equation over a transport of another transported term.
+%\LT[noinline]{try to finish it if we have time.}
+
+\subsection{Normalisation by evaluation, and the logical predicate interpretation} \label{sec:nbe}
+We implement normalisation by evaluation (NbE) for the substitution calculus.
+Following the approach for type theory~\cite{Altenkirch2017}, we define inductive-recursively both the type of variables (with their embedding into terms) and the type of renamings (with their embedding into substitutions).
+The implementation is straightforward, so we omit the details here.
+In the end, this yields a normalisation function that produces, for every term, a de Bruijn variable and computes:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{normalise}\;\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Conid{Σ[}\;\Varid{tⁿ}\;\Varid{∈}\;\Conid{NeTm}\;\Conid{Γ}\;\mskip1.5mu]\;\Varid{t}\;\mathop\equiv\;\Varid{⌜}\;\Varid{tⁿ}\;\Varid{⌝}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Compared to NbE for substitution calculus using QIITs, our formalisation is simpler: no transports appear at all, because variables and terms are not indexed by their types.
+
+The picture is very different for the logical predicate interpretation.
+Although NbE works cleanly, the logical predicate interpretation --- often considered a benchmark challenge~\cite{Abel2019} for language formalisation --- remains at least as difficult as in the QIIT-based setting, even for substitution calculus.
+
+To see why, recall that the motives for \ensuremath{\Conid{Ctx}} and \ensuremath{\Conid{Ty}} in the logical predicate interpretation are given by
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{Ctxᴾ}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{ctxᴾ}\;\mathbin{:}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{wkᴾ}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Conid{Sub}\;\Varid{ctxᴾ}\;\Conid{Γ}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Conid{Tyᴾ}\;\mathbin{:}\;\Conid{Ctxᴾ}\;\Conid{Γ}\;\Varid{→}\;\Conid{Ty}\;\Conid{Γ}\;\Varid{→}\;{}\<[24]%
+\>[24]{}\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Conid{Tyᴾ}\;\Conid{Γᴾ}\;\Conid{A}\;\mathrel{=}\;\Conid{Ty}\;(\Varid{ctxᴾ}\;\Conid{Γᴾ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{wkᴾ}\;\Conid{Γᴾ}\;\kern-1.5pt]_{\text{T}}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Here the typing constraint \ensuremath{(\Varid{ctxᴾ}\;\Conid{Γᴾ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{wkᴾ}\;\Conid{Γᴾ}\;\kern-1.5pt]_{\text{T}})} already shares the familiar shape of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}}, but with an additional complication: the index explicitly demands a type substitution.
+Since the QIIRT representation only provides equality constructors for type substitutions, the development quickly results in repeated and tedious use of transports.
+
+In short, NbE benefits directly from removing typing indices and avoids transports altogether, whereas the logical predicate interpretation still inherits the need for coercions with type substitutions.
+We did not bother to finish all cases of the logical predicate interpretation, as Altenkirch and Kaposi~\cite{Altenkirch2016a} have already shown that such tedious use of transports is possible (but impractical) in theory.
+
+\subsection{Strictification} \label{sec:strictify}
+\FNF[inline]{Work further on the idea of strictification? (R2)}
+Instead, we turn our attention to \emph{strictification}~\cite{Donko2022,Kaposi2025}: given a model of type theory, certain equality constructors can be made strict (i.e., made to hold judgementally) to form a new model.
+A familiar analogy is the transition from lists to difference lists, where a list is represented by a list-appending function and list concatenation is function composition, which is associative judgementally.
+
+In the same spirit, we may attempt to `strictify' the category part of substitution calculus using the Yoneda embedding, so that the unit laws and associativity law hold strictly, \emph{modulo} the property of naturality.
+Given any \ensuremath{\Conid{SC}}-algebra, we define a presheaf interpretation of \ensuremath{\Conid{Sub}} by
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{Subʸ}\;(\Conid{Γ}\;\Conid{Δ}\;\mathbin{:}\;\Conid{Ctx})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{constructor}\;\Varid{\char95 ,\char95 }{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{y}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;\Varid{∀}\;\{\mskip1.5mu \Conid{Θ}\mskip1.5mu\}\;\Varid{→}\;\Conid{Sub}\;\Conid{Θ}\;\Conid{Γ}\;\Varid{→}\;\Conid{Sub}\;\Conid{Θ}\;\Conid{Δ}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{nat}\;{}\<[10]%
+\>[10]{}\mathbin{:}\;(\Varid{τ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Ξ}\;\Conid{Θ})\;(\Varid{δ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Θ}\;\Conid{Γ})\;\Varid{→}\;\Varid{y}\;\Varid{δ}\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;\Varid{y}\;(\Varid{δ}\;\Varid{∘}\;\Varid{τ}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+By set-truncation, any two morphisms \ensuremath{\Varid{σ}} and \ensuremath{\Varid{σ'}} in the presheaf category are equal whenever their functionals agree:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{≡ʸ→≡}\;\mathbin{:}\;\{\mskip1.5mu \Varid{σ}\;\Varid{σ'}\;\mathbin{:}\;\Conid{Subʸ}\;\Conid{Γ}\;\Conid{Δ}\mskip1.5mu\}\;\Varid{→}\;\Varid{σ}\;\Varid{≡ʸ}\;\Varid{σ'}\;\Varid{→}\;\Varid{σ}\;\mathop\equiv\;\Varid{σ'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where \ensuremath{\Varid{\char95 ≡ʸ\char95 }} denotes the path type between their functionals.
+Completing the Yoneda embedding then gives us strict unit and associativity laws, up to \ensuremath{\Varid{≡ʸ→≡}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Yoneda}\;\Varid{.idS∘\char95 }\;{}\<[17]%
+\>[17]{}\anonymous \;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Varid{≡ʸ→≡}\;\Varid{refl}{}\<[E]%
+\\
+\>[B]{}\Conid{Yoneda}\;\Varid{.\char95 ∘idS}\;{}\<[17]%
+\>[17]{}\anonymous \;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Varid{≡ʸ→≡}\;\Varid{refl}{}\<[E]%
+\\
+\>[B]{}\Conid{Yoneda}\;\Varid{.assocS}\;{}\<[17]%
+\>[17]{}\anonymous \;\anonymous \;\anonymous \;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Varid{≡ʸ→≡}\;\Varid{refl}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We can adapt the local universe construction~\cite{Lumsdaine2015,Donko2022} for \ensuremath{\Conid{M}\;\mathbin{:}\;\Conid{SC}} to further strictify the laws for type substitutions for substitution.
+Types under the context $\Gamma$ in the local universe construction are interpreted as a triple consisting of a context $V$ as the local universe, a type under the context $V$, and a substitution from $\Gamma$ to $V$:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{Ty³}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Ctx})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{constructor}\;\Varid{ty³}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{V}\;\mathbin{:}\;\Conid{Ctx}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{E}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Conid{Ty}\;\Conid{V}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{⌜\char95 ⌝}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{V}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Type substitution is defined by `delaying' the substitution, i.e., by viewing the substitution \ensuremath{\Varid{⌜}\;\Conid{A}\;\Varid{⌝}} as an accumulator parameter.
+Then, we can show that the laws about type substitution boil down the right unit and the associativity laws for substitutions:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}(\Conid{M}\;\mathbin{!})\;\Varid{.[idS]T}\;\{\mskip1.5mu \Conid{Γ}\mskip1.5mu\}\;\{\mskip1.5mu \Conid{A}\mskip1.5mu\}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{cong}\;(\Varid{ty³}\;(\Conid{A}\;\Varid{.V})\;(\Conid{E}\;\Conid{A}))\;\highlight{addition}{(\Varid{sym}\;(\Varid{⌜}\;\Conid{A}\;\Varid{⌝}\;\Varid{∘idS}))}{}\<[E]%
+\\
+\>[3]{}(\Conid{M}\;\mathbin{!})\;\Varid{.[∘]T}\;\Conid{A}\;\Varid{σ}\;\Varid{τ}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{cong}\;(\Varid{ty³}\;(\Conid{A}\;\Varid{.V})\;(\Conid{E}\;\Conid{A}))\;\highlight{addition}{(\Varid{assocS}\;\Varid{σ}\;\Varid{τ}\;\Varid{⌜}\;\Conid{A}\;\Varid{⌝})}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+If \ensuremath{\Conid{M}} is strictified by the Yoneda embedding, then the laws for identity substitution and substitution composition in \ensuremath{\Conid{M}\;\mathbin{!}} will be strict up to \ensuremath{\Varid{≡ʸ→≡}}.
+Hence, combining both techniques, we can construct models where both the category laws and substitution laws are strict after applying \ensuremath{\Varid{≡ʸ→≡}}.
+
+Nevertheless, strictification does \emph{not} resolve our difficulties with the logical predicate interpretation.
+In particular, the lack of strict propositions (or just definitional UIP) in \CA prevents \ensuremath{\Varid{σ}\;\Varid{≡ʸ}\;\Varid{σ'}} from being strictly equal, since the paths between their properties must still be identified manually using set-truncation.
+Although \Agda provides a form of strict propositions \ensuremath{\Conid{Prop}}, it is not designed to work with \CA and interacts poorly with cubical primitives for now~\cite{Agda-issue2022}.
+As a consequence, coercions along equations identified by set-truncation remain unavoidable.
+
+\section{Discussion and conclusion}
+\label{sec:discussion}
+
+%\FNF{Also discuss alternative choices such as basing the formalisation on (split) comprehension categories, see Brunerie and de Boer~\cite{Boer2020,Brunerie2019} (R2)}
+
+It is well known that type theory in type theory is possible in theory, but in practice its formalisation often requires giving up some of the support and safety checks provided by proof assistants.
+From one point of view, our work addresses the following question: is there any existing type-theoretic proof assistant that can formalise the intrinsic representation of type theory using QIITs reliably, without compromise?
+Based on our experimental formalisation in \CA, our answer is regrettably: not yet.
+
+\paragraph{Comparison with intrinsic approaches}
+Previous intrinsic formalisations~\cite{Kaposi2019,Kaposi2017,Altenkirch2016a,Altenkirch2017} based on the CwF semantics of type theory have resorted to using postulated constructors and custom rewrite rules to manually define QIITs and their eliminators.
+However, this comes at a cost: the proof assistant no longer performs strict positivity, coverage, or termination checks for functions defined from quotient inductive types, nor does it supports dependent pattern matching, program extraction, and interactive theorem proving.
+The loss of coverage check for inductive types is mitigated by using hand-crafted eliminators (see \Cref{sec:tt:elim}), since the coverage check is also in effect for record types.
+
+The Ehrhard's style of presentation~\cite{Coquand2020,Ehrhard1988}, whereas the set of terms remains indexed by types but without any transport in the middle of terms for specifying equations, can be stated by using dependent paths and has been adopted by a recent formalisation~\cite{Altenkirch2026} in \CA.
+
+Our approach leads to no transports appearing in its own syntax and also avoids the use of dependent paths at all.
+However, the same transports (and the same equations for them) seem to have a tendency to come back in concrete model constructions, as explained in \cref{sec:meta}.
+
+We remark that the lack of transports is an advantage for avoiding strict positivity issues in the current implementation of \CA.
+By using native features such as higher inductive types, rather than postulates in ordinary \Agda, we do get computational interpretations.
+For example, our implementation of normalization by evaluation in \cref{sec:nbe} can actually normalise terms and could be potentially extracted as Haskell programs with the cubical information explicitly erased using \Agda's \text{\ttfamily \char45{}\char45{}erased\char45{}cubical} feature.
+
+It is not yet clear if other initial models of type theory, such as (split) comprehension categories~\cite{Jacobs1993} and contextual categories~\cite{Cartmell1986}, can be defined as QIITs or QIIRTs.
+
+\paragraph{Strictification}
+Kaposi and Pujet~\cite{Kaposi2025} have shown how strictification techniques can simplify proofs, but this is an orthogonal issue.
+Even though strictification turns most of the equality constructors about substitution to strict equalities, this does not help with transports appearing in the QIIT definition of terms and the resulting strict positivity issues, as strictification can only be applied \emph{after} the inductive types are defined.
+In \cref{sec:strictify}, we have sketched how also our notion of models can be strictified using a simpler idea.
+However, a proper strictification may require a different metatheory than `just' the support of QII(R)Ts (see below).
+
+% Untyped version might still be easiest to work with, with current proof assistant technology
+
+\paragraph{A proper syntax for HII(R)Ts}
+The syntax for declaring higher inductive-inductive types and QIITs in \CA falls short of their theoretical capabilities~\cite{Kaposi2020b,Kaposi2019}.
+As shown in \Cref{sec:tt:cwf}, the legitimate definition of type theory based on CwF with transports violates the syntactic restriction of strict positivity in \CA.
+Even though transport hell is better avoided in practice, this is not a justification for excluding otherwise valid definitions.
+The alternative definition, based on natural models, does not violate strict positivity but still requires forward declarations (\Cref{sec:tt:mutual}) to overcome the inconvenience of syntax.
+
+\paragraph{A theory of QIIRTs}
+We work around the problem by defining type theory using QIIRTs, but this raises another question: what is the theory of QIIRTs, anyway?
+
+Different forms of QIITs and inductive-recursive types (IRTs) have been used to develop type theory in type theory.
+Large and infinitary IRTs are used for the standard model of universe; small inductive-recursive types~\cite{Hancock2013} for implementing normalisation by evaluation~\cite{Altenkirch2017}; QIITs and QIIRTs for the intrinsic representations of type theory.
+Note further that the target of \ensuremath{\Varid{tyOf}} is the inductive type \ensuremath{\Conid{Ty}} being defined simultaneously, so the QIIRTs used to define our representation is neither large nor small.
+Nevertheless, we expect that the encoding of small inductive-recursive types as indexed inductive types still applies to this `self-referring' induction-recursion, as we have derived our representation by `undoing' the encoding.
+
+A framework to encompass all variants will be convenient for proof assistant implementation.
+Such a scheme may be developed by extending the type theory of QIITs~\cite{Kaposi2019,Kovacs2020a} with an additional construct to distinguish strict and weak equalities and a coverage condition for function clauses.
+
+\paragraph{The support of interleaved mutual definitions}
+Another challenge concerns interleaved mutual definitions.
+Since constructors of QIITs may be mutually interleaved, the elaboration from dependent pattern matching to eliminators need to take this into account.
+Our workaround, using forward declarations to lift function clauses to fix the dependency, sacrifices their computational behaviour.
+Furthermore, our definitions appear to reach the limits of the termination checker: even seemingly harmless functions when defining the recursion principle fail to pass termination checking.
+
+% \paragraph{The need for strict propositions}
+\paragraph{Strict propositions, and observational type theory}
+The recent work on strictified syntax~\cite{Kaposi2025} addresses transport hell using observational type theory (OTT)~\cite{Pujet2022,Pujet2023,Pujet2024}, with strict propositions in the metatheory playing a central role.
+While \Agda does support strict propositions~\cite{Gilbert2019}, this feature was not designed to work with \CA~\cite{Agda-issue2022}.
+%
+%\paragraph{The implementation of OTT or XTT}
+To formalise the metatheory of type theory using QIITs with as few transports as possible, it seems inevitable to use a different metatheory rather than the off-the-shelf metatheory offered by \CA.
+In particular, regularity and definitional UIP, supported by OTT (see \cite{Altenkirch2019,Pujet2024} for the discussion on regularity) and by its cubical variant XTT~\cite{Sterling2022}, would immediately simplify our standard model~(\Cref{sec:standard-model}) and make formalising metatheory more feasible using Kaposi and Pujet's strictification.
+
+The use of QIITs in OTT~\cite{Kaposi2025} in \Agda requires the user themselves to implement the coercion rules for inductive types~\cite{Pujet2024} as well as their elimination principles.
+Quotient inductive types are not supported in the implementation of OTT in Rocq~\cite{Pujet2024a} and its theory is still being developed~\cite{Felicissimo2025a}.
+
+The option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} in the forthcoming \Agda 2.9.0~\cite{Agda-issue2025} disables the \ensuremath{\Conid{Glue}} type in cubical mode and, in principle, yields a cubical type theory compatible with UIP~\cite{Agda-issue2019}.
+Since Cubical Agda already provides support for HIITs (with the earlier caveats), realising a type theory with QIITs, strict propositions, and computational UIP, as a variant of Cubical Agda may now be within reach~\cite{Tan2025,Tan2025a}.
+
+\paragraph{The Ford transformation and definitional UIP}
+The Ford transformation is known to work well with definitional UIP~\cite{Altenkirch2006}.
+So far, we have only `Forded' the \text{\ttfamily Tm} constructors, but what if every constructor were Forded, with indices removed entirely?
+The resulting representation of type theory would remain intrinsic, but all typing constraints would appear as equality proofs, which are definitionally proof-irrelevant in XTT or OTT.
+This could provide a way to escape transport hell without relying on strictification.
+
+Of course, explicitly transforming these typing constraints to equality proofs would still be tedious and error-prone, so an elaboration from QIITs to their Forded presentation using QIIRTs would be useful.
+The connection between QIITs and its Forded definition with the index eliminated echoes the notion of reornament~\cite{Dagand2014,Ko2016,Dagand2017}, so this translation itself may be of interest in general.
+
+\paragraph{Conclusion}
+There are still gaps between a pen-and-paper type theory and a fully formalised type theory in a proof assistant, on both the theoretical and practical sides.
+Without further advances in the technology of proof assistants, formalising type theory intrinsically within a proof assistant remains a difficult task.
+We hope that the lessons learned here can help the design of future proof assistants, so that one day we may implement a proof assistant within a proof assistant without (too much) sweat and tears.
+
+\begin{acks}
+  We appreciate the constructive comments from the anonymous reviewers.
+  Our syntax for displayed equations in \cref{sec:tt:elim} is inspired by the syntax of equational reasoning for displayed categories on 1Lab.\footnote{\url{https://1lab.dev/Cat.Displayed.Base.html}}
+  We are also grateful to Shu-Hung You for his comments on the early draft.
+  The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.
+\end{acks}
+
+%\IfFileExists{./reference.bib}{\bibliography{reference}}{\bibliography{ref}}
+\bibliography{ref}
 
@@
 
-%\begin{acks}
-%To Robert, for the bagels and explaining CMYK and color spaces.
-%\end{acks}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main.tex"
+%%% eval: (add-hook 'after-save-hook (lambda () (shell-command "make tex")) t)
+%%% End:

### 2025-12-03 tex/CPP26/main.tex
Remove leftover comment

@@
 
-% FNF (Fri 5 Sep)
-
 Internalising the syntax and semantics of type theory in type theory is a long-standing problem which stretches the limits of the theory~\cite{Dybjer1996,Danielsson2006,Chapman2009,McBride2010,Altenkirch2016a}.
@@
 
-Early approaches to formalising type theory, e.g.\ McKinna and Pollack~\cite{McKinna1999}, dealt with untyped terms that were later refined by a typing relation, setoid equality, or quotients, and thus had to prove a lot of congruence lemmas by hand~\cite{Danielsson2006,Chapman2009,Brunerie2019,Boer2020}.
+Early approaches to formalising type theory, e.g.\ McKinna and Pollack~\cite{McKinna1999}, dealt with untyped terms that were later refined by a typing relation or setoid equality, and thus had to prove a lot of congruence lemmas by hand~\cite{Danielsson2006,Chapman2009}.
 %
@@
 %
-Indeed, in a natural model, the indexing of terms over their types $\mathsf{Tm}_{\Gamma} : \mathsf{Ty}(\Gamma) \to \mathsf{Set}$ (as in a category with families) is replaced by a `fibred' perspective where each term instead \emph{has} a type, as picked out by a function $\mathsf{tyOf} : \mathsf{Tm}(\Gamma) \to \mathsf{Ty}(\Gamma)$.
-%
-Terms and types are still indexed by contexts $\Gamma$, but since most `type mismatches' arise from equations between types, not equations between contexts (indeed many formulations of type theory does not even have a notion of context equality), this should mean that many uses of transports can be avoided.
+Indeed, in a natural model, the family of terms indexed by their types $\mathsf{Tm}_{\Gamma} : \mathsf{Ty}(\Gamma) \to \mathsf{Set}$ (as in a category with families) is replaced by a `fibred' perspective where each term instead \emph{has} a type, as picked out by a function $\mathsf{tyOf} : \mathsf{Tm}(\Gamma) \to \mathsf{Ty}(\Gamma)$.
+%
+Terms and types are still indexed by contexts $\Gamma$, but since most `type mismatches' arise from equations between types, not equations between contexts (indeed many formulations of type theory do not even have a notion of context equality), this should mean that many uses of transports can be avoided.
 
@@
 %
-\CA is particularly a good fit for such a project, because not only does it support QIITs, it also supports inductive-recursive types~\cite{Dybjer1999}, which are needed to simultaneously define the recursive $\mathsf{tyOf}$ function together with the inductively defined types $\mathsf{Tm}(\Gamma)$ and $\mathsf{Ty}(\Gamma)$.
+\CA is a particularly good fit for such a project, because not only does it support QIITs, it also supports inductive-recursive types~\cite{Dybjer1999}, which are needed to simultaneously define the recursive $\mathsf{tyOf}$ function together with the inductively defined types $\mathsf{Tm}(\Gamma)$ and $\mathsf{Ty}(\Gamma)$.
 %
@@
 %
-Strictification~\cite{Donko2022,Kaposi2025}, a recent technique for strictifying equations, cannot be applied directly because of the lack of strict proposition in the metatheory.
+Strictification~\cite{Donko2022,Kaposi2025} --- a recent technique for making certain equations hold up to definitional equality --- partially helps, but does not completely eliminate the problem.
 %
@@
 \item We discuss strictification constructions on models, and show that they also apply to our notion of natural models (\cref{sec:strictify}).
-\item We develop normalisation by evaluation for substitution calculus~(\cref{sec:nbe}) as a proof of concept: our development is carried out in \CA, which has a computational implementation of QIITs and principles such as function extensionality, so the resulting normaliser computes, and can potentially be extracted as a verified program.
+\item We develop normalisation by evaluation for the substitution calculus~(\cref{sec:nbe}) as a proof of concept: our development is carried out in \CA, which has a computational implementation of QIITs and principles such as function extensionality, so the resulting normaliser computes, and can potentially be extracted as a verified program.
 \item We discuss pros and cons of our approach compared to other approaches, and which features of a proof assistant and its metatheory would make future formalisation more feasible (\cref{sec:discussion}).
@@
 
-%\LT{the idea of using natural model appears at least in 2024 \cite{Bense2024}, and it is a natural idea to formalise type theory in this way.}
 \section{Setting and metatheory}
-% FNF (Sun 7 Sep)
 
@@
 Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0.
+It also typechecks with \Agda version 2.8.0.
 %
@@
 
-In this section, we will explain why Altenkirch and Kaposi's representation~\cite{Altenkirch2016a} is hard to use in practice and rejected by \CA arising from transports in its definition.
+In this section, we explain why Altenkirch and Kaposi's representation~\cite{Altenkirch2016a} is hard to use in practice, and rejected by \CA due to transports in its definition.
 Then, we show how their representation can be transformed to a representation based on Awodey's natural models.
@@
 Of course, one could bypass the strict positivity check, but doing so would undermine the general trustworthiness of formalisation.
-Another possibility is to fix the syntactic restriction for HIITs, but it is unclear what conditions should be.
-Therefore, we seek for an equivalent definition without transports first.
-
-Another source of transports arises from equations over equations, but this can be avoided by using dependent paths.
+Another possibility is to fix the syntactic restriction for HIITs, but it is unclear what the conditions should be.
+Therefore, we seek an equivalent definition without transports first.
+
+One option is to use \emph{dependent paths} to encode equations over equations.
 For example, the fact that the identity term substitution really acts as an identity is introduced as an equality constructor \ensuremath{[\mathsf{idS}]_{\text{t}}}, defined over the highlighted equality constructor \ensuremath{[\mathsf{idS}]_{\text{T}}\;\mathbin{:}\;\Conid{A}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} for the identity type substitution:
@@
 \end{hscode}\resethooks
-Although equations over equations are in principle more manageable, it quickly leads us to \emph{equations over equations over yet more equations} in their elimination rules.
+This is in principle managable --- indeed, Altenkirch, Kaposi and Xie~\cite{Altenkirch2026} successfully use dependent paths and some encoding tricks to represent the groupoid syntax of type theory in \CA --- but the \ensuremath{\Conid{PathP}} type is not part of a standard presentation of type theory, but instead rather \CA specific (of course, dependent paths can also be encoded using transports, but then inheriting the problems described above).
+Dependent paths also quickly leads to equations over equations over yet more equations in their elimination rules.
 Hence, it is still preferable to avoid them if possible.
@@
 To avoid transports in the definition itself, we note that the index \ensuremath{\Conid{A}} of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}} often needs an explicit proof for the typing constraint --- for example, that the term \ensuremath{\Varid{t}} in the substitution \ensuremath{(\Varid{σ}\;\Varid{,}\;\Varid{t})} has type \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} --- if this does not happen to hold strictly (i.e., up to judgemental equality), so enforcing this constraint in the index of \ensuremath{\Conid{Tm}} just shoots ourselves in the foot.
-Hence, we apply the `Ford transformation'~\cite{McBride1999} (`You can have any index you want, as long as it is equal to the specified one') to move the constraint on its index to its argument as an equality proof (highlighted below):
+Hence, we apply the `Ford transformation'~\cite{McBride1999} (``You can have any index you want, as long as it is equal to the specified one'') to move the constraint on its index to its argument as an equality proof (highlighted below):
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any information but only defers the proof obligation.
+As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any additional information but only defers the proof obligation.
 %This redundant argument can be removed later when defining its eliminator (\Cref{sec:tt:elim}).
@@
 
-\subsection{Representing substitution calculus using QIIRT}
-Building on the changes described in \Cref{sec:tt:terms-without-indices}, we now spell out our version of substitution calculus.
+\subsection{Representing the substitution calculus using QIIRT}
+Building on the changes described in \Cref{sec:tt:terms-without-indices}, we now spell out our version of the substitution calculus.
 The following types are defined simultaneously with a recursive function (changes compared to the QIIT version highlighted):
@@
 \end{hscode}\resethooks
-We write \ensuremath{\Varid{wk}} for the `weakening' substitution \ensuremath{\Varid{wk}\;\mathrel{=}\;\Varid{π₁}\;\Varid{idS}\;\mathbin{:}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})\;\Conid{Γ}}, and \ensuremath{\Varid{vz}} for the `top variable' \ensuremath{\Varid{vz}\;\mathrel{=}\;\Varid{π₂}\;\Varid{idS}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})}.
-This works well except that we have to interleave the function clauses of \ensuremath{\Varid{tyOf}} with constructors.
+We write \ensuremath{\Varid{wk}} for the `weakening' substitution \ensuremath{\Varid{wk}\;\mathrel{=}\;\Varid{π₁}\;\Varid{idS}\;\mathbin{:}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})\;\Conid{Γ}}, and \ensuremath{\Varid{vz}} for the most recently bound `zeroth variable' \ensuremath{\Varid{vz}\;\mathrel{=}\;\Varid{π₂}\;\Varid{idS}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})}.
+
+The above definition works well, except that we have to interleave the function clauses of \ensuremath{\Varid{tyOf}} with constructors.
 For example, we need define the function clause for \ensuremath{\Varid{π₂}\;\Varid{σ}} before the $\eta$-law for substitution:
@@
 Yet, as interleaving function clauses with inductive types is also not supported, the strict equality \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathrel{=}\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} is not available at this point for \ensuremath{\Varid{p}}.
-We use forward declarations to introduce the required equalities \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} and \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{idS})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} to be defined later as \ensuremath{\Varid{refl}} and \ensuremath{[\circ]_{\text{T}}}, see \Cref{sec:tt:mutual} for further details.}
+We again use forward declarations to introduce the required equalities \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{σ})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{π₁}\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} and \ensuremath{\Varid{tyOf}\;(\Varid{π₂}\;\Varid{idS})\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{π₁}\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} to be defined later as \ensuremath{\Varid{refl}} and \ensuremath{[\circ]_{\text{T}}}, see \Cref{sec:tt:mutual} for further details.}
 
@@
 Apart from the extra clauses of \ensuremath{\Varid{tyOf}}, the main change happens in the constructor \ensuremath{\Varid{app}}.
-The constraint that \ensuremath{\Varid{t}} is of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} is enforced there, but every other constructor remains as almost the same as their QIIT definition.
+The constraint that \ensuremath{\Varid{t}} is of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} is enforced there, but every other constructor remains almost the same as their QIIT definition.
 We have formulated application and abstraction as an isomorphism between terms of type \ensuremath{\Conid{B}} in context \ensuremath{\Conid{Γ}\;\Varid{,}\;\Conid{A}} and terms of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} in context \ensuremath{\Conid{Γ}}, but we can also derive ordinary application:
@@
 
-As an example, we can write the identity function at type \ensuremath{\Conid{A}} as \ensuremath{\Varid{id}\;\Conid{A}\;\mathrel{=}\;\Varid{abs}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;(\Varid{π₂}\;\Varid{idS})}, and the identity function at type \ensuremath{\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\kern-1.5pt]_{\text{T}})} applied to the identity function at type \ensuremath{\Conid{A}} is represented by
+As an example, we can write the identity function at type \ensuremath{\Conid{A}} as \ensuremath{\Varid{id}\;\Conid{A}\;\mathrel{=}\;\Varid{abs}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;\Varid{vz}}, and the identity function at type \ensuremath{\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\kern-1.5pt]_{\text{T}})} applied to the identity function at type \ensuremath{\Conid{A}} is represented by
 \begin{hscode}\SaveRestoreHook
@@
 
-For the type \ensuremath{\Conid{𝔹}} of Boolean, its code \ensuremath{\Varid{𝕓}} is introduced with a type equality \ensuremath{\Conid{El𝕓}} such that the elements of \ensuremath{\Varid{𝕓}} are exactly \ensuremath{\Conid{𝔹}}:
+For the type \ensuremath{\Conid{𝔹}} of Booleans, its code \ensuremath{\Varid{𝕓}} is introduced with a type equality \ensuremath{\Conid{El𝕓}} such that the elements of \ensuremath{\Varid{𝕓}} are exactly \ensuremath{\Conid{𝔹}}:
 \begin{hscode}\SaveRestoreHook
@@
 
-In the end, we emphasise that the introduction of superfluous equality proofs and constructors only makes sense because of set-truncation.
+In the end, we emphasise that the introduction of superfluous equality proofs and constructors only makes sense because of set truncation.
 These additional arguments are essentially unique and thus do not add any new laws to type theory, but merely serve as devices to meet the syntactic restriction of strict positivity in the current implementation of \CA.
+After the definition, one can introduce `smart' constructors that supply canonical constructions for the superfluous equality proofs, if one so wishes.
 
@@
 The recursion and elimination principles make this property concrete.
-Here, we only discuss the part for substitution calculus, since other type formers are addressed similarly.
+Here, we only discuss the part for the substitution calculus, since other type formers are addressed similarly.
 For the interested reader, see our formalisation.
@@
 The signature for an algebra is packed in a record type~\ensuremath{\Conid{SC}} (short for Substitution Calculus).
-Inductive types and the function \ensuremath{\Varid{tyOf}} are interpreted as indexed types and a function between sets.
-Constructors of our syntax, except superfluous ones, correspond to function fields in this record, including equality constructors and clauses of \ensuremath{\Varid{tyOf}}.
+Inductive types and the function \ensuremath{\Varid{tyOf}} are interpreted as indexed sets and a function between sets, respectively, with explicit set truncation assumptions.
+Constructors of our syntax
+correspond to function fields in this record, including equality constructors and clauses of \ensuremath{\Varid{tyOf}}.
 \begin{hscode}\SaveRestoreHook
@@
 
-Superfluous equality constructors, like \ensuremath{\Conid{S.tyOfπ₂idS}}, are not required as fields in the record.
-Instead, their semantic counterparts are defined within any given model using the other methods.
-For example,
+One could think that it would be advantageous to leave out superfluous
+equality constructors as fields in the record, as their semantic
+counterparts can be defined within any given model using the other
+methods, and thus reduce the burden for users defining models.
+For example, the semantic interpretation of \ensuremath{\Conid{S.tyOfπ₂idS}} can be given as
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-This simplifies the definition of models, as we only need to provide interpretations for the essential constructors.
+However, some models might allow stricter and thus simpler
+implementations of such `derived' constructors, which in turn might
+simplify later definitions\footnote{For example, we were struggling to
+  finish the construction of the standard \ensuremath{\Conid{Set}} model including
+  $\Pi$-types and a universe, described in \Cref{sec:standard-model},
+  before we changed to a more specialised interpretation of some
+  derived constructors. This made our remaining proof obligations
+  definitionally true.}, and so we \emph{do} include them in the
+record.  For derived equality constructors such as \ensuremath{\Varid{tyOfπ₂idS}} above,
+their implementation is essentially unique due to set truncation
+assumption, but for derived point constructors, we also ask for a
+proof that the constructor is equal to its the canonical
+interpretation.
 
@@
 
+Unfortunately, the termination checker of \CA is especially brittle and does not see the above as obviously terminating, perhaps (among other things) because of the seemingly unrelated recursive call \ensuremath{\Varid{recTy}\;\Conid{B}} in the type of \ensuremath{\Varid{recTyOf}}.
+Since all recursive calls in clauses defining the recursive functions are on structurally smaller terms, we are confident that the definition is indeed terminating, and mark it as such using the \texttt{\{-\# TERMINATING \#-\}} pragma.
+
 For the elimination principle, we consider the notion of displayed algebras over an \ensuremath{\Conid{SC}}-algebra \ensuremath{\Conid{M}}, as a record type \ensuremath{\mathsf{SC}^{\bullet}} parametric in \ensuremath{\Conid{M}}, and later instantiate \ensuremath{\Conid{M}} to the term algebra, i.e.\ the syntax.
@@
 As motives are indexed by their underlying model, we will have equations over equations of the underlying model.
-It is convenient to specialise dependent paths for them, e.g.,
+Inspired by the syntax of equational reasoning for displayed categories in the 1Lab~\cite{1lab:displayed}, we introduce some convenient syntax for specialised dependent paths, e.g.,
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-The signature for \ensuremath{\mathsf{SC}^{\bullet}}-algebras is similar to the one for \ensuremath{\Conid{SC}}-algebras.
-The difference here is that each displayed operation is indexed by their underlying operation and thus equations become equations over equations.
+The signature for \ensuremath{\mathsf{SC}^{\bullet}}-algebras is similar to the one for \ensuremath{\Conid{SC}}-algebras, except
+that each displayed operation is indexed by their underlying operation (leading to equations over equations).
 \begin{hscode}\SaveRestoreHook
@@
 
-The elimination principle is stated similarly to the recursion principle but indexed over the term algebra~(\Cref{sec:meta:term}),
+The elimination principle is stated similarly to the recursion principle but indexed over the term algebra~(\Cref{sec:meta:term}):
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
+Just like for the recursors, we again mark the eliminators as terminating using a pragma.
 
@@
 \end{hscode}\resethooks
-We also use set-truncation to identify the highlighted \ensuremath{\Varid{p}\;\Varid{∙}\;\Varid{q}} with the desired underlying equation in special-purpose equational reasoning combinators such as the following:
+We also use set truncation to identify the highlighted \ensuremath{\Varid{p}\;\Varid{∙}\;\Varid{q}} with the desired underlying equation in special-purpose equational reasoning combinators such as the following:
 \begin{hscode}\SaveRestoreHook
@@
 These steps give us an equation over \ensuremath{\Varid{ηπ}\;\Varid{∙}\;(\Varid{refl}\;\Varid{∙}\;\Varid{refl})}, and \ensuremath{\Varid{beginSub[}\;\Varid{ηπ}\;\Varid{]\char95 }} gives us an easy way to correct this to an equation over \ensuremath{\Varid{ηπ}} as desired instead.
-
-\FNF{Comment on the trick of including extra fields in the record type of models}
 
@@
 This translation from QIITs in theory to their actual definitions in \CA should be sufficient to define mutually interleaved QIITs.
-Indeed, pedagogical presentations of type theory typically introduces one type former at a time, together with its formation, introduction, elimination, and equality rules (see, e.g.~\cite{Hofmann1997}), rather than presenting all type formers at once using a few large monolithic sets of rules.
+Indeed, pedagogical presentations of type theory typically introduces one type former at a time, together with its formation, introduction, elimination, and equality rules (see, e.g., Hofmann~\cite{Hofmann1997}), rather than presenting all type formers at once using a few large monolithic sets of rules.
 
@@
 
-In our formalisation, we fall back on forward declarations alone with coercions.
+In our formalisation, we fall back on forward declarations along with coercions.
 We are still investigating the root cause of this behaviour, but it may point to a design flaw.
 
-\LT[inline]{TERMINATION checker? (R1, R3)}
+%\LT[inline]{TERMINATION checker? (R1, R3)}
 
@@
 Coherence conditions are then verified using standard properties of transport.
-We have formalised the standard model for type theory with $\Pi$-types, Booleans, and a Tarski universe, except the case for \ensuremath{\Varid{π[]}}.
+We have formalised the standard model for type theory with $\Pi$-types, Booleans, and a Tarski universe.
 
@@
 If regularity were available, this would collapse to the trivial reflexivity proof.
-%The case |π[]| above involves an equation over a transport of another transported term.
-%\LT[noinline]{try to finish it if we have time.}
 
@@
 The picture is very different for the logical predicate interpretation.
-Although NbE works cleanly, the logical predicate interpretation --- often considered a benchmark challenge~\cite{Abel2019} for language formalisation --- remains at least as difficult as in the QIIT-based setting, even for substitution calculus.
+Although NbE works cleanly, the logical predicate interpretation --- often considered a benchmark challenge~\cite{Abel2019} for language formalisation --- remains at least as difficult as in the QIIT-based setting, even for the substitution calculus.
 
@@
 \subsection{Strictification} \label{sec:strictify}
-\FNF[inline]{Work further on the idea of strictification? (R2)}
+
 Instead, we turn our attention to \emph{strictification}~\cite{Donko2022,Kaposi2025}: given a model of type theory, certain equality constructors can be made strict (i.e., made to hold judgementally) to form a new model.
@@
 \>[5]{}\Varid{nat}\;{}\<[10]%
-\>[10]{}\mathbin{:}\;(\Varid{τ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Ξ}\;\Conid{Θ})\;(\Varid{δ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Θ}\;\Conid{Γ})\;\Varid{→}\;\Varid{y}\;\Varid{δ}\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;\Varid{y}\;(\Varid{δ}\;\Varid{∘}\;\Varid{τ}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-By set-truncation, any two morphisms \ensuremath{\Varid{σ}} and \ensuremath{\Varid{σ'}} in the presheaf category are equal whenever their functionals agree:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{≡ʸ→≡}\;\mathbin{:}\;\{\mskip1.5mu \Varid{σ}\;\Varid{σ'}\;\mathbin{:}\;\Conid{Subʸ}\;\Conid{Γ}\;\Conid{Δ}\mskip1.5mu\}\;\Varid{→}\;\Varid{σ}\;\Varid{≡ʸ}\;\Varid{σ'}\;\Varid{→}\;\Varid{σ}\;\mathop\equiv\;\Varid{σ'}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-where \ensuremath{\Varid{\char95 ≡ʸ\char95 }} denotes the path type between their functionals.
-Completing the Yoneda embedding then gives us strict unit and associativity laws, up to \ensuremath{\Varid{≡ʸ→≡}}:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[10]{}\mathbin{:}\;(\Varid{τ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Ξ}\;\Conid{Θ})\;(\Varid{δ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Θ}\;\Conid{Γ})\;{}\<[E]%
+\\
+\>[10]{}\Varid{→}\;\{\mskip1.5mu \Varid{δ'}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Ξ}\;\Conid{Γ}\mskip1.5mu\}\;\Varid{→}\;\Varid{δ}\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;\Varid{δ'}\;\Varid{→}\;\Varid{y}\;\Varid{δ}\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;\Varid{y}\;\Varid{δ'}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We are thankful to one of our anonymous reviewers for pointing out
+that introducing \ensuremath{\Varid{δ'}} and a proof of \ensuremath{\Varid{δ}\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;\Varid{δ'}} rather than working
+with \ensuremath{\Varid{δ}\;\Varid{∘}\;\Varid{τ}} directly gives a little bit more slack, which gives us
+strict unit and associativity laws:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{17}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Yoneda}\;\Varid{.idS}\;{}\<[24]%
+\>[24]{}\mathrel{=}\;(\Varid{λ}\;\Varid{γ}\;\Varid{→}\;\Varid{γ})\;\Varid{,}\;(\Varid{λ}\;\Varid{τ}\;\Varid{δ}\;\Varid{r}\;\Varid{→}\;\Varid{r}){}\<[E]%
+\\
+\>[B]{}\Conid{Yoneda}\;\Varid{.\char95 ∘\char95 }\;(\Varid{σ}\;\Varid{,}\;\Varid{pσ})\;(\Varid{σ'}\;\Varid{,}\;\Varid{pσ'})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathrel{=}\;(\Varid{λ}\;\Varid{γ}\;\Varid{→}\;\Varid{σ}\;(\Varid{σ'}\;\Varid{γ}))\;\Varid{,}\;(\Varid{λ}\;\Varid{τ}\;\Varid{δ}\;\Varid{r}\;\Varid{→}\;\Varid{pσ}\;\Varid{τ}\;(\Varid{σ'}\;\Varid{δ})\;(\Varid{pσ'}\;\Varid{τ}\;\Varid{δ}\;\Varid{r})){}\<[E]%
+\\
 \>[B]{}\Conid{Yoneda}\;\Varid{.idS∘\char95 }\;{}\<[17]%
 \>[17]{}\anonymous \;{}\<[24]%
-\>[24]{}\mathrel{=}\;\Varid{≡ʸ→≡}\;\Varid{refl}{}\<[E]%
+\>[24]{}\mathrel{=}\;\Varid{refl}{}\<[E]%
 \\
@@
 \>[17]{}\anonymous \;{}\<[24]%
-\>[24]{}\mathrel{=}\;\Varid{≡ʸ→≡}\;\Varid{refl}{}\<[E]%
+\>[24]{}\mathrel{=}\;\Varid{refl}{}\<[E]%
 \\
@@
 \>[17]{}\anonymous \;\anonymous \;\anonymous \;{}\<[24]%
-\>[24]{}\mathrel{=}\;\Varid{≡ʸ→≡}\;\Varid{refl}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-We can adapt the local universe construction~\cite{Lumsdaine2015,Donko2022} for \ensuremath{\Conid{M}\;\mathbin{:}\;\Conid{SC}} to further strictify the laws for type substitutions for substitution.
+\>[24]{}\mathrel{=}\;\Varid{refl}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+To further strictify the laws for type substitutions for the substitution calculus, we can adapt the local universe construction~\cite{Lumsdaine2015,Donko2022} for \ensuremath{\Conid{M}\;\mathbin{:}\;\Conid{SC}}.
 Types under the context $\Gamma$ in the local universe construction are interpreted as a triple consisting of a context $V$ as the local universe, a type under the context $V$, and a substitution from $\Gamma$ to $V$:
@@
 Type substitution is defined by `delaying' the substitution, i.e., by viewing the substitution \ensuremath{\Varid{⌜}\;\Conid{A}\;\Varid{⌝}} as an accumulator parameter.
-Then, we can show that the laws about type substitution boil down the right unit and the associativity laws for substitutions:
+Then, we can show that the functor laws for type substitution boil down to the right unit and the associativity laws for substitutions:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-If \ensuremath{\Conid{M}} is strictified by the Yoneda embedding, then the laws for identity substitution and substitution composition in \ensuremath{\Conid{M}\;\mathbin{!}} will be strict up to \ensuremath{\Varid{≡ʸ→≡}}.
-Hence, combining both techniques, we can construct models where both the category laws and substitution laws are strict after applying \ensuremath{\Varid{≡ʸ→≡}}.
-
+If \ensuremath{\Conid{M}} is strictified by the Yoneda embedding, then the laws for identity substitution and substitution composition in \ensuremath{\Conid{M}\;\mathbin{!}} will be strict.
+Hence, combining both techniques, we can construct models with strict category laws and substitution functor laws.
 Nevertheless, strictification does \emph{not} resolve our difficulties with the logical predicate interpretation.
-In particular, the lack of strict propositions (or just definitional UIP) in \CA prevents \ensuremath{\Varid{σ}\;\Varid{≡ʸ}\;\Varid{σ'}} from being strictly equal, since the paths between their properties must still be identified manually using set-truncation.
-Although \Agda provides a form of strict propositions \ensuremath{\Conid{Prop}}, it is not designed to work with \CA and interacts poorly with cubical primitives for now~\cite{Agda-issue2022}.
-As a consequence, coercions along equations identified by set-truncation remain unavoidable.
+For example, the above strictification will not make the substitution rule for $\Pi$-types strict, and thus transports are still needed.
 
@@
 \label{sec:discussion}
-
-%\FNF{Also discuss alternative choices such as basing the formalisation on (split) comprehension categories, see Brunerie and de Boer~\cite{Boer2020,Brunerie2019} (R2)}
 
@@
 
-The Ehrhard's style of presentation~\cite{Coquand2020,Ehrhard1988}, whereas the set of terms remains indexed by types but without any transport in the middle of terms for specifying equations, can be stated by using dependent paths and has been adopted by a recent formalisation~\cite{Altenkirch2026} in \CA.
-
-Our approach leads to no transports appearing in its own syntax and also avoids the use of dependent paths at all.
+Ehrhard~\cite{Ehrhard1988,Coquand2020} presented a variation of the substitution calculus, which, as Altenkirch, Kaposi and Xie~\cite{Altenkirch2026} point out, is well suited for formalisation in \CA, since all use of transports can be replaced by the use of dependent paths.
+%
+Our approach leads to no transports appearing in the syntax and also avoids the use of dependent paths at all.
 However, the same transports (and the same equations for them) seem to have a tendency to come back in concrete model constructions, as explained in \cref{sec:meta}.
@@
 
-It is not yet clear if other initial models of type theory, such as (split) comprehension categories~\cite{Jacobs1993} and contextual categories~\cite{Cartmell1986}, can be defined as QIITs or QIIRTs.
+For other notions of models of type theory, such as locally Cartesian closed categories~\cite{seely1984}, (split) comprehension categories~\cite{Jacobs1993} or contextual categories~\cite{Cartmell1986}, it is not yet clear if initial models can be defined as QIITs or QIIRTs.
+Brunerie and de Boer~\cite{Brunerie2019,Boer2020} work with contextual categories in order to prove the Initiality Conjecture~\cite{streicher1991} for Martin-L\"of Type Theory, but uses an extrinsic representation of terms.
 
@@
 Different forms of QIITs and inductive-recursive types (IRTs) have been used to develop type theory in type theory.
-Large and infinitary IRTs are used for the standard model of universe; small inductive-recursive types~\cite{Hancock2013} for implementing normalisation by evaluation~\cite{Altenkirch2017}; QIITs and QIIRTs for the intrinsic representations of type theory.
-Note further that the target of \ensuremath{\Varid{tyOf}} is the inductive type \ensuremath{\Conid{Ty}} being defined simultaneously, so the QIIRTs used to define our representation is neither large nor small.
+Large and infinitary IRTs are used for the standard model of universes, while small inductive-recursive types~\cite{Hancock2013} have been used for implementing normalisation by evaluation~\cite{Altenkirch2017}.
+Meanwhile QIITs and QIIRTs can be used for intrinsic representations of type theory.
+However, note that in our representation, the target of \ensuremath{\Varid{tyOf}} is the inductive type \ensuremath{\Conid{Ty}} being defined simultaneously, which is not a form of definition supported by current theories of large or small induction-recursion.
 Nevertheless, we expect that the encoding of small inductive-recursive types as indexed inductive types still applies to this `self-referring' induction-recursion, as we have derived our representation by `undoing' the encoding.
@@
 A framework to encompass all variants will be convenient for proof assistant implementation.
-Such a scheme may be developed by extending the type theory of QIITs~\cite{Kaposi2019,Kovacs2020a} with an additional construct to distinguish strict and weak equalities and a coverage condition for function clauses.
+Such a scheme might possibly be developed by extending the type theory of QIITs~\cite{Kaposi2019,Kovacs2020a} with an additional construct to distinguish strict and weak equalities and a coverage condition for function clauses.
 
@@
 Our workaround, using forward declarations to lift function clauses to fix the dependency, sacrifices their computational behaviour.
-Furthermore, our definitions appear to reach the limits of the termination checker: even seemingly harmless functions when defining the recursion principle fail to pass termination checking.
-
-% \paragraph{The need for strict propositions}
+Another option is to use `Xie's Trick'~\cite{xieTrick} and reduce the mutual definition to a two-sorted QIIT with a base sort of tags.
+Furthermore, our definitions appear to reach the limits of the termination checker: even seemingly harmless functions when defining the recursion principle fail to be seen as terminating.
+
 \paragraph{Strict propositions, and observational type theory}
@@
 To formalise the metatheory of type theory using QIITs with as few transports as possible, it seems inevitable to use a different metatheory rather than the off-the-shelf metatheory offered by \CA.
-In particular, regularity and definitional UIP, supported by OTT (see \cite{Altenkirch2019,Pujet2024} for the discussion on regularity) and by its cubical variant XTT~\cite{Sterling2022}, would immediately simplify our standard model~(\Cref{sec:standard-model}) and make formalising metatheory more feasible using Kaposi and Pujet's strictification.
+In particular, regularity and definitional UIP --- supported by OTT (see Altenkirch, Boulier, Kaposi and Tabareau~\cite{Altenkirch2019} and Pujet and Tabareau~\cite{Pujet2024} for discussion of regularity) and by its cubical variant XTT~\cite{Sterling2022} --- would simplify our standard model in \Cref{sec:standard-model} considerably.
 
@@
 The option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} in the forthcoming \Agda 2.9.0~\cite{Agda-issue2025} disables the \ensuremath{\Conid{Glue}} type in cubical mode and, in principle, yields a cubical type theory compatible with UIP~\cite{Agda-issue2019}.
-Since Cubical Agda already provides support for HIITs (with the earlier caveats), realising a type theory with QIITs, strict propositions, and computational UIP, as a variant of Cubical Agda may now be within reach~\cite{Tan2025,Tan2025a}.
+Since \CA already provides support for HIITs (with the earlier caveats), realising a type theory with QIITs, strict propositions, and computational UIP, as a variant of Cubical Agda may now be within reach~\cite{Tan2025,Tan2025a}.
 
@@
 \begin{acks}
-  We appreciate the constructive comments from the anonymous reviewers.
-  Our syntax for displayed equations in \cref{sec:tt:elim} is inspired by the syntax of equational reasoning for displayed categories on 1Lab.\footnote{\url{https://1lab.dev/Cat.Displayed.Base.html}}
-  We are also grateful to Shu-Hung You for his comments on the early draft.
-  The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.
+  We appreciate the comments from the anonymous reviewers, which improved the results of the paper and the formalisation in several ways.
+  We are also grateful to Shu-Hung You for his comments on an early draft.
+  The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3, and the Engineering and Physical Sciences Research Council
+  (EP/Y000455/2).
+
+  All data underpinning this publication are openly available from
+  \href{https://archive.softwareheritage.org/swh:1:dir:e403ad5f13c4f758df6a40c7a2eb072a6b24f1b3;origin=https://github.com/L-TChen/TTasQIIRT;visit=swh:1:snp:cb2e64c8f15931cd5c5a9673aae3aacfc6add86e;anchor=swh:1:rev:1d92dcc6b37babf3b304b33470f288c230bf98bd}{swh:1:rev:1d92dcc6b37babf3b304b33470f288c230bf98bd}.
 \end{acks}

### 2025-12-03 tex/CPP26/final/main.tex
submitted camera-ready version

--- 2025-12-03 tex/CPP26/main.tex
@@
-\documentclass[sigplan,screen,natbib,pbalance]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-\setcopyright{acmlicensed}
-\copyrightyear{2025}
-\acmYear{2025}
-\acmDOI{XXXXXXX.XXXXXXX}
-\acmConference[CPP '26]{Certified Programs and Proofs}{January 12--13, 2026}{Rennes, France}
-%%
-%%  Uncomment \acmBooktitle if the title of the proceedings is different
-%%  from ``Proceedings of ...''!
-%%
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-\acmISBN{978-1-4503-XXXX-X/2018/06}
+\documentclass[sigplan,screen,natbib,balance]{acmart}
+
+%%%%% Article setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% The following is specific to CPP '26 and the paper
+%%% 'Can We Formalise Type Theory Intrinsically without Any Compromise? A Case Study in Cubical Agda'
+%%% by Liang-Ting Chen, Fredrik Nordvall Forsberg, and Tzu-Chun Tsai.
+%%%
+\setcopyright{cc}
+\setcctype{by}
+\acmDOI{10.1145/3779031.3779090}
+\acmYear{2026}
+\copyrightyear{2026}
+\acmISBN{979-8-4007-2341-4/2026/01}
+\acmConference[CPP '26]{Proceedings of the 15th ACM SIGPLAN International Conference on Certified Programs and Proofs}{January 12--13, 2026}{Rennes, France}
+\acmBooktitle{Proceedings of the 15th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '26), January 12--13, 2026, Rennes, France}
+\received{2025-09-13}
+\received[accepted]{2025-11-13}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \usepackage[UKenglish]{babel}
-\usepackage{newunicodechar}
+%\usepackage{newunicodechar}
 \usepackage{xspace}
@@
 \let\Bbbk\relax
-\usepackage{amsmath,amssymb,mathbbol}
+\usepackage{amsmath,amssymb}
+%\usepackage{mathbbol} % not supported by ACM
 \usepackage{bbold}
 \usepackage[inline]{enumitem}
-
-%% Remove the following if there are no todo items.
-\setlength {\marginparwidth }{2cm}
-\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
 \usepackage[capitalise]{cleveref}
-\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
-\newcommand{\FNF}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
-
 \usepackage{microtype}
-\usepackage{newunicodechar}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \RequirePackage{amsfonts}
-\RequirePackage{newunicodechar}
+%\RequirePackage{newunicodechar}
 
@@
 \DeclareUnicodeCharacter{03C9}{\ensuremath{\omega}}
-\newunicodechar{β}{\ensuremath{\beta}}
-\newunicodechar{η}{\ensuremath{\eta}}
-\newunicodechar{Γ}{\ensuremath{\Gamma}}
-\newunicodechar{Δ}{\ensuremath{\Delta}}
-\newunicodechar{Θ}{\ensuremath{\Theta}}
-\newunicodechar{Ξ}{\ensuremath{\Xi}}
-\newunicodechar{Π}{\ensuremath{\Pi}}
-\newunicodechar{τ}{\ensuremath{\tau}}
-\newunicodechar{γ}{\ensuremath{\gamma}}
-\newunicodechar{δ}{\ensuremath{\delta}}
-\newunicodechar{∅}{\ensuremath{\emptyset}}
-\newunicodechar{≡}{\ensuremath{\equiv}}
-\newunicodechar{∘}{\ensuremath{\circ}}
-\newunicodechar{𝔹}{\ensuremath{\mathbb{B}}}
-\newunicodechar{𝕓}{\ensuremath{\mathbb{b}}}
-\newunicodechar{∎}{\ensuremath{\blacksquare}}
-\newunicodechar{∈}{\ensuremath{\in}}
-\newunicodechar{⋆}{\ensuremath{\star}}
-\newunicodechar{⌜}{\ensuremath{\ulcorner}}
-\newunicodechar{⌝}{\ensuremath{\urcorner}}
-\newunicodechar{ʸ}{\ensuremath{{}^y}}
-\newunicodechar{ⁿ}{\ensuremath{{}^n}}
-\newunicodechar{ᴾ}{\ensuremath{{}^P}}
+
+\DeclareUnicodeCharacter{03B2}{\ensuremath{\beta}}
+\DeclareUnicodeCharacter{03B7}{\ensuremath{\eta}}
+\DeclareUnicodeCharacter{0393}{\ensuremath{\Gamma}}
+\DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
+\DeclareUnicodeCharacter{0398}{\ensuremath{\Theta}}
+\DeclareUnicodeCharacter{039E}{\ensuremath{\Xi}}
+\DeclareUnicodeCharacter{03A0}{\ensuremath{\Pi}}
+\DeclareUnicodeCharacter{03C4}{\ensuremath{\tau}}
+\DeclareUnicodeCharacter{03B3}{\ensuremath{\gamma}}
+\DeclareUnicodeCharacter{03B4}{\ensuremath{\delta}}
+\DeclareUnicodeCharacter{2205}{\ensuremath{\emptyset}}
+\DeclareUnicodeCharacter{2261}{\ensuremath{\equiv}}
+\DeclareUnicodeCharacter{2218}{\ensuremath{\circ}}
+\DeclareUnicodeCharacter{1D539}{\ensuremath{\mathbb{B}}}
+\DeclareUnicodeCharacter{1D553}{\ensuremath{\mathbb{b}}}
+\DeclareUnicodeCharacter{220E}{\ensuremath{\blacksquare}}
+\DeclareUnicodeCharacter{2208}{\ensuremath{\in}}
+\DeclareUnicodeCharacter{22C6}{\ensuremath{\star}}
+\DeclareUnicodeCharacter{231C}{\ensuremath{\ulcorner}}
+\DeclareUnicodeCharacter{231D}{\ensuremath{\urcorner}}
+\DeclareUnicodeCharacter{02B8}{\ensuremath{{}^y}}
+\DeclareUnicodeCharacter{207F}{\ensuremath{{}^n}}
+\DeclareUnicodeCharacter{1D3E}{\ensuremath{{}^P}}
+\DeclareUnicodeCharacter{2219}{\ensuremath{\cdot}}
 
@@
 \DeclareUnicodeCharacter{1D57}{\ensuremath{^t}}
-\newunicodechar{∙}{\ensuremath{\cdot}}
 
@@
 \renewcommand\Varid[1]{\mathord{\textsf{#1}}}
-%\let\Conid\Varid
-%\newcommand\Keyword[1]{\textsf{\textbf{#1}}}
 
@@
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
 \title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
-\subtitle{A Case Study in \CA}
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
+\subtitle{A Case Study in Cubical Agda}
+
 \author{Liang-Ting Chen}
+\orcid{0000-0002-3250-1331}
 \affiliation{%
@@
   \city{Taipei}
-  \country{Taiwan}}
+  \country{Taiwan}
+}
 \email{liangtingchen@as.edu.tw}
@@
 \author{Fredrik Nordvall Forsberg}
+\orcid{0000-0001-6157-9288}
 \affiliation{%
@@
   \city{Glasgow}
-  \country{United Kingdom}}
+  \country{United Kingdom}
+}
 \email{fredrik.nordvall-forsberg@strath.ac.uk}
@@
 \author{Tzu-Chun Tsai}
+\orcid{0009-0008-2855-5717}
 \affiliation{%
@@
   \city{Taipei}
-  \country{Taiwan}}
-%\affiliation{%
-%  \institution{University of Amsterdam}
-%  \city{Amsterdam}
-%  \country{Netherlands}}
-
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-%\renewcommand{\shortauthors}{Trovato et al.}
+  \country{Taiwan}
+}
+\email{gene0905@icloud.com}
 
@@
 \keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}
-%% A "teaser" image appears between the author and affiliation
-%% information and the body of the document, and typically spans the
-%% page.
-
-%\received{20 February 2007}
-%\received[revised]{12 March 2009}
-%\received[accepted]{5 June 2009}
 
@@
 \section{Introduction}
-
-%\LT[inline]{Clearly articulate what has been formalised and what has not, when appropriate? Ideally finish formalisation, otherwise scale back and discuss what has been done. (R1, R3)}
-%\LT[inline]{Alternative formulation? E.g., \url{https://web.archive.org/web/20241004151846/https://lists.chalmers.se/pipermail/agda/2019/011176.html} (R2)}
-
-%\listoftodos
 
@@
 
-\section{Setting and metatheory}
-
-%\LT{Should we discharge this assumption in our formalisation, eg set-truncate where needed instead? (R1, R2)}
+\paragraph{Formalisation} Our formalisation can be found at \url{https://github.com/L-TChen/TTasQIIRT}, or archived with DOI \href{https://doi.org/10.5281/zenodo.17802827}{10.5281/zenodo.17802827}.
+
+\section{Setting and Metatheory}
+
 Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0.
@@
 %
-The constant path \ensuremath{\Varid{refl}} defined as \ensuremath{\Varid{λ}\;\anonymous \;\Varid{→}\;\Varid{a}} witnesses that equality is reflexive \ensuremath{\Varid{a}\;\mathop\equiv\;\Varid{a}}, and paths can be lifted to type families in the sense that there is a transport operation \ensuremath{\Varid{subst}\;\mathbin{:}\;(\Conid{P}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type})\;\Varid{→}\;\Varid{x}\;\mathop\equiv\;\Varid{y}\;\Varid{→}\;\Conid{P}\;\Varid{x}\;\Varid{→}\;\Conid{P}\;\Varid{y}}.
+The constant path \ensuremath{\Varid{refl}}, defined as \ensuremath{\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{a}}, witnesses that equality is reflexive \ensuremath{\Varid{a}\;\mathop\equiv\;\Varid{a}}, and paths can be lifted to type families in the sense that there is a transport operation \ensuremath{\Varid{subst}\;\mathbin{:}\;(\Conid{P}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type})\;\Varid{→}\;\Varid{x}\;\mathop\equiv\;\Varid{y}\;\Varid{→}\;\Conid{P}\;\Varid{x}\;\Varid{→}\;\Conid{P}\;\Varid{y}}.
 %
@@
 
-\section{Type theory as quotient inductive types} \label{sec:tt}
-
-%\FNF{Discuss ``another approach is staying completely indexed and always using PathP instead of transports'' here, or in Discussion (R2)}
+\section{Type Theory as Quotient Inductive Types} \label{sec:tt}
 
@@
 
-%This section's aim is to exhibit that Altenkirch and Kaposi's representation, which contains the transport hells and violates the syntactic restriction imposed by \CA, can be transformed to a representation based on Awodey's natural model, which is free from transports and accepted by \CA.
-%Then, we will show how other type formers can be represented in this way.
-%In the reminder of this section, we will give its elimination principle and explain how these definitions are formalised in \CA.
-
-\subsection{Type theory as the initial CwF model} \label{sec:tt:cwf}
+\subsection{Type Theory as the Initial CwF Model} \label{sec:tt:cwf}
 We briefly recall the QIIT representation given by Altenkirch and Kaposi.
@@
 The inhabitants of these types are valid derivations in type theory, because their validity is enforced by typing constraints.
-The four types of judgements in type theory are represented inductive-inductively %and indexed by their context and by their types for terms
+The four types of judgements in type theory are represented inductive-inductively
 as
@@
 
-When formulating the corresponding rule for the interaction between \ensuremath{\Varid{\char95 ∘\char95 }} and \ensuremath{\Varid{\char95 ,\char95 }}, we encounter a type mismatch that needs to be resolved by a transport (highlighted in \highlight{addition}{\text{green}}): %, leading to the transport hell when reasoning with this equality:
+When formulating the corresponding rule for the interaction between \ensuremath{\Varid{\char95 ∘\char95 }} and \ensuremath{\Varid{\char95 ,\char95 }}, we encounter a type mismatch that needs to be resolved by a transport (highlighted in \highlight{addition}{\text{green}}):
 \begin{hscode}\SaveRestoreHook
@@
 
-\subsection{The `Ford transformation' and index elimination} \label{sec:tt:terms-without-indices}
+\subsection{The `Ford Transformation' and Index Elimination} \label{sec:tt:terms-without-indices}
 
@@
 As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any additional information but only defers the proof obligation.
-%This redundant argument can be removed later when defining its eliminator (\Cref{sec:tt:elim}).
 
@@
 
-\subsection{Representing the substitution calculus using QIIRT}
+\subsection{Representing the Substitution Calculus using QIIRT}
 Building on the changes described in \Cref{sec:tt:terms-without-indices}, we now spell out our version of the substitution calculus.
@@
 
-\subsection{The type of Booleans}
+\subsection{The Type of Booleans}
 
@@
 
-\subsection{A Tarski universe} \label{sec:tt:univ}
+\subsection{A Tarski Universe} \label{sec:tt:univ}
 Using the same idiom described previously, a Tarski universe of types is introduced to our type theory in the same vein.
@@
 
-\subsection{Recursion and elimination principles} \label{sec:tt:elim}
+\subsection{Recursion and Elimination Principles} \label{sec:tt:elim}
 We turn to recursion and elimination principles for our syntax.
@@
 
-\subsection{Practical workarounds for mutual definitions}  \label{sec:tt:mutual}
+\subsection{Practical Workarounds for Mutual Definitions}  \label{sec:tt:mutual}
 
@@
 In practice, however, limitations (and occasional mysterious bugs) of the proof assistant require us to adopt certain workarounds in order to implement the intended definitions.
-%This shows that the current design of \CA is not yet fully aligned with the theory of quotient inductive-inductive types~\cite{Kaposi2019}.
-
-\paragraph{Mutually interleaved QIITs}
+
+\paragraph{Mutually Interleaved QIITs}
 Constructors of QIITs currently can not be interleaved in \CA~\cite{Agdaissue2021}, even within an \ensuremath{\Keyword{interleaved}\;\Keyword{mutual}} block.
@@
 
-\paragraph{Mutual interleaved QIIRTs}
+\paragraph{Mutual Interleaved QIIRTs}
 Interleaving function clauses with inductive types is a different matter, since we cannot forward declare a function clause together with its computational behaviour.\footnote{Custom rewrite rules are not allowed in \CA.}
@@
 
-\paragraph{Mutually-defined functions}
-%\LT[noinline]{(Low) Create an Agda issue?}
+\paragraph{Mutually-Defined Functions}
 Since the constructors of QII(R)Ts can be mutually interleaved, their recursion and elimination principles also need to be given in the same vein.
@@
 
-%\LT[inline]{TERMINATION checker? (R1, R3)}
-
 \section{Metatheory}\label{sec:meta}
@@
 
-%We find that reasoning with this definition \emph{per se} in \CA is hard even with substitution calculus: the lack of strict equalities in type theory is known to cause the transport hell~\cite{Altenkirch2016a} and even worse it cannot be mitigated by using custom rewriting rules~\cite{Cockx2020,Cockx2021} in \CA.
-%Nevertheless, programming with type theory seems doable, as normalisation can be
-
-\subsection{Term model} \label{sec:meta:term}
+\subsection{Term Model} \label{sec:meta:term}
 The term model is a self-interpretation of syntax, allowing displayed models to be instantiated over it and ensuring that the elimination rule computes.
@@
 
-\subsection{Standard model} \label{sec:standard-model}
+\subsection{Standard Model} \label{sec:standard-model}
 
@@
 
-\subsection{Normalisation by evaluation, and the logical predicate interpretation} \label{sec:nbe}
+\subsection{Normalisation by Evaluation, and the Logical Predicate Interpretation} \label{sec:nbe}
 We implement normalisation by evaluation (NbE) for the substitution calculus.
@@
 \end{hscode}\resethooks
-Compared to NbE for substitution calculus using QIITs, our formalisation is simpler: no transports appear at all, because variables and terms are not indexed by their types.
+Compared to NbE for the substitution calculus using QIITs, our formalisation is simpler: no transports appear at all, because variables and terms are not indexed by their types.
 
@@
 
-In the same spirit, we may attempt to `strictify' the category part of substitution calculus using the Yoneda embedding, so that the unit laws and associativity law hold strictly, \emph{modulo} the property of naturality.
+In the same spirit, we may attempt to `strictify' the category part of the substitution calculus using the Yoneda embedding, so that the unit laws and associativity law hold strictly, \emph{modulo} the property of naturality.
 Given any \ensuremath{\Conid{SC}}-algebra, we define a presheaf interpretation of \ensuremath{\Conid{Sub}} by
@@
 
-\section{Discussion and conclusion}
+\section{Discussion and Conclusion}
 \label{sec:discussion}
@@
 
-\paragraph{Comparison with intrinsic approaches}
+\paragraph{Comparison with Intrinsic Approaches}
 Previous intrinsic formalisations~\cite{Kaposi2019,Kaposi2017,Altenkirch2016a,Altenkirch2017} based on the CwF semantics of type theory have resorted to using postulated constructors and custom rewrite rules to manually define QIITs and their eliminators.
@@
 
-% Untyped version might still be easiest to work with, with current proof assistant technology
-
-\paragraph{A proper syntax for HII(R)Ts}
+\paragraph{A Proper Syntax for HII(R)Ts}
 The syntax for declaring higher inductive-inductive types and QIITs in \CA falls short of their theoretical capabilities~\cite{Kaposi2020b,Kaposi2019}.
@@
 
-\paragraph{A theory of QIIRTs}
+\paragraph{A Theory of QIIRTs}
 We work around the problem by defining type theory using QIIRTs, but this raises another question: what is the theory of QIIRTs, anyway?
@@
 
-\paragraph{The support of interleaved mutual definitions}
+\paragraph{The Support of Interleaved Mutual Definitions}
 Another challenge concerns interleaved mutual definitions.
@@
 
-\paragraph{Strict propositions, and observational type theory}
+\paragraph{Strict Propositions, and Observational Type Theory}
 The recent work on strictified syntax~\cite{Kaposi2025} addresses transport hell using observational type theory (OTT)~\cite{Pujet2022,Pujet2023,Pujet2024}, with strict propositions in the metatheory playing a central role.
@@
 %
-%\paragraph{The implementation of OTT or XTT}
 To formalise the metatheory of type theory using QIITs with as few transports as possible, it seems inevitable to use a different metatheory rather than the off-the-shelf metatheory offered by \CA.
@@
 
-\paragraph{The Ford transformation and definitional UIP}
+\paragraph{The Ford Transformation and Definitional UIP}
 The Ford transformation is known to work well with definitional UIP~\cite{Altenkirch2006}.
@@
   All data underpinning this publication are openly available from
-  \href{https://archive.softwareheritage.org/swh:1:dir:e403ad5f13c4f758df6a40c7a2eb072a6b24f1b3;origin=https://github.com/L-TChen/TTasQIIRT;visit=swh:1:snp:cb2e64c8f15931cd5c5a9673aae3aacfc6add86e;anchor=swh:1:rev:1d92dcc6b37babf3b304b33470f288c230bf98bd}{swh:1:rev:1d92dcc6b37babf3b304b33470f288c230bf98bd}.
+  \href{https://doi.org/10.5281/zenodo.17802827}{doi:10.5281/zenodo.17802827}.
 \end{acks}
 
-%\IfFileExists{./reference.bib}{\bibliography{reference}}{\bibliography{ref}}
 \bibliography{ref}

### 2025-12-03 tex/CPP26/final/main.tex
paper source

@@
 %% Please copy and paste the code instead of the example below.
-\begin{CCSXML}\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ccs2012>}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
+\begin{CCSXML}
+<ccs2012>
    <concept>

### 2025-12-03 tex/CPP26/final/main.tex
overfull hboxes

@@
 \begin{CCSXML}
-<ccs2012>
    <concept>
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{18}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\highlight{addition}{\Varid{\char95 ,\char95 ∶[\char95 ]}}\;{}\<[18]%
-\>[18]{}\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\>[18]{}\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[E]%
+\\
+\>[18]{}\Varid{→}\;(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
 \\
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{16}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{\char95 ↑\char95 }\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;\Varid{→}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\>[B]{}\Varid{\char95 ↑\char95 }\;{}\<[6]%
+\>[6]{}\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Δ})\;{}\<[E]%
+\\
+\>[6]{}\Varid{→}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
 \\
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{16}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\highlight{addition}{\Varid{app}}\;{}\<[16]%
-\>[16]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;(\Conid{B}\;\mathbin{:}\;\Conid{Ty}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}))\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;\Conid{B})}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}){}\<[E]%
+\>[16]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;(\Conid{B}\;\mathbin{:}\;\Conid{Ty}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}))\;{}\<[E]%
+\\
+\>[16]{}\Varid{→}\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;\Conid{B})}\;\Varid{→}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A}){}\<[E]%
 \\
@@
 \\
-\>[11]{}\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{wk}\;\Varid{,}\;\Varid{vz}\;\Varid{∶[}\;\Conid{𝔹[]₂}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\Varid{pb₂}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
+\>[11]{}\mathop\equiv\;\Conid{P}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{wk}\;\Varid{,}\;\Varid{vz}\;\Varid{∶[}\;\Conid{𝔹[]₂}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}}\;{}\<[E]%
+\\
+\>[11]{}\hsindent{4}{}\<[15]%
+\>[15]{}[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\Varid{pb₂}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{T}})\;{}\<[E]%
 \\
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{27}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \\
-\>[B]{}\Varid{...}{}\<[E]%
-\\
-\>[B]{}\Varid{recSub}\;(\Varid{σ}\;\Conid{S.,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;{}\<[27]%
-\>[27]{}\mathrel{=}\;\Varid{recSub}\;\Varid{σ}\;\Varid{,}\;\Varid{recTm}\;\Varid{t}\;\Varid{∶[}\;\Varid{recTyOf}\;\Varid{t}\;\Varid{pt}\;\mskip1.5mu]{}\<[E]%
+\>[B]{}\Varid{recSub}\;(\Varid{σ}\;\Conid{S.,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathrel{=}\;\Varid{recSub}\;\Varid{σ}\;\Varid{,}\;\Varid{recTm}\;\Varid{t}\;\Varid{∶[}\;\Varid{recTyOf}\;\Varid{t}\;\Varid{pt}\;\mskip1.5mu]{}\<[E]%
 \\
@@
 \>[14]{}\mathbin{:}\;\mathsf{tyOf}^{\bullet}\;{}\<[23]%
-\>[23]{}(\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;\Varid{≡Ty[}\;\Varid{tyOf[]}\;\mskip1.5mu]\;(\mathsf{tyOf}^{\bullet}\;\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{T}}){}\<[E]%
+\>[23]{}(\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;{}\<[E]%
+\\
+\>[23]{}\Varid{≡Ty[}\;\Varid{tyOf[]}\;\mskip1.5mu]\;(\mathsf{tyOf}^{\bullet}\;\Varid{t}^{\bullet}\;[\kern-1.5pt\;\Varid{σ}^{\bullet}\;\kern-1.5pt]^{\bullet}_{\text{T}}){}\<[E]%
 \\
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\Varid{p}\;\Varid{i}))\;\Varid{x'}\;\Varid{y'}\;\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\Varid{q}\;\Varid{i}))\;\Varid{y'}\;\Varid{z'}\;{}\<[E]%
+\>[3]{}\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\Varid{p}\;\Varid{i}))\;\Varid{x'}\;\Varid{y'}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{→}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{B}\;(\Varid{q}\;\Varid{i}))\;\Varid{y'}\;\Varid{z'}\;{}\<[E]%
 \\
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\Varid{beginSub[\char95 ]\char95 }\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\mathbin{:}\;(\{\mskip1.5mu \Varid{p}\mskip1.5mu\}\;\Varid{q}\;\mathbin{:}\;\Varid{σ}\;\mathop\equiv\;\Varid{τ})\;\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{p}\;\mskip1.5mu]\;\Varid{τ}^{\bullet}\;\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{q}\;\mskip1.5mu]\;\Varid{τ}^{\bullet}{}\<[E]%
+\>[3]{}\Varid{beginSub[\char95 ]\char95 }\;\mathbin{:}\;(\{\mskip1.5mu \Varid{p}\mskip1.5mu\}\;\Varid{q}\;\mathbin{:}\;\Varid{σ}\;\mathop\equiv\;\Varid{τ})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{p}\;\mskip1.5mu]\;\Varid{τ}^{\bullet}\;\Varid{→}\;\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{q}\;\mskip1.5mu]\;\Varid{τ}^{\bullet}{}\<[E]%
 \\
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{25}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \\
-\>[3]{}\Varid{std}\;\Varid{.\char95 ,\char95 ∶[\char95 ]}\;\Varid{σ}\;(\Conid{A}\;\Varid{,}\;\Varid{t})\;\Varid{p}\;\Varid{γ}\;\mathrel{=}\;(\Varid{σ}\;\Varid{γ}\;\Varid{,}\;\highlight{addition}{\Varid{transport}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{p}\;\Varid{i}\;\Varid{γ})\;(\Varid{t}\;\Varid{γ})}){}\<[E]%
+\>[3]{}\Varid{std}\;\Varid{.\char95 ,\char95 ∶[\char95 ]}\;\Varid{σ}\;(\Conid{A}\;\Varid{,}\;\Varid{t})\;\Varid{p}\;\Varid{γ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathrel{=}\;(\Varid{σ}\;\Varid{γ}\;\Varid{,}\;\highlight{addition}{\Varid{transport}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{p}\;\Varid{i}\;\Varid{γ})\;(\Varid{t}\;\Varid{γ})}){}\<[E]%
 \ColumnHook

### 2025-12-03 tex/CPP26/final/main.tex
ACM wants us to cite the formalisation

@@
 
-\paragraph{Formalisation} Our formalisation can be found at \url{https://github.com/L-TChen/TTasQIIRT}, or archived with DOI \href{https://doi.org/10.5281/zenodo.17802827}{10.5281/zenodo.17802827}.
+\paragraph{Formalisation} Our formalisation~\cite{formalisation} can be found at \url{https://github.com/L-TChen/TTasQIIRT}, or archived with DOI \href{https://doi.org/10.5281/zenodo.17802827}{10.5281/zenodo.17802827}.
 
@@
   (EP/Y000455/2).
+\end{acks}
+
+\section*{Data-Availability Statement}
 
 All data underpinning this publication are openly available from
-  \href{https://doi.org/10.5281/zenodo.17802827}{doi:10.5281/zenodo.17802827}.
-\end{acks}
+Zenodo~\cite{formalisation} with DOI \href{https://doi.org/10.5281/zenodo.17802827}{10.5281/zenodo.17802827}.
 
@@
 \end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main.tex"
-%%% eval: (add-hook 'after-save-hook (lambda () (shell-command "make tex")) t)
-%%% End:

### 2025-12-04 tex/CPP26/final/main.tex
ACM cannot deal with subtitles together with question marks

@@
 
-\title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
-\subtitle{A Case Study in Cubical Agda}
+\title[Can We Formalise Type Theory Intrinsically without Any Compromise?]{Can We Formalise Type Theory Intrinsically without Any Compromise? A Case Study in Cubical Agda}
 
@@
 \\
-\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
-\\
+\>[8]{}\Varid{→}\;{}\<[11]%
+\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;{}\<[11]%
+\>[11]{}\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
+\\[\blanklineskip]%
 \>[B]{}\hsindent{3}{}\<[3]%
@@
 \>[8]{}\mathbin{:}\;{}\<[11]%
-\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[35]%
-\>[35]{}\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
+\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
 \\
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{≡Sub[}\;\Varid{refl}\;\Varid{]⟨}\;\Varid{cong}\;(\Varid{λ}\;\Varid{z}\;\Varid{→}\;\Varid{...}\;\Varid{,}\;\Varid{...}\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{z}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;(\Conid{Ty∙-is-set}\;{}\<[71]%
-\>[71]{}\anonymous \;\anonymous )\;\Varid{⟩}\;{}\<[E]%
+\>[5]{}\Varid{≡Sub[}\;\Varid{refl}\;\Varid{]⟨}\;\Varid{cong}\;{}\<[25]%
+\>[25]{}(\Varid{λ}\;\Varid{z}\;\Varid{→}\;\Varid{...}\;\Varid{,}\;\Varid{...}\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{z}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;{}\<[E]%
+\\
+\>[25]{}(\Conid{Ty∙-is-set}\;{}\<[38]%
+\>[38]{}\anonymous \;\anonymous )\;\Varid{⟩}\;{}\<[E]%
 \\
